### 2015-10-29 paper/pldi16-submission/main.tex
init pldi16 paper

%
% LaTeX template for prepartion of submissions to PLDI'15
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'15 web site.
% 
\documentclass[pldi,indentedstyle=false]{sigplanconf-pldi15}

%
% the following standard packages may be helpful, but are not required
%
\usepackage{SIunits}            % typset units correctly
\usepackage{courier}            % standard fixed width font
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

\begin{document}

%
% any author declaration will be ignored  when using 'plid' option (for double blind review)
%

\title{NanoMaLy}

\authorinfo{Eric L. Seidel\and Ranjit Jhala}
           {UC San Diego}
           {{eseidel,rjhala}@cs.ucsd.edu}

\authorinfo{Westley Weimer}
           {University of Virginia}
           {weimer@viginia.edu}

\maketitle
\begin{abstract}
\end{abstract}

% 11 pages total (not including bib)
% ----------------------------------
%  1p : intro
%  2p : overview
%  2p : type-carrying semantics
%  1p : evaluation: quickchecking type errors
%  2p : interactive semantics
%  1p : evaluation: nanomaly
%  1p : related work
%  1p : conclusion

\section{Introduction}          % 1 page
\section{Overview}              % 2 pages?
\section{Type-carrying Semantics} % 2 pages
\begin{itemize}
\item values are tagged with their types, just like ``untyped'' langs
\item special ``hole'' value whose type is not yet known, used for function args
\item on-the-fly unification to determine ``correct'' type for holes
\end{itemize}

% Algorithm:
%   Input: ML Program (let f1 = e1 in let f2 = e2 in ... e)
%   Output: 'safe' or 'fn v1 .. vn' where 'fn v1 .. vn' gets stuck

\section{Evalution: Recasting Type Errors as Runtime Errors} % 1 page
\label{sec:eval-recast-type}
\begin{itemize}
\item benchmarks: our data + seminal data
\item both cases: \textbf{random} search sufficient to trigger runtime crash in 80\% of programs
\item how many of the ``safe'' programs are actually safe??
\end{itemize}

\section{``Interactive'' Semantics \& Trace Exploration} % 2 pages
\begin{itemize}
\item extend operational semantics to collect reduction graph
\item nodes are terms, edges indicate ``steps-to'' and ``sub-term'' relations
\item visualize path through reduction graph
\item expand edges to reveal more fine-grained steps (step/jump forward/backward)
\item never lose context (unlike traditional debugger)
\end{itemize}

\section{Evaluation: NanoMaLy}                % 1 page
\section{Related Work}              % 1 pages

\subsection{Type Error Localization}
\label{sec:type-error-local}
\begin{itemize}
\item seminal
\item bayesian (Zhang+Myers)
\item MaxSMT
\item counterfactual (Chen+Erwig)
\end{itemize}
\subsection{Debugging \& Program Understanding}
\label{sec:debugging}

\section{Conclusion}                % 1 page

\end{document}

### 2015-11-10 paper/pldi16-submission/evaluation.tex
texing

\begin{figure}
\judgementHead{Evaluation}{\stepsto{e, \su}{e, \su}}
$$
\inference[\recontext]{\stepsto{e, \su}{e_1, \su_1}}{\stepsto{C[e], \su}{C[e_1], \su_1}}
$$

$$
\inference[\restuck]{}{\stepsto{C[\stuck], \su}{\stuck, \su}}
$$

$$
\inference[\replusgood]
  {n_1, \su_2 = \force{v_1}{\tnat} \\
   n_2, \su_3 = \force{v_2}{\tnat} \\ 
   n = \eplus{n_1}{n_2}}
  {\stepsto{\eplus{v_1}{v_2}, \su_1}{n, \su_1\su_2\su_3}}
$$

$$
\inference[\replusbadone]
  {\stuck, \su_2 = \force{v_1}{\tnat}}
  {\stepsto{\eplus{v_1}{v_2}, \su_1}{\stuck, \su_1\su_2}}
$$

$$
\inference[\replusbadtwo]
  {\stuck, \su_2 = \force{v_2}{\tnat}}
  {\stepsto{\eplus{v_1}{v_2}, \su_1}{\stuck, \su_1\su_2}}
$$

$$
\inference[\reifgoodone]
  {\etrue, \su_2 = \force{v}{\tbool}}
  {\stepsto{\eif{v}{e_1}{e_2}, \su_1}{e_1, \su_1\su_2}}
$$

$$
\inference[\reifgoodtwo]
  {\efalse, \su_2 = \force{v}{\tbool}}
  {\stepsto{\eif{v}{e_1}{e_2}, \su_1}{e_2, \su_1\su_2}}
$$

$$
\inference[\reifbad]
  {\stuck, \su_2 = \force{v}{\tbool}}
  {\stepsto{\eif{v}{e_1}{e_2}, \su_1}{\stuck, \su_1\su_2}}
$$
\caption{Evaluation relation}
\label{fig:evaluation}
\end{figure}

### 2015-11-10 paper/pldi16-submission/semantics.tex
texing

\begin{itemize}
\item how do we run ill-typed programs?
\item for a lang like ocaml, dynamic semantics are independent of static
  semantics, just lambda calculus. so no problem to run ill-typed
  program
\item but what about functions? what type of arguments should we pass? consider

\begin{lstlisting}
let f x = 
  let y = 1 + x in
    1. +. y
\end{lstlisting}

does \texttt{f} take an int, float, string? int and float are both
somewhat plausible, but string or anything else is ``clearly'' bogus. so
we cannot provide \emph{completely arbitrary} inputs to
\texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
\hole{}, which is a placeholder for a value whose type we have not
yet determined. As we execute the program, we instantiate holes with
concrete values as demanded by the primitive operations in the
program. For example, the hole we pass to f will be instantiated to an
int when we reach the \lstinline{1 + x} term. Thus, y will be an int as
well, and the program will get stuck at \lstinline{1. +. y}. \ES{this
  reads more like overview text..}

Next, we formalize our search for witnesses to type-errors. We present
the syntax and dynamic semantics of a simple lambda calculus with
integers and booleans, extended with our notion of holes, as well as our
search algorithm. We prove that our system \emph{soundly} finds
witnesses, \ie if we find a witness then there is no possible typing for
the input program.

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}

\subsection{Semantics}
\label{sec:semantics}
\input{evaluation.tex}

\item values are tagged with their types, just like ``untyped'' langs
\item special ``hole'' value whose type is not yet known, used for function args
\item on-the-fly unification to determine ``correct'' type for holes
\end{itemize}

% Algorithm:
%   Input: ML Program (let f1 = e1 in let f2 = e2 in ... e)
%   Output: 'safe' or 'fn v1 .. vn' where 'fn v1 .. vn' gets stuck

### 2015-11-11 paper/pldi16-submission/semantics.tex
separate expr and type holes
texified proof sketch

@@
 \texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
-\hole{}, which is a placeholder for a value whose type we have not
+\ehole{}, which is a placeholder for a value whose type we have not
 yet determined. As we execute the program, we instantiate holes with
@@
 
+% \item values are tagged with their types, just like ``untyped'' langs
+% \item special ``hole'' value whose type is not yet known, used for function args
+% \item on-the-fly unification to determine ``correct'' type for holes
+\end{itemize}
+
 Next, we formalize our search for witnesses to type-errors. We present
-the syntax and dynamic semantics of a simple lambda calculus with
-integers and booleans, extended with our notion of holes, as well as our
+the syntax and dynamic semantics of \lang -- a simple lambda calculus with
+integers and booleans, extended with our notion of holes -- as well as our
 search algorithm. We prove that our system \emph{soundly} finds
@@
 
+\autoref{fig:syntax} describes the syntax of \lang, a simple lambda
+calculus with numbers and booleans. As we are specifically interested in
+programs that \emph{do} go wrong, we include an explicit \stuck state in
+our syntax.
+
+\paragraph{Holes}
+\label{sec:holes}
+The main novelty in our system is the notion of a ``hole'', written
+\ehole{}, which represents an uninstantiated value. Importantly, we do
+not even know what type the value should have. Holes may also appear in
+types, where they may be thought of as type variables that we will not
+generalize over.
+
 \subsection{Semantics}
 \label{sec:semantics}
-\input{evaluation.tex}
+\input{operational.tex}
 
-\item values are tagged with their types, just like ``untyped'' langs
-\item special ``hole'' value whose type is not yet known, used for function args
-\item on-the-fly unification to determine ``correct'' type for holes
-\end{itemize}
+\autoref{fig:operational} describes the small-step contextual reduction
+semantics for \lang. The evaluation relation is parameterized by a pair
+of \forcesym and \gensym functions that handle runtime type-checking and
+instantiation of holes respectively. The relation must also maintain a
+substitution \su, mapping holes to concrete values, so that we do not
+instantiate the same hole with different values in different contexts,
+and importantly, so that we can report a concrete witness to any
+discovered type errors.
+
+Note that each primitive reduction step -- addition, if-elimination, and
+function application -- uses \forcesym to ensure that values have the
+appropriate type (and that holes are instantiated) before continuing the
+computation. Additionally, beta-reduction \emph{does not} type-check its
+argument, it only ensures that the function can be applied.
+
+\begin{thm}
+\label{thm:all-reduce}
+  Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
+\end{thm}
+% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
+%   Simple induction on the evaluation relation.
+% \end{proof}
+
+\subsection{Soundness}
+\label{sec:soundness}
+We now show that our evaluation relation conservatively instantiates
+holes, \ie if, given a function $f$, we report that 
+$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$,
+then for any type $t$ you assign to $f$, there exists an input \hastype{v}{t} such that
+$\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
+
+Intuitively, we can think of the evaluation of \eapp{f}{\ehole{}} as
+computing a \emph{partial type} -- a type that may contain holes -- for
+$f$. We can extract this type from the result of evaluation as follows.
+
+\begin{defn}
+\label{defn:partial-type}
+  If \stepi{i}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su}, then the $i$th
+  \emph{partial type} of f, written \ptype{i}{f},
+  is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
+
+  We will omit the subscript when we wish to refer to the final partial
+  type. \ES{need to define typeof..}
+\end{defn}
+
+We also define a refinement relation between types.
+
+\begin{defn}
+\label{defn:subtyping}
+  A type $s$ is a \emph{refinement} of a type $t$, written \tsub{s}{t}, if
+  $\exists \su.\ t = \subst{\su}{s}$.
+\end{defn}
+
+Given these two definitions, we show that each evaluation step
+refines the partial type of $f$.
+
+\begin{lem}
+\label{lem:refine-partial}
+  For all $k$, \tsub{\ptype{k+1}{f}}{\ptype{k}{f}}.
+\end{lem}
+
+Finally, we show that any value that is not compatible with the $k$th
+partial type of $f$ will cause $f$ to get stuck in \emph{at most} $k$
+steps.
+
+\begin{lem}
+\label{lem:k-stuck}
+  For any \hastype{v}{t} that is not compatible with
+  the input of \ptype{k}{f}, \stepi{k}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
+\end{lem}
+\begin{proof}
+  By induction on $k$.
+
+  Let $s_k = \ptype{k}{f}$. Suppose \hastype{v}{\tnsub{t}{s_{k+1}}}, we
+  will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
+
+  \begin{description}
+  \item[Case \tnsub{t}{s_k}:] 
+    The inductive hypothesis applies.
+  \item[Case \tsub{t}{s_k} but \tnsub{t}{s_{k+1}}:]
+    By \autoref{cor:force} we must have called \forcesym at step $k+1$. A
+    case analysis of the applicable rules shows that \forcesym cannot
+    have succeeded.
+  \end{description}
+\end{proof}
+
+Now we can prove our soundness theorem.
+
+\begin{thm}
+\label{thm:soundness}
+  For any function $f$, if \steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
+  then
+
+$$
+  \forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}
+$$
+\end{thm}
+\begin{proof}
+  Let $\tfun{s}{\_} = \ptype{}{f}$.
+  \begin{description}
+  \item[Case \tsub{s}{t} or \tsub{t}{s}:] Our witness is already valid.
+  \item[Case \tincompat{s}{t}:] We can apply Lemma~\ref{lem:k-stuck} and
+    use \emph{any} \hastype{v}{t} as a witness for $f$.
+  \end{description}
+\end{proof}
+
+% \item values are tagged with their types, just like ``untyped'' langs
+% \item special ``hole'' value whose type is not yet known, used for function args
+% \item on-the-fly unification to determine ``correct'' type for holes
+% \end{itemize}
 

### 2015-11-12 paper/pldi16-submission/semantics.tex
s/nat/int/g

@@
 Next, we formalize our search for witnesses to type-errors. We present
-the syntax and dynamic semantics of \lang -- a simple lambda calculus with
+the syntax and operational semantics of \lang -- a simple lambda calculus with
 integers and booleans, extended with our notion of holes -- as well as our
@@
 
-\autoref{fig:syntax} describes the syntax of \lang, a simple lambda
+Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
 calculus with numbers and booleans. As we are specifically interested in
@@
 
-\autoref{fig:operational} describes the small-step contextual reduction
-semantics for \lang. The evaluation relation is parameterized by a pair
-of \forcesym and \gensym functions that handle runtime type-checking and
-instantiation of holes respectively. The relation must also maintain a
-substitution \su, mapping holes to concrete values, so that we do not
-instantiate the same hole with different values in different contexts,
-and importantly, so that we can report a concrete witness to any
+Figure~\ref{fig:operational} describes the small-step contextual
+reduction semantics for \lang. The evaluation relation is parameterized
+by a pair of \forcesym and \gensym functions that handle runtime
+type-checking and instantiation of holes respectively. The relation must
+also maintain a substitution \su, mapping holes to concrete values, so
+that we do not instantiate the same hole with different values in
+different contexts, and so that we can report a concrete witness to any
 discovered type errors.
@@
 computation. Additionally, beta-reduction \emph{does not} type-check its
-argument, it only ensures that the function can be applied.
+argument, it only ensures that the value being applied is a function.
 
@@
   Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
+  \ES{do we really need to state this, or is it obvious?}
 \end{thm}
@@
 $\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$,
-then for any type $t$ you assign to $f$, there exists an input \hastype{v}{t} such that
+then for any type $\tfun{s}{t}$ you assign to $f$, there exists an input \hastype{v}{s} such that
 $\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
 
-Intuitively, we can think of the evaluation of \eapp{f}{\ehole{}} as
+We can think of the evaluation of \eapp{f}{\ehole{}} as
 computing a \emph{partial type} -- a type that may contain holes -- for

### 2015-11-12 paper/pldi16-submission/semantics.tex
compatible instead of refinement

@@
 
-We also define a refinement relation between types.
+We also define a compatibility relation between types.
 
 \begin{defn}
-\label{defn:subtyping}
-  A type $s$ is a \emph{refinement} of a type $t$, written \tsub{s}{t}, if
-  $\exists \su.\ t = \subst{\su}{s}$.
+\label{defn:type-compat}
+  A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
+  $\exists \su.\ t = \subst{\su}{s} \lor s = \subst{\su}{t}$.
 \end{defn}
@@
 \label{lem:refine-partial}
-  For all $k$, \tsub{\ptype{k+1}{f}}{\ptype{k}{f}}.
+  For all $k$, \tcompat{\ptype{k}{f}}{\ptype{k+1}{f}}.
 \end{lem}
+
+\begin{lem}
+\label{lem:force-inst}
+  For all $k$, if $\ptype{k}{f} \neq \ptype{k+1}{f}$, then \forcesym must
+  have been called at step $k+1$.
+\end{lem}
+\begin{proof}
+  By case analysis on the evaluation rules. If 
+  $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in 
+  $f$'s argument must have been instantiated with a concrete value at step $k+1$.
+  An examination of the rules shows that only place this happens is 
+  in the first case of \forcesym.
+\end{proof}
 
@@
 
-  Let $s_k = \ptype{k}{f}$. Suppose \hastype{v}{\tnsub{t}{s_{k+1}}}, we
+  Let $s_k = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
   will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
@@
   \begin{description}
-  \item[Case \tnsub{t}{s_k}:] 
+  \item[Case \tincompat{t}{s_k}:] 
     The inductive hypothesis applies.
-  \item[Case \tsub{t}{s_k} but \tnsub{t}{s_{k+1}}:]
-    By \autoref{cor:force} we must have called \forcesym at step $k+1$. A
-    case analysis of the applicable rules shows that \forcesym cannot
-    have succeeded.
+  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:] 
+    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
+    $k+1$. A case analysis of the applicable rules shows that \forcesym
+    cannot have succeeded.
   \end{description}
@@
   For any function $f$, if \steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
-  then
-
-$$
-  \forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}
-$$
+  then $\forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
 \end{thm}
@@
   \begin{description}
-  \item[Case \tsub{s}{t} or \tsub{t}{s}:] Our witness is already valid.
+  \item[Case \tcompat{s}{t}:] Our witness is already valid.
   \item[Case \tincompat{s}{t}:] We can apply Lemma~\ref{lem:k-stuck} and

### 2015-11-12 paper/pldi16-submission/semantics.tex
formatting

@@
 
-Next, we formalize our search for witnesses to type-errors. We present
-the syntax and operational semantics of \lang -- a simple lambda calculus with
-integers and booleans, extended with our notion of holes -- as well as our
-search algorithm. We prove that our system \emph{soundly} finds
-witnesses, \ie if we find a witness then there is no possible typing for
-the input program.
-
+Next, we formalize our search for witnesses to type-errors.
+%
+We present the syntax and operational semantics of \lang -- a simple
+lambda calculus with integers and booleans, extended with our notion of
+holes -- as well as our search algorithm.
+%
+We prove that our system \emph{soundly} finds witnesses, \ie if we find
+a witness then there is no possible typing for the input program.
+%
 \subsection{Syntax}
@@
 \input{syntax.tex}
-
+%
 Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
-calculus with numbers and booleans. As we are specifically interested in
+calculus with numbers and booleans.
+%
+As we are specifically interested in
 programs that \emph{do} go wrong, we include an explicit \stuck state in
 our syntax.
-
+%
 \paragraph{Holes}
@@
 The main novelty in our system is the notion of a ``hole'', written
-\ehole{}, which represents an uninstantiated value. Importantly, we do
-not even know what type the value should have. Holes may also appear in
-types, where they may be thought of as type variables that we will not
-generalize over.
-
+\ehole{}, which represents an uninstantiated value.
+%
+Importantly, we do not even know what type the value should have.
+%
+Holes may also appear in types, where they may be thought of as type
+variables that we will not generalize over.
+%
 \subsection{Semantics}
@@
 \input{operational.tex}
-
+%
 Figure~\ref{fig:operational} describes the small-step contextual
-reduction semantics for \lang. The evaluation relation is parameterized
-by a pair of \forcesym and \gensym functions that handle runtime
-type-checking and instantiation of holes respectively. The relation must
-also maintain a substitution \su, mapping holes to concrete values, so
-that we do not instantiate the same hole with different values in
-different contexts, and so that we can report a concrete witness to any
-discovered type errors.
+reduction semantics for \lang.
+%
+We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
+$e$ is $e_1$, $e'$ is $e_j$ and $\forall i,j, 1 \leq i < j$, we have
+\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
+%
+We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
+$\stepi{j}{e}{\su}{e'}{\su'}$.
+%
+The evaluation relation is parameterized by a pair of \forcesym and
+\gensym functions that handle runtime type-checking and instantiation of
+holes respectively.
+%
+The relation must also maintain a substitution \su, mapping holes to
+concrete values, so that we do not instantiate the same hole with
+different values in different contexts, and so that we can report a
+concrete witness to any discovered type errors.
 
@@
 appropriate type (and that holes are instantiated) before continuing the
-computation. Additionally, beta-reduction \emph{does not} type-check its
-argument, it only ensures that the value being applied is a function.
-
+computation.
+%
+Additionally, beta-reduction \emph{does not} type-check its argument, it
+only ensures that the value being applied is a function.
+%
 \begin{thm}
@@
 % \end{proof}
-
+%
 \subsection{Soundness}
@@
 $f$. We can extract this type from the result of evaluation as follows.
-
+%
 \begin{defn}
@@
 \end{defn}
-
+%
 We also define a compatibility relation between types.
-
+%
 \begin{defn}
@@
 \end{defn}
-
+%
 Given these two definitions, we show that each evaluation step
 refines the partial type of $f$.
-
+%
 \begin{lem}
@@
 \end{lem}
-
+%
 \begin{lem}
@@
 \begin{proof}
-  By case analysis on the evaluation rules. If 
-  $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in 
-  $f$'s argument must have been instantiated with a concrete value at step $k+1$.
+  By case analysis on the evaluation rules. 
+  % 
+  If $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in $f$'s
+  argument must have been instantiated with a concrete value at step
+  $k+1$.
+  %
   An examination of the rules shows that only place this happens is 
@@
 \end{proof}
-
+%
 Finally, we show that any value that is not compatible with the $k$th
@@
 steps.
-
+%
 \begin{lem}
@@
   will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
-
   \begin{description}
@@
 \end{proof}
-
+%
 Now we can prove our soundness theorem.
-
+%
 \begin{thm}

### 2015-11-12 paper/pldi16-submission/semantics.tex
proof of lem:refine-partial

@@
-\begin{itemize}
-\item how do we run ill-typed programs?
-\item for a lang like ocaml, dynamic semantics are independent of static
-  semantics, just lambda calculus. so no problem to run ill-typed
-  program
-\item but what about functions? what type of arguments should we pass? consider
-
-\begin{lstlisting}
-let f x = 
-  let y = 1 + x in
-    1. +. y
-\end{lstlisting}
-
-does \texttt{f} take an int, float, string? int and float are both
-somewhat plausible, but string or anything else is ``clearly'' bogus. so
-we cannot provide \emph{completely arbitrary} inputs to
-\texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
-\ehole{}, which is a placeholder for a value whose type we have not
-yet determined. As we execute the program, we instantiate holes with
-concrete values as demanded by the primitive operations in the
-program. For example, the hole we pass to f will be instantiated to an
-int when we reach the \lstinline{1 + x} term. Thus, y will be an int as
-well, and the program will get stuck at \lstinline{1. +. y}. \ES{this
-  reads more like overview text..}
-
-% \item values are tagged with their types, just like ``untyped'' langs
-% \item special ``hole'' value whose type is not yet known, used for function args
-% \item on-the-fly unification to determine ``correct'' type for holes
-\end{itemize}
-
+% \begin{itemize}
+% \item how do we run ill-typed programs?
+% \item for a lang like ocaml, dynamic semantics are independent of static
+%   semantics, just lambda calculus. so no problem to run ill-typed
+%   program
+% \item but what about functions? what type of arguments should we pass? consider
+%
+% \begin{lstlisting}
+% let f x = 
+%   let y = 1 + x in
+%     1. +. y
+% \end{lstlisting}
+%
+% does \texttt{f} take an int, float, string? int and float are both
+% somewhat plausible, but string or anything else is ``clearly'' bogus. so
+% we cannot provide \emph{completely arbitrary} inputs to
+% \texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
+% \ehole{}, which is a placeholder for a value whose type we have not
+% yet determined. As we execute the program, we instantiate holes with
+% concrete values as demanded by the primitive operations in the
+% program. For example, the hole we pass to f will be instantiated to an
+% int when we reach the \lstinline{1 + x} term. Thus, y will be an int as
+% well, and the program will get stuck at \lstinline{1. +. y}. \ES{this
+%   reads more like overview text..}
+%
+% % \item values are tagged with their types, just like ``untyped'' langs
+% % \item special ``hole'' value whose type is not yet known, used for function args
+% % \item on-the-fly unification to determine ``correct'' type for holes
+% \end{itemize}
+%
 Next, we formalize our search for witnesses to type-errors.
@@
 \end{lem}
+\begin{proof}
+  By case analysis on the evaluation rules. 
+  %
+  Note that all rules are type-preserving with the exception of when
+  \forcesym is called on a hole, which case we instantiate the hole with
+  a concrete value.
+  %
+  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
+\end{proof}
+%
+Furthermore, only a call to \forcesym can change the partial type of $f$.
 %
@@
     By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
-    $k+1$. A case analysis of the applicable rules shows that \forcesym
-    cannot have succeeded.
+    $k+1$. 
+    %
+    A case analysis of the applicable rules shows that \forcesym cannot
+    have succeeded.
   \end{description}

### 2015-11-12 paper/pldi16-submission/semantics.tex
define typeof

@@
 
-We can think of the evaluation of \eapp{f}{\ehole{}} as
-computing a \emph{partial type} -- a type that may contain holes -- for
-$f$. We can extract this type from the result of evaluation as follows.
+We can think of the evaluation of \eapp{f}{\ehole{}} as computing a
+\emph{partial type} -- a type that may contain holes -- for $f$.
+%
+We can extract this type from the result of evaluation as follows.
 %
@@
   type. \ES{need to define typeof..}
+\end{defn}
+%
+The \typeof{} function is an approximation of the type of an expression.
+\begin{defn}
+\label{def:typeof}
+  \[
+  \begin{array}{lcll}
+    \typeof{n}   & \defeq & \tint & \\
+    \typeof{b}   & \defeq & \tbool & \\
+    \typeof{\efun{x}{e}} & \defeq & \tfun{\thole{i}}{\typeof{e}}, & \quad \text{$i$ is fresh} \\
+    \typeof{e} & \defeq & \thole{i}, & \quad \text{$i$ is fresh} \\
+  \end{array}
+  \]
 \end{defn}

### 2015-11-12 paper/pldi16-submission/semantics.tex
misc

@@
   We will omit the subscript when we wish to refer to the final partial
-  type. \ES{need to define typeof..}
+  type.
 \end{defn}
@@
 
-% \item values are tagged with their types, just like ``untyped'' langs
-% \item special ``hole'' value whose type is not yet known, used for function args
-% \item on-the-fly unification to determine ``correct'' type for holes
-% \end{itemize}
+\subsection{Search Algorithm}
+\label{sec:search-algorithm}
+%
+We have shown how to find a witness for a function of a single argument,
+but in a language with higher-order functions and currying it may not be
+clear \emph{syntactically} how many arguments a function takes.
+%
+Thus, we wrap our operational semantics for \lang in a search loop that
+supplies an increasing number of arguments until the function returns a
+value that is not a lambda.
 

### 2015-11-13 paper/pldi16-submission/semantics.tex
misc
search loop

@@
 
-  Let $s_k = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
+  Let $\tfun{s_k}{\_} = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
   will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
@@
 
+The top-level search loop takes as input an open program -- a sequence
+of binders -- and searches for an expression that closes the input
+program and gets stuck. Concretely, given an input
+%
+\begin{code}
+  let f1 = e1 in
+  let f2 = e2 in
+  ...
+  let fn = en in
+\end{code}
+%
+we will search for an expression of the form:
+%
+\begin{code}
+  fn v1 ... vn
+\end{code}
+%
+Figure~\ref{fig:expression-api} describes a small API for manipulating
+and evaluating \lang expressions, which we will use to define our search
+algorithm.
+%
+The bulk of the search is performed by @eval@, which
+implements our operational semantics from \S\ref{sec:semantics}.
+%
+The operational semantics is non-deterministic due to \gensym,
+thus @eval@ returns a list of possible results.
+%
+\begin{figure}[t]
+  \centering
+  \begin{mcode}
+  -- transitive small-step reduction, 
+  -- returning a list of results
+  eval :: ($e$, $\su$) -> [($v$, $\su$)]
+
+  -- manipulating expressions
+  subst   :: $\su$ -> $v$ -> $v$
+  mkApps  :: $e$ -> [$e$] -> $e$
+  mkLets  :: [($x$, $e$)] -> $e$ -> $e$
+  isStuck :: $v$ -> Bool
+  \end{mcode}
+  \caption{Expression API}
+  \label{fig:expression-api}
+\end{figure}
+%
+We also define a few helper functions for manipulating expressions:
+\begin{itemize}
+\item @subst@ applies a substitution of holes to a value,
+\item @mkApps@ creates a nested sequence of applications in the usual
+  left-associative style,
+\item @mkLets@ takes a list of binders and a body expression, and
+  creates a sequence of nested let-binders, and
+\item @isStuck@ tests whether a value is the \stuck term.
+\end{itemize}
+%
+\begin{figure*}[t]
+  \centering
+  \begin{mcode}
+  check :: [($x$, $e$)] -> Result
+  check bnds =
+    -- (2) search for a witness
+    case find (isStuck . fst) results of
+      Nothing      -> Safe
+      Just (_, su) -> Unsafe (mkApps f (subst su args))
+    where
+      (args, results) = loop []
+      f               = snd (last bnds)
+      build args      = mkLets bnds (mkApps f args)
+
+      -- (1) find the correct number of arguments
+      loop :: [$v$] -> ([$v$], [($v$, $\su$)])
+      loop args = case eval (build args, []) of 
+        ($\efun{x}{e}$, _) : _ -> loop (args `snoc` $\ehole{}$)
+        results      -> (args, results)  
+  \end{mcode}
+  \caption{Implementing our counter-example search in terms of the
+    operational semantics. 
+    \ES{1 and 2 are backwards, ugh...}
+    \ES{should address case where output types of successive runs dont match}
+  }
+  \label{fig:search-algo}
+\end{figure*}
+%
+Figure~\ref{fig:search-algo} summarizes the overall implementation of
+our search for witnesses, which takes as input a sequences of binders
+and returns either @Safe@ if no witness could be found, or |Unsafe $e$|
+where $e$ is a term that causes the input program to get stuck.
+
 % Algorithm:

### 2015-11-13 paper/pldi16-submission/evaluation-witness.tex
evaluation

\section{Evaluation: Recasting Type Errors as Runtime Errors}
\label{sec:eval-witness}
%
The immediate question is ``What fraction of type errors admit
witnesses?''
%
To answer this question we implemented a prototype of our search
procedure for the pure subset of \ocaml, \ie \lang extended with
algebraic datatypes and records. 
%
In our implementation we instantiated \gensym with a simple random
generation of values, which we will show is more than sufficient for the
majority of type errors.
%
We evaluated our implementation of two sets of known-bad programs, \ie
programs that were rejected by the \ocaml compiler because of a type
error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at UC San Diego (IRB \#XXXXX). 
%
We recorded each interaction with the \ocaml top-level system over the
course of the first three assignments, from which we extracted XXXX
distinct, ill-typed \ocaml programs.
%
The second dataset -- widely used in the literature -- comes from a
similar course at the University of Washington~\cite{XXXXX}, from which
we extracted XXXX ill-typed programs.

We ran our search algorithm on each program with the entry point set to
the function that \ocaml had identified as containing a type error. 
%
Due to the possibility of non-termination we set a limit on the number
of reductions to perform, increasing in 1,000-step increments from 1,000
steps to 10,000 steps total.
%
We also added a na\"ive check for infinite recursion; at each recursive
function call we check whether the new arguments are identical to the
current arguments.
%
If so, the function cannot possibly terminate and we report an error.
%
While not a \emph{type error}, infinite recursion is still a clear bug
in the program, and thus valuable feedback for the user.

\begin{figure}[t]
  \centering

  Have a table or graph! \ES{graph of cumulative errors per timeout? \ie
    monotonically increasing, with a cap of 90\% or so}
\caption{Experimental results}
\label{fig:results-witness}
\end{figure}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figure~\ref{fig:results-witness}.
%
In both datasets our tool was able to find a witness for at least XX\%
of the programs.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, X\% triggered an unbound variable error and X\%
triggered an infinite recursion error.
%
XX programs were deemed safe and XX timed out even at 10,000 steps, \ie
we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg dynamic-symbolic execution,
could likely trigger more of the type errors, our experiments show that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

% \begin{itemize}
% \item benchmarks: our data + seminal data
% \item both cases: \textbf{random} search sufficient to trigger runtime crash in 80\% of programs
% \item how many of the ``safe'' programs are actually safe??
% \end{itemize}

### 2015-11-13 paper/pldi16-submission/main.tex
texing
operational semantics
texified proof sketch
s/nat/int/g
search loop
evaluation

@@
 % 
-\documentclass[pldi,indentedstyle=false]{sigplanconf-pldi15}
+\documentclass[10pt,pldi,indentedstyle=false]{sigplanconf-pldi15}
 
@@
 %
-\usepackage{SIunits}            % typset units correctly
+\usepackage{amsmath}
+\usepackage{amssymb}
+% \usepackage{SIunits}            % typset units correctly
 \usepackage{courier}            % standard fixed width font
@@
 \newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
+\usepackage{xspace}
+\usepackage{commands}
+
+\usepackage{amsthm}
+% \newtheorem{lemma}{Lemma}
+% \newtheorem{corollary}{Corollary}
+% \newtheorem{theorem}{Theorem}
+% \newtheorem{definition}{Definition}
+\theoremstyle{plain}% default 
+\newtheorem{thm}{Theorem} % [section] 
+\newtheorem{lem}[thm]{Lemma} 
+\newtheorem{prop}[thm]{Proposition} 
+\newtheorem*{cor}{Corollary} 
+\theoremstyle{definition}
+\newtheorem{defn}{Definition} % [section]
+
+\usepackage[inference]{semantic}
+
+\input{haskellListings}
 
@@
 \section{Overview}              % 2 pages?
-\section{Type-carrying Semantics} % 2 pages
-\begin{itemize}
-\item values are tagged with their types, just like ``untyped'' langs
-\item special ``hole'' value whose type is not yet known, used for function args
-\item on-the-fly unification to determine ``correct'' type for holes
-\end{itemize}
+% \section{Type-carrying Semantics} % 2 pages
+\input{semantics.tex}
 
-% Algorithm:
-%   Input: ML Program (let f1 = e1 in let f2 = e2 in ... e)
-%   Output: 'safe' or 'fn v1 .. vn' where 'fn v1 .. vn' gets stuck
-
-\section{Evalution: Recasting Type Errors as Runtime Errors} % 1 page
-\label{sec:eval-recast-type}
-\begin{itemize}
-\item benchmarks: our data + seminal data
-\item both cases: \textbf{random} search sufficient to trigger runtime crash in 80\% of programs
-\item how many of the ``safe'' programs are actually safe??
-\end{itemize}
+% \section{Evalution: Recasting Type Errors as Runtime Errors} % 1 page
+\input{evaluation-witness.tex}
 

### 2015-11-13 paper/pldi16-submission/interactive.tex
start interactive semantics

\section{Explaining Type Errors With Traces}
\label{sec:explaining}
%
We have shown how to reliably find witnesses to type errors in \ocaml,
but this not fully address our original goal -- to \emph{explain} the
errors.
%
Having identified an input vector that triggers a bug, a common next
step is to step through the program with a \emph{debugger} to observe
how the program evolves.
%
The existing debuggers and interpreters for \ocaml assume a type-correct
program, so unfortunately we cannot use them off-the-shelf.
%
Instead we return to our semantics for \lang and extend the evaluation
rules to collect a trace that we can present to the user, demonstrating
precisely \emph{how} their program goes wrong.

The trace takes the form of a \emph{reduction graph}, where the nodes
are terms and the edges represent either the single-step $\hookrightarrow$
or a ``sub-term'' relation.

\begin{itemize}
\item extend operational semantics to collect reduction graph
\item nodes are terms, edges indicate ``steps-to'' and ``sub-term'' relations
\item visualize path through reduction graph
\item expand edges to reveal more fine-grained steps (step/jump forward/backward)
\item never lose context (unlike traditional debugger)
\end{itemize}

### 2015-11-14 paper/pldi16-submission/interactive.tex
ugh traces
todo

@@
 rules to collect a trace that we can present to the user, demonstrating
-precisely \emph{how} their program goes wrong.
+precisely \emph{how} their program went wrong.
 
 The trace takes the form of a \emph{reduction graph}, where the nodes
-are terms and the edges represent either the single-step $\hookrightarrow$
-or a ``sub-term'' relation.
+are terms and the edges represent either the single-step
+$\hookrightarrow$ or a ``sub-term'' relation. For example, evaluating
+the expression @1 + 2 + 3@ would produce the graph in
+Figure~\ref{fig:simple-reduction}.
+%
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\linewidth]{simple.png}
+\caption{The reduction graph for \texttt{1 + 2 + 3}.}
+\label{fig:simple-reduction}
+\end{figure}
+%
 

### 2015-11-15 paper/pldi16-submission/interactive.tex
interactive semantics

@@
 %
+We choose this graph representation instead of a simple, linear sequence
+of expressions because it will allow us to express a variety of
+traversals, such as ``step into'' and ``step over'' -- commonly found in
+traditional debuggers.
 
-\begin{itemize}
-\item extend operational semantics to collect reduction graph
-\item nodes are terms, edges indicate ``steps-to'' and ``sub-term'' relations
-\item visualize path through reduction graph
-\item expand edges to reveal more fine-grained steps (step/jump forward/backward)
-\item never lose context (unlike traditional debugger)
-\end{itemize}
+\subsection{Interactive Semantics}
+\label{sec:inter-semant}
+%
+\begin{figure*}[t]
+\relDescription{Trace Syntax}
+$$
+\begin{array}{rrcl}
+  & \tr & ::= & \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr
+\end{array}
+$$
+\\
+\relDescription{\subtermssym}
+\begin{gather*}
+\begin{array}{lcl}
+\subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
+\subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
+\subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
+                                &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\ 
+                                &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
+\subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
+                                &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
+\subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
+\subterms{e}                 & \defeq & \bullet
+\end{array}
+\end{gather*}
+\judgementHead{Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
+\begin{gather*}
+\inference[\recontext]
+  {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
+  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \subterms{C[e_1]}; \tr_1}} 
+\\ \\
+\inference[\reappgood]
+  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}}
+  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
+         {e\sub{x}{v_2}}{\su_1\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
+\\ \\
+\inference[\reappbad]
+  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}}
+  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
+\end{gather*}
+\caption{A selection of the operational semantics from
+  Figure~\ref{fig:operational}, extended to collect a full reduction
+  graph.}
+\label{fig:interactive}
+\end{figure*}
+%
+The changes to the operational semantics of \S\ref{sec:semantics} are
+mechanical, so we will not reproduce them in full; instead we describe
+the procedure for extending a transition rule and provide a selection of
+examples in Figure~\ref{fig:interactive}.
+
+First, we extend the transition relation to collect a set of edges from
+which we will construct the graph. 
+%
+An edge between two expressions is either a ``steps-to'' edge -- written
+\singlestep{e_1}{e_2} -- indicating that $e_1$ transitions to $e_2$ in a
+single step, or a ``sub-term'' edge -- written \subterm{e_1}{e_2} --
+indicating that $e_1$ contains $e_2$ as a sub-expression.
+%
+Collecting the steps-to edges is a simple matter of recording the
+consequent of each original rule in the trace; each original judgment
+\step{e_1}{\su_1}{e_2}{\su_2} becomes 
+\stepg{e_1}{\su_1}{\tr_1}{e_2}{\su_2}{\singlestep{e_1}{e_2}; \tr_1}.
+%
+The sub-term edges can be delegated to a new \subtermssym helper
+function, which adds edges from an expression to each of its
+\emph{immediate} sub-expressions.
+%
+We collect new sub-term edges after each transition, thus the final
+template for the small-step relation is:
+$$
+\stepg{e_1}{\su_1}{\tr_1}{e_2}{\su_2}{\singlestep{e_1}{e_2}; \subterms{e_2}; \tr_1}
+$$
+
+% \begin{itemize}
+% \item extend operational semantics to collect reduction graph
+% \item nodes are terms, edges indicate ``steps-to'' and ``sub-term'' relations
+% \item visualize path through reduction graph
+% \item expand edges to reveal more fine-grained steps (step/jump forward/backward)
+% \item never lose context (unlike traditional debugger)
+% \end{itemize}

### 2015-11-15 paper/pldi16-submission/semantics.tex
evaluation
confusing terminology

@@
+\section{Searching for Type-Error Witnesses}
+\label{sec:searching-witness}
 % \begin{itemize}
@@
   %
-  Note that all rules are type-preserving with the exception of when
+  Note that all rules preserve partial types with the exception of when
   \forcesym is called on a hole, which case we instantiate the hole with
@@
 where $e$ is a term that causes the input program to get stuck.
+%
+The search is split into two phases: (1) we supply an increasing number
+of arguments until we find a saturated application, and (2) we search
+through the list of results from the non-deterministic evaluator for a
+witness.
 

### 2015-11-15 paper/pldi16-submission/interactive.tex
more interactive

@@
 template for the small-step relation is:
-$$
+\[
 \stepg{e_1}{\su_1}{\tr_1}{e_2}{\su_2}{\singlestep{e_1}{e_2}; \subterms{e_2}; \tr_1}
-$$
+\]
 
+\subsection{Traversing the Reduction Graph}
+\label{sec:traversing-graph}
+Given a reduction graph $G$ and an initial path
+$p = \singlestep{e_1}{\singlestep{e_2}{e_n}}$, such that
+\stepsg{e_1}{\emptysu}{\bullet}{e_n}{\su}{\tr}, we define the following
+traversals in Figure~\ref{fig:traversing-graph}:
+%
+\begin{itemize}
+\item \stepforwardsym takes a single step forward.
+\item \stepbackwardsym takes a single step backward.
+\item \jumpforwardsym takes a ``big'' step forward to the next function call.
+\item \jumpbackwardsym takes a ``big'' step backward to the previous function call.
+\item \stepintosym steps into a function call in a sub-term, isolating it from the context.
+\item \stepoversym steps over a function call in a sub-term.
+\end{itemize}
+%
+The initial path $p$ is required for the backward-steps as a node may
+have multiple incoming $\leadsto$ edges, \eg
+\singlestep{\eplus{1}{2}}{3} and \singlestep{\eplus{2}{1}}{3}.
+%
+The sub-term edges $\searrow$ allow us to decompose an expression into a
+sub-expression and the surrounding context, thus enabling the \stepintosym
+and \stepoversym traversals.
+
+\begin{figure*}[t]
+\centering
+\[
+\begin{array}{lcl}
+\stepforward{e_i}  &\defeq& {e_{i+1}} \\ \\
+\stepbackward{e_i} &\defeq& {e_{i-1}} \\ \\
+\jumpforward{e_i} &\defeq& \left\{\begin{array}{ll}
+                         e_{i+1}, & \text{if } e_{i+1} = \eapp{v_1}{v_2} \\
+                         \jumpforward{e_{i+1}}, & \text{otherwise}
+                         \end{array}\right\} \\ \\
+\jumpbackward{e_i} &\defeq& \left\{\begin{array}{ll}
+                         e_{i-1}, & \text{if } e_{i-1} = \eapp{v_1}{v_2} \\
+                         \jumpbackward{e_{i-1}}, & \text{otherwise}
+                         \end{array}\right\} \\ \\
+\stepinto{e_i} &\defeq& \left\{\begin{array}{ll}
+                         e\sub{x}{v_2}, & \text{if } e_i = C[\eapp{v_1}{v_2}] \text{ and } \singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}
+                         \end{array}\right\} \\ \\
+\stepover{e_i} &\defeq& \left\{\begin{array}{ll}
+                         C[v], & \text{if } e_i = C[\eapp{v_1}{v_2}] \text{ and } \multistep{\eapp{v_1}{v_2}}{v}
+                         \end{array}\right\}
+\end{array}
+\]
+\caption{Rules for traversing the reduction graph. \stepintosym and \stepoversym
+  require a traversal of the sub-term edges to decompose $e_i$ into the
+  target expression \eapp{v_1}{v_2} and the context $C$.}
+\label{fig:traversing-graph}
+\end{figure*}
 % \begin{itemize}

### 2015-11-16 paper/pldi16-submission/evaluation-nanomaly.tex
text about nanomaly and eval

\section{Evaluation: \nanomaly}
\label{sec:evaluation-nanomaly}
We have evaluated our prototype of \nanomaly in two phases:
%
\begin{enumerate}
\item We investigate the \emph{complexity} of the generated traces by various size metrics.
\item We investigate the \emph{usefulness} of the generated traces compared to the standard compiler errors by performing an A/B study.
\end{enumerate}

\subsection{Trace Complexity}
\label{sec:trace-complexity}
For each of the ill-typed programs from \S~\ref{sec:eval-witness} for
which we could find a witness, we measure the complexity of the
generated trace according to the following metrics:
%
\begin{enumerate}
\item The number of single-step reductions along the path from the
  witness to the stuck term. This can be thought of as a worst-case
  complexity, \ie ``How big is the fully-expanded trace?''
\item The number forward (or backward) jumps that can be taken along the
  path from the witness to the stuck term. These abstract many details
  of the computation that are often uninteresting or irrelevant to the
  explanation.
\item \ES{others?}
\end{enumerate}
%
\begin{figure}[t]
\centering
Have a graph!  
\caption{Complexity of the generated traces.}
\label{fig:results-complexity}
\end{figure}
%
The results of the experiment are summarized in
Figure~\ref{fig:results-complexity}.
%
The average number of single-step reductions is XX with a minimum of XX
and a maximum of XX. The average number of jumps is YY with a minimum of
YY and a maximium of YY.

\subsection{User Study}
\label{sec:user-study}
We also performed an A/B study at the University of Virginia (IRB
\#XXXX). 
%
Participants were recruited from the graduate-level Programming
Languages course CSYYY, with the possibility of winning a \$50 gift
certificate upon completion.
%
We selected a random sample of nine ill-typed programs from our
benchmark dataset and from submissions to our public-facing demo of
\nanomaly.
%
For each program, we presented participants with the program and
alternatingly, either \ocaml's type error message or our interactive
visualization of a preset trace.
%
In order to alleviate any potential bias in problem selection, we also
alternated whether participants began with \ocaml or \nanomaly.
%
Participants were asked to identify the \emph{source} of the error,
\emph{explain} the error, and finally \emph{fix} it.

We collected the time taken for each problem and recruited a colleague
to grade the submissions, hiding whether a given solution came from the
\ocaml or \nanomaly group.

\paragraph{Results}
\label{sec:study-results}
\ES{TODO!}

### 2015-11-16 paper/pldi16-submission/nanomaly.tex
text about nanomaly and eval

\section{Implementation: \nanomaly}
\label{sec:impl-nanomaly}
We have implemented our interactive semantics for the pure subset of
\ocaml with a web-based visualization of the generated traces in a tool
called \nanomaly.
%
It takes as input an \ocaml program and a function to check, searches
for a witness as described in \S~\ref{sec:searching-witness}, and
produces as output an interactive trace of the (possibly buggy)
execution~\footnote{If no witness could be found we simply return a trace
  of the last attempt.}.
%
We initially display a big-step reduction, \ie the source term followed
by the stuck term (or the final value).
%
The user can then progressively fill in the hidden steps of the
computation by selecting a visible term and choosing one of the
applicable traversal strategies -- described in
\S~\ref{sec:traversing-graph} -- to insert another term into the
visualization.

A sample interaction of the student program in
Figure~\ref{fig:palindrome} can be seen in
Figure~\ref{fig:nanomaly-palindrome}.
%
The initial state of the visualization tells us that the witness is
@palindrome []@, and that after some number of steps -- the thick arrow
denotes a multi-step transition -- we try to compare a function to a
list, which is not allowed in \ocaml.
%
Based on our semantics for \lang one would expect the final state to be
the \stuck term, but in practice we simply halt execution at the point
where an expression would transition to \stuck, as this final state is
more informative.
%
Note also that the equality test is colored black while the contextual
if-expression is faded out, this tells the user that the program got
stuck while trying to evaluate the sub-expression.

Upon seeing the stuck term, one might wonder where the @function@
came from. 
%
To investigate we select the stuck term and click the ``jump backward''
button to search backwards from the stuck term for the most recent
function call, which brings us to @listReverse [] = w@, in the same
context as before.
%
Uncontent with the explanation so far, we ``step forward'' twice from
the @listReverse []@ term, bringing us to @helper [] = w@.
%
At this point it is clear that the @helper@ function is not defined
correctly, we have supplied it with the single argument we expected and
yet it still returned a @function@.

The problem is that the @function@ keyword in \ocaml defines an
anonymous function that takes a single argument and immediately does a
case-analysis without giving the argument a name.
%
The solution is to replace @function@ with an explicit @match xs with@
-- naming the value we wish to case-analyse.
%
After applying our fix, \nanomaly -- and more importantly \ocaml --
decide that @listReverse@ is safe to run.
\ES{these last few paragraphs probably belong in the overview}
%
\begin{figure}[t]
\centering
\begin{code}
  let listReverse l =
    let rec helper xs = function 
      | [] -> xs 
      | hd::tl -> helper (hd :: xs) tl 
    in helper []

  let palindrome w =
    if listReverse w = w 
    then true 
    else false
\end{code}
\caption{Ill-typed \texttt{palindrome} function. The error lies in the
  use of the \texttt{function} construct which introduces an extra,
  undesired lambda abstraction, but manifests itself in the equality
  test on line XX. The fix is to replace \texttt{function} with
  \texttt{match xs with}. \ES{I think we should use this example in the
    intro/overview, leaving it here for now until they exist..}}
\label{fig:palindrome}
\end{figure}
%
\begin{figure}[t]
% \centering
Initial state: \\
\includegraphics[width=\linewidth]{palindrome} \\
After ``jumping backward'' from the stuck term: \\
\includegraphics[width=\linewidth]{palindrome2} \\
After ``stepping forward'' twice from @listReverse []@: \\
\includegraphics[width=\linewidth]{palindrome3} \\
\caption{Sample interaction}
\label{fig:nanomaly-palindrome}
\end{figure}

### 2015-11-16 paper/pldi16-submission/haskellListings.tex

\usepackage{listings}

\usepackage{color}
% uncomment next line to restore colors
% \def\withcolor{}

\ifdefined\withcolor
	\definecolor{haskellblue}{rgb}{0.0, 0.0, 1.0}
	\definecolor{haskellstr}{rgb}{0.2, 0.2, 0.6}
	\definecolor{haskellred}{rgb}{1.0, 0.0, 0.0}
	\definecolor{gray_ulisses}{gray}{0.55}
	\definecolor{castanho_ulisses}{rgb}{0.71,0.33,0.14}
	\definecolor{preto_ulisses}{rgb}{0.41,0.20,0.04}
	\definecolor{green_ulisses}{rgb}{0.0,0.4,0.0}
\else
	\definecolor{haskellblue}{gray}{0.1}
	\definecolor{haskellstr}{gray}{0.1}
	\definecolor{haskellred}{gray}{0.1}
	\definecolor{gray_ulisses}{gray}{0.1}
	\definecolor{castanho_ulisses}{gray}{0.1}
	\definecolor{preto_ulisses}{gray}{0.1}
	\definecolor{green_ulisses}{gray}{0.1}
\fi

\def\codesize{\normalsize}

\lstdefinelanguage{HaskellUlisses}{
	basicstyle=\codesize\ttfamily,
	sensitive=true,
	morecomment=[l][\color{gray_ulisses}\ttfamily\codesize]{--},
	morecomment=[s][\color{gray_ulisses}\ttfamily\codesize]{\{-}{-\}},
	morestring=[b]",
	stringstyle=\color{haskellstr},
	showstringspaces=false,
	numberstyle=\codesize,
	numberblanklines=true,
	showspaces=false,
	breaklines=true,
	showtabs=false,
        % mathescape=true,
	emph=
	{[1]
		FilePath,IOError,abs,acos,acosh,and,any,appendFile,approxRational,asTypeOf,asin,
		asinh,atan,atan2,atanh,basicIORun,break,catch,ceiling,chr,compare,concat,concatMap,
		const,cos,cosh,curry,cycle,decodeFloat,denominator,digitToInt,div,divMod,drop,
		dropWhile,either,elem,encodeFloat,enumFrom,enumFromThen,enumFromThenTo,enumFromTo,
		error,even,exp,exponent,fail,filter,flip,floatDigits,floatRadix,floatRange,floor,
		fmap,foldl,foldl1,foldr,foldr1,fromDouble,fromEnum,fromInt,fromInteger,
		fromRational,fst,gcd,getChar,getContents,getLine,head,id,inRange,index,init,intToDigit,
		interact,ioError,isAlpha,isAlphaNum,isAscii,isControl,isDenormalized,isDigit,isHexDigit,
		isIEEE,isInfinite,isLower,isNaN,isNegativeZero,isOctDigit,isPrint,isSpace,isUpper,iterate,
		last,lcm,length,lex,lexDigits,lexLitChar,lines,log,logBase,lookup,map,mapM,mapM_,max,
		maxBound,maximum,maybe,min,minBound,minimum,mod,negate,not,notElem,numerator,odd,
		or,pi,primExitWith,print,product,properFraction,putChar,putStr,putStrLn,quot,
		quotRem,range,rangeSize,read,readDec,readFile,readFloat,readHex,readIO,readInt,readList,readLitChar,
		readLn,readOct,readParen,readSigned,reads,readsPrec,realToFrac,recip,rem,repeat,replicate,
		reverse,round,scaleFloat,scanl,scanl1,scanr,scanr1,seq,sequence,sequence_,show,showChar,showInt,
		showList,showLitChar,showParen,showSigned,showString,shows,showsPrec,significand,signum,sin,
		sinh,snd,span,splitAt,sqrt,subtract,succ,sum,tail,take,takeWhile,tan,tanh,threadToIOResult,toEnum,
		toInt,toInteger,toLower,toRational,toUpper,truncate,uncurry,undefined,unlines,until,unwords,unzip,
		unzip3,userError,words,writeFile,zip,zip3,zipWith,zipWith3,listArray,doParse,for,initTo,
                create,get,set,div,rescale,add,delete,insert,prop_focus_left_master,average,best,insert,union,split,size,fromList,copy,group,good,bad,foo,explode,singleton,difference,fromJust,sort,unfold,
                target, query, decode, encode, check, refuteSMT, binder,
                subst,unapply,apply,proxy,refinement,fresh,guard,constrain,oneOf, 
                queryList,queryCtor,queryField,ctors,decodeCtor,whichOf,ctorArity,eval,
                mkCtor,gCtors,gEncode,gEncodeFields,gDecode,gDecodeFields,reproxyRep,empty,splitCtor,checkField,scanM, 
                padAverage,focusUp,execute,checkSMT,inputTypes,outputType,toReft,app
                %, inTypes, inputTypes, outputType, execute, smtFindModel, smtRefuteModel 
	},
	emphstyle={[1]\color{haskellblue}},
	emph=
	{[2]
		OkMap,OkRBT,OkStackSet,TTrue,Map,Bool,Char,Double,Either,Float,IO,Integer,Int,Maybe,Ordering,Rational,Ratio,ReadS,ShowS,String,Word8,Nat,Pos,Rng,Score,
                Ptr,ForeignPtr,CSize,InPacket,Tree,Prop,TreeEq,TreeLt,Vec,
                NullTerm,IncrList,DecrList,UniqList,BST,MinHeap,MaxHeap,
                PtrN,ByteStringN,ByteStringEq,VO,ByteStringsEq,ByteStringNE,OrdList,Var,RType,Constrain,Gen,Var,Proxy,SMT,Targetable,RefType,Refinement,Ctor,C1,Rep,Rec0,U1,
                GCtors,GDecode,GDecodeFields,GEncode,GEncodeFields,OrdMap,MinusKey,
                len,isBH,isBal,bh,isRB,keys,List,Sorted,RBT,Col,isBlack,OrdRBT,Set,sz,
                StackSet,NoDuplicates,Data,RBTree,XMonad,Generic,true
	},
	emphstyle={[2]\color{castanho_ulisses}},
	emph=
	{[3]
		case,class,data,deriving,do,else,if,return,def,import,in,infixl,infixr,instance,let,tmapM,for2M,forM,zipWithM,otherwise,
		module,measure,pred,predicate,of,primitive,then,type,where,lazy,throw,when
	},
	emphstyle={[3]\color{preto_ulisses}\textbf},
	emph=
	{[4]
		quot,rem,div,mod,elem,notElem,seq
	},
	emphstyle={[4]\color{castanho_ulisses}\textbf},
	emph=
	{[5]
		PS,Tip,Node,Black,Red,EQ,False,GT,Just,LT,Left,Nothing,Right,True,Show,Eq,Ord,Num,C,N,Leaf,Bin,CounterExample
	},
	emphstyle={[5]\color{green_ulisses}},
	emph=
	{[6]
		patError, irrefutPatError, nonExhaustiveGuardsError, recSelError, errorOut, 
		noMethodBinding
	},
	emphstyle={[6]\color{haskellred}}
}

%%%ORIG
%%%\lstnewenvironment{code}
%%%{\textbf{Haskell Code} \hspace{1cm} \hrulefill \lstset{language=HaskellUlisses}}
%%%{\hrule\smallskip}

%V1
%\lstnewenvironment{code}
%{\smallskip \lstset{language=HaskellUlisses}}
%{\smallskip}

\lstdefinelanguage{HaskellUlissesMath}[]{HaskellUlisses}{mathescape=true}

\lstnewenvironment{code}
{\lstset{language=HaskellUlisses}}
{}

\lstnewenvironment{mcode}
{\lstset{language=HaskellUlissesMath}}
{}

\lstnewenvironment{fcode}
{\lstset{language=HaskellUlisses, frame=L}}
{}

\lstMakeShortInline[language=HaskellUlisses]@
\lstMakeShortInline[language=HaskellUlissesMath]|

### 2015-11-16 paper/pldi16-submission/interactive.tex
add syntax for substitutions

@@
   {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
-         {e\sub{x}{v_2}}{\su_1\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
+         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
 \\ \\
@@
   {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}}
-  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
+  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
 \end{gather*}
@@
 and \stepoversym traversals.
+%
 
@@
   require a traversal of the sub-term edges to decompose $e_i$ into the
-  target expression \eapp{v_1}{v_2} and the context $C$.}
+  target expression \eapp{v_1}{v_2} and the context $C$.
+  \ES{need to mention $G$ and $p$ in these rules}
+}
 \label{fig:traversing-graph}
 \end{figure*}
+
 % \begin{itemize}

### 2015-11-17 paper/pldi16-submission/fac.py
intro

@@ -1,0 +1,5 @@
+def factorial(n):
+  if n <= 0:
+    return True
+  else:
+    return n * factorial(n - 1)

### 2015-11-17 paper/pldi16-submission/main.tex
start interactive semantics
ugh traces
text about nanomaly and eval
intro
tweak

@@
 %
-\documentclass[10pt,pldi,indentedstyle=false]{sigplanconf-pldi15}
+\documentclass[10pt,pldi,nocopyrightspace,indentedstyle=false]{sigplanconf-pldi15}
 
@@
 \usepackage{commands}
+\usepackage{graphicx}
 
@@
 
-\section{``Interactive'' Semantics \& Trace Exploration} % 2 pages
-\begin{itemize}
-\item extend operational semantics to collect reduction graph
-\item nodes are terms, edges indicate ``steps-to'' and ``sub-term'' relations
-\item visualize path through reduction graph
-\item expand edges to reveal more fine-grained steps (step/jump forward/backward)
-\item never lose context (unlike traditional debugger)
-\end{itemize}
+%\section{``Interactive'' Semantics \& Trace Exploration} % 2 pages
+\input{interactive.tex}
 
-\section{Evaluation: NanoMaLy}                % 1 page
+\input{nanomaly.tex}
+
+% \section{Evaluation: \nanomaly}                % 1 page
+\input{evaluation-nanomaly.tex}
+
 \section{Related Work}              % 1 pages

### 2015-11-17 paper/pldi16-submission/semantics.tex
add syntax for substitutions
tweak

@@
-\section{Searching for Type-Error Witnesses}
+\section{Type-Error Witnesses}
 \label{sec:searching-witness}
+
+Our goal is to find concrete values that demonstrate how a
+program ``goes wrong''.
+
+\paragraph{Problem: Which inputs are bad?}
+%
+One approach is to randomly generate input values and use
+them to execute the program until we find one that causes
+the program to go wrong. However, to see why this approach
+is naive, consider the following example:
+%
+\begin{lstlisting}
+  let f x =
+    let y = 1 + x in
+      1. +. y
+\end{lstlisting}
+%
+What \emph{types} of inputs should we test \texttt{f} with?
+Values of type \texttt{int} and \texttt{float} are fair game,
+but values of type say, \texttt{string} or \texttt{int list}
+will cause the program to go wrong in an \emph{irrelevant}
+manner.
+
+\paragraph{Solution:} \rj{STOP}
+
+% we cannot provide \emph{completely arbitrary} inputs to
+% \texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
+% \ehole{}, which is a placeholder for a value whose type we have not
+% yet determined. As we execute the program, we instantiate holes with
+% concrete values as demanded by the primitive operations in the
+% program. For example, the hole we pass to f will be instantiated to an
+% int when we reach the \lstinline{1 + x} term. Thus, y will be an int as
+% well, and the program will get stuck at \lstinline{1. +. y}. \ES{this
+%   reads more like overview text..}
+%
+
 % \begin{itemize}
@@
 % \end{itemize}
+
+Next, we formalize the notion of type error witnesses as follows.
 %
-Next, we formalize our search for witnesses to type-errors.
+First, we define a core calculus within which we will work~\S~\ref{sec:syntax}.
 %
-We present the syntax and operational semantics of \lang -- a simple
-lambda calculus with integers and booleans, extended with our notion of
-holes -- as well as our search algorithm.
+Second, we develop a (non-deterministic) operational semantics
+for ill-typed programs that precisely defines the notion
+of a \emph{witness}~\S~\ref{sec:semantics},
 %
-We prove that our system \emph{soundly} finds witnesses, \ie if we find
-a witness then there is no possible typing for the input program.
+Third, we formalize and prove a notion of \emph{generality} for
+witnesses, which states, intuitively that if we find a
+single witness then for \emph{every possible} type
+assignment, there exist inputs that are guaranteed to make
+the program ``go wrong''~\S~\ref{sec:soundness}, and finally,
 %
+Fourth, we refine the operational semantics into a
+\emph{search procedure} that returns concrete (general)
+witnesses for ill-typed programs~\S~\ref{sec:search-algorithm}.
+
 \subsection{Syntax}
@@
 Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
-calculus with numbers and booleans.
+calculus with integers and booleans.
 %
-As we are specifically interested in
-programs that \emph{do} go wrong, we include an explicit \stuck state in
-our syntax.
-%
+As we are specifically interested in programs that \emph{do} go wrong,
+we include an explicit \stuck\ state in our syntax.
+
 \paragraph{Holes}
 \label{sec:holes}
-The main novelty in our system is the notion of a ``hole'', written
-\ehole{}, which represents an uninstantiated value.
 %
-Importantly, we do not even know what type the value should have.
+Recall that a key challenge in our setting is to find witnesses
+that are meaningful and do not arise from choosing values from
+irrelevant types.
 %
-Holes may also appear in types, where they may be thought of as type
-variables that we will not generalize over.
+We solve this problem be equipping our term language with a notion
+of a \emph{hole}, written \ehole{}, which represent \emph{unconstrained}
+values that may replaced with values from \emph{any} type.
 %
+Dually, holes may also appear in types, where they may be thought
+of as type variables that we will not generalize over.
+%
+
+\paragraph{Substitutions}
+%
+Our semantics ensure the generality of witnesses by incrementally
+\emph{refining} holes filling in just as much information as is
+needed locally to make progress (inspired by the manner in
+which SmallCheck uses lazy evaluation~\cite{smallcheck}).
+%
+We track how the holes are incrementally filled in, by using
+(value) \emph{substitutions} that map holes (denoted by their index)
+to values which may, in turn, contain holes.
+%
+
 \subsection{Semantics}
@@
 %
+A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
+and a substitution $\vsu$.
+%
+We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
+transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
+%
 We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
-$e$ is $e_1$, $e'$ is $e_j$ and $\forall i,j, 1 \leq i < j$, we have
+$e$ is $e_0$, $e'$ is $e_j$ and $\forall 0 \leq i < j$, we have
 \step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
@@
 $\stepi{j}{e}{\su}{e'}{\su'}$.
-%
+
 The evaluation relation is parameterized by a pair of \forcesym and
@@
   $\exists \su.\ t = \subst{\su}{s} \lor s = \subst{\su}{t}$.
+  \ES{we're abusing the \su notation here..}
 \end{defn}
@@
 %   Output: 'safe' or 'fn v1 .. vn' where 'fn v1 .. vn' gets stuck
+
+% !TEX root = main.tex

### 2015-11-17 paper/pldi16-submission/introduction.tex

\section{Introduction}
\label{sec:introduction}

% \paragraph{Problem}
Type errors are a common stumbling block for students trying to learn
typed functional languages like \ocaml and \haskell.
%
The compiler imposes a strict discipline that prevents novices from
running their programs until the type-checker has been appeased.
%
In effect this prioritizes the \emph{static} semantics of the language
over the \emph{dynamic} semantics.
%
This is fine for expert users who are already familiar with the language
and type system, but seems backwards for novices, who are typically
introduced to the dynamic semantics first.

Consider the following ill-typed @factorial@ function, where we return
@true@ in the base case instead of @1@.
%
\begin{code}
  let rec factorial n =
    if n <= 0 then
      true
    else
      n * factorial (n-1)
\end{code}
%
\ocaml's response is to tell us that on line 5, the recursive call to
@factorial@ returns a @bool@, but we needed an @int@ for the
multiplication.
%
\begin{verbatim}
File "fac.ml", line 5, characters 8-23:
Error: This expression has type bool
       but an expression was expected
       of type int
\end{verbatim}
%
\ocaml is correct, but we may ask \emph{why} @factorial@ returns a
@bool@ instead of an @int@.

Contrast this \emph{static} type error to the \emph{dynamic} type error
that arises in the equivalent \ruby function, when we call
@factorial(1)@.
%
\begin{verbatim}
fac.rb:5:in `*': true can't be coerced
                 into Fixnum (TypeError)
  from fac.rb:5:in `factorial'
  from fac.rb:10:in `<main>'
\end{verbatim}
%
The same basic information is there, @*@ was given a @bool@ (@true@)
when it expected an @int@ (@FixNum@), but since the error occurred at
runtime, \ruby is able to provide a stack trace that provides important
contextual information about the error.
%
More importantly, being able to \emph{run} the bad program means we can
also test it on other inputs, \eg @0@, or we can load it into a debugger
and step through the execution to see concretely how the error arises.

To make matters worse, it is well known that type inference algorithms
based on unification have a penchant for \emph{reporting} errors far
from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much work has been done on pinpointing the \emph{source} of a type
error~\cite{zhang-myers, maxsmt, counterfactual}, but an accurate source
location still does not explain \emph{why} the program is wrong.

% \paragraph{Contributions}
In this paper we propose exposing \emph{static} type errors for the
\emph{runtime} crashes that they are.
%
Our first contribution is an algorithm for searching for
\emph{witnesses} to type errors, \ie input vectors that cause a program
to crash with a type error.
%
This is not as trivial as it sounds.
%
While one can indeed execute an \ocaml program without type-checking it
-- the dynamic semantics do not depend on the static semantics -- what
types of values should one pass to the program?
%
We do not want to generate \emph{spurious} witnesses, inputs that the
type-checker would have excluded, so we introduce a semantics for
executing \ocaml programs with ``holes'', values whose type we do not
yet know.
%
% A hole can be passed around through the program until it is demanded by
% a primitive operation, \eg addition, function application, branching,
% etc., at which point we know from the context what type it must have,
% and can instantiate it with a concrete value.
%
Our semantics \emph{conservatively} instantiates holes with concrete
values, if we can find a witness to a type error then there is no valid
typing for the program.

Given a witness to a type error, the prospective \ocaml student may
still be at a loss.
%
The standard \ocaml interpreter and debugging infrastructure expect
well-typed programs, so they cannot be used to investigate \emph{how}
the witness causes the program to crash.
%
Thus, our second contribution is an interactive visualization of the
execution of \ocaml programs (well-typed or not).
%
We present the execution as a \emph{reduction graph}, in which the nodes
are expressions and the edges indicate either a single-step reduction or
a sub-term relation.
%
This graph is easy to generate during the execution and admits many of
the standard debugging steps, \eg ``step forward'', ``step into'', and
``step over'', as well as backwards motion and jumping to the next (or
previous) function call.
%
The visualization initially shows the source term (\ie the witness) and
the stuck term (or the final value), and allows the novice to
iteratively expand the computation to show intermediate steps by
selecting an expression and choosing an strategy for picking the next
node to insert into the graph.
%
This process of iterative expansion of the trace ensures that the
context of intermediate steps is not lost; whereas traditional debuggers
force one to move between statements, in our visualization you never
lose sight of where you came from.

\paragraph{Outline}
We begin the rest of the paper with an overview of our search algorithm
and trace visualization (\S~\ref{sec:overview}).
%
Then, we formalize our algorithm and prove it sound
(\S~\ref{sec:searching-witness}), and evaluate it on a large set of real
student programs (\S~\ref{sec:eval-witness}).
%
Next, we extend the search algorithm to collect the reduction graph and
formalize a set of traversals (\S~\ref{sec:explaining}), and measure the
complexity of these traces on our benchmarks
(\S~\ref{sec:evaluation-nanomaly}).
%
Finally, we present the results of a preliminary user study on the full
system (\S~\ref{sec:user-study}).

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2015-11-17 paper/pldi16-submission/overview.tex

--- 2015-11-17 paper/pldi16-submission/introduction.tex
@@
-\section{Introduction}
-\label{sec:introduction}
-
-% \paragraph{Problem}
-Type errors are a common stumbling block for students trying to learn
-typed functional languages like \ocaml and \haskell.
-%
-The compiler imposes a strict discipline that prevents novices from
-running their programs until the type-checker has been appeased.
-%
-In effect this prioritizes the \emph{static} semantics of the language
-over the \emph{dynamic} semantics.
-%
-This is fine for expert users who are already familiar with the language
-and type system, but seems backwards for novices, who are typically
-introduced to the dynamic semantics first.
-
-Consider the following ill-typed @factorial@ function, where we return
-@true@ in the base case instead of @1@.
-%
-\begin{code}
-  let rec factorial n =
-    if n <= 0 then
-      true
-    else
-      n * factorial (n-1)
-\end{code}
-%
-\ocaml's response is to tell us that on line 5, the recursive call to
-@factorial@ returns a @bool@, but we needed an @int@ for the
-multiplication.
-%
-\begin{verbatim}
-File "fac.ml", line 5, characters 8-23:
-Error: This expression has type bool
-       but an expression was expected
-       of type int
-\end{verbatim}
-%
-\ocaml is correct, but we may ask \emph{why} @factorial@ returns a
-@bool@ instead of an @int@.
-
-Contrast this \emph{static} type error to the \emph{dynamic} type error
-that arises in the equivalent \ruby function, when we call
-@factorial(1)@.
-%
-\begin{verbatim}
-fac.rb:5:in `*': true can't be coerced
-                 into Fixnum (TypeError)
-  from fac.rb:5:in `factorial'
-  from fac.rb:10:in `<main>'
-\end{verbatim}
-%
-The same basic information is there, @*@ was given a @bool@ (@true@)
-when it expected an @int@ (@FixNum@), but since the error occurred at
-runtime, \ruby is able to provide a stack trace that provides important
-contextual information about the error.
-%
-More importantly, being able to \emph{run} the bad program means we can
-also test it on other inputs, \eg @0@, or we can load it into a debugger
-and step through the execution to see concretely how the error arises.
-
-To make matters worse, it is well known that type inference algorithms
-based on unification have a penchant for \emph{reporting} errors far
-from their source.
-%
-This further increases the novice's confusion and can actively mislead
-them to focus their investigation on an irrelevant piece of code.
-%
-Much work has been done on pinpointing the \emph{source} of a type
-error~\cite{zhang-myers, maxsmt, counterfactual}, but an accurate source
-location still does not explain \emph{why} the program is wrong.
-
-% \paragraph{Contributions}
-In this paper we propose exposing \emph{static} type errors for the
-\emph{runtime} crashes that they are.
-%
-Our first contribution is an algorithm for searching for
-\emph{witnesses} to type errors, \ie input vectors that cause a program
-to crash with a type error.
-%
-This is not as trivial as it sounds.
-%
-While one can indeed execute an \ocaml program without type-checking it
--- the dynamic semantics do not depend on the static semantics -- what
-types of values should one pass to the program?
-%
-We do not want to generate \emph{spurious} witnesses, inputs that the
-type-checker would have excluded, so we introduce a semantics for
-executing \ocaml programs with ``holes'', values whose type we do not
-yet know.
-%
-% A hole can be passed around through the program until it is demanded by
-% a primitive operation, \eg addition, function application, branching,
-% etc., at which point we know from the context what type it must have,
-% and can instantiate it with a concrete value.
-%
-Our semantics \emph{conservatively} instantiates holes with concrete
-values, if we can find a witness to a type error then there is no valid
-typing for the program.
-
-Given a witness to a type error, the prospective \ocaml student may
-still be at a loss.
-%
-The standard \ocaml interpreter and debugging infrastructure expect
-well-typed programs, so they cannot be used to investigate \emph{how}
-the witness causes the program to crash.
-%
-Thus, our second contribution is an interactive visualization of the
-execution of \ocaml programs (well-typed or not).
-%
-We present the execution as a \emph{reduction graph}, in which the nodes
-are expressions and the edges indicate either a single-step reduction or
-a sub-term relation.
-%
-This graph is easy to generate during the execution and admits many of
-the standard debugging steps, \eg ``step forward'', ``step into'', and
-``step over'', as well as backwards motion and jumping to the next (or
-previous) function call.
-%
-The visualization initially shows the source term (\ie the witness) and
-the stuck term (or the final value), and allows the novice to
-iteratively expand the computation to show intermediate steps by
-selecting an expression and choosing an strategy for picking the next
-node to insert into the graph.
-%
-This process of iterative expansion of the trace ensures that the
-context of intermediate steps is not lost; whereas traditional debuggers
-force one to move between statements, in our visualization you never
-lose sight of where you came from.
-
-\paragraph{Outline}
-We begin the rest of the paper with an overview of our search algorithm
-and trace visualization (\S~\ref{sec:overview}).
-%
-Then, we formalize our algorithm and prove it sound
-(\S~\ref{sec:searching-witness}), and evaluate it on a large set of real
-student programs (\S~\ref{sec:eval-witness}).
-%
-Next, we extend the search algorithm to collect the reduction graph and
-formalize a set of traversals (\S~\ref{sec:explaining}), and measure the
-complexity of these traces on our benchmarks
-(\S~\ref{sec:evaluation-nanomaly}).
-%
-Finally, we present the results of a preliminary user study on the full
-system (\S~\ref{sec:user-study}).
+\section{Overview}
+\label{sec:overview}
 

### 2015-11-17 paper/pldi16-submission/semantics.tex

@@
 
-\paragraph{Solution:} \rj{STOP}
+\paragraph{Solution:} \RJ{STOP}
 
@@
 %
+The substitutions let us ensure that we instantiate each hole
+with the same (partially defined) value, regardless of the multiple
+contexts in which the hole appears, and hence, ensures we can
+report a concrete (and general) witness for any (dynamically)
+discovered type errors.
 
@@
 $\stepi{j}{e}{\su}{e'}{\su'}$.
+%
+The evaluation relation is parameterized by a pair of functions:
+called \emph{narrow} (\forcesym) and \emph{generate} (\gensym),
+that ``dynamically'' perform type-checking and hole-filling
+respectively.
 
-The evaluation relation is parameterized by a pair of \forcesym and
-\gensym functions that handle runtime type-checking and instantiation of
-holes respectively.
+\paragraph{Narrowing Types} The function $\force{e}{t}$ takes as
+input an term \texttt{e} and refines it to have the type \texttt{t}
+by yielding a pair of either the same expression and a value substitution
+the refines the holes in \texttt{e} to ensure it has type \texttt{t},
+or yields the stuck term if no such refinement is possible:
 %
-The relation must also maintain a substitution \su, mapping holes to
-concrete values, so that we do not instantiate the same hole with
-different values in different contexts, and so that we can report a
-concrete witness to any discovered type errors.
+$$
+\begin{array}{lcl}
+\forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
+\force{\ehole{i}}{t} & \defeq & \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}} \\
+\force{v}{\ehole{}}  & \defeq & \pair{v}{\emptysu} \\
+\force{n}{\tint}     & \defeq & \pair{n}{\emptysu} \\
+\force{v}{\tint}     & \defeq & \pair{\stuck}{\emptysu} \\
+\force{b}{\tbool}    & \defeq & \pair{b}{\emptysu} \\
+\force{v}{\tbool}    & \defeq & \pair{\stuck}{\emptysu} \\
+\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
+\force{v}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\stuck}{\emptysu} \\
+\end{array}
+$$
+
+\paragraph{Generating Values} The function $\gen{t}$ takes as input a type
+\texttt{t} and returns \RJ{HEREHERE}
+
+$$
+\begin{array}{lcll}
+\gensym       & ::      & t \rightarrow v \\
+\gen{\tint}   & \defeq  & n & \quad \text{non-deterministic} \\
+\gen{\tbool}  & \defeq  & b & \quad \text{non-deterministic} \\
+\gen{\tfun{t_1}{t_2}}   & \defeq & \efun{x}{\ehole{i}}, & \quad \text{$i$ is fresh} \\
+\gen{\thole{}} & \defeq & \ehole{i}, & \quad \text{$i$ is fresh} \\
+\end{array}
+$$
+
+%
 
@@
 \end{thm}
+
+%
 % \begin{proof}%[Proof of \autoref{thm:all-reduce}]

### 2015-11-17 paper/pldi16-submission/plots.py
add some graphs, finally..

@@ -1,0 +1,75 @@
+import csv
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+
+BUCKETS = range(500, 3001, 500)
+COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
+
+SAFE = ['S', 'T']
+SAFE_L = ['Safe', 'Timeout']
+UNSAFE = ['U', 'B', 'D'] #, 'O']
+UNSAFE_L = ['Unsafe', 'Unbound', 'Diverge'] #, 'Output']
+ALL = UNSAFE + SAFE
+ALL_L = UNSAFE_L + SAFE_L
+
+def read_csv(f):
+    with open(f) as f:
+        return list(csv.reader(f))
+
+def cumulative_coverage(data):
+    headers = data[0]
+    data = data[1:]
+    return [(l, len([r for r in data
+                     if int(r[1]) <= l
+                     and r[4] in UNSAFE])
+             / float(len(data)))
+            for l in BUCKETS]
+
+def plot_coverage(data):
+    xy = cumulative_coverage(data)
+
+    N = len(xy)
+    ind = np.arange(N)    # the x locations for the groups
+    width = 0.5       # the width of the bars: can also be len(x) sequence
+
+    p1 = plt.bar(ind, [r[1] for r in xy], width,
+                 color=COLORS[0])
+
+    plt.xlabel('Timeout (steps)')
+    plt.ylabel('% witnesses found')
+    plt.title('Cumulative Coverage')
+    plt.xticks(ind + width/2.0, [r[0] for r in xy])
+    plt.yticks(np.arange(0.0, 1.1, 0.1))
+    plt.legend((p1[0],), ('Seminal',))
+    # plt.legend((p1[0], p2[0]), ('Men', 'Women'))
+
+    plt.show()
+
+def plot_distrib(data, label):
+    data = data[1:]
+    rs = [len([r for r in data if r[4] == o])
+          for o in ALL]
+
+    # N = len(xy)
+    # ind = np.arange(N)    # the x locations for the groups
+    # width = 0.5       # the width of the bars: can also be len(x) sequence
+
+    plt.axes(aspect=1)
+    p1 = plt.pie(rs, labels=ALL_L,
+                 autopct='%.1f%%',
+                 colors=COLORS,
+                 shadow=True)
+
[…]

### 2015-11-17 paper/pldi16-submission/plots.py
trace size graph for seminal

@@ -46,6 +46,44 @@ def plot_coverage(data):
 
     plt.show()
 
+def plot_trace_size(data, label):
+    # xy = cumulative_coverage(data)
+
+    # N = len(xy)
+    # ind = np.arange(N)    # the x locations for the groups
+    # width = 0.5       # the width of the bars: can also be len(x) sequence
+
+    # p1 = plt.bar(ind, [r[1] for r in xy], width,
+    #              color=COLORS[0])
+
+    step = [int(r[5]) for r in data[1:] if r[4] in UNSAFE and int(r[5]) > 0]
+    jump = [int(r[6]) for r in data[1:] if r[4] in UNSAFE and int(r[6]) > 0]
+    
+    bins = [0, 5, 10, 20, 50, 100, 1000]
+    binlabels = ['<5', '5-9', '10-19', '20-49', '50-99', '>=100']
+    ind = np.arange(1, len(bins))
+    width = 0.35
+
+    fig, ax = plt.subplots()
+
+    y,binEdges=np.histogram(step,bins=bins)
+    p1 = ax.bar(ind, y, label='Steps', width=width, color=COLORS[0])
+
+    y,binEdges=np.histogram(jump,bins=bins)
+    p2 = ax.bar(ind+width, y, label='Jumps', width=width, color=COLORS[1])
+    # p1 = plt.bar(0.5*(binEdges[1:]+binEdges[:-1]), y, label='Steps')
+    # p1 = plt.hist([step,jump], bins=bins, label=['Steps', 'Jumps'], range=(0,300), color=COLORS[:2])
+
+    # plt.xlabel('Size')
+    plt.ylabel('Traces')
+    plt.title('Size of generated traces (%s)' % label)
+    plt.xticks(ind + width, binlabels)
+    # plt.yticks(np.arange(0.0, 1.1, 0.1))
+    # plt.legend((p1[0],), ('Seminal',))
+    plt.legend((p1[0], p2[0]), ('Steps', 'Jumps'))
+
+    plt.show()
+
 def plot_distrib(data, label):
     data = data[1:]
     rs = [len([r for r in data if r[4] == o])
@@ -58,6 +96,8 @@ def plot_distrib(data, label):
     plt.axes(aspect=1)
     p1 = plt.pie(rs, labels=ALL_L,
                  autopct='%.1f%%',
+                 pctdistance=1.15,
+                 labeldistance=10,
                  colors=COLORS,
[…]

### 2015-11-17 paper/pldi16-submission/plots.py
dpi

@@ -33,6 +33,7 @@ def plot_coverage(data):
     ind = np.arange(N)    # the x locations for the groups
     width = 0.5       # the width of the bars: can also be len(x) sequence
 
+    fig = plt.figure()
     p1 = plt.bar(ind, [r[1] for r in xy], width,
                  color=COLORS[0])
 
@@ -44,7 +45,8 @@ def plot_coverage(data):
     plt.legend((p1[0],), ('Seminal',))
     # plt.legend((p1[0], p2[0]), ('Men', 'Women'))
 
-    plt.show()
+    # plt.show()
+    fig.savefig('coverage.png', dpi=300)
 
 def plot_trace_size(data, label):
     # xy = cumulative_coverage(data)
@@ -82,7 +84,8 @@ def plot_trace_size(data, label):
     # plt.legend((p1[0],), ('Seminal',))
     plt.legend((p1[0], p2[0]), ('Steps', 'Jumps'))
 
-    plt.show()
+    # plt.show()
+    fig.savefig('trace_size_%s.png' % label.lower(), dpi=300)
 
 def plot_distrib(data, label):
     data = data[1:]
@@ -92,7 +95,7 @@ def plot_distrib(data, label):
     # N = len(xy)
     # ind = np.arange(N)    # the x locations for the groups
     # width = 0.5       # the width of the bars: can also be len(x) sequence
-
+    fig = plt.figure()
     plt.axes(aspect=1)
     p1 = plt.pie(rs, labels=ALL_L,
                  autopct='%.1f%%',
@@ -112,5 +115,6 @@ def plot_distrib(data, label):
     # plt.legend((p1[0],), ('Seminal',))
     # plt.legend((p1[0], p2[0]), ('Men', 'Women'))
 
-    plt.show()
+    # plt.show()
+    fig.savefig('distrib_%s.png' % label.lower(), dpi=300)
 

### 2015-11-18 paper/pldi16-submission/plots.py
embed graphs as pdfs for better scaling

@@ -17,13 +17,21 @@ def read_csv(f):
     with open(f) as f:
         return list(csv.reader(f))
 
+def autolabel(ax, rects):
+    # attach some text labels
+    for rect in rects:
+        height = rect.get_height()
+        ax.text(rect.get_x() + rect.get_width()/2., height,
+                '%d' % int(height),
+                ha='center', va='bottom')
+
 def cumulative_coverage(data):
     headers = data[0]
     data = data[1:]
-    return [(l, len([r for r in data
-                     if int(r[1]) <= l
-                     and r[4] in UNSAFE])
-             / float(len(data)))
+    return [(l, int(100 * len([r for r in data
+                               if int(r[1]) <= l
+                               and r[4] in UNSAFE])
+                          / float(len(data))))
             for l in BUCKETS]
 
 def plot_coverage(data):
@@ -41,14 +49,15 @@ def plot_coverage(data):
     plt.ylabel('% witnesses found')
     plt.title('Cumulative Coverage')
     plt.xticks(ind + width/2.0, [r[0] for r in xy])
-    plt.yticks(np.arange(0.0, 1.1, 0.1))
+    plt.yticks(np.arange(0, 101, 10))
     plt.legend((p1[0],), ('Seminal',))
     # plt.legend((p1[0], p2[0]), ('Men', 'Women'))
+    autolabel(plt, p1)
 
     # plt.show()
-    fig.savefig('coverage.png', dpi=300)
+    fig.savefig('coverage.pdf', dpi=300)
 
-def plot_trace_size(data, label):
+def plot_trace_size(data, label, steporjump):
     # xy = cumulative_coverage(data)
 
     # N = len(xy)
@@ -63,29 +72,45 @@ def plot_trace_size(data, label):
     
     bins = [0, 5, 10, 20, 50, 100, 1000]
     binlabels = ['<5', '5-9', '10-19', '20-49', '50-99', '>=100']
-    ind = np.arange(1, len(bins))
-    width = 0.35
+    ind = np.arange(0, len(binlabels))
+    width = 0.5
+    # plt.figure(figsize=(10,5))
 
-    fig, ax = plt.subplots()
+    # fig, ax = plt.subplots()
+    ax = plt.subplot(2,1,1)
[…]

### 2015-11-18 paper/pldi16-submission/interactive.tex

@@
 \begin{array}{lcl}
-\stepforward{e_i}  &\defeq& {e_{i+1}} \\ \\
-\stepbackward{e_i} &\defeq& {e_{i-1}} \\ \\
-\jumpforward{e_i} &\defeq& \left\{\begin{array}{ll}
-                         e_{i+1}, & \text{if } e_{i+1} = \eapp{v_1}{v_2} \\
-                         \jumpforward{e_{i+1}}, & \text{otherwise}
+\stepforward{G}{p}{e_i}  &\defeq& \left\{\begin{array}{ll}
+    e_j, & \text{where } \singlestep{e_i}{e_j} \in G
                          \end{array}\right\} \\ \\
-\jumpbackward{e_i} &\defeq& \left\{\begin{array}{ll}
-                         e_{i-1}, & \text{if } e_{i-1} = \eapp{v_1}{v_2} \\
-                         \jumpbackward{e_{i-1}}, & \text{otherwise}
+\stepbackward{G}{p}{e_i}  &\defeq& \left\{\begin{array}{ll}
+    e_j, & \text{where } \singlestep{e_j}{e_i} \in G \text{ and } e_j \in p
                          \end{array}\right\} \\ \\
-\stepinto{e_i} &\defeq& \left\{\begin{array}{ll}
+\jumpforward{G}{p}{e_i} &\defeq& \text{let } e_j = \stepforward{G}{p}{e_i} \text{ in }
+                         \left\{\begin{array}{ll}
+                         e_j, & \text{if } e_j = \eapp{v_1}{v_2} \\
+                         \jumpforward{G}{p}{e_{j}}, & \text{otherwise}
+                         \end{array}\right\} \\ \\
+\jumpbackward{G}{p}{e_i} &\defeq& \text{let } e_j = \stepbackward{G}{p}{e_i} \text{ in }
+                         \left\{\begin{array}{ll}
+                         e_j, & \text{if } e_j = \eapp{v_1}{v_2} \\
+                         \jumpbackward{G}{p}{e_{j}}, & \text{otherwise}
+                         \end{array}\right\} \\ \\
+\stepinto{G}{p}{e_i} &\defeq& \left\{\begin{array}{ll}
                          e\sub{x}{v_2}, & \text{if } e_i = C[\eapp{v_1}{v_2}] \text{ and } \singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}
                          \end{array}\right\} \\ \\
-\stepover{e_i} &\defeq& \left\{\begin{array}{ll}
+\stepover{G}{p}{e_i} &\defeq& \left\{\begin{array}{ll}
                          C[v], & \text{if } e_i = C[\eapp{v_1}{v_2}] \text{ and } \multistep{\eapp{v_1}{v_2}}{v}
@@
 \]
-\caption{Rules for traversing the reduction graph. \stepintosym and \stepoversym
-  require a traversal of the sub-term edges to decompose $e_i$ into the
-  target expression \eapp{v_1}{v_2} and the context $C$.
-  \ES{need to mention $G$ and $p$ in these rules}
-}
+\caption{Rules for traversing the reduction graph given a path and
+  node. \stepintosym and \stepoversym require a traversal of the
+  sub-term edges to decompose $e_i$ into the target expression
+  \eapp{v_1}{v_2} and the context $C$.  \ES{these rules are quite ugly and waste space..} }
 \label{fig:traversing-graph}

### 2015-11-18 paper/pldi16-submission/semantics.tex

@@
 %
-Figure~\ref{fig:operational} describes the small-step contextual
-reduction semantics for \lang.
-%
-A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
-and a substitution $\vsu$.
-%
-We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
-transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
-%
-We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
-$e$ is $e_0$, $e'$ is $e_j$ and $\forall 0 \leq i < j$, we have
-\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
-%
-We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
-$\stepi{j}{e}{\su}{e'}{\su'}$.
-%
+
 The evaluation relation is parameterized by a pair of functions:
@@
 
-\paragraph{Narrowing Types} The function $\force{e}{t}$ takes as
+\paragraph{Narrowing Types} The procedure $\force{e}{t}$ takes as
 input an term \texttt{e} and refines it to have the type \texttt{t}
@@
 \begin{array}{lcl}
-\forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
+% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
+% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
 \force{\ehole{i}}{t} & \defeq & \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}} \\
@@
 
-\paragraph{Generating Values} The function $\gen{t}$ takes as input a type
-\texttt{t} and returns \RJ{HEREHERE}
+\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
+takes as input a type \texttt{t} and returns a value of that type. For base
+types, the procedure returns an arbitrary value. For functions, it returns
+a lambda with a \emph{new} hole denoting the return value, and unconstrained
+types (denoted by $\thole{}$) yield fresh, unconstrained values
+(denoted by $\ehole{i}$).
 
@@
 \begin{array}{lcll}
-\gensym       & ::      & t \rightarrow v \\
+% \gensym       & ::      & t \rightarrow v \\
 \gen{\tint}   & \defeq  & n & \quad \text{non-deterministic} \\
@@
 
+\paragraph{Steps and Traces}
 %
+Figure~\ref{fig:operational} describes the small-step contextual
+reduction semantics for \lang.
+%
+A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
+and a substitution $\vsu$.
+%
+We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
+transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
+%
+A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
+$\pair{e_0}{\vsu_0}, \ldots, \pair{e_n}{\vsu_n}$ such that
+$\forall 0 \leq i < n$, we have \step{e_i}{\vsu_i}{e_{i+1}}{\vsu_{i+1}}.
+%
+We write \steptr{\trace}{e}{\vsu}{e'}{\vsu'} if $\trace$ is a trace of the form
+$\pair{e}{\vsu},\ldots,\pair{e'}{\vsu'}$.
+%
+We write \steps{e}{\vsu}{e'}{\vsu'} if \steps{e}{\vsu}{e'}{\vsu'}
+for some trace $\trace$.
 
-Note that each primitive reduction step -- addition, if-elimination, and
+\paragraph{Primitive Reductions}
+Primitive reduction steps -- addition, if-elimination, and
 function application -- uses \forcesym to ensure that values have the
 appropriate type (and that holes are instantiated) before continuing the
-computation.
-%
-Additionally, beta-reduction \emph{does not} type-check its argument, it
-only ensures that the value being applied is a function.
-%
-\begin{thm}
-\label{thm:all-reduce}
-  Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
-  \ES{do we really need to state this, or is it obvious?}
-\end{thm}
+computation. Importantly, beta-reduction \emph{does not} type-check its
+argument, it only ensures that the value being applied is a function.
 
-%
+%% \begin{thm}
+%% \label{thm:all-reduce}
+%%   Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
+%% \ES{do we really need to state this, or is it obvious?}
+%% \end{thm}
+
 % \begin{proof}%[Proof of \autoref{thm:all-reduce}]
@@
 %
-\subsection{Soundness}
-\label{sec:soundness}
-We now show that our evaluation relation conservatively instantiates
-holes, \ie if, given a function $f$, we report that
-$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$,
-then for any type $\tfun{s}{t}$ you assign to $f$, there exists an input \hastype{v}{s} such that
-$\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
+\subsection{Generality}\label{sec:soundness}
 
-We can think of the evaluation of \eapp{f}{\ehole{}} as computing a
-\emph{partial type} -- a type that may contain holes -- for $f$.
-%
-We can extract this type from the result of evaluation as follows.
-%
-\begin{defn}
-\label{defn:partial-type}
-  If \stepi{i}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su}, then the $i$th
-  \emph{partial type} of f, written \ptype{i}{f},
-  is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
+Recall that a key technical challenge in generating witnesses is
+that we have no (static) type information to rely upon. Thus, we
+must avoid the trap of generating \emph{spurious} witnesses that
+arise from picking irrelevant values, when instead there exist
+perfectly good values of a \emph{different} type, under which
+the program would not go wrong.
 
-  We will omit the subscript when we wish to refer to the final partial
-  type.
-\end{defn}
-%
-The \typeof{} function is an approximation of the type of an expression.
-\begin{defn}
-\label{def:typeof}
+\paragraph{Witness Generality} We can now define
+We now show that our evaluation relation instantiates holes in a
+\emph{general} manner. That is, given a function $f$, if we have
+$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$, then for every
+possible input type $t$, we can find a value $\hastype{v}{t}$
+such that $\eapp{f}{v}$ goes wrong.
+
+\begin{thm}{\textbf{[Witness Generality]}}
+\label{thm:soundness}
+  For any function $f$, if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
+  then for every type $t$ there exists $\hastype{v}{t}$ such that
+  \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
+\end{thm}
+
+We need to develop some machinery in order to prove this theorem.
+First, we show how our evaluation rules encode a dynamic form of
+type inference, and next, we show that the types inferred via
+evaluation are indeed maximally general.
+
+\paragraph{The Type of an Expression} The (dynamic) type of an
+expression $e$ is defined as a function $\typeof{e}$ defined as:
   \[
@@
   \]
-\end{defn}
+
+\paragraph{Dynamic Type Inference}
+We can think of the evaluation of \eapp{f}{\ehole{}} as (dynamically)
+inferring a \emph{partial type} --- a type that may contain holes ---
+for $f$.
 %
-We also define a compatibility relation between types.
+We can extract this type from evaluation traces.
 %
-\begin{defn}
-\label{defn:type-compat}
-  A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
-  $\exists \su.\ t = \subst{\su}{s} \lor s = \subst{\su}{t}$.
-  \ES{we're abusing the \su notation here..}
-\end{defn}
+Formally, we say that if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su},
+then the \emph{partial type of} $f$ upto $\trace$, written \ptype{\trace}{f},
+is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
 %
+%We will omit the subscript when we wish to refer to the final partial
+%type, \ie\ at the step where the expression has been reduced to a value
+%(or stuck.)
+
+\paragraph{Type Compatibility}
+Two types are compatible if one of them can be obtained by
+suitably filling the holes of the other (\ie by suitably ``instantiating''
+the type variables appearing in the other). Formally, a type $s$ is
+\emph{compatible} with a type $t$, written \tcompat{s}{t}, if
+$\exists \tsu.\ t = \subst{\tsu}{s} \lor s = \subst{\tsu}{t}$.
+
+\paragraph{Preservation}
 Given these two definitions, we show that each evaluation step
-refines the partial type of $f$.
+refines the partial type of $f$, \ie preserves type compatibility.
 %
@@
 \label{lem:refine-partial}
-  For all $k$, \tcompat{\ptype{k}{f}}{\ptype{k+1}{f}}.
+If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
+$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is single
+step extension of $\trace$)
+%
+%The partial type of $f$ upto $\trace$ is compatible
+%with the partial type upto $\trace'$, \ie\
+%
+then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
 \end{lem}
 \begin{proof}
-  By case analysis on the evaluation rules.
+  By induction on the length of $\trace$ and case analysis on the evaluation rules.
   %
@@
 \end{proof}
+
+\paragraph{Narrowing}
 %
@@
 \label{lem:force-inst}
-  For all $k$, if $\ptype{k}{f} \neq \ptype{k+1}{f}$, then \forcesym must
-  have been called at step $k+1$.
+If
+$\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu}, \ldots,\pair{e}{\vsu}$
+and
+$\trace' = \trace, \steps{e}{\vsu}{e'}{\vsu'}$
+and
+$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
+then the final step $\steps{e}{\vsu}{e'}{\vsu'}$ invokes \forcesym.
 \end{lem}
+
 \begin{proof}
@@
   %
-  If $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in $f$'s
-  argument must have been instantiated with a concrete value at step
-  $k+1$.
+  If $\ptype{\trace}{f} \neq \ptype{\trace}{f}$ then one of the holes in $f$'s
+  argument must have been instantiated with a concrete value at the last step.
   %
@@
 \end{proof}
+
+\paragraph{Incompatible Values Are Wrong}
 %
-Finally, we show that any value that is not compatible with the $k$th
-partial type of $f$ will cause $f$ to get stuck in \emph{at most} $k$
-steps.
+We show that \emph{any} value that is \emph{incompatible} with
+the partial type upto $\trace$ will cause $f$ to get stuck
+in \emph{at most} $k+1$ steps where $k$ is the length of the
+trace $\trace$.
 %
@@
 \label{lem:k-stuck}
-  For any \hastype{v}{t} that is not compatible with
-  the input of \ptype{k}{f}, \stepi{k}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
+  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
+     not \tcompat{t}{\ptype{\trace}{f}} and
+     \hastype{v}{t}
+  then
+     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
 \end{lem}
+
 \begin{proof}
-  By induction on $k$.
-
-  Let $\tfun{s_k}{\_} = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
-  will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
+  By induction on $k$, the length of $\trace$.
+  Let $\tfun{s_k}{\_} = \ptype{\trace}{f}$.
+  Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
+  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
+  in at most $k+1$ steps:
   \begin{description}
@@
 \end{proof}
+
+\paragraph{Proof of Theorem~\ref{thm:soundness}}
 %
-Now we can prove our soundness theorem.
+Now we can prove our soundness theorem. Intuitively, suppose $\trace$
+witnesses that $f$ gets stuck, and let $t$ be the partial type of $f$
+upto $\trace$. We show that \emph{all} types $s$ have stuck-inducing
+values by splitting cases on whether the type is compatible with the
+partial type upto $\trace$.
 %
-\begin{thm}
-\label{thm:soundness}
-  For any function $f$, if \steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
-  then $\forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
-\end{thm}
-\begin{proof}
-  Let $\tfun{s}{\_} = \ptype{}{f}$.
 \begin{description}
-  \item[Case \tcompat{s}{t}:] Our witness is already valid.
-  \item[Case \tincompat{s}{t}:] We can apply Lemma~\ref{lem:k-stuck} and
-    use \emph{any} \hastype{v}{t} as a witness for $f$.
+  \item[Case \tcompat{s}{t}:]
+   The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
+   gets stuck.
+  \item[Case \tincompat{s}{t}:]
+   By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
+   $\eapp{f}{v}$ gets stuck.
 \end{description}
-\end{proof}
 
@@
 %
-We have shown how to find a witness for a function of a single argument,
-but in a language with higher-order functions and currying it may not be
-clear \emph{syntactically} how many arguments a function takes.
+So far, we have seen how a trace leading to a stuck configuration yields
+a general witness demonstrating that the program is ill-typed (\ie\ goes
+wrong for all inputs.)
+In particular, we have shown how to non-deterministically find a witnesses
+for a function of a \emph{single} argument.
+
+In order to convert the semantics into a \emph{procedure} for finding
+witnesses, we must address two challenges.
 %
-Thus, we wrap our operational semantics for \lang in a search loop that
-supplies an increasing number of arguments until the function returns a
-value that is not a lambda.
+First, we must resolve the non-determinism introduced by \gensym.
+%
+Second, in the presence of higher-order functions and currying,
+we must determine how many concrete values to generate to make
+execution go wrong (as we cannot rely upon static typing to
+provide this information.)
 
-The top-level search loop takes as input an open program -- a sequence
-of binders -- and searches for an expression that closes the input
-program and gets stuck. Concretely, given an input
+The witness generation procedure @genWitness@ is formalized in
+Figure~\ref{fig:algo-gen-witness}.
+%
+Next, we describe what it takes as input and output, and how it
+addresses the above challenges to search the space of possible
+executions for general type error witnesses.
+
+\paragraph{Inputs and Outputs}
+%
+Obviously, the problem of generating inputs is undecidable in general.
+%
+The witness generation procedure takes two inputs.
+%
+First, a search bound @n@ which is used to define the \emph{number} of
+traces to explore. (We assume, without loss of generality, that all
+traces are finite.)
+%
+Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
+denoting the list of \emph{top-level definitions}
 %
@@
 %
-we will search for an expression of the form:
+The witness generation procedure returns as output a list of (general)
+witness expressions $e$ each of which is of the form @fn v1 ... vn@.
 %
-\begin{code}
-  fn v1 ... vn
-\end{code}
+The \emph{empty} list is returned when no witness can be found after
+exploring $n$ traces.
 %
-Figure~\ref{fig:expression-api} describes a small API for manipulating
-and evaluating \lang expressions, which we will use to define our search
-algorithm.
+In the sequel, we refer to @fn@ as the \emph{target} of witness generation.
+
+\paragraph{Modeling Semantics}
+%: \texttt{eval} and \texttt{isStuck}}
 %
-The bulk of the search is performed by @eval@, which
-implements our operational semantics from \S\ref{sec:semantics}.
+We resolve the non-determinism in the operational semantics
+(\S~\ref{sec:semantics}) via the procedure:
 %
-The operational semantics is non-deterministic due to \gensym,
-thus @eval@ returns a list of possible results.
+\begin{mcode}
+  eval :: ($e$, $\su$) -> [($v$, $\su$)]
+\end{mcode}
 %
-\begin{figure}[t]
+Due to the non-determinism introduced by \gensym, a call @eval($e$, $\vsu$)@
+returns the \emph{list} possible results of the form $\pair{e'}{\vsu'}$ such
+that $\steps{e}{\vsu}{e'}{\vsu'}$.
+%
+We determine whether a given value is stuck by using the predicate
+%
+\begin{mcode}
+  isStuck :: $v$ -> Bool
+\end{mcode}
+
+\paragraph{Generating Target Arguments}
+%
+We address the issue of currying by defining a procedure @genArgs@ such that
+@genArgs [(f1,e1),...,(fn, en)]@ yields a list of holes @[v1,...,vk]@
+such that @fn v1 ... vk@ \emph{does not} evaluate to a lambda.
+%
+This is achieved via a simple @loop@ that keeps adding holes until
+@eval@uating the closed term yields a non-lambda value.
+%
+The helper functions @witness@ and @close@ are used to respectively
+apply the parameters to the target and close the result under the top-level
+let-binders, prior to invoking @eval@.
+%
+That is, @mkApps@ creates a nested sequence of applications in
+the usual left-associative style, and @mkLets@ takes a list of
+binders and a body expression, and creates a sequence of nested
+let-binders that close the body expression.
+
+\begin{figure}[ht]
 \centering
 \begin{mcode}
-  -- transitive small-step reduction,
-  -- returning a list of results
-  eval :: ($e$, $\su$) -> [($v$, $\su$)]
+genArgs :: [($x$, $e$)] -> [$v$]
+genArgs bs = loop []
+  where
+  loop vs  = case eval (close bs vs) of
+               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\ehole{}$:vs)
+               _        -> vs
 
-  -- manipulating expressions
-  subst   :: $\su$ -> $v$ -> $v$
-  mkApps  :: $e$ -> [$e$] -> $e$
-  mkLets  :: [($x$, $e$)] -> $e$ -> $e$
-  isStuck :: $v$ -> Bool
+close :: [($x$, $e$)] -> [$v$] -> ($e$, $\vsu$)
+close bs vs = (e, $\emptysu$)
+  where
+    e       = mkLets bs (witness bs vs)
+
+witness :: [($x$, $e$)] -> [$v$] -> $e$
+witness bs vs = mkApps target vs
+  where
+    target    = snd (last bs)
 \end{mcode}
-  \caption{Expression API}
-  \label{fig:expression-api}
+\caption{Find number of arguments for target function}
+\label{fig:algo-gen-args}
 \end{figure}
+
+\paragraph{Generating Witnesses}
+% Procedure \texttt{genWitness}}
 %
-We also define a few helper functions for manipulating expressions:
-\begin{itemize}
-\item @subst@ applies a substitution of holes to a value,
-\item @mkApps@ creates a nested sequence of applications in the usual
-  left-associative style,
-\item @mkLets@ takes a list of binders and a body expression, and
-  creates a sequence of nested let-binders, and
-\item @isStuck@ tests whether a value is the \stuck term.
-\end{itemize}
+Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
+implementation of our search for witnesses, in procedure @genWitness@
+which takes as input a bound @n@ and a list of top level binders, and
+returns a list of witness expressions $e$ which demonstrate how the input
+program gets stuck.
 %
-\begin{figure*}[t]
+The search proceeds as follows.
+%
+@(1)@ we invoke @genArgs@ to determine the number of parameters required
+by the (curried) target function.
+%
+@(2)@ we obtain a closed expression $e$ by closing the target with the
+computed holes.
+%
+@(3)@ we take the first @n@ traces returned by @eval@ on the closed $e$
+and
+@(4)@ we extract the substitutions corresponding to the $\stuck$ traces,
+and use them to return the list of witnesses.
+%
+We obtain the following as a corrollary of Theorem~\ref{thm:soundness}:
+
+\begin{thm}{\textbf{[Witness Generation]}}
+\label{thm:generation}
+  If @genWitness k@ $[(f_1, e_1), \ldots, (f_n, e_n)]$ is non-empty,
+  then for every type $t$ there exists $\hastype{v}{t}$ such that
+  \steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
+\end{thm}
+
+\begin{figure}[ht]
 \centering
 \begin{mcode}
-  check :: [($x$, $e$)] -> Result
-  check bnds =
-    -- (2) search for a witness
-    case find (isStuck . fst) results of
-      Nothing      -> Safe
-      Just (_, su) -> Unsafe (mkApps f (subst su args))
+genWitness :: Nat -> [($x$, $e$)] -> [$v$]
+genWitness n bs
+       = [witness bs ($\vsu$ vs) | $\vsu$ <- $\vsu$s]
   where
-      (args, results) = loop []
-      f               = snd (last bnds)
-      build args      = mkLets bnds (mkApps f args)
+   vs  = genArgs defs []      -- (1)
+   $e$    = close defs vs        -- (2)
+   res = take n (eval e)      -- (3)
+   $\vsu$s   = [$\vsu$ | $\pair{\stuck}{\vsu}$ <- res]  -- (4)
+\end{mcode}
+\caption{Generating Witnesses}
+\label{fig:algo-gen-witness}
+\end{figure}
 
-      -- (1) find the correct number of arguments
-      loop :: [$v$] -> ([$v$], [($v$, $\su$)])
-      loop args = case eval (build args, []) of
-        ($\efun{x}{e}$, _) : _ -> loop (args `snoc` $\ehole{}$)
-        results      -> (args, results)
-  \end{mcode}
-  \caption{Implementing our counter-example search in terms of the
-    operational semantics.
-    \ES{1 and 2 are backwards, ugh...}
-    \ES{should address case where output types of successive runs dont match}
-  }
-  \label{fig:search-algo}
-\end{figure*}
+%% \begin{figure}[t]
+  %% \centering
+  %% \begin{mcode}
+  %% -- transitive small-step reduction, returning a list of results
+  %% eval :: ($e$, $\su$) -> [($v$, $\su$)]
+%%
+  %% -- | is a value stuck?
+  %% isStuck :: $v$ -> Bool
+%%
+  %% mkApps  :: $e$ -> [$e$] -> $e$
+  %% mkLets  :: [($x$, $e$)] -> $e$ -> $e$
+  %% \end{mcode}
+  %% \caption{Expression API}
+  %% \label{fig:expression-api}
+%% \end{figure}
+%% %
+%% We also define a few helper functions for manipulating expressions:
+%% \begin{itemize}
+%% \item @subst@ applies a substitution of holes to a value,
+%% \item @mkApps@ creates a nested sequence of applications in the usual
+  %% left-associative style,
+%% \item @mkLets@ takes a list of binders and a body expression, and
+  %% creates a sequence of nested let-binders, and
+%% \item @isStuck@ tests whether a value is the \stuck term.
+%% \end{itemize}
+
+%%%  \begin{figure*}[t]
+  %%%  \centering
+  %%%  \begin{mcode}
+  %%%  check :: [($x$, $e$)] -> Result
+  %%%  check bnds =
+    %%%  -- (2) search for a witness
+    %%%  case find (isStuck . fst) results of
+      %%%  Nothing      -> Safe
+      %%%  Just (_, su) -> Unsafe (mkApps f (subst su args))
+    %%%  where
+      %%%  (args, results) = loop []
+      %%%  f               = snd (last bnds)
+      %%%  build args      = mkLets bnds (mkApps f args)
+%%%
+      %%%  -- (1) find the correct number of arguments
+      %%%  loop :: [$v$] -> ([$v$], [($v$, $\su$)])
+      %%%  loop args = case eval (build args, []) of
+        %%%  ($\efun{x}{e}$, _) : _ -> loop (args `snoc` $\ehole{}$)
+        %%%  results      -> (args, results)
+  %%%  \end{mcode}
+  %%%  \caption{A procedure for generating witnesses}
+    %%%  \ES{should address case where output types of successive runs dont match}
+  %%%  }
+  %%%  \label{fig:search-algo}
+%%%  \end{figure*}
 %
-Figure~\ref{fig:search-algo} summarizes the overall implementation of
-our search for witnesses, which takes as input a sequences of binders
-and returns either @Safe@ if no witness could be found, or |Unsafe $e$|
-where $e$ is a term that causes the input program to get stuck.
-%
-The search is split into two phases: (1) we supply an increasing number
-of arguments until we find a saturated application, and (2) we search
-through the list of results from the non-deterministic evaluator for a
-witness.
-
-% Algorithm:
-%   Input: ML Program (let f1 = e1 in let f2 = e2 in ... e)
-%   Output: 'safe' or 'fn v1 .. vn' where 'fn v1 .. vn' gets stuck
 

### 2015-11-18 paper/pldi16-submission/semantics.tex
narrow should take the current substitution.
it needs to apply the su to a hole if possible, to preserve sharing

@@
 
-\paragraph{Narrowing Types} The procedure $\force{e}{t}$ takes as
+\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
 input an term \texttt{e} and refines it to have the type \texttt{t}
@@
 % \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
-\force{\ehole{i}}{t} & \defeq & \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}} \\
-\force{v}{\ehole{}}  & \defeq & \pair{v}{\emptysu} \\
-\force{n}{\tint}     & \defeq & \pair{n}{\emptysu} \\
-\force{v}{\tint}     & \defeq & \pair{\stuck}{\emptysu} \\
-\force{b}{\tbool}    & \defeq & \pair{b}{\emptysu} \\
-\force{v}{\tbool}    & \defeq & \pair{\stuck}{\emptysu} \\
-\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
-\force{v}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\stuck}{\emptysu} \\
+\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex} \begin{array}{l}
+    \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
+         \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
+  \end{array} \\
+\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
+\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
+\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
+\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
+\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
+\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
+\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
 \end{array}
@@
 % \gensym       & ::      & t \rightarrow v \\
-\gen{\tint}   & \defeq  & n & \quad \text{non-deterministic} \\
-\gen{\tbool}  & \defeq  & b & \quad \text{non-deterministic} \\
+\gen{\tint}   & \defeq  & n, & \quad \text{non-deterministic} \\
+\gen{\tbool}  & \defeq  & b, & \quad \text{non-deterministic} \\
 \gen{\tfun{t_1}{t_2}}   & \defeq & \efun{x}{\ehole{i}}, & \quad \text{$i$ is fresh} \\
@@
   then for every type $t$ there exists $\hastype{v}{t}$ such that
-  \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
+  $\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
 \end{thm}
@@
 %
-Due to the non-determinism introduced by \gensym, a call @eval($e$, $\vsu$)@
+Due to the non-determinism introduced by \gensym, a call \hbox{|eval($e$, $\vsu$)|}
 returns the \emph{list} possible results of the form $\pair{e'}{\vsu'}$ such
@@
 %
-We address the issue of currying by defining a procedure @genArgs@ such that
-@genArgs [(f1,e1),...,(fn, en)]@ yields a list of holes @[v1,...,vk]@
-such that @fn v1 ... vk@ \emph{does not} evaluate to a lambda.
+We address the issue of currying by defining a procedure @genArgs@ such that \\
+\hbox{@genArgs [(f1,e1),...,(fn,en)]@} yields a list of holes @[v1,...,vk]@
+such that \hbox{@fn v1 ... vk@} \emph{does not} evaluate to a lambda.
 %
@@
   then for every type $t$ there exists $\hastype{v}{t}$ such that
-  \steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
+  $\steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
 \end{thm}

### 2015-11-18 paper/pldi16-submission/semantics.tex
minor tweaks to semantics

@@
 \paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
-input an term \texttt{e} and refines it to have the type \texttt{t}
-by yielding a pair of either the same expression and a value substitution
-the refines the holes in \texttt{e} to ensure it has type \texttt{t},
-or yields the stuck term if no such refinement is possible:
+input an term \texttt{e} and refines it to have the type \texttt{t} by
+yielding a pair of either the same expression and a value substitution
+the refines the holes in \texttt{e} to ensure it has type \texttt{t}, or
+yields the stuck term if no such refinement is possible. In the case
+where \texttt{e} is a hole, it first checks in the given $\vsu$ to see
+if the hole has already been instantiated and, if so, returns the
+existing instantiation:
 %
@@
 Primitive reduction steps -- addition, if-elimination, and
-function application -- uses \forcesym to ensure that values have the
+function application -- use \forcesym to ensure that values have the
 appropriate type (and that holes are instantiated) before continuing the
@@
 
-\paragraph{Witness Generality} We can now define
+\paragraph{Witness Generality}
 We now show that our evaluation relation instantiates holes in a
@@
 Formally, we say that if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su},
-then the \emph{partial type of} $f$ upto $\trace$, written \ptype{\trace}{f},
+then the \emph{partial type} of $f$ upto $\trace$, written \ptype{\trace}{f},
 is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
@@
   Note that all rules preserve partial types with the exception of when
-  \forcesym is called on a hole, which case we instantiate the hole with
+  \forcesym is called on a hole, in which case we instantiate the hole with
   a concrete value.
@@
 and
-$\trace' = \trace, \steps{e}{\vsu}{e'}{\vsu'}$
+$\trace' \defeq \trace, \step{e}{\vsu}{e'}{\vsu'}$
 and
 $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
-then the final step $\steps{e}{\vsu}{e'}{\vsu'}$ invokes \forcesym.
+then the final step $\step{e}{\vsu}{e'}{\vsu'}$ invokes \forcesym.
 \end{lem}
@@
   If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
-     not \tcompat{t}{\ptype{\trace}{f}} and
-     \hastype{v}{t}
+     \tincompat{t}{\ptype{\trace}{f}}, and
+     \hastype{v}{t},
   then
@@
 \end{lem}
-
+\ES{self-todo: take a close look at the $\trace$'s vs k's}
 \begin{proof}
   By induction on $k$, the length of $\trace$.
-  Let $\tfun{s_k}{\_} = \ptype{\trace}{f}$.
+
+  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$.
   Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
@@
   \begin{description}
-  \item[Case \tincompat{t}{s_k}:]
+  \item[Case \tincompat{t}{s_{\trace}}:]
     The inductive hypothesis applies.
-  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:]
+  \item[Case \tcompat{t}{s_{\trace}} but \tincompat{t}{s_{\trace'}}:]
     By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
@@
 Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
-implementation of our search for witnesses, in procedure @genWitness@
+implementation of our search for witnesses, in the procedure @genWitness@,
 which takes as input a bound @n@ and a list of top level binders, and
-returns a list of witness expressions $e$ which demonstrate how the input
+returns a list of witness expressions $e$ that demonstrate how the input
 program gets stuck.

### 2015-11-18 paper/pldi16-submission/related.tex
ugh

\section{Related Work}
\label{sec:related-work}

\subsection{Type Error Localization}
\label{sec:type-error-local}
\begin{itemize}
\item \cite{lerner_searching_2007}
\item \cite{zhang_toward_2014}
\item \cite{pavlinovic_finding_2014}
\item \cite{chen_counter-factual_2014}
\end{itemize}

\subsection{Testing}
\label{sec:testing}
\begin{itemize}
\item \cite{claessen_quickcheck:_2000}
\item \cite{runciman_smallcheck_2008}
\item \cite{godefroid_dart:_2005}
\item \cite{cadar_klee:_2008}
\item \cite{tillmann_pexwhite_2008}
\end{itemize}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2015-11-18 paper/pldi16-submission/related.tex
working on the related work

@@
 \label{sec:type-error-local}
+It is well known that unification-based type inference procedures can
+produce poor error messages, and in particular, can misidentify the
+\emph{source} of the type error.
+%
+
 \begin{itemize}
-\item \cite{lerner_searching_2007}
-\item \cite{zhang_toward_2014}
-\item \cite{pavlinovic_finding_2014}
-\item \cite{chen_counter-factual_2014}
+\item Lerner \etal~\cite{lerner_searching_2007} attempt to localize the
+  type error and suggest a fix by replacing expressions (or removing
+  them entirely) with alternatives based on the surrounding program
+  context.
+\item Chen and Erwig~\cite{chen_counter-factual_2014} use a variational
+  type system represent the possibility that the inferred type of an
+  expression may be incorrect, because the expression may be the source
+  of the error. It then attempts to deduce the error source by searching
+  for an expression whose type can be changed such that type inference
+  would succeed. In contrast to Seminal, which searches for changes at
+  the value-level, by searching at the type level the search is complete
+  due the finite universe of types in a program.
+\item Neubauer and Thiemann~\cite{neubauer_discriminative_2003} present
+  a decidable type system based on discriminative sum types, in which
+  all terms are typeable and type derivations contain all type errors in
+  a program. They then use the typing derivation to slice out the parts
+  of the expression related to each error.
+\item Zhang and Myers~\cite{zhang_toward_2014} present an algorithm for
+  identifying the most likely culprit in a system of unsatisfiable
+  constraints (\eg type equalities), based on Bayesian reasoning.
+\item Pavlinovic \etal~\cite{pavlinovic_finding_2014} translate the
+  error localization problem to a MaxSMT optimization problem, using
+  compiler-provided weights to rank the possible sources.
+% \item \cite{chen_error-tolerant_2012}
+% \item \cite{okeefe_type_1992}
+% \item \cite{gomard_partial_1990}
+% \item \cite{thatte_type_1988}
 \end{itemize}
+
+In contrast to these approaches, we do not attempt to localize or fix
+the type error. Instead we try to explain it to the novice user using a
+dynamic witness that demonstrates how their program is not just
+ill-typed but truly wrong. In addition, allowing users to run their
+program (even knowing that it is wrong) enables experimentation and the
+use of debuggers to step through the program and investigate its
+evolution.
 
@@
 \label{sec:testing}
+\nanomaly is at its heart a test generator, and builds on a rich line of
+work.
+%
+Our use of holes to represent unknown values is inspired by the work of
+Runciman, Naylor, and Lindblad~\cite{runciman_smallcheck_2008,naylor_finding_2007,lindblad_property_2007},
+%
+who use lazy evaluation to drastically reduce the search space for
+exhaustive test generation, by grouping together equivalent inputs by
+the set of values they force. An exhaustive search is complete (up to
+the depth bound), if a witness exists it will be found, but due to the
+exponential blowup in the search space the depth bound can be quite
+limited without advanced grouping and filtering techniques.
+%
+Our search is not exhaustive; instead we use random generation to fill
+in holes on demand.
+%
+Random test generation~\cite{claessen_quickcheck:_2000,csallner_jcrasher:_2004,pacheco_feedback-directed_2007}
+%
+is by its nature incomplete, but is able to check larger inputs than
+exhaustive testing as a result.
+
+Instead of enumerating values, which may trigger the same path through
+the program, one might enumerate paths. 
+%
+Dynamic-symbolic execution~\cite{godefroid_dart:_2005,cadar_klee:_2008,tillmann_pex_2008}
+%
+enables this by symbolic execution (to track which path a given input
+vector triggers) with concrete execution (to ensure test failures are
+not spurious). The system collects a path condition during execution,
+which tracks symbolically what conditions must be met to trigger the
+current path. Upon successfully completing a test run, it negates the
+path condition and queries a solver for another set of inputs that
+satisfy the negated path condition, \ie inputs that will not trigger the
+same path. Thus, dynamic-symbolic execution can prune the search space
+much faster than techniques based on enumerating values, but is limited
+by the expressiveness of the underlying solver. Our operational
+semantics is amenable to dynamic-symbolic execution, one would just need
+to collect the path condition and replace our implementation of \gensym
+by a call to the solver. We chose to use lazy, random generation instead
+because it does not incur the overhead of an external solver, and
+produces high coverage for our domain of novice programs.
+
+% \begin{itemize}
+% \item \cite{claessen_quickcheck:_2000}
+% \item 
+% \item \cite{godefroid_dart:_2005}
+% \item \cite{cadar_klee:_2008}
+% \item \cite{tillmann_pex_2008}
+% \end{itemize}
+
+\subsection{Misc}
 \begin{itemize}
-\item \cite{claessen_quickcheck:_2000}
-\item \cite{runciman_smallcheck_2008}
-\item \cite{godefroid_dart:_2005}
-\item \cite{cadar_klee:_2008}
-\item \cite{tillmann_pexwhite_2008}
+\item \cite{vytiniotis_equality_2012} extend the Haskell compiler GHC to
+  support compiling ill-typed programs, but their intent is rather
+  different from ours. Their goal was to allow programmers to
+  incrementally test refactorings, which often cause type errors in
+  distant pieces of code. They replace any expression that fails to
+  type check with a \emph{runtime} error, but do not perform
+  any type checking at runtime.
+\item \cite{perera_functional_2012}
 \end{itemize}

### 2015-11-18 paper/pldi16-submission/evaluation-nanomaly.tex
tweak
comment
trace size graph for seminal
move some text to the overview

@@
 \label{sec:evaluation-nanomaly}
-We have evaluated our prototype of \nanomaly in two phases:
+We have implemented our interactive semantics for the pure subset of
+\ocaml with a web-based visualization of the generated traces in a tool
+called \nanomaly.
+%
+It takes as input an \ocaml program and a function to check, searches
+for a witness as described in \S~\ref{sec:searching-witness}, and
+produces as output an interactive trace of the (possibly buggy)
+execution~\footnote{If no witness could be found we simply return a trace
+  of the last attempt.}.
+
+We have not really (\ES{huh?}) evaluated our prototype of \nanomaly in two phases:
 %
@@
 %
-\begin{figure}[t]
+\begin{figure*}[t]
 \centering
-Have a graph!  
+\begin{minipage}{0.49\linewidth}
+\includegraphics[width=\linewidth]{trace_size_seminal.pdf} % TODO: switch to ucsd
+\end{minipage}
+\begin{minipage}{0.49\linewidth}
+\includegraphics[width=\linewidth]{trace_size_seminal.pdf}
+\end{minipage}
 \caption{Complexity of the generated traces.}
 \label{fig:results-complexity}
-\end{figure}
+\end{figure*}
 %

### 2015-11-18 paper/pldi16-submission/overview.tex
move some text to the overview

@@
 \label{sec:overview}
+
+Our goal is to find concrete values that demonstrate how a
+program ``goes wrong''.
+
+\paragraph{Problem: Which inputs are bad?}
+%
+One approach is to randomly generate input values and use
+them to execute the program until we find one that causes
+the program to go wrong. However, to see why this approach
+is naive, consider the following example:
+%
+\begin{code}
+  let f x =
+    let y = 1 + x in
+      1. +. y
+\end{code}
+%
+What \emph{types} of inputs should we test \texttt{f} with?
+Values of type \texttt{int} and \texttt{float} are fair game,
+but values of type say, \texttt{string} or \texttt{int list}
+will cause the program to go wrong in an \emph{irrelevant}
+manner.
+
+\paragraph{Solution:} \RJ{STOP}
+
+\subsection{Visualizing Witnesses}
+\label{sec:visual-witness}
+% We have implemented our interactive semantics for the pure subset of
+% \ocaml with a web-based visualization of the generated traces in a tool
+% called \nanomaly.
+% %
+% It takes as input an \ocaml program and a function to check, searches
+% for a witness as described in \S~\ref{sec:searching-witness}, and
+% produces as output an interactive trace of the (possibly buggy)
+% execution~\footnote{If no witness could be found we simply return a trace
+%   of the last attempt.}.
+% %
+We initially display a big-step reduction, \ie the source term followed
+by the stuck term (or the final value).
+%
+The user can then progressively fill in the hidden steps of the
+computation by selecting a visible term and choosing one of the
+applicable traversal strategies -- described in
+\S~\ref{sec:traversing-graph} -- to insert another term into the
+visualization.
+
+A sample interaction of the student program in
+Figure~\ref{fig:palindrome} can be seen in
+Figure~\ref{fig:nanomaly-palindrome}.
+%
+\begin{figure}[ht]
+\centering
+\begin{code}
+  let listReverse l =
+    let rec helper xs = function
+      | []     -> xs
+      | hd::tl -> helper (hd::xs) tl
+    in helper []
+
+  let palindrome w =
+    listReverse w = w
+\end{code}
+\caption{Ill-typed \texttt{palindrome} function. The error lies in the
+  use of the \texttt{function} construct which introduces an extra,
+  undesired lambda abstraction, but manifests itself in the equality
+  test on line XX. The fix is to replace \texttt{function} with
+  \texttt{match xs with}. \ES{I think we should use this example in the
+    intro/overview, leaving it here for now until they exist..}}
+\label{fig:palindrome}
+\end{figure}
+%
+\begin{figure}[ht]
+% \centering
+% Initial state: \\
+% \includegraphics[width=\linewidth]{palindrome} \\
+% After ``jumping backward'' from the stuck term: \\
+% \includegraphics[width=\linewidth]{palindrome2} \\
+% After ``stepping forward'' twice from @listReverse []@: \\
+% \includegraphics[width=\linewidth]{palindrome3} \\
+\caption{Sample interaction}
+\label{fig:nanomaly-palindrome}
+\end{figure}
+%
+The initial state of the visualization tells us that the witness is
+@palindrome []@, and that after some number of steps -- the thick arrow
+denotes a multi-step transition -- we try to compare a function to a
+list, which is not allowed in \ocaml.
+%
+Based on our semantics for \lang one would expect the final state to be
+the \stuck term, but in practice we simply halt execution at the point
+where an expression would transition to \stuck, as this final state is
+more informative.
+%
+Note also that the equality test is colored black while the contextual
+if-expression is faded out, this tells the user that the program got
+stuck while trying to evaluate the sub-expression.
+
+Upon seeing the stuck term, one might wonder where the @function@
+came from.
+%
+To investigate we select the stuck term and click the ``jump backward''
+button to search backwards from the stuck term for the most recent
+function call, which brings us to @listReverse [] = w@, in the same
+context as before.
+%
+Uncontent with the explanation so far, we ``step forward'' twice from
+the @listReverse []@ term, bringing us to @helper [] = w@.
+%
+At this point it is clear that the @helper@ function is not defined
+correctly, we have supplied it with the single argument we expected and
+yet it still returned a @function@.
+
+The problem is that the @function@ keyword in \ocaml defines an
+anonymous function that takes a single argument and immediately does a
+case-analysis without giving the argument a name.
+%
+The solution is to replace @function@ with an explicit @match xs with@
+-- naming the value we wish to case-analyse.
+%
+After applying our fix, \nanomaly -- and more importantly \ocaml --
+decide that @listReverse@ is safe to run.
+\ES{these last few paragraphs probably belong in the overview}
+%
 

### 2015-11-18 paper/pldi16-submission/semantics.tex
move some text to the overview

@@
 
-Our goal is to find concrete values that demonstrate how a
-program ``goes wrong''.
+% Our goal is to find concrete values that demonstrate how a
+% program ``goes wrong''.
 
-\paragraph{Problem: Which inputs are bad?}
-%
-One approach is to randomly generate input values and use
-them to execute the program until we find one that causes
-the program to go wrong. However, to see why this approach
-is naive, consider the following example:
-%
-\begin{lstlisting}
-  let f x =
-    let y = 1 + x in
-      1. +. y
-\end{lstlisting}
-%
-What \emph{types} of inputs should we test \texttt{f} with?
-Values of type \texttt{int} and \texttt{float} are fair game,
-but values of type say, \texttt{string} or \texttt{int list}
-will cause the program to go wrong in an \emph{irrelevant}
-manner.
+% \paragraph{Problem: Which inputs are bad?}
+% %
+% One approach is to randomly generate input values and use
+% them to execute the program until we find one that causes
+% the program to go wrong. However, to see why this approach
+% is naive, consider the following example:
+% %
+% \begin{lstlisting}
+%   let f x =
+%     let y = 1 + x in
+%       1. +. y
+% \end{lstlisting}
+% %
+% What \emph{types} of inputs should we test \texttt{f} with?
+% Values of type \texttt{int} and \texttt{float} are fair game,
+% but values of type say, \texttt{string} or \texttt{int list}
+% will cause the program to go wrong in an \emph{irrelevant}
+% manner.
 
-\paragraph{Solution:} \RJ{STOP}
+% \paragraph{Solution:} \RJ{STOP}
 

### 2015-11-18 paper/pldi16-submission/related.tex

@@
 \label{sec:related-work}
+In this section we connect our work to related efforts on type errors,
+testing, and program exploration.
 
-\subsection{Type Error Localization}
-\label{sec:type-error-local}
+\subsection{Type Errors}
+\label{sec:type-error}
+
+\paragraph{Diagnosis and Repair}
+\label{sec:diagnosis-repair}
 It is well known that unification-based type inference procedures can
@@
 %
+Thus, many groups have explored techniques to pinpoint the true source
+of the error and recommend fixes.
 
-\begin{itemize}
-\item Lerner \etal~\cite{lerner_searching_2007} attempt to localize the
-  type error and suggest a fix by replacing expressions (or removing
-  them entirely) with alternatives based on the surrounding program
-  context.
-\item Chen and Erwig~\cite{chen_counter-factual_2014} use a variational
-  type system represent the possibility that the inferred type of an
-  expression may be incorrect, because the expression may be the source
-  of the error. It then attempts to deduce the error source by searching
-  for an expression whose type can be changed such that type inference
-  would succeed. In contrast to Seminal, which searches for changes at
-  the value-level, by searching at the type level the search is complete
-  due the finite universe of types in a program.
-\item Neubauer and Thiemann~\cite{neubauer_discriminative_2003} present
-  a decidable type system based on discriminative sum types, in which
-  all terms are typeable and type derivations contain all type errors in
-  a program. They then use the typing derivation to slice out the parts
-  of the expression related to each error.
-\item Zhang and Myers~\cite{zhang_toward_2014} present an algorithm for
+Lerner \etal~\cite{lerner_searching_2007} attempt to localize the type
+error and suggest a fix by replacing expressions (or removing them
+entirely) with alternatives based on the surrounding program context.
+%
+Chen and Erwig~\cite{chen_counter-factual_2014} use a variational type
+system represent the possibility that the inferred type of an expression
+may be incorrect, because the expression may be the source of the
+error. It then attempts to deduce the error source by searching for an
+expression whose type can be changed such that type inference would
+succeed. In contrast to Seminal, which searches for changes at the
+value-level, by searching at the type level the search is complete due
+the finite universe of types in a program.
+%
+Neubauer and Thiemann~\cite{neubauer_discriminative_2003} present a
+decidable type system based on discriminative sum types, in which all
+terms are typeable and type derivations contain all type errors in a
+program. They then use the typing derivation to slice out the parts of
+the expression related to each error.
+%
+Zhang and Myers~\cite{zhang_toward_2014} present an algorithm for
 identifying the most likely culprit in a system of unsatisfiable
 constraints (\eg type equalities), based on Bayesian reasoning.
-\item Pavlinovic \etal~\cite{pavlinovic_finding_2014} translate the
-  error localization problem to a MaxSMT optimization problem, using
+%
+Pavlinovic \etal~\cite{pavlinovic_finding_2014} translate the error
+localization problem to a MaxSMT optimization problem, using
 compiler-provided weights to rank the possible sources.
@@
 % \item \cite{thatte_type_1988}
-\end{itemize}
 
@@
 evolution.
+
+\paragraph{Running Ill-Typed Programs}
+\label{sec:running-ill-typed}
+Vytiniotis \etal~\cite{vytiniotis_equality_2012} extend the Haskell
+compiler GHC to support compiling ill-typed programs, but their intent
+is rather different from ours. Their goal was to allow programmers to
+incrementally test refactorings, which often cause type errors in
+distant pieces of code. They replace any expression that fails to type
+check with a \emph{runtime} error, but do not perform any type checking
+at runtime.
 
@@
 
-\subsection{Misc}
-\begin{itemize}
-\item \cite{vytiniotis_equality_2012} extend the Haskell compiler GHC to
-  support compiling ill-typed programs, but their intent is rather
-  different from ours. Their goal was to allow programmers to
-  incrementally test refactorings, which often cause type errors in
-  distant pieces of code. They replace any expression that fails to
-  type check with a \emph{runtime} error, but do not perform
-  any type checking at runtime.
-\item \cite{perera_functional_2012}
-\end{itemize}
+\subsection{Program Exploration}
+Perera \etal~\cite{perera_functional_2012} present a tracing semantics
+for functional programs that tags values with their provenance, enabling
+a form of backwards program slicing from a final value to the sequence
+of reductions that produced it. Notably, they allow the user to supply a
+\emph{partial value} -- a value containing holes -- and in turn present
+a partial slice, containing only those steps that affected the specified
+components of the partial value. This system is designed to answer
+questions of the form ``Where did this value come from?'' and thus is
+focused on backward exploration. In contrast, our trace visualization
+supports forward \emph{and} backward exploration, though our backward
+steps are more limited than Perera's. Specifically, we do not support
+selecting a value and inserting the intermediate terms that preceded it
+while ignoring unrelated computation steps; this would be interesting
+future work.
+
+\ES{todo: more on program exploration}
 

### 2015-11-18 paper/pldi16-submission/evaluation-witness.tex
experimenting with histogram plot while i wait for data...
add some graphs, finally..
trace size graph for seminal
graphs

@@
 Due to the possibility of non-termination we set a limit on the number
-of reductions to perform, increasing in 1,000-step increments from 1,000
-steps to 10,000 steps total.
+of reductions to perform, increasing in 500-step increments from 500
+steps to 3,000 steps total.
 %
@@
 \centering
-
-  Have a table or graph! \ES{graph of cumulative errors per timeout? \ie
-    monotonically increasing, with a cap of 90\% or so}
-\caption{Experimental results}
+\begin{minipage}{\linewidth}
+\includegraphics[width=\linewidth]{coverage.pdf}
+\end{minipage}
+\begin{minipage}{\linewidth}
+\includegraphics[width=\linewidth]{distrib.pdf}
+\end{minipage}
+\vspace{-8ex}
+\caption{Experimental results. Our random search successfully finds
+  witnesses for over 85\% of the programs at the lowest timeout and
+  improves slightly to 88\% at the next level. In both datasets we
+  detect actual type errors about 82\% of the time, unbound variables or
+  constructors 3-4\% of the time, and diverging loops 2-3\% of the
+  time. For the remaining 11-12\% of the programs we are unable to
+  provide any useful feedback. \ES{TODO: the timeout number for ucsd is
+    a bit misleading due to a mistake in the benchmarking code. re-running to fix..}
+}
 \label{fig:results-witness}
@@
 %
-XX programs were deemed safe and XX timed out even at 10,000 steps, \ie
+XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
 we could not provide any useful feedback for XX\% of the total programs.

### 2015-11-18 paper/pldi16-submission/introduction.tex
misc

@@
 Much work has been done on pinpointing the \emph{source} of a type
-error~\cite{zhang-myers, maxsmt, counterfactual}, but an accurate source
-location still does not explain \emph{why} the program is wrong.
+error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
+but an accurate source location still does not explain \emph{why} the
+program is wrong.
 

### 2015-11-18 paper/pldi16-submission/evaluation-witness.tex

@@
 The first dataset comes from the Spring 2014 undergraduate Programming
-Languages course at UC San Diego (IRB \#XXXXX). 
+Languages course at UC San Diego (IRB \#140608). 
 %
@@
 The second dataset -- widely used in the literature -- comes from a
-similar course at the University of Washington~\cite{XXXXX}, from which
-we extracted XXXX ill-typed programs.
+similar course at the University of
+Washington~\cite{lerner_seminal:_2006}, from which we extracted 284
+ill-typed programs.
 

### 2015-11-18 paper/pldi16-submission/plots.py
graphs

@@ -54,7 +54,7 @@ def plot_coverage(seminal, ucsd):
     plt.title('Cumulative Coverage')
     plt.xticks(ind + width/2.0, [r[0] for r in xy_s])
     plt.yticks(np.arange(0, 101, 10))
-    plt.legend((p1[0], p2[0]), ('Seminal', 'UCSD'))
+    plt.legend((p1[0], p2[0]), ('Seminal', 'UCSD'), 'lower right')
     # plt.legend((p1[0], p2[0]), ('Men', 'Women'))
     autolabel(plt, p1)
     autolabel(plt, p2)
@@ -90,6 +90,7 @@ def plot_trace_size(data, label):
     ax.legend((p1[0],), ('Steps',))
     plt.ylabel('Traces')
     ax.set_xlim(0,6)
+    ax.set_ylim(0,len(step))
     plt.xticks(ind + width/2.0, binlabels)
     # autolabel(ax, p1)
 
@@ -101,6 +102,7 @@ def plot_trace_size(data, label):
     ax.legend((p2[0],), ('Jumps',))
     plt.ylabel('Traces')
     ax.set_xlim(0,6)
+    ax.set_ylim(0,len(step))
     plt.xticks(ind + width/2.0, binlabels)
     # autolabel(ax, p2)
 

### 2015-11-19 paper/pldi16-submission/plots.py
sigh, more graph nonsense

@@ -3,7 +3,7 @@ import matplotlib
 import matplotlib.pyplot as plt
 import numpy as np
 
-BUCKETS = range(500, 3001, 500)
+BUCKETS = [0.1, 0.2, 1.0, 10.0, 60.0 ] # range(500, 3001, 500)
 COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
 
 SAFE = ['S', 'T']
@@ -12,6 +12,10 @@ UNSAFE = ['U', 'B', 'D'] #, 'O']
 UNSAFE_L = ['Unsafe', 'Unbound', 'Diverge'] #, 'Output']
 ALL = UNSAFE + SAFE
 ALL_L = UNSAFE_L + SAFE_L
+
+ALL_D = [ ['S', 'T'], ['U'], ['B'], ['D'] ]
+ALL_DL = [ 'No Witness', 'Witness', 'Unbound', 'Diverge' ] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
+
 
 def read_csv(f):
     with open(f) as f:
@@ -29,7 +33,7 @@ def cumulative_coverage(data):
     headers = data[0]
     data = data[1:]
     return [(l, int(100 * len([r for r in data
-                               if int(r[1]) <= l
+                               if float(r[2]) <= l
                                and r[4] in UNSAFE])
                           / float(len(data))))
             for l in BUCKETS]
@@ -49,20 +53,21 @@ def plot_coverage(seminal, ucsd):
     p2 = plt.bar(ind + width, [r[1] for r in xy_u], width,
                  color=COLORS[1])
 
-    plt.xlabel('Timeout (steps)')
-    plt.ylabel('% witnesses found')
-    plt.title('Cumulative Coverage')
-    plt.xticks(ind + width/2.0, [r[0] for r in xy_s])
-    plt.yticks(np.arange(0, 101, 10))
-    plt.legend((p1[0], p2[0]), ('Seminal', 'UCSD'), 'lower right')
+    plt.xlabel('Timeout (seconds)', fontsize=20)
+    plt.ylabel('% witnesses found', fontsize=20)
+    plt.title('Cumulative Coverage', fontsize=24)
+    plt.xticks(ind + width, [r[0] for r in xy_s], fontsize='large')
+    plt.yticks(np.arange(0, 101, 10), fontsize='large')
[…]

### 2015-11-19 paper/pldi16-submission/evaluation.tex
consolidate evaluation sections into one

@@
-\begin{figure}
-\judgementHead{Evaluation}{\stepsto{e, \su}{e, \su}}
-$$
-\inference[\recontext]{\stepsto{e, \su}{e_1, \su_1}}{\stepsto{C[e], \su}{C[e_1], \su_1}}
-$$
+% \section{Evaluation: Recasting Type Errors as Runtime Errors}
+\section{Evaluation}
+\label{sec:evaluation}
 
-$$
-\inference[\restuck]{}{\stepsto{C[\stuck], \su}{\stuck, \su}}
-$$
+%
+The immediate question is ``What fraction of type errors admit
+witnesses?''
+%
+To answer this question we implemented a prototype of our search
+procedure for the pure subset of \ocaml, \ie \lang extended with
+algebraic datatypes and records. 
+%
+In our implementation we instantiated \gensym with a simple random
+generation of values, which we will show is more than sufficient for the
+majority of type errors.
+%
+We evaluated our implementation of two sets of known-bad programs, \ie
+programs that were rejected by the \ocaml compiler because of a type
+error.
+%
+The first dataset comes from the Spring 2014 undergraduate Programming
+Languages course at UC San Diego (IRB \#140608). 
+%
+We recorded each interaction with the \ocaml top-level system over the
+course of the first three assignments, from which we extracted 4,407
+distinct, ill-typed \ocaml programs.
+%
+The second dataset -- widely used in the literature -- comes from a
+similar course at the University of
+Washington~\cite{lerner_seminal:_2006}, from which we extracted 284
+ill-typed programs.
 
-$$
-\inference[\replusgood]
-  {n_1, \su_2 = \force{v_1}{\tnat} \\
-   n_2, \su_3 = \force{v_2}{\tnat} \\ 
-   n = \eplus{n_1}{n_2}}
-  {\stepsto{\eplus{v_1}{v_2}, \su_1}{n, \su_1\su_2\su_3}}
-$$
+\subsection{Witness Coverage}
+\label{sec:eval:witness-coverage}
+%
+We ran our search algorithm on each program for 1,000 iterations, with
+the entry point set to the function that \ocaml had identified as
+containing a type error.
+%
+Due to the possibility of non-termination we set a timeout of 10 minutes
+per program.
+% Due to the possibility of non-termination we set a limit on the number
+% of reductions to perform, increasing in 500-step increments from 500
+% steps to 3,000 steps total.
+%
+We also added a na\"ive check for infinite recursion; at each recursive
+function call we check whether the new arguments are identical to the
+current arguments.
+%
+If so, the function cannot possibly terminate and we report an error.
+%
+While not a \emph{type error}, infinite recursion is still a clear bug
+in the program, and thus valuable feedback for the user.
 
-$$
-\inference[\replusbadone]
-  {\stuck, \su_2 = \force{v_1}{\tnat}}
-  {\stepsto{\eplus{v_1}{v_2}, \su_1}{\stuck, \su_1\su_2}}
-$$
+\begin{figure}[t]
+\centering
+\begin{minipage}{\linewidth}
+\includegraphics[width=\linewidth]{coverage.pdf}
+\end{minipage}
+\begin{minipage}{\linewidth}
+\includegraphics[width=\linewidth]{distrib.pdf}
+\end{minipage}
+% \vspace{-8ex}
+\caption{Results of our coverage testing and the distribution of test
+  outcomes. Our random search successfully finds witnesses for over 85\%
+  of the programs at the lowest timeout and improves slightly to 88\% at
+  the next level. In both datasets we detect actual type errors about
+  82\% of the time, unbound variables or constructors 3-4\% of the time,
+  and diverging loops 2-3\% of the time. For the remaining 11-12\% of
+  the programs we are unable to provide any useful feedback.  }
+\label{fig:results-witness}
+\end{figure}
 
-$$
-\inference[\replusbadtwo]
-  {\stuck, \su_2 = \force{v_2}{\tnat}}
-  {\stepsto{\eplus{v_1}{v_2}, \su_1}{\stuck, \su_1\su_2}}
-$$
+\paragraph{Results}
+\label{sec:results-witness}
+The results of our experiments are summarized in
+Figure~\ref{fig:results-witness}.
+%
+In both datasets our tool was able to find a witness for at least 87\%
+of the programs in under 10 seconds.
+%
+Interestingly, while the vast majority of witnesses corresponded to a
+type-error, as expected, 3-4\% triggered an unbound variable error and
+2-3\% triggered an infinite recursion error.
+%
+For the remaining 12\% of programs we were unable to provide any useful
+feedback as they either completed 1,000 tests successfully, or timed out
+after 10 minutes.
+%
+% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
+% we could not provide any useful feedback for XX\% of the total programs.
+%
+While a more advanced search procedure, \eg dynamic-symbolic execution,
+could likely trigger more of the type errors, our experiments show that
+type errors are coarse enough (or that novice programs are \emph{simple}
+enough) that these techniques are not necessary.
 
-$$
-\inference[\reifgoodone]
-  {\etrue, \su_2 = \force{v}{\tbool}}
-  {\stepsto{\eif{v}{e_1}{e_2}, \su_1}{e_1, \su_1\su_2}}
-$$
+\subsection{Trace Complexity}
+\label{sec:trace-complexity}
+For each of the ill-typed programs for which we could find a witness, we
+measure the complexity of the generated trace according to the following
+metrics:
+%
+\begin{enumerate}
+\item The number of single-step reductions along the path from the
+  witness to the stuck term. This can be thought of as a worst-case
+  complexity, \ie ``How big is the fully-expanded trace?''
+\item The number forward (or backward) jumps that can be taken along the
+  path from the witness to the stuck term. This is not quite the same as
+  a stack-trace as it shows \emph{all} function calls, including those
+  that returned successfully.
+  % These abstract many details of the computation that are often
+  % uninteresting or irrelevant to the explanation.
+% \item \ES{others?}
+\end{enumerate}
+%
+\begin{figure}[t]
+\centering
+% \includegraphics[width=\linewidth]{trace_size.pdf}
+% \begin{minipage}{0.49\linewidth}
+\includegraphics[width=\linewidth]{trace_size_step.pdf}
+% \end{minipage}
+% \begin{minipage}{0.49\linewidth}
+\includegraphics[width=\linewidth]{trace_size_jump.pdf}
+% \end{minipage}
+\caption{Complexity of the generated traces. At least 90\% of the traces
+  have a jump complexity of less than 10, with an average of 3.}
+\label{fig:results-complexity}
+\end{figure}
+%
 
-$$
-\inference[\reifgoodtwo]
-  {\efalse, \su_2 = \force{v}{\tbool}}
-  {\stepsto{\eif{v}{e_1}{e_2}, \su_1}{e_2, \su_1\su_2}}
-$$
+\paragraph{Results}
+\label{sec:results-complexity}
+The results of the experiment are summarized in
+Figure~\ref{fig:results-complexity}.
+%
+The average number of single-step reductions per trace is 26 for the
+UCSD dataset (27 for the Seminal dataset) with a maximum of 2,744 (659
+for Seminal). The average number of jumps per trace is 3 (also 3 for
+Seminal) with a maximium of 177 (52 for Seminal). In both datasets at
+least 90\% of the traces contain fewer than 10 jumps.
+%
 
-$$
-\inference[\reifbad]
-  {\stuck, \su_2 = \force{v}{\tbool}}
-  {\stepsto{\eif{v}{e_1}{e_2}, \su_1}{\stuck, \su_1\su_2}}
-$$
-\caption{Evaluation relation}
-\label{fig:evaluation}
-\end{figure}
+\subsection{Discussion}
+\label{sec:discussion}
+To summarize, our experiments demonstrate that \nanomaly finds witnesses
+to type errors: (1) with high coverage in a timespan amenable to
+compile-time analysis, and (2) with traces that have a low average
+complexity of 3 function calls.
+
+There are, of course, drawbacks to our approach.
+
+\paragraph{Random Generation}
+Random test generation has difficulty generating highly constrained
+values, \eg red-black trees or a pair of equal integers. If the type
+error is hidden behind a complex branch condition \nanomaly may not be
+able to trigger it. Exhaustive testing and dynamic-symbolic execution
+can address this short-coming by performing an exhaustive search for
+inputs (\emph{resp}. paths through the program). As our experiments
+show, however, novice programs do not appear to require more advanced
+search techniques, likely because the novice programs tend to be simple.
+
+\paragraph{Infinite Types}
+Our algorithm does check for infinite types inside \forcesym, but there
+are some degenerate cases where it is unable to detect them. Consider,
+for example, the following buggy version of @replicate@ which is
+supposed to produce a list containing @n@ copies of @x@.
+%
+\begin{code}
+  let rec replicate n x =
+    if n <= 0 then
+      []
+    else
+      replicate (n-1) [x]
+\end{code}
+%
+This code produces a nested list (with @n@ levels of nesting) containing
+a single copy of @x@. \ocaml detects a cyclic \hbox{@'a = 'a list@}
+constraint in the recursive call and throws a type error, whereas
+\nanomaly happily recurses as many times as necessary. Strictly
+speaking, this function itself cannot ``go wrong'', the program would
+not get stuck until a \emph{client} attempted to use the result
+expecting a flat list. But this is not very satisfying as @replicate@ is
+clearly to blame. Furthermore, in our experience, infinite-type errors
+are some of the more difficult ones to debug (and to explain to
+novices), so better support for this scenario would be useful.
+
+\paragraph{Trace Explosion}
+Though the average complexity of our generated traces is low in terms of
+jumps, there are some extreme outliers. We cannot reasonably expect a
+novice user to explore a trace containing 50+ terms and draw a
+conclusion about which pieces contributed to the bug in their
+program. Enhancing our visualization to slice out program paths relevant
+to specific values, \eg in the style of~ \cite{perera_functional_2012},
+would likely help alleviate this issue. This would allow users to
+highlight a confusing value and ask: ``Where did this come from?''
+
+% \begin{itemize}
+% \item benchmarks: our data + seminal data
+% \item both cases: \textbf{random} search sufficient to trigger runtime crash in 80\% of programs
+% \item how many of the ``safe'' programs are actually safe??
+% \end{itemize}

### 2015-11-19 paper/pldi16-submission/introduction.tex
consolidate evaluation sections into one

@@
 %
+% Then, we formalize our algorithm and prove it sound
+% (\S~\ref{sec:searching-witness}), and evaluate it on a large set of real
+% student programs (\S~\ref{sec:eval}).
+% %
+% Next, we extend the search algorithm to collect the reduction graph and
+% formalize a set of traversals (\S~\ref{sec:explaining}), and measure the
+% complexity of these traces on our benchmarks
+% (\S~\ref{sec:evaluation-nanomaly}).
 Then, we formalize our algorithm and prove it sound
-(\S~\ref{sec:searching-witness}), and evaluate it on a large set of real
-student programs (\S~\ref{sec:eval-witness}).
+(\S~\ref{sec:searching-witness}).
 %
-Next, we extend the search algorithm to collect the reduction graph and
-formalize a set of traversals (\S~\ref{sec:explaining}), and measure the
-complexity of these traces on our benchmarks
-(\S~\ref{sec:evaluation-nanomaly}).
+Next, we extend the algorithm to collect a reduction graph and formalize
+a set of traversals over the graph (\S~\ref{sec:explaining}).
 %
-Finally, we present the results of a preliminary user study on the full
-system (\S~\ref{sec:user-study}).
+Finally, we evaluate our search algorithm and trace generation on a
+large set of real student programs (\S~\ref{sec:evaluation}).
+%
+% Finally, we present the results of a preliminary user study on the full
+% system (\S~\ref{sec:user-study}).
 

### 2015-11-19 paper/pldi16-submission/plots.py

@@ -124,9 +124,13 @@ def plot_trace_size(seminal, ucsd):
     # ax.set(adjustable='box-forced', aspect=4)
 
     y,binEdges=np.histogram(jump_s,bins=bins)
+    foo = y / float(len(jump_s)) 
+    print(foo[0] + foo[1])
     p1 = plt.bar(ind, y / float(len(jump_s)), label='Seminal', width=width, color=COLORS[0])
 
     y,binEdges=np.histogram(jump_u,bins=bins)
+    foo = y / float(len(jump_u)) 
+    print(foo[0] + foo[1])
     p2 = plt.bar(ind + width, y / float(len(jump_u)), label='UCSD', width=width, color=COLORS[1])
     plt.legend((p1[0],p2[0]), ('Seminal','UCSD'), fontsize=16)
     plt.xlabel('Jumps', fontsize=20)
@@ -189,7 +193,7 @@ def plot_distrib(seminal, ucsd):
     #plt.tight_layout()
 
     plt.suptitle('Distribution of Results', fontsize=24, y=0.9)
-    plt.figlegend(p1[0], ALL_DL, 'lower right', fontsize=16)
+    plt.figlegend(p1[0], ALL_DL, 'lower center', fontsize=16, ncol=4)
 
 
     # p2 = plt.pie(rs, labels=ALL_L,

### 2015-11-19 paper/pldi16-submission/evaluation.tex
one more take on the graphs

@@
 program. Enhancing our visualization to slice out program paths relevant
-to specific values, \eg in the style of~ \cite{perera_functional_2012},
-would likely help alleviate this issue. This would allow users to
+to specific values, \eg in the style of~\cite{perera_functional_2012},
+would likely help alleviate this issue, allowing users to
 highlight a confusing value and ask: ``Where did this come from?''

### 2015-11-19 paper/pldi16-submission/plots.py

@@ -54,7 +54,7 @@ def plot_coverage(seminal, ucsd):
                  color=COLORS[1])
 
     plt.xlabel('Timeout (seconds)', fontsize=20)
-    plt.ylabel('% witnesses found', fontsize=20)
+    plt.ylabel('Witnesses Found (% of total programs)', fontsize=20)
     plt.title('Cumulative Coverage', fontsize=24)
     plt.xticks(ind + width, [r[0] for r in xy_s], fontsize='large')
     plt.yticks(np.arange(0, 101, 10), fontsize='large')
@@ -103,7 +103,8 @@ def plot_trace_size(seminal, ucsd):
     y,binEdges=np.histogram(step_u,bins=bins)
     p2 = plt.bar(ind + width, y / float(len(step_u)), label='UCSD', width=width, color=COLORS[1])
     plt.legend((p1[0],p2[0]), ('Seminal','UCSD'), fontsize=16)
-    plt.xlabel('Steps', fontsize=20)
+    plt.title('Trace Complexity', fontsize=24)
+    plt.xlabel('Total Steps', fontsize=20)
     plt.ylabel('Traces (%)', fontsize=20)
     # ax.set_xlim(0,6)
     # ax.set_ylim(0,len(step))
@@ -133,7 +134,10 @@ def plot_trace_size(seminal, ucsd):
     print(foo[0] + foo[1])
     p2 = plt.bar(ind + width, y / float(len(jump_u)), label='UCSD', width=width, color=COLORS[1])
     plt.legend((p1[0],p2[0]), ('Seminal','UCSD'), fontsize=16)
-    plt.xlabel('Jumps', fontsize=20)
+    # plt.title('Complexity of Traces (in jumps)', fontsize=24)
+    plt.xlabel('Total Jumps', fontsize=20)
+    # plt.xlabel('Jumps', fontsize=20)
+    plt.ylabel('Traces (%)', fontsize=20)
     # plt.ylabel('Traces')
     # ax.set_xlim(0,6)
     # fig.set_ylim(0.0,1.0)

### 2015-11-19 paper/pldi16-submission/plots.py
done with eval

@@ -53,7 +53,7 @@ def plot_coverage(seminal, ucsd):
     p2 = plt.bar(ind + width, [r[1] for r in xy_u], width,
                  color=COLORS[1])
 
-    plt.xlabel('Timeout (seconds)', fontsize=20)
+    plt.xlabel('Witness found in <x seconds', fontsize=20)
     plt.ylabel('Witnesses Found (% of total programs)', fontsize=20)
     plt.title('Cumulative Coverage', fontsize=24)
     plt.xticks(ind + width, [r[0] for r in xy_s], fontsize='large')

### 2015-11-19 paper/pldi16-submission/introduction.tex
complete draft of overview and abstract

@@
 
-Consider the following ill-typed @factorial@ function, where we return
+Consider the following ill-typed @fac@ function, where we return
 @true@ in the base case instead of @1@.
 %
+\begin{figure}[h!]
 \begin{code}
-  let rec factorial n =
+  let rec fac n =
     if n <= 0 then
@@
     else
-      n * factorial (n-1)
+      n * fac (n-1)
 \end{code}
-%
+\caption{Ill-typed \texttt{fac} function that mistakenly
+  returns \texttt{true} in the base case instead of \texttt{1}.}
+\label{fig:factorial}
+\end{figure}
+
 \ocaml's response is to tell us that on line 5, the recursive call to
-@factorial@ returns a @bool@, but we needed an @int@ for the
+@fac@ returns a @bool@, but we needed an @int@ for the
 multiplication.
@@
 %
-\ocaml is correct, but we may ask \emph{why} @factorial@ returns a
+\ocaml is correct, but we may ask \emph{why} @fac@ returns a
 @bool@ instead of an @int@.
@@
 that arises in the equivalent \ruby function, when we call
-@factorial(1)@.
+@fac(1)@.
 %
@@
                  into Fixnum (TypeError)
-  from fac.rb:5:in `factorial'
+  from fac.rb:5:in `fac'
   from fac.rb:10:in `<main>'
@@
 Our semantics \emph{conservatively} instantiates holes with concrete
-values, if we can find a witness to a type error then there is no valid
+values; if we can find a witness to a type error then there is no valid
 typing for the program.
@@
 %
-We present the execution as a \emph{reduction graph}, in which the nodes
-are expressions and the edges indicate either a single-step reduction or
-a sub-term relation.
+We present the execution as a \emph{reduction graph}, which records
+all of the small-step reductions in the context in which they occur.
+%
+%  in which the nodes
+% are expressions and the edges indicate either a single-step reduction or
+% a sub-term relation.
 %
@@
 The visualization initially shows the source term (\ie the witness) and
-the stuck term (or the final value), and allows the novice to
-iteratively expand the computation to show intermediate steps by
-selecting an expression and choosing an strategy for picking the next
-node to insert into the graph.
+the stuck term (or the final value), and allows the user to iteratively
+expand the computation to expose intermediate steps, by selecting an
+expression and choosing a traversal strategy (\eg ``step forward'').
 %

### 2015-11-19 paper/pldi16-submission/interactive.tex
remove text from interactive.tex

@@
 %
-We have shown how to reliably find witnesses to type errors in \ocaml,
-but this not fully address our original goal -- to \emph{explain} the
-errors.
+% We have shown how to reliably find witnesses to type errors in \ocaml,
+% but this not fully address our original goal -- to \emph{explain} the
+% errors.
+% %
+% Having identified an input vector that triggers a bug, a common next
+% step is to step through the program with a \emph{debugger} to observe
+% how the program evolves.
+% %
+% The existing debuggers and interpreters for \ocaml assume a type-correct
+% program, so unfortunately we cannot use them off-the-shelf.
+% %
+% Instead we return to our semantics for \lang and extend the evaluation
+% rules to collect a trace that we can present to the user, demonstrating
+% precisely \emph{how} their program went wrong.
+
+% The trace takes the form of a \emph{reduction graph}, where the nodes
+% are terms and the edges represent either the single-step
+% $\hookrightarrow$ or a ``sub-term'' relation. For example, evaluating
+% the expression @1 + 2 + 3@ would produce the graph in
+% Figure~\ref{fig:simple-reduction}.
+% %
+% \begin{figure}[t]
+%   \centering
+%   \includegraphics[width=\linewidth]{simple.png}
+% \caption{The reduction graph for \texttt{1 + 2 + 3}.}
+% \label{fig:simple-reduction}
+% \end{figure}
+% %
+% We choose this graph representation instead of a simple, linear sequence
+% of expressions because it will allow us to express a variety of
+% traversals, such as ``step into'' and ``step over'' -- commonly found in
+% traditional debuggers.
+
+In this section we formalize the collection and traversal of the reduction graph.
 %
-Having identified an input vector that triggers a bug, a common next
-step is to step through the program with a \emph{debugger} to observe
-how the program evolves.
+First, we extend our semantics from \S~\ref{sec:semantics} to collect
+the edges in the graph (\S~\ref{sec:inter-semant}).
 %
-The existing debuggers and interpreters for \ocaml assume a type-correct
-program, so unfortunately we cannot use them off-the-shelf.
-%
-Instead we return to our semantics for \lang and extend the evaluation
-rules to collect a trace that we can present to the user, demonstrating
-precisely \emph{how} their program went wrong.
+Then, we express a set of common debugging steps as graph traversals
+(\S~\ref{sec:traversing-graph}).
 
-The trace takes the form of a \emph{reduction graph}, where the nodes
-are terms and the edges represent either the single-step
-$\hookrightarrow$ or a ``sub-term'' relation. For example, evaluating
-the expression @1 + 2 + 3@ would produce the graph in
-Figure~\ref{fig:simple-reduction}.
-%
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\linewidth]{simple.png}
-\caption{The reduction graph for \texttt{1 + 2 + 3}.}
-\label{fig:simple-reduction}
-\end{figure}
-%
-We choose this graph representation instead of a simple, linear sequence
-of expressions because it will allow us to express a variety of
-traversals, such as ``step into'' and ``step over'' -- commonly found in
-traditional debuggers.
-
-\subsection{Interactive Semantics}
+\subsection{Tracing Semantics}
 \label{sec:inter-semant}
@@
 \begin{array}{lcl}
+\subtermssym                 & \dcolon & e \to \tr \\
 \subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
@@
 \inference[\reappgood]
-  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}}
+  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
   {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
@@
 \inference[\reappbad]
-  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}}
+  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
   {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
@@
 %
-The changes to the operational semantics of \S\ref{sec:semantics} are
+The changes to the operational semantics of \S~\ref{sec:semantics} are
 mechanical, so we will not reproduce them in full; instead we describe
@@
 \stepg{e_1}{\su_1}{\tr_1}{e_2}{\su_2}{\singlestep{e_1}{e_2}; \subterms{e_2}; \tr_1}
+\]
+
+After evaluation the reduction graph can be constructed directly from
+the trace $\tr$ as follows:
+\[
+G(\tr) = \pair{\{e \spmid e \in \tr\}}{\tr}
 \]

### 2015-11-19 paper/pldi16-submission/evaluation.tex

@@
 %
-The immediate question is ``What fraction of type errors admit
-witnesses?''
+% The immediate question is ``What fraction of type errors admit
+% witnesses?''
 %
-To answer this question we implemented a prototype of our search
-procedure for the pure subset of \ocaml, \ie \lang extended with
-algebraic datatypes and records. 
+We have implemented a prototype of our search procedure and trace
+visualization for the pure subset of \ocaml, in a tool called \nanomaly.
+% , \ie \lang extended with
+% algebraic datatypes and records. 
 %
@@
 majority of type errors.
-%
-We evaluated our implementation of two sets of known-bad programs, \ie
+
+There are two questions we seek to answer with our evaluation:
+\begin{enumerate}
+\item What fraction of type errors \emph{admit} witnesses?
+\item How \emph{complex} are the traces our witnesses produce?
+\end{enumerate}
+
+We answer these questions on two sets of known-bad programs, \ie
 programs that were rejected by the \ocaml compiler because of a type
@@
 The first dataset comes from the Spring 2014 undergraduate Programming
-Languages course at UC San Diego (IRB \#140608). 
+Languages course at UC San Diego. 
 %
 We recorded each interaction with the \ocaml top-level system over the
-course of the first three assignments, from which we extracted 4,407
-distinct, ill-typed \ocaml programs.
+course of the first three assignments (IRB \#140608), from which we
+extracted 4,407 distinct, ill-typed \ocaml programs.
 %
@@
 \caption{Results of our coverage testing and the distribution of test
-  outcomes. Our random search successfully finds witnesses for over 85\%
-  of the programs at the lowest timeout and improves slightly to 88\% at
-  the next level. In both datasets we detect actual type errors about
-  82\% of the time, unbound variables or constructors 3-4\% of the time,
-  and diverging loops 2-3\% of the time. For the remaining 11-12\% of
-  the programs we are unable to provide any useful feedback.  }
+  outcomes. Our random search successfully finds witnesses for 83\% of
+  the programs in under one second, improving slightly to 87\% in under
+  10 seconds. In both datasets we detect actual type errors about 82\%
+  of the time, unbound variables or constructors 3-4\% of the time, and
+  diverging loops 2-3\% of the time. For the remaining 11-12\% of the
+  programs we are unable to provide any useful feedback.  }
 \label{fig:results-witness}
@@
 %
-In both datasets our tool was able to find a witness for at least 87\%
-of the programs in under 10 seconds.
+In both datasets our tool was able to find a witness for 83\% of the
+programs in under one second, \ie fast enough to be integrated as a
+compile-time check. If we extend our tolerance to a 10 second timeout,
+we hit a maximum of 87\% coverage.
 %
@@
 \item The number forward (or backward) jumps that can be taken along the
-  path from the witness to the stuck term. This is not quite the same as
+  path from the witness to the stuck term\footnote{This is not quite the same as
   a stack-trace as it shows \emph{all} function calls, including those
-  that returned successfully.
+  that returned successfully.}.
   % These abstract many details of the computation that are often
@@
 
-There are, of course, drawbacks to our approach.
+There are, of course, drawbacks to our approach. Three that stand out
+are: (1) coverage limits due to random generation, (2) the inability to
+handle certain instances of infinite types, and (3) dealing with an
+explosion in the size of generated traces.
 
@@
 \paragraph{Infinite Types}
-Our algorithm does check for infinite types inside \forcesym, but there
-are some degenerate cases where it is unable to detect them. Consider,
-for example, the following buggy version of @replicate@ which is
-supposed to produce a list containing @n@ copies of @x@.
+Our implementation does check for infinite types inside \forcesym, but
+there are some degenerate cases where it is unable to detect
+them. Consider, for example, the following buggy version of @replicate@,
+which is supposed to produce a list containing @n@ copies of @x@.
 %
@@
 constraint in the recursive call and throws a type error, whereas
-\nanomaly happily recurses as many times as necessary. Strictly
-speaking, this function itself cannot ``go wrong'', the program would
-not get stuck until a \emph{client} attempted to use the result
-expecting a flat list. But this is not very satisfying as @replicate@ is
-clearly to blame. Furthermore, in our experience, infinite-type errors
-are some of the more difficult ones to debug (and to explain to
-novices), so better support for this scenario would be useful.
+\nanomaly happily recurses @n@ times to produce the nested list.
+Strictly speaking, this function itself cannot ``go wrong'', the
+program would not get stuck until a \emph{client} attempted to use the
+result expecting a flat list. But this is not very satisfying as
+@replicate@ is clearly to blame. Furthermore, in our experience,
+infinite-type errors are some of the more difficult ones to debug (and
+to explain to novices), so better support for this scenario would be
+useful.
 

### 2015-11-19 paper/pldi16-submission/main.tex
add some graphs, finally..
ugh
consolidate evaluation sections into one

@@
 %
-\documentclass[10pt,pldi,nocopyrightspace,indentedstyle=false]{sigplanconf-pldi15}
+\documentclass[blind,preprint,nocopyrightspace,10pt,squareay,clearpagebib,explicitsize]{sigplanconf-pldi15}
 
@@
 \usepackage{graphicx}
+\usepackage{pdfpages}
+%\usepackage[doi]{natbib}
 
@@
 \newtheorem{defn}{Definition} % [section]
+
+\usepackage{pgfplots}
+\usepackage{pgfkeys}
+\pgfplotsset{compat=1.12}
 
@@
 \begin{abstract}
+  Type errors are a common stumbling block for newcomers to typed
+  functional languages.
+  % 
+  Compilers prevent standard debugging techniques by refusing to run
+  ill-typed programs.
+  %
+  We demonstrate how to expose static type errors for the runtime
+  crashes they hide, by searching for a witness that will trigger the
+  error.
+  %
+  Our search procedure collects an execution trace for which we have
+  developed an interactive visualization to demonstrate precisely
+  \emph{how} the witness makes the program ``go wrong''.
+  %
+  We have implemented our system for \ocaml and demonstrate its efficacy
+  on a large set of real student programs.
 \end{abstract}
@@
 
-\section{Introduction}          % 1 page
-\section{Overview}              % 2 pages?
+% \section{Introduction}          % 1 page
+\input{introduction.tex}
+
+% \section{Overview}              % 2 pages?
+\input{overview.tex}
+
 % \section{Type-carrying Semantics} % 2 pages
@@
 % \section{Evalution: Recasting Type Errors as Runtime Errors} % 1 page
-\input{evaluation-witness.tex}
+% \input{evaluation-witness.tex}
 
-%\section{``Interactive'' Semantics \& Trace Exploration} % 2 pages
+% \section{``Interactive'' Semantics \& Trace Exploration} % 2 pages
 \input{interactive.tex}
 
-\input{nanomaly.tex}
+% \input{nanomaly.tex}
 
 % \section{Evaluation: \nanomaly}                % 1 page
-\input{evaluation-nanomaly.tex}
+% \input{evaluation-nanomaly.tex}
+\input{evaluation.tex}
 
-\section{Related Work}              % 1 pages
+% \section{Related Work}              % 1 pages
+\input{related.tex}
 
-\subsection{Type Error Localization}
-\label{sec:type-error-local}
-\begin{itemize}
-\item seminal
-\item bayesian (Zhang+Myers)
-\item MaxSMT
-\item counterfactual (Chen+Erwig)
-\end{itemize}
-\subsection{Debugging \& Program Understanding}
-\label{sec:debugging}
+% \section{Conclusion}                % 1 page
 
-\section{Conclusion}                % 1 page
+{
+\bibliographystyle{abbrvnat}
+\bibliography{main}
+}
 

### 2015-11-19 paper/pldi16-submission/overview.tex

--- 2015-11-17 paper/pldi16-submission/overview.tex
@@
 \label{sec:overview}
+We start with an overview of our contributions.
+%
+First we illustrate why generating suitable inputs to functions in the
+absense of type information is tricky. 
+%
+Then we describe our solution to the problem and highlight the
+similarity to static type inference.
+%
+Finally, we demonstrate our visualization of the witnesses we produce.
+
+\subsection{Generating Witnesses}
+\label{sec:generating-witnesses}
+Our goal is to find concrete values that demonstrate how a
+program ``goes wrong''.
+
+\paragraph{Problem: Which inputs are bad?}
+%
+One approach is to randomly generate input values and use them to
+execute the program until we find one that causes the program to go
+wrong. However, to see why this approach is naive, consider the
+@fac@ function from Figure~\ref{fig:factorial}, reproduced here.
+
+\begin{code}
+  let rec fac n =
+    if n <= 0 then
+      true
+    else
+      n * fac (n-1)
+\end{code}
+\ES{having two copies of \texttt{fac} seems silly, but the back-reference across a page boundary is no good either...}
+
+What \emph{types} of inputs should we test @fac@ with?
+%
+Values of type @int@ are fair game, but values of type, say,
+@string@ or @int list@ will cause the program to go wrong
+in an \emph{irrelevant} manner.
+%
+Concretely, we want to avoid testing @fac@ with any type other
+than @int@ because any other type would cause @fac@ to get stuck
+immediately in the @n <= 0@ test.
+
+\paragraph{Solution: Don't generate inputs until forced.}
+Our solution is to avoid generating a concrete value for the input at
+all, until we can be sure of its type.
+%
+The intuition is that we want to be as lenient as possible in our tests,
+so we make no assumptions about types until it becomes clear from the
+context what type an input must have.
+%
+This is actually quite similar in spirit to type inference.
+
+To defer input generation, we borrow the notion of a ``hole'' from
+SmallCheck~\cite{runciman_smallcheck_2008}.
+%
+A hole -- written \ehole{} -- is a \emph{placeholder} for a value; in
+our case we do not even know its type.
+%
+We leave all inputs as uninstantiated holes until they are demanded by
+the program, \eg due to a primitive operation like the @<=@ test.
+
+\paragraph{Narrowing Input Types}
+Primitive operations, data construction, and case-analysis \emph{narrow}
+the types of values. 
+%
+For concrete values this amounts to a runtime type check, we ensure that
+the value has a type compatible with the expected type.
+%
+For holes, this means we now know the type it should
+have (or in the case of compound data we know \emph{more} about the
+type) so we can instantiate the hole with a value.
+%
+The value may itself contain more holes, corresponding to components
+whose type we still do not know.
+%
+Consider the @fst@ function:
+%
+\begin{code}
+  let fst p = match p with
+    (a, b) -> a
+\end{code}
+%
+The case analysis tells us that @p@ must be a pair, but it says
+\emph{nothing} about the contents of the pair.
+%
+Thus, upon reaching the case-analysis we would generate a pair
+containing fresh holes for the @fst@ and @snd@ component. 
+%
+Notice the similarity between instantiation of type variables and
+instantiation of holes.
+%
+We can compute an approximate type for @fst@ by approximating the types
+of the (instantiated) input and output, which would give us:
+%
+\begin{mcode}
+  fst : ($\thole{1}$ * $\thole{2}$) -> $\thole{1}$
+\end{mcode}
+%
+We call this type approximate because we only see a single path through
+the program, and thus will miss narrowing points that only occur in
+other paths.
+
+Returning to @fac@, given a hole as input we will narrow the hole
+to an @int@ upon reaching the @<=@ test.
+%
+At this point we choose a random @int@ for the instantiation and
+concrete execution takes over entirely leading us to the expected crash
+in the multiplication.
+
+As we show in \S~\ref{sec:soundness} our lazy instantiation of holes is
+\emph{conservative}.
+%
+We prove that if calling a function with a hole as input causes the
+function to ``go wrong'', then there is \emph{no possible} type for the
+function.
+%
+In other words, for any type you might assign to the function, there
+exist valid inputs that will cause it to crash.
+
+\paragraph{Problem: How many inputs does a function take?}
+There is another wrinkle though, how did we know that @fac@ takes
+a single argument instead of two (or none)?
+%
+It is clear, syntactically, that @fac@ takes \emph{at least} one
+argument, but in a higher-order language with currying syntax can be
+deceiving. 
+%
+Consider the following definition:
+%
+\begin{code}
+  let incAllByOne = List.map (+ 1)
+\end{code}
+%
+Is @incAllByOne@ a function?
+%
+If so, how many arguments does it take?
+%
+The \ocaml compiler deduces that @incAllByOne@ takes a single argument
+because @List.map@'s \emph{type} says it takes two arguments, and it is
+partially applied to @(+ 1)@.
+%
+As we are dealing with ill-typed programs we do not have the luxury of
+typing information, so we must deduce the number of arguments by
+experimentation.
+
+\paragraph{Solution: Search for saturated application.}
+Luckily the solution is straightforward, we add arguments one-by-one
+until we reach a \emph{saturated} application, \ie until evaluating the
+application returns a value other than a lambda.
+
+\subsection{Visualizing Witnesses}
+\label{sec:visual-witness}
+We have shown how to reliably find witnesses to type errors in \ocaml,
+but this does not fully address our original goal -- to \emph{explain}
+the errors.
+%
+Having identified an input vector that triggers a bug, a common next
+step is to step through the program with a \emph{debugger} to observe
+how the program evolves.
+%
+The existing debuggers and interpreters for \ocaml assume a type-correct
+program, so unfortunately we cannot use them off-the-shelf.
+%
+Instead we extend our search for witnesses to produce an execution
+trace.
+
+\paragraph{Reduction Graph}
+Our trace takes the form of a reduction graph, which records small-step
+reductions in the context in which they occur.
+%
+These graphs have two types of edges: 
+%
+(1) ``steps-to'' edges that capture the small-step transition between
+two terms, and
+%
+(2) ``sub-term'' edges that capture the containment relation between two
+terms.
+
+For example, evaluating the expression @1+2+3@ would produce the
+graph in Figure~\ref{fig:simple-reduction-hi}.
+%
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\linewidth]{simple.pdf}
+  \caption{The reduction graph for \texttt{1+2+3}. The two edges
+    produced by the transition from \texttt{1+2+3} to \hbox{\texttt{3+3}}
+    are highlighted.}
+\label{fig:simple-reduction-hi}
+\end{figure}
+%
+Notice that when we transition from @1+2+3@ to @3+3@ we collect
+both that edge \emph{and} an edge from the sub-term @1+2@ to @3@.
+%
+These additional edges allow us to implement two common debugging
+operations \emph{post-hoc}: ``step into'' to zoom in on a specific
+function call, and ``step over'' to skip over an uninteresting
+sub-computation.
+
+\paragraph{Interacting with the graph}
+The reduction graph is useful for formulating and executing traversals,
+but displaying it all at once would quickly become overwhelming.
+%
+Our interaction begins by displaying a big-step reduction, \ie the
+witness followed by the stuck term.
+%
+The user can then progressively fill in the hidden steps of the
+computation by selecting a visible term and choosing one of the
+applicable traversal strategies -- formalized in
+\S~\ref{sec:traversing-graph} -- to insert another term into the
+visualization.
+
+A sample interaction with the trace of @fac 1@ can be seen in
+Figure~\ref{fig:nanomaly-factorial}.
+%
+\begin{figure*}[t]
+\centering
+\includegraphics[width=0.8\linewidth]{fac-steps.png}
+\caption{A sequence of interactions with the trace of 
+  \texttt{fac 1}. The stuck term is red, in each node the redex is
+  highlighted. Thick arrows denote a multi-step transition, thin arrows
+  denote a single-step transition. We start in step 1. In step 2 we jump
+  forward from the witness to the next function call. In step 3 we step
+  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
+  of execution. In step 4 we take a single step forward from 
+  \texttt{fac 0} (hiding the context for space).}
+\label{fig:nanomaly-factorial}
+\end{figure*}
+%
+The initial state of the visualization tells us that after some number
+of steps -- the thick arrow denotes a multi-step transition -- we try to
+multiply @1@ by @true@.
+
+Upon seeing the stuck term, we might wonder where @true@ came from.
+%
+To investigate we select the stuck term and click the ``jump backward''
+button to search backwards from the stuck term for the most recent
+function call, which brings us to @1 * fac 0@. Notice at this point that
+@fac 0@ is highlighted while @1 *@ is grayed out. This tells us that
+@fac 0@ is the redex in this term.
+
+@fac 0@ seems like the right thing to do so we choose to ``step into''
+it, which inserts a new multi-step transition from @fac 0@ to @true@.
+%
+Finally, we take a ``step forward'' from @fac 0@, bringing us to @fac@'s
+body. If we mouse over the body term we will see a popup with the
+environment at this point, notably telling us that @n = 0@. At this point
+it is clear that @fac@ handles the @n <= 0@ case incorrectly and should
+instead return an @int@.
+
+% Upon seeing the stuck term, one might wonder where the @function@
+% came from.
+% %
+% To investigate we select the stuck term and click the ``jump backward''
+% button to search backwards from the stuck term for the most recent
+% function call, which brings us to @listReverse [] = w@, in the same
+% context as before.
+% %
+% Uncontent with the explanation so far, we ``step forward'' twice from
+% the @listReverse []@ term, bringing us to @helper [] = w@.
+% %
+% At this point it is clear that the @helper@ function is not defined
+% correctly, we have supplied it with the single argument we expected and
+% yet it still returned a @function@.
+
+% The problem is that the @function@ keyword in \ocaml defines an
+% anonymous function that takes a single argument and immediately does a
+% case-analysis without giving the argument a name.
+% %
+% The solution is to replace @function@ with an explicit @match xs with@
+% -- naming the value we wish to case-analyse.
+% %
+% After applying our fix, \nanomaly -- and more importantly \ocaml --
+% decide that @listReverse@ is safe to run.
+% \ES{these last few paragraphs probably belong in the overview}
+%
 

### 2015-11-19 paper/pldi16-submission/semantics.tex
down to 11pp

@@
 % \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
-\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex} \begin{array}{l}
-    \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
-         \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
-  \end{array} \\
+\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
+\begin{cases}
+  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
+  \pair{v}{\ehole{i} \mapsto v}   & \mbox{o.w.} v = \gen{t}\\
+\end{cases} \\
+ % \begin{array}{l}
+    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
+         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
+  % \end{array} \\
 \force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
@@
   Note that all rules preserve partial types with the exception of when
-  \forcesym is called on a hole, in which case we instantiate the hole with
+  \forcesym\ is called on a hole, in which case we instantiate the hole with
   a concrete value.
@@
 %
-We show that \emph{any} value that is \emph{incompatible} with
-the partial type upto $\trace$ will cause $f$ to get stuck
-in \emph{at most} $k+1$ steps where $k$ is the length of the
-trace $\trace$.
+\emph{Any} value that is \emph{incompatible} with
+the partial type upto trace $\trace$ will cause $f$ to get stuck
+in \emph{at most} $k+1$ steps where $k$ is the length of $\trace$.
 %
@@
   By induction on $k$, the length of $\trace$.
-
   Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$.
@@
   in at most $k+1$ steps:
-  \begin{description}
-  \item[Case \tincompat{t}{s_{\trace}}:]
+%
+  %\begin{description}
+  % \item
+  [Case \tincompat{t}{s_{\trace}}:]
     The inductive hypothesis applies.
-  \item[Case \tcompat{t}{s_{\trace}} but \tincompat{t}{s_{\trace'}}:]
+  %\item
+  [Case \tcompat{t}{s_{\trace}} but \tincompat{t}{s_{\trace'}}:]
     By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
@@
     have succeeded.
-  \end{description}
+  %\end{description}
 \end{proof}
@@
 %
-Now we can prove our soundness theorem. Intuitively, suppose $\trace$
-witnesses that $f$ gets stuck, and let $t$ be the partial type of $f$
-upto $\trace$. We show that \emph{all} types $s$ have stuck-inducing
-values by splitting cases on whether the type is compatible with the
-partial type upto $\trace$.
+Suppose $\trace$ witnesses that $f$ gets stuck,
+and let $t$ be the partial type of $f$ upto $\trace$.
+We show that \emph{all} types $s$ have stuck-inducing
+values by splitting cases on whether the type is
+compatible with the partial type upto $\trace$.
 %
-\begin{description}
-  \item[Case \tcompat{s}{t}:]
+% \begin{description}
+  %\item
+  [Case \tcompat{s}{t}:]
    The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
    gets stuck.
-  \item[Case \tincompat{s}{t}:]
+  %\item
+  [Case \tincompat{s}{t}:]
    By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
    $\eapp{f}{v}$ gets stuck.
-\end{description}
+%\end{description}
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
 \centering

### 2015-11-19 paper/pldi16-submission/haskellListings.tex
add some keywords
down to 11pp

@@
 % uncomment next line to restore colors
-% \def\withcolor{}
+\def\withcolor{}
 
@@
                 create,get,set,div,rescale,add,delete,insert,prop_focus_left_master,average,best,insert,union,split,size,fromList,copy,group,good,bad,foo,explode,singleton,difference,fromJust,sort,unfold,
-                target, query, decode, encode, check, refuteSMT, binder,
                 subst,unapply,apply,proxy,refinement,fresh,guard,constrain,oneOf,
@@
                 mkCtor,gCtors,gEncode,gEncodeFields,gDecode,gDecodeFields,reproxyRep,empty,splitCtor,checkField,scanM,
-                padAverage,focusUp,execute,checkSMT,inputTypes,outputType,toReft,app
+                padAverage,focusUp,execute,checkSMT,inputTypes,outputType,toReft,app,
+								genArgs, genWitness, close, witness, mkApps, mkLets, isStuck
                 %, inTypes, inputTypes, outputType, execute, smtFindModel, smtRefuteModel
@@
 		case,class,data,deriving,do,else,if,return,def,import,in,infixl,infixr,instance,let,tmapM,for2M,forM,zipWithM,otherwise,
-		module,measure,pred,predicate,of,primitive,then,type,where,lazy,throw,when
+		module,measure,pred,predicate,of,primitive,then,type,where,lazy,throw,when,
+                rec,function,fun,match,with
 	},

### 2015-11-19 paper/pldi16-submission/interactive.tex
down to 11pp

@@
 \label{sec:explaining}
+
+A trace, on its own, is likely too detailed to be a good
+explanation of the type error. One approach is to use the
+witness input to step through the program with a
+\emph{debugger} to observe how the program evolves.
 %
-% We have shown how to reliably find witnesses to type errors in \ocaml,
-% but this not fully address our original goal -- to \emph{explain} the
-% errors.
-% %
-% Having identified an input vector that triggers a bug, a common next
-% step is to step through the program with a \emph{debugger} to observe
-% how the program evolves.
-% %
-% The existing debuggers and interpreters for \ocaml assume a type-correct
-% program, so unfortunately we cannot use them off-the-shelf.
-% %
-% Instead we return to our semantics for \lang and extend the evaluation
-% rules to collect a trace that we can present to the user, demonstrating
-% precisely \emph{how} their program went wrong.
+This route is problematic for two reasons.
+%
+First, existing debuggers and interpreters for
+typed languages (\eg\ \ocaml) typically require
+a type-correct program as input.
+%
+Second, we wish to have a quicker way to get
+to the essence of the error, \eg\ by skipping
+over irrelevant sub-computations, and focusing
+on the important ones.
+
+Next, we present a novel way to debug executions.
+%
+First, we develop a notion of a
+\emph{reduction graph}~(\S~\ref{sec:reduction-graph}).
+%
+Second, we extend our semantics so that they incrementally
+collect the edges in the graph (\S~\ref{sec:inter-semant}).
+%
+Finally, we express a set of common debugging steps as
+graph traversals (\S~\ref{sec:traversing-graph}), yielding
+an novel visual interactive debugger that allows the user
+to effectively visualize \emph{how} the program goes (wrong.)
 
@@
 % traditional debuggers.
-
-In this section we formalize the collection and traversal of the reduction graph.
-%
-First, we extend our semantics from \S~\ref{sec:semantics} to collect
-the edges in the graph (\S~\ref{sec:inter-semant}).
-%
-Then, we express a set of common debugging steps as graph traversals
-(\S~\ref{sec:traversing-graph}).
 

### 2015-11-19 paper/pldi16-submission/interactive.tex
down to 11pp

@@
 
-A trace, on its own, is likely too detailed to be a good
-explanation of the type error. One approach is to use the
-witness input to step through the program with a
-\emph{debugger} to observe how the program evolves.
+A trace, on its own, is likely too detailed to be
+a good explanation of the type error. One approach
+is to use the witness input to step through the
+program with a \emph{debugger} to observe how
+the program evolves.
 %
@@
 %
-Second, we extend our semantics so that they incrementally
-collect the edges in the graph (\S~\ref{sec:inter-semant}).
+Second, we extend our semantics with a form of \emph{tracing}
+so that they incrementally collect the edges in the graph
+(\S~\ref{sec:inter-semant}).
 %
-Finally, we express a set of common debugging steps as
-graph traversals (\S~\ref{sec:traversing-graph}), yielding
+Finally, we express a set of common \emph{interactive debugging}
+steps as graph traversals (\S~\ref{sec:traversing-graph}), yielding
 an novel visual interactive debugger that allows the user
@@
 % traditional debuggers.
+
+\subsection{Reduction Graphs}
 
@@
 
-\subsection{Traversing the Reduction Graph}
+% \subsection{Traversing the Reduction Graph}
+\subsection{Interactive Debugging}
 \label{sec:traversing-graph}
+
 Given a reduction graph $G$ and an initial path

### 2015-11-19 paper/pldi16-submission/interactive.tex
down to 11pp

@@
 
-A trace, on its own, is likely too detailed to be
+A trace, on its own, is too detailed to be
 a good explanation of the type error. One approach
@@
 %
-First, we develop a notion of a
-\emph{reduction graph}~(\S~\ref{sec:reduction-graph}).
+First, we develop a notion of a \emph{reduction graphs}
+and extend our semantics with a form of \emph{tracing}
+so that they incrementally collect the edges
+in the graph (\S~\ref{sec:inter-semant}).
 %
-Second, we extend our semantics with a form of \emph{tracing}
-so that they incrementally collect the edges in the graph
-(\S~\ref{sec:inter-semant}).
-%
-Finally, we express a set of common \emph{interactive debugging}
+Next, we express a set of common \emph{interactive debugging}
 steps as graph traversals (\S~\ref{sec:traversing-graph}), yielding
-an novel visual interactive debugger that allows the user
+an novel interactive debugger that allows the user
 to effectively visualize \emph{how} the program goes (wrong.)
@@
 
-\subsection{Reduction Graphs}
-
 \subsection{Tracing Semantics}
@@
 \begin{figure*}[t]
-\relDescription{Trace Syntax}
-$$
-\begin{array}{rrcl}
-  & \tr & ::= & \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr
-\end{array}
-$$
-\\
-\relDescription{\subtermssym}
+% \relDescription{Trace Syntax}
+% $$
+% \begin{array}{rrcl}
+  % & \tr & ::= & \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr
+% \end{array}
+% $$
+% \\
+\relDescription{Computing Sub-Terms}
 \begin{gather*}
@@
 \end{gather*}
-\judgementHead{Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
+\judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
 \begin{gather*}
@@
 \end{figure*}
+
+\paragraph{Reduction Graphs}
 %
-The changes to the operational semantics of \S~\ref{sec:semantics} are
-mechanical, so we will not reproduce them in full; instead we describe
-the procedure for extending a transition rule and provide a selection of
-examples in Figure~\ref{fig:interactive}.
+A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
+intuitively indicates that $e_1$ steps, in a single step, to $e_2$.
+%
+A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
+intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
+%
+A \emph{reduction graph} is a set of edges:
+$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr$$
 
-First, we extend the transition relation to collect a set of edges from
-which we will construct the graph.
+\paragraph{Tracing Semantics}
 %
-An edge between two expressions is either a ``steps-to'' edge -- written
-\singlestep{e_1}{e_2} -- indicating that $e_1$ transitions to $e_2$ in a
-single step, or a ``sub-term'' edge -- written \subterm{e_1}{e_2} --
-indicating that $e_1$ contains $e_2$ as a sub-expression.
+We extend the transition relation (\S~\ref{sec:semantics}) to
+collect the set of edges corresponding to the reduction graph.
 %
-Collecting the steps-to edges is a simple matter of recording the
-consequent of each original rule in the trace; each original judgment
-\step{e_1}{\su_1}{e_2}{\su_2} becomes
-\stepg{e_1}{\su_1}{\tr_1}{e_2}{\su_2}{\singlestep{e_1}{e_2}; \tr_1}.
+Concretely, we extend the operational semantics to
+a relation of the form $\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\tr'}$
+where $\tr'$ collects the edges of the transition.
+
+\paragraph{Collecting Edges}
 %
-The sub-term edges can be delegated to a new \subtermssym helper
-function, which adds edges from an expression to each of its
+Next, we describe the general recipe for extending a transition
+rule to collect edges, and provide a selection of examples
+in Figure~\ref{fig:interactive}.
+%
+The steps-to edges are collected by recording the consequent of
+each original rule in the trace. That is, each original judgment
+\step{e}{\vsu}{e'}{\vsu'} becomes
+\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \tr}.
+%
+The sub-term edges are delegated to a helper function \subtermssym\
+which adds edges from an expression to each of its
 \emph{immediate} sub-expressions.
 %
-We collect new sub-term edges after each transition, thus the final
-template for the small-step relation is:
+We collect \subtermssym edges after each transition,
+to get the following template for the small-step relation:
 \[
-\stepg{e_1}{\su_1}{\tr_1}{e_2}{\su_2}{\singlestep{e_1}{e_2}; \subterms{e_2}; \tr_1}
+\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \subterms{e'}; \tr}
 \]
+\RJ{shouldn't this be $\subterms{e}$?}
 
-After evaluation the reduction graph can be constructed directly from
-the trace $\tr$ as follows:
-\[
-G(\tr) = \pair{\{e \spmid e \in \tr\}}{\tr}
-\]
+% After evaluation the reduction graph can be constructed directly from
+% the trace $\tr$ as follows:
+% \[
+% G(\tr) = \pair{\{e \spmid e \in \tr\}}{\tr}
+% \]
 

### 2015-11-19 paper/pldi16-submission/semantics.tex
down to 11pp

@@
   \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
-  \pair{v}{\ehole{i} \mapsto v}   & \mbox{o.w.} v = \gen{t}\\
+  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
 \end{cases} \\
@@
 
-Recall that a key technical challenge in generating witnesses is
+A key technical challenge in generating witnesses is
 that we have no (static) type information to rely upon. Thus, we
@@
 %
-We obtain the following as a corrollary of Theorem~\ref{thm:soundness}:
+We obtain the following corollary of Theorem~\ref{thm:soundness}:
 

### 2015-11-19 paper/pldi16-submission/interactive.tex
down to 11pp

@@
 \]
-\RJ{shouldn't this be $\subterms{e}$?}
 
@@
 
-Given a reduction graph $G$ and an initial path
-$p = \singlestep{e_1}{\singlestep{e_2}{e_n}}$, such that
-\stepsg{e_1}{\emptysu}{\bullet}{e_n}{\su}{\tr}, we define the following
-traversals in Figure~\ref{fig:traversing-graph}:
+Next, we show how to build an visual interactive debugger
+from the traced semantics, by describing the visualization
+\emph{state} \ie\ what the user sees at any given moment,
+the set of \emph{commands} available to user, and what
+they do, and finally how we use a command to \emph{update}
+the visualization state. In the sequel, for clarity of
+exposition, we assume we have a (global) trace:
+$\stepg{e}{\emptysu}{\bullet}{e'}{\vsu'}{\tr}$.
+
+\paragraph{State}
+A \emph{visualization state} is ...
+The \emph{initial state} is the graph containing the single edge
+${e \hookrightarrow \stuck}$.
++ Visualization = ExprGraph + invariants (unique predecessor) -- link to figure.
+
+\paragraph{Commands}
+\begin{itemize}
+\item \stepforwardsym\ takes a single step forward.
+\item \stepbackwardsym\ takes a single step backward.
+\item \jumpforwardsym\ takes a ``big'' step forward to the next function call.
+\item \jumpbackwardsym\ takes a ``big'' step backward to the previous function call.
+\item \stepintosym\ steps into a function call in a sub-term, isolating it from the context.
+\item \stepoversym\ steps over a function call in a sub-term.
+\end{itemize}
+
+\paragraph{Update}
+Figure~\ref{fig:traversing-graph} shows how we update the state for each
+command...
 %
-\begin{itemize}
-\item \stepforwardsym takes a single step forward.
-\item \stepbackwardsym takes a single step backward.
-\item \jumpforwardsym takes a ``big'' step forward to the next function call.
-\item \jumpbackwardsym takes a ``big'' step backward to the previous function call.
-\item \stepintosym steps into a function call in a sub-term, isolating it from the context.
-\item \stepoversym steps over a function call in a sub-term.
-\end{itemize}
+find   :: Vis -> Cmd -> Expr
+update :: Vis -> Cmd -> Vis
 %
@@
 %
-The sub-term edges $\searrow$ allow us to decompose an expression into a
+Note that the sub-term edges $\searrow$ allow us to decompose an expression into a
 sub-expression and the surrounding context, thus enabling the \stepintosym
 and \stepoversym traversals.
-%
+
+data Sel = Expr
+
+data Vis = { graph :: TraceGraph, edges = ExprGraph  }
+
+path :: Vis -> Expr -> [Expr]
+stepX :: Sel -> Vis -> Vis(edited)
+
+path v e = graphPath (graph v,  root (v, e),  e)
+
+putBefore, putAfter :: EGraph -> Expr -> Expr -> EGraph
+putRoot  :: EGraph -> Expr -> EGraph
+getRoot  :: EGraph -> Expr -> Expr
+
+pathToRoot  ::
 
@@
 \end{figure*}
-
-% \begin{itemize}
-% \item extend operational semantics to collect reduction graph
-% \item nodes are terms, edges indicate ``steps-to'' and ``sub-term'' relations
-% \item visualize path through reduction graph
-% \item expand edges to reveal more fine-grained steps (step/jump forward/backward)
-% \item never lose context (unlike traditional debugger)
-% \end{itemize}

### 2015-11-19 paper/pldi16-submission/interactive.tex
down to 11pp

@@
 exposition, we assume we have a (global) trace:
-$\stepg{e}{\emptysu}{\bullet}{e'}{\vsu'}{\tr}$.
+$\stepg{e_0}{\emptysu}{\bullet}{e_n}{\_}{\tr}$, where
+$e_0$ and $e_n$ are the \emph{initial} and \emph{final}
+expressions respectively.
 
-\paragraph{State}
-A \emph{visualization state} is ...
-The \emph{initial state} is the graph containing the single edge
-${e \hookrightarrow \stuck}$.
-+ Visualization = ExprGraph + invariants (unique predecessor) -- link to figure.
+\paragraph{Visualization State}
+%
+A \emph{visualization state} $\vstate$ is a \emph{directed graph}
+whose vertices are expressions and whose edges are such
+that each vertex has at most one predecessor and at most one
+successor. In other words, the visualization state looks
+like a set of linear lists of expressions as shown in Figure~\ref{fig:FIXTHIS}.
+%
+The \emph{initial state} is the graph containing a single
+edge linking the initial and final expressions.
 
 \paragraph{Commands}
+Our debugger supports the following \emph{commands}, each of which
+is parameterized by a single expression (vertex) selected from the
+(current) visualization state:
+%
 \begin{itemize}
-\item \stepforwardsym\ takes a single step forward.
-\item \stepbackwardsym\ takes a single step backward.
-\item \jumpforwardsym\ takes a ``big'' step forward to the next function call.
-\item \jumpbackwardsym\ takes a ``big'' step backward to the previous function call.
-\item \stepintosym\ steps into a function call in a sub-term, isolating it from the context.
-\item \stepoversym\ steps over a function call in a sub-term.
+%
+\item \stepforwardsym, \stepbackwardsym:
+      show result of a single step forward or backward respectively,
+%
+\item \jumpforwardsym:
+      show result of taking multiple steps (a \emph{``big''} step)
+      upto the first beta-reduction forward or backward respectively,
+%
+\item \stepintosym:
+      show result of stepping into a function call in a sub-term,
+      isolating it from the context,
+
+\item \stepoversym:
+      show result of skipping over a function call in a sub-term.
 \end{itemize}
@@
 \paragraph{Update}
-Figure~\ref{fig:traversing-graph} shows how we update the state for each
-command...
+Figure~\ref{fig:traversing-graph} shows how we update the state
+for each command.
 %
-find   :: Vis -> Cmd -> Expr
-update :: Vis -> Cmd -> Vis
+The procedure $\findExpr{\vstate}{\cmd}{e}$ traverses the $\tr$
+and the current visualization state to compute the new
+expression that should be added to the visualization state
+and $\updState{\vstate}{\cmd}{e}$ then updates the graph
+by inserting the new expression appropriately, using one of
+the following functions.
 %
-The initial path $p$ is required for the backward-steps as a node may
-have multiple incoming $\leadsto$ edges, \eg
-\singlestep{\eplus{1}{2}}{3} and \singlestep{\eplus{2}{1}}{3}.
+$\putBefore{\vstate}{e}{e'}$ (resp. $\putAfter{\vstate}{e}{e'}$)
+returns the modified version of $\vstate$ where $e$ is the
+immediate predecessor of $e'$ (resp. $e'$ the immediate successor of $e$);
 %
-Note that the sub-term edges $\searrow$ allow us to decompose an expression into a
-sub-expression and the surrounding context, thus enabling the \stepintosym
-and \stepoversym traversals.
+$\putRoot{\vstate}{e}$ returns the modified version of $\vstate$ extended
+with a new vertex $e$ that has no predecessor or successor, and,
+%
+$\getRoot{\vstate}{e'}$ be the vertex (expression) obtained
+by transitively following the predecessors of $e'$ until a
+source vertex.
+%
+Note that the sub-term edges $\searrow$ allow us to decompose an
+expression into a sub-expression and the surrounding context, thus
+enabling the \stepintosym\ and \stepoversym\ traversals.
 
-data Sel = Expr
+\RJ{EXAMPLE of interaction from overview goes here}
 
-data Vis = { graph :: TraceGraph, edges = ExprGraph  }
-
-path :: Vis -> Expr -> [Expr]
-stepX :: Sel -> Vis -> Vis(edited)
-
-path v e = graphPath (graph v,  root (v, e),  e)
-
-putBefore, putAfter :: EGraph -> Expr -> Expr -> EGraph
-putRoot  :: EGraph -> Expr -> EGraph
-getRoot  :: EGraph -> Expr -> Expr
-
-pathToRoot  ::
-
-\begin{figure*}[t]
-\centering
-\[
-\begin{array}{lcl}
-\stepforward{G}{p}{e_i}  &\defeq& \left\{\begin{array}{ll}
-    e_j, & \text{where } \singlestep{e_i}{e_j} \in G
-                         \end{array}\right\} \\ \\
-\stepbackward{G}{p}{e_i}  &\defeq& \left\{\begin{array}{ll}
-    e_j, & \text{where } \singlestep{e_j}{e_i} \in G \text{ and } e_j \in p
-                         \end{array}\right\} \\ \\
-\jumpforward{G}{p}{e_i} &\defeq& \text{let } e_j = \stepforward{G}{p}{e_i} \text{ in }
-                         \left\{\begin{array}{ll}
-                         e_j, & \text{if } e_j = \eapp{v_1}{v_2} \\
-                         \jumpforward{G}{p}{e_{j}}, & \text{otherwise}
-                         \end{array}\right\} \\ \\
-\jumpbackward{G}{p}{e_i} &\defeq& \text{let } e_j = \stepbackward{G}{p}{e_i} \text{ in }
-                         \left\{\begin{array}{ll}
-                         e_j, & \text{if } e_j = \eapp{v_1}{v_2} \\
-                         \jumpbackward{G}{p}{e_{j}}, & \text{otherwise}
-                         \end{array}\right\} \\ \\
-\stepinto{G}{p}{e_i} &\defeq& \left\{\begin{array}{ll}
-                         e\sub{x}{v_2}, & \text{if } e_i = C[\eapp{v_1}{v_2}] \text{ and } \singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}
-                         \end{array}\right\} \\ \\
-\stepover{G}{p}{e_i} &\defeq& \left\{\begin{array}{ll}
-                         C[v], & \text{if } e_i = C[\eapp{v_1}{v_2}] \text{ and } \multistep{\eapp{v_1}{v_2}}{v}
-                         \end{array}\right\}
-\end{array}
-\]
-\caption{Rules for traversing the reduction graph given a path and
-  node. \stepintosym and \stepoversym require a traversal of the
-  sub-term edges to decompose $e_i$ into the target expression
-  \eapp{v_1}{v_2} and the context $C$.  \ES{these rules are quite ugly and waste space..} }
-\label{fig:traversing-graph}
-\end{figure*}
+%
+%
+%The initial path $p$ is required for the backward-steps as a node may
+%have multiple incoming $\leadsto$ edges, \eg
+%\singlestep{\eplus{1}{2}}{3} and \singlestep{\eplus{2}{1}}{3}.
+%

### 2015-11-19 paper/pldi16-submission/haskellListings.tex
down to 11pp
done with abstract

@@
 		isIEEE,isInfinite,isLower,isNaN,isNegativeZero,isOctDigit,isPrint,isSpace,isUpper,iterate,
-		last,lcm,length,lex,lexDigits,lexLitChar,lines,log,logBase,lookup,map,mapM,mapM_,max,
+		last,lcm,length,lex,lexDigits,lexLitChar,lines,log,logBase,lookup,mapM,mapM_,max,
 		maxBound,maximum,maybe,min,minBound,minimum,mod,negate,not,notElem,numerator,odd,
@@
 		quotRem,range,rangeSize,read,readDec,readFile,readFloat,readHex,readIO,readInt,readList,readLitChar,
-		readLn,readOct,readParen,readSigned,reads,readsPrec,realToFrac,recip,rem,repeat,replicate,
+		readLn,readOct,readParen,readSigned,reads,readsPrec,realToFrac,recip,rem,repeat,
 		reverse,round,scaleFloat,scanl,scanl1,scanr,scanr1,seq,sequence,sequence_,show,showChar,showInt,
@@
                 GCtors,GDecode,GDecodeFields,GEncode,GEncodeFields,OrdMap,MinusKey,
-                len,isBH,isBal,bh,isRB,keys,List,Sorted,RBT,Col,isBlack,OrdRBT,Set,sz,
+                len,isBH,isBal,bh,isRB,keys,Sorted,RBT,Col,isBlack,OrdRBT,Set,sz,
                 StackSet,NoDuplicates,Data,RBTree,XMonad,Generic,true

### 2015-11-19 paper/pldi16-submission/main.tex
done with abstract

--- 2015-11-17 paper/pldi16-submission/main.tex
@@
 
+\usepackage{pgfplots}
+\usepackage{pgfkeys}
+\pgfplotsset{compat=1.12}
+
 \usepackage[inference]{semantic}
@@
 
-\authorinfo{Eric L. Seidel\and Ranjit Jhala}
+\authorinfo{Eric L. Seidel}
            {UC San Diego}
-           {{eseidel,rjhala}@cs.ucsd.edu}
+           {eseidel@cs.ucsd.edu}
+
+\authorinfo{Ranjit Jhala}
+           {UC San Diego}
+           {jhala@cs.ucsd.edu}
 
@@
 \maketitle
-\begin{abstract}
-\end{abstract}
 
-% 11 pages total (not including bib)
-% ----------------------------------
-%  1p : intro
-%  2p : overview
-%  2p : type-carrying semantics
-%  1p : evaluation: quickchecking type errors
-%  2p : interactive semantics
-%  1p : evaluation: nanomaly
-%  1p : related work
-%  1p : conclusion
+\input{abstract}
+\input{introduction}
+\input{overview}
+\input{semantics}
+% \section{interactive}
+\input{evaluation}
+\input{related}
 
-\section{Introduction}          % 1 page
-\section{Overview}              % 2 pages?
-% \section{Type-carrying Semantics} % 2 pages
-\input{semantics.tex}
-
-% \section{Evalution: Recasting Type Errors as Runtime Errors} % 1 page
-\input{evaluation-witness.tex}
-
-%\section{``Interactive'' Semantics \& Trace Exploration} % 2 pages
-\input{interactive.tex}
-
-\input{nanomaly.tex}
-
-% \section{Evaluation: \nanomaly}                % 1 page
-\input{evaluation-nanomaly.tex}
-
-\section{Related Work}              % 1 pages
-
-\subsection{Type Error Localization}
-\label{sec:type-error-local}
-\begin{itemize}
-\item seminal
-\item bayesian (Zhang+Myers)
-\item MaxSMT
-\item counterfactual (Chen+Erwig)
-\end{itemize}
-\subsection{Debugging \& Program Understanding}
-\label{sec:debugging}
-
-\section{Conclusion}                % 1 page
+{
+\bibliographystyle{abbrvnat}
+\bibliography{main}
+}
 

### 2015-11-19 paper/pldi16-submission/overview.tex
done with abstract

@@
 \label{sec:overview}
-We start with an overview of our contributions.
+
+We start with an overview of our approach to
+explaining (static) type errors using \emph{witnesses}
+that (dynamically) show how the program goes wrong.
 %
-First we illustrate why generating suitable inputs to functions in the
-absense of type information is tricky. 
+We illustrate why generating suitable inputs
+to functions is tricky in the absence of type
+information.
 %
-Then we describe our solution to the problem and highlight the
-similarity to static type inference.
+Then we describe our solution to the problem
+and highlight the similarity to static type
+inference,
 %
@@
 \label{sec:generating-witnesses}
-Our goal is to find concrete values that demonstrate how a
-program ``goes wrong''.
+Our goal is to find concrete values
+that demonstrate how a program ``goes wrong''.
 
@@
 %
-One approach is to randomly generate input values and use them to
-execute the program until we find one that causes the program to go
-wrong. However, to see why this approach is naive, consider the
-@fac@ function from Figure~\ref{fig:factorial}, reproduced here.
+One approach is to randomly generate input values and
+use them to execute the program until we find one that
+causes the program to go wrong.
+%
+Unfortunately, this approach quickly runs aground.
+Consider the @fac@ function from Figure~\ref{fig:factorial},
+reproduced here.
 
@@
 %
-A hole -- written \ehole{} -- is a \emph{placeholder} for a value; in
+A hole --- written \ehole{} --- is a \emph{placeholder} for a value; in
 our case we do not even know its type.
@@
 
-As we show in \S~\ref{sec:soundness} our lazy instantiation of holes is
-\emph{conservative}.
+\paragraph{Witness Generality}
+We show in \S~\ref{sec:soundness} our lazy instantiation of holes
+produces \emph{general witnesses}.
 %
-We prove that if calling a function with a hole as input causes the
-function to ``go wrong'', then there is \emph{no possible} type for the
-function.
+That is, we show that if ``executing''
+a function with a hole as input causes the
+function to ``go wrong'', then there is
+\emph{no possible} type for the function.
 %
-In other words, for any type you might assign to the function, there
-exist valid inputs that will cause it to crash.
+In other words, for \emph{any} type you might
+assign to the function, there exist inhabitants
+that will cause the function to go wrong.
 
 \paragraph{Problem: How many inputs does a function take?}
-There is another wrinkle though, how did we know that @fac@ takes
-a single argument instead of two (or none)?
+%
+There is another wrinkle though, how did we know
+that @fac@ takes a single argument instead of two
+(or none)?
 %
@@
 As we are dealing with ill-typed programs we do not have the luxury of
-typing information, so we must deduce the number of arguments by
-experimentation.
+typing information.
 
 \paragraph{Solution: Search for saturated application.}
-Luckily the solution is straightforward, we add arguments one-by-one
-until we reach a \emph{saturated} application, \ie until evaluating the
-application returns a value other than a lambda.
+We solve this problem by deducing the number of arguments
+via an iterative process. We add arguments one-by-one
+until we reach a \emph{saturated} application, \ie\
+until evaluating the application returns a value
+other than a lambda.
 
@@
 computation by selecting a visible term and choosing one of the
-applicable traversal strategies -- formalized in
-\S~\ref{sec:traversing-graph} -- to insert another term into the
+applicable traversal strategies --- formalized in
+\S~\ref{sec:explaining} --- to insert another term into the
 visualization.

### 2015-11-19 paper/pldi16-submission/abstract.tex
done with abstract

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we first symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  %
  Second, we prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} input types, there exist
  inhabitants that can make the function go wrong.
  %
  Third, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Fourth, we evaluate our approach on two data sets
  comprising more than 5000 ill-typed student programs.
  Our technique is able to generate witnesses for
  nearly 90\% of the programs, and our reduction graph
  is yields small counterexamples for 90\% of the witnesses.
\end{abstract}

### 2015-11-20 paper/pldi16-submission/interactive.tex
update graph api/update figure

@@
 %
-A \emph{visualization state} $\vstate$ is a \emph{directed graph}
+A \emph{visualization state} @VState@ is a \emph{directed graph}
 whose vertices are expressions and whose edges are such
@@
 
+\input{traversals.tex}
+
 \paragraph{Update}
@@
 %
-The procedure $\findExpr{\vstate}{\cmd}{e}$ traverses the $\tr$
+The procedure @findExpr vs cmd e@ traverses the $\tr$
 and the current visualization state to compute the new
 expression that should be added to the visualization state
-and $\updState{\vstate}{\cmd}{e}$ then updates the graph
+and @updState vs cmd e@ then updates the graph
 by inserting the new expression appropriately, using one of
@@
 %
-$\putBefore{\vstate}{e}{e'}$ (resp. $\putAfter{\vstate}{e}{e'}$)
-returns the modified version of $\vstate$ where $e$ is the
-immediate predecessor of $e'$ (resp. $e'$ the immediate successor of $e$);
+@putBefore vs e e'@ \hbox{(resp. @putAfter vs e e'@)}
+returns the modified version of @vs@ where @e'@ is the
+immediate predecessor of @e@ (resp. the immediate successor of @e@);
 %
-$\putRoot{\vstate}{e}$ returns the modified version of $\vstate$ extended
-with a new vertex $e$ that has no predecessor or successor, and,
+@putRoot vs e@ returns the modified version of @vs@ extended
+with a new vertex @e@ that has no predecessor or successor, and,
 %
-$\getRoot{\vstate}{e'}$ be the vertex (expression) obtained
-by transitively following the predecessors of $e'$ until a
+@getRoot vs e@ returns the vertex (expression) obtained
+by transitively following the predecessors of @e@ until a
 source vertex.
+%
+@getSubterms vs e@ traverses the sub-term edges to decompose an
+expression into a list of sub-expressions paired with their context.
+%
+@applyCtx vs e ctx@ applies @ctx@ to @e@, traversing the sub-term edges
+in reverse to find the super-term of @e@.
+%
+@findApp vs e@ builds on top of @getSubterms@ to find the first
+application sub-term (if any), \ie the first sub-term that looks like
+$\eapp{v_1}{v_2}$.
+%
+@findValue vs e@ traverses the single-step edges to find the final value
+that @e@ reduces to.
 %

### 2015-11-20 paper/pldi16-submission/evaluation.tex
hide UCSD for anonymity

@@
 The first dataset comes from the Spring 2014 undergraduate Programming
-Languages course at UC San Diego. 
+Languages course at our institution.
 %
 We recorded each interaction with the \ocaml top-level system over the
-course of the first three assignments (IRB \#140608), from which we
-extracted 4,407 distinct, ill-typed \ocaml programs.
+course of the first three assignments (IRB \# hidden for blind review), % \#140608),
+from which we extracted 4,407 distinct, ill-typed \ocaml programs.
 %
@@
 %
-The average number of single-step reductions per trace is 26 for the
-UCSD dataset (27 for the Seminal dataset) with a maximum of 2,744 (659
-for Seminal). The average number of jumps per trace is 3 (also 3 for
-Seminal) with a maximium of 177 (52 for Seminal). In both datasets at
+The average number of single-step reductions per trace is 26 for our
+dataset (27 for the UW dataset) with a maximum of 2,744 (659 for
+UW). The average number of jumps per trace is 3 (also 3 for
+UW) with a maximium of 177 (52 for UW). In both datasets at
 least 90\% of the traces contain fewer than 10 jumps.

### 2015-11-20 paper/pldi16-submission/plots.py

@@ -3,6 +3,8 @@ import matplotlib
 import matplotlib.pyplot as plt
 import numpy as np
 
+UCSD = 'AUTHORS'
+
 BUCKETS = [0.1, 0.2, 1.0, 10.0, 60.0 ] # range(500, 3001, 500)
 COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
 
@@ -58,7 +60,7 @@ def plot_coverage(seminal, ucsd):
     plt.title('Cumulative Coverage', fontsize=24)
     plt.xticks(ind + width, [r[0] for r in xy_s], fontsize='large')
     plt.yticks(np.arange(0, 101, 10), fontsize='large')
-    plt.legend((p1[0], p2[0]), ('Seminal', 'UCSD'), 'lower right', fontsize=16)
+    plt.legend((p1[0], p2[0]), ('UW', UCSD), 'lower right', fontsize=16)
     # plt.legend((p1[0], p2[0]), ('Men', 'Women'))
     autolabel(plt, p1)
     autolabel(plt, p2)
@@ -98,11 +100,11 @@ def plot_trace_size(seminal, ucsd):
 
     fig = plt.figure()
     y,binEdges=np.histogram(step_s,bins=bins)
-    p1 = plt.bar(ind, y / float(len(step_s)), label='Seminal', width=width, color=COLORS[0])
+    p1 = plt.bar(ind, y / float(len(step_s)), label='UW', width=width, color=COLORS[0])
 
     y,binEdges=np.histogram(step_u,bins=bins)
-    p2 = plt.bar(ind + width, y / float(len(step_u)), label='UCSD', width=width, color=COLORS[1])
-    plt.legend((p1[0],p2[0]), ('Seminal','UCSD'), fontsize=16)
+    p2 = plt.bar(ind + width, y / float(len(step_u)), label=UCSD, width=width, color=COLORS[1])
+    plt.legend((p1[0],p2[0]), ('UW',UCSD), fontsize=16)
     plt.title('Trace Complexity', fontsize=24)
     plt.xlabel('Total Steps', fontsize=20)
     plt.ylabel('Traces (%)', fontsize=20)
@@ -127,13 +129,13 @@ def plot_trace_size(seminal, ucsd):
     y,binEdges=np.histogram(jump_s,bins=bins)
     foo = y / float(len(jump_s)) 
     print(foo[0] + foo[1])
[…]

### 2015-11-20 paper/pldi16-submission/abstract.tex
done with abstract

@@
   that can make the program go wrong.
-  %
-  Second, we prove that our procedure synthesizes
+  We prove that our procedure synthesizes
   \emph{general witnesses} in that if a witness is
@@
   %
-  Third, we show how to extend the above procedure to
+  Second, we show how to extend the above procedure to
   produce a \emph{reduction graph} that can be used to
@@
   %
-  Fourth, we evaluate our approach on two data sets
-  comprising more than 5000 ill-typed student programs.
+  Third, we evaluate our approach on two data sets
+  comprising more than 4500 ill-typed student programs.
   Our technique is able to generate witnesses for

### 2015-11-20 paper/pldi16-submission/evaluation.tex
done with abstract

--- 2015-11-19 paper/pldi16-submission/evaluation.tex
@@
 course of the first three assignments (IRB \#140608), from which we
-extracted 4,407 distinct, ill-typed \ocaml programs.
+extracted \ucsdsize\ distinct, ill-typed \ocaml programs.
 %

### 2015-11-20 paper/pldi16-submission/introduction.tex
done with abstract

@@
 
-% \paragraph{Problem}
-Type errors are a common stumbling block for students trying to learn
-typed functional languages like \ocaml and \haskell.
+Type errors are a common stumbling block for students
+trying to learn typed functional languages like \ocaml\
+and \haskell.
 %
-The compiler imposes a strict discipline that prevents novices from
-running their programs until the type-checker has been appeased.
+Consider the ill-typed @fac@ function on the left in
+Figure~\ref{fig:factorial}.
 %
-In effect this prioritizes the \emph{static} semantics of the language
-over the \emph{dynamic} semantics.
+The function returns @true@ in the base case (instead of @1@).
+and so \ocaml's responds with the error message:
 %
-This is fine for expert users who are already familiar with the language
-and type system, but seems backwards for novices, who are typically
-introduced to the dynamic semantics first.
+\begin{verbatim}
+  File "fac.ml", line 5, characters 8-23:
+  Error: This expression has type bool
+         but an expression was expected
+         of type int
+\end{verbatim}
 
-Consider the following ill-typed @fac@ function, where we return
-@true@ in the base case instead of @1@.
+This message is makes perfect sense to an expert who is familiar
+with the language and has a good mental model of how the type
+system works.
 %
-\begin{figure}[h!]
+However, it may perplex a novice who is still to develop such a
+mental model.
+%
+To make matters worse, unification based type inference algorithms
+often report errors far removed from their source.
+%
+This further increases the novice's confusion and can actively mislead
+them to focus their investigation on an irrelevant piece of code.
+%
+Much recent work has focused on analyzing unification constraints
+to properly \emph{localize} a type error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
+but an accurate source location still does not explain \emph{why} the
+program is wrong.
+
+\begin{figure}[t]
+\centering
+\begin{minipage}{.49\linewidth}
 \begin{code}
@@
 \end{code}
-\caption{Ill-typed \texttt{fac} function that mistakenly
-  returns \texttt{true} in the base case instead of \texttt{1}.}
+\end{minipage}
+\hspace{0.3in}
+\begin{minipage}{.3\linewidth}
+  \includegraphics[height=2.5in]{fac-overview.png}
+\end{minipage}
+\caption{(L) An ill-typed \texttt{fac} function; (R) Dynamically witnessing the type error in \texttt{fac}}
 \label{fig:factorial}
@@
 
-\ocaml's response is to tell us that on line 5, the recursive call to
-@fac@ returns a @bool@, but we needed an @int@ for the
-multiplication.
+In this paper we propose a new approach to explaining type
+\emph{static} type errors: by \emph{dynamically} witnessing
+how exactly an ill-typed program goes wrong.
 %
-\begin{verbatim}
-File "fac.ml", line 5, characters 8-23:
-Error: This expression has type bool
-       but an expression was expected
-       of type int
-\end{verbatim}
+We have developed \toolname\ an interactive tool which uses
+the source of the ill-typed function to automatically synthesize
+the result on the right in Figure~\ref{fig:factorial}, which
+shows how the recursive calls reduce to a configuration where
+the program ``goes wrong'' \ie\ the @int@ value @1@ is to be
+multiplied with the @bool@ value @true@.
+We achieve this via three concrete contributions.
+
+\paragraph{1. Finding Witnesses}
+Our first contribution is an algorithm for searching for
+\emph{witnesses} to type errors, \ie\ inputs that cause a
+program to go wrong~(\S~\ref{sec:semantics}).
 %
-\ocaml is correct, but we may ask \emph{why} @fac@ returns a
-@bool@ instead of an @int@.
-
-Contrast this \emph{static} type error to the \emph{dynamic} type error
-that arises in the equivalent \ruby function, when we call
-@fac(1)@.
+This problem is surprisingly tricky when we cannot rely on
+static type information.
 %
-\begin{verbatim}
-fac.rb:5:in `*': true can't be coerced
-                 into Fixnum (TypeError)
-  from fac.rb:5:in `fac'
-  from fac.rb:10:in `<main>'
-\end{verbatim}
+In particular, we must avoid the trap of \emph{spurious} inputs
+that cause irrelevant problems that would be avoided by picking
+values of a different, relevant type.
 %
-The same basic information is there, @*@ was given a @bool@ (@true@)
-when it expected an @int@ (@FixNum@), but since the error occurred at
-runtime, \ruby is able to provide a stack trace that provides important
-contextual information about the error.
-%
-More importantly, being able to \emph{run} the bad program means we can
-also test it on other inputs, \eg @0@, or we can load it into a debugger
-and step through the execution to see concretely how the error arises.
-
-To make matters worse, it is well known that type inference algorithms
-based on unification have a penchant for \emph{reporting} errors far
-from their source.
-%
-This further increases the novice's confusion and can actively mislead
-them to focus their investigation on an irrelevant piece of code.
-%
-Much work has been done on pinpointing the \emph{source} of a type
-error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
-but an accurate source location still does not explain \emph{why} the
-program is wrong.
-
-% \paragraph{Contributions}
-In this paper we propose exposing \emph{static} type errors for the
-\emph{runtime} crashes that they are.
-%
-Our first contribution is an algorithm for searching for
-\emph{witnesses} to type errors, \ie input vectors that cause a program
-to crash with a type error.
-%
-This is not as trivial as it sounds.
-%
-While one can indeed execute an \ocaml program without type-checking it
--- the dynamic semantics do not depend on the static semantics -- what
-types of values should one pass to the program?
-%
-We do not want to generate \emph{spurious} witnesses, inputs that the
-type-checker would have excluded, so we introduce a semantics for
-executing \ocaml programs with ``holes'', values whose type we do not
-yet know.
-%
-% A hole can be passed around through the program until it is demanded by
-% a primitive operation, \eg addition, function application, branching,
-% etc., at which point we know from the context what type it must have,
-% and can instantiate it with a concrete value.
+We solve this problem by developing a novel operational semantics
+that corresponds to executing the program with \emph{holes}
+values whose type is unknown.
 %
 Our semantics \emph{conservatively} instantiates holes with concrete
-values; if we can find a witness to a type error then there is no valid
-typing for the program.
+values; thereby dynamically inferring the type of the input type
+until the point the program goes wrong.
+%
+We prove that our procedure synthesizes \emph{general witnesses},
+which means, intuitively, that if a witness is found for a given
+ill-typed function, then \emph{for all} input types, there exist
+inhabitants that can make the function go wrong.
 
-Given a witness to a type error, the prospective \ocaml student may
-still be at a loss.
+Given a witness to a type error, the novice may still be at a loss.
 %
-The standard \ocaml interpreter and debugging infrastructure expect
+The standard \ocaml\ interpreter and debugging infrastructure expect
 well-typed programs, so they cannot be used to investigate \emph{how}
@@
 %
-Thus, our second contribution is an interactive visualization of the
-execution of \ocaml programs (well-typed or not).
+More importantly, the execution itself may be quite long and may contain
+details not relevant to the actual error.
+
+\paragraph{2. Visualizing Witnesses}
+Our second contribution is a novel interactive visualization of the
+execution of \ocaml\ programs well-typed or not~(\S~\ref{sec:interactive}).
 %
-We present the execution as a \emph{reduction graph}, which records
-all of the small-step reductions in the context in which they occur.
+We extend the semantics to also build a \emph{reduction graph}
+which records all of the small-step reductions and the context
+in which they occur.
 %
-%  in which the nodes
-% are expressions and the edges indicate either a single-step reduction or
-% a sub-term relation.
+The graph lets us build a visualization that shows a sequence of
+steps from the source witness to the stuck term and allows the user
+to interactively expand the  computation to expose intermediate steps,
+by selecting an expression and choosing a traversal strategy.
 %
-This graph is easy to generate during the execution and admits many of
-the standard debugging steps, \eg ``step forward'', ``step into'', and
-``step over'', as well as backwards motion and jumping to the next (or
-previous) function call.
+The strategies include many of the standard debugging moves, \eg\
+stepping \emph{forward} or \emph{into} or \emph{over} calls as well
+stepping or jumping \emph{backward} to understand how a particular
+value was created, while preserving a context of the intermediate
+steps that allow the user to keep track of a term's provenance.
+
+% This process of iterative expansion of the trace ensures that
+% the context of intermediate steps is not lost; whereas traditional
+% debuggers force one to move between statements, in our visualization
+% you never lose sight of where you came from.
+
+Obviously, the problem of finding witnesses is undecidable.
+In fact, due to the necessarily conservative nature of static
+typing, there may not even exist. Thus, our heuristic approach
+is only useful if it can find \emph{compact} witnesses for
+\emph{real-world} programs.
+
+\paragraph{3. Evaluating Witnesses}
 %
-The visualization initially shows the source term (\ie the witness) and
-the stuck term (or the final value), and allows the user to iteratively
-expand the computation to expose intermediate steps, by selecting an
-expression and choosing a traversal strategy (\eg ``step forward'').
+Our third contribution is an extensive evaluation of our approach
+on two different sets of ill-typed programs obtained by instrumenting
+compilers used in beginner's classes~(\S~\ref{sec:eval}).
 %
-This process of iterative expansion of the trace ensures that the
-context of intermediate steps is not lost; whereas traditional debuggers
-force one to move between statements, in our visualization you never
-lose sight of where you came from.
+The first is the \uwbench\ data set~\cite{lerner_searching_2007}
+comprising \uwsize\ ill-typed programs.
+%
+The second is a new \ucsdbench\ data set, comprising \ucsdsize\
+ill-typed programs.
+%
+We show that for both data sets, our technique is able to generate
+witnesses for nearly 90\% of the programs, in under a second the
+vast majority of cases.
+%
+Furthermore, we show that a simple interactive strategy yields
+compact counterexample traces with fewer than 5 steps for 60\%
+of the programs, and under 10 steps for 90\% of the programs.
 
-\paragraph{Outline}
-We begin the rest of the paper with an overview of our search algorithm
-and trace visualization (\S~\ref{sec:overview}).
-%
+In short, our results show that at least for beginners, ill-typed
+programs \emph{do} go wrong, in the vast majority of cases, which,
+in turn, opens the door to a completely different, dynamic way
+to explain, understand, and appreciate the benefits of static typing.
+
+%% Then, we formalize our algorithm and prove it sound
+%% (\S~\ref{sec:searching-witness}).
+%% %
+%% Next, we extend the algorithm to collect a reduction graph and formalize
+%% a set of traversals over the graph (\S~\ref{sec:explaining}).
+%% %
+%% Finally, we evaluate our search algorithm and trace generation on a
+%% large set of real student programs (\S~\ref{sec:evaluation}).
+%%
+%% \paragraph{Outline}
+%% We begin the rest of the paper with an overview of our search algorithm
+%% and trace visualization (\S~\ref{sec:overview}).
+
 % Then, we formalize our algorithm and prove it sound
@@
 % (\S~\ref{sec:evaluation-nanomaly}).
-Then, we formalize our algorithm and prove it sound
-(\S~\ref{sec:searching-witness}).
-%
-Next, we extend the algorithm to collect a reduction graph and formalize
-a set of traversals over the graph (\S~\ref{sec:explaining}).
-%
-Finally, we evaluate our search algorithm and trace generation on a
-large set of real student programs (\S~\ref{sec:evaluation}).
 %

### 2015-11-20 paper/pldi16-submission/interactive.tex
ugh, more graph apis

@@
 %
-The procedure @findExpr vs cmd e@ traverses the $\tr$
+The procedure @findExpr vs cmd e@ traverses $\tr$
 and the current visualization state to compute the new
@@
 %
-@putRoot vs e@ returns the modified version of @vs@ extended
-with a new vertex @e@ that has no predecessor or successor, and,
+@putRoot vs e e'@ returns the modified version of @vs@ extended
+with a new root vertex @e@ and its successor @e'@, and,
 %
-@getRoot vs e@ returns the vertex (expression) obtained
-by transitively following the predecessors of @e@ until a
-source vertex.
+% @getRoot vs e@ returns the vertex (expression) obtained
+% by transitively following the predecessors of @e@ until a
+% source vertex.
+%
+@getNext vs e@ returns the immediate successor of @e@ in $\tr$.
+%
+@getPrev vs e@ computes the path @p@ between @e@ and its immediate
+predecessor in the current visualization, and then returns @e@'s immediate
+predecessor along @p@.
 %
@@
 %
-@findApp vs e@ builds on top of @getSubterms@ to find the first
+\hbox{@findApp vs e@} builds on top of @getSubterms@ to find the first
 application sub-term (if any), \ie the first sub-term that looks like
@@
 %
-@findValue vs e@ traverses the single-step edges to find the final value
+@findVal vs e@ traverses the single-step edges to find the final value
 that @e@ reduces to.

### 2015-11-20 paper/pldi16-submission/introduction.tex
done with intro

@@
 
-% This process of iterative expansion of the trace ensures that
-% the context of intermediate steps is not lost; whereas traditional
-% debuggers force one to move between statements, in our visualization
-% you never lose sight of where you came from.
-
 Obviously, the problem of finding witnesses is undecidable.
@@
 on two different sets of ill-typed programs obtained by instrumenting
-compilers used in beginner's classes~(\S~\ref{sec:eval}).
+compilers used in beginner's classes~(\S~\ref{sec:evaluation}).
 %
@@
 
+\smallskip
 In short, our results show that at least for beginners, ill-typed
 programs \emph{do} go wrong, in the vast majority of cases, which,
-in turn, opens the door to a completely different, dynamic way
-to explain, understand, and appreciate the benefits of static typing.
-
-%% Then, we formalize our algorithm and prove it sound
-%% (\S~\ref{sec:searching-witness}).
-%% %
-%% Next, we extend the algorithm to collect a reduction graph and formalize
-%% a set of traversals over the graph (\S~\ref{sec:explaining}).
-%% %
-%% Finally, we evaluate our search algorithm and trace generation on a
-%% large set of real student programs (\S~\ref{sec:evaluation}).
-%%
-%% \paragraph{Outline}
-%% We begin the rest of the paper with an overview of our search algorithm
-%% and trace visualization (\S~\ref{sec:overview}).
-
-% Then, we formalize our algorithm and prove it sound
-% (\S~\ref{sec:searching-witness}), and evaluate it on a large set of real
-% student programs (\S~\ref{sec:eval}).
-% %
-% Next, we extend the search algorithm to collect the reduction graph and
-% formalize a set of traversals (\S~\ref{sec:explaining}), and measure the
-% complexity of these traces on our benchmarks
-% (\S~\ref{sec:evaluation-nanomaly}).
-%
-% Finally, we present the results of a preliminary user study on the full
-% system (\S~\ref{sec:user-study}).
+in turn, opens the door to a novel dynamic way to explain, understand,
+and appreciate the benefits of static typing.
 

### 2015-11-20 paper/pldi16-submission/evaluation.tex
done with intro

@@
 The average number of single-step reductions per trace is 26 for the
-UCSD dataset (27 for the Seminal dataset) with a maximum of 2,744 (659
-for Seminal). The average number of jumps per trace is 3 (also 3 for
-Seminal) with a maximium of 177 (52 for Seminal). In both datasets at
+UCSD dataset (27 for the \uwbench\ dataset) with a maximum of 2,744 (659
+for \uwbench). The average number of jumps per trace is 3 (also 3 for
+\uwbench) with a maximium of 177 (52 for \uwbench). In both datasets at
 least 90\% of the traces contain fewer than 10 jumps.

### 2015-11-20 paper/pldi16-submission/introduction.tex
done with intro

@@
 
-Obviously, the problem of finding witnesses is undecidable.
-In fact, due to the necessarily conservative nature of static
-typing, there may not even exist. Thus, our heuristic approach
-is only useful if it can find \emph{compact} witnesses for
+\paragraph{3. Evaluating Witnesses}
+%
+Of course, the problem of finding witnesses is
+undecidable. In fact, due to the necessarily
+conservative nature of static typing, there
+may not even exist any witnesses for a given
+ill-typed program.
+%
+Thus, our approach is a heuristic that is only useful
+if it can find \emph{compact} witnesses for
 \emph{real-world} programs.
-
-\paragraph{3. Evaluating Witnesses}
 %
@@
 \smallskip
-In short, our results show that at least for beginners, ill-typed
-programs \emph{do} go wrong, in the vast majority of cases, which,
-in turn, opens the door to a novel dynamic way to explain, understand,
-and appreciate the benefits of static typing.
+Our results show that in the vast majority of
+cases, (novices') ill-typed programs \emph{do} go wrong.
+This, in turn, opens the door to a novel dynamic way to
+explain, understand, and appreciate the benefits of static typing.
 

### 2015-11-20 paper/pldi16-submission/evaluation.tex

--- 2015-11-20 paper/pldi16-submission/evaluation.tex
@@
 course of the first three assignments (IRB \# hidden for blind review), % \#140608),
-from which we extracted 4,407 distinct, ill-typed \ocaml programs.
+from which we extracted \ucsdsize\ distinct, ill-typed \ocaml programs.
 %

### 2015-11-20 paper/pldi16-submission/introduction.tex

--- 2015-11-20 paper/pldi16-submission/introduction.tex
@@
 
+\smallskip
 In short, our results show that at least for beginners, ill-typed

### 2015-11-20 paper/pldi16-submission/overview.tex

@@
 %
-Finally, we demonstrate our visualization of the witnesses we produce.
+Finally, we demonstrate our visualization of
+the synthesized witnesses.
 
@@
 Unfortunately, this approach quickly runs aground.
-Consider the @fac@ function from Figure~\ref{fig:factorial},
-reproduced here.
-
-\begin{code}
-  let rec fac n =
-    if n <= 0 then
-      true
-    else
-      n * fac (n-1)
-\end{code}
-\ES{having two copies of \texttt{fac} seems silly, but the back-reference across a page boundary is no good either...}
-
+Recall the erroneous @fac@ function from Figure~\ref{fig:factorial}.
+%~\S~\ref{sec:introduction}:
+%
+% \begin{code}
+  % let rec fac n =
+    % if n <= 0 then
+      % true
+    % else
+      % n * fac (n-1)
+% \end{code}
+% \ES{having two copies of \texttt{fac} seems silly, but the back-reference across a page boundary is no good either...}
+%
 What \emph{types} of inputs should we test @fac@ with?

### 2015-11-20 paper/pldi16-submission/evaluation.tex

@@
 The average number of single-step reductions per trace is 26 for our
-dataset (27 for the UW dataset) with a maximum of 2,744 (659 for
-UW). The average number of jumps per trace is 3 (also 3 for
-UW) with a maximium of 177 (52 for UW). In both datasets at
+dataset (27 for the \uwbench\ dataset) with a maximum of 2,744 (659 for
+\uwbench). The average number of jumps per trace is 3 (also 3 for
+\uwbench) with a maximium of 177 (52 for \uwbench). In both datasets at
 least 90\% of the traces contain fewer than 10 jumps.

### 2015-11-20 paper/pldi16-submission/introduction.tex
tweak intro

--- 2015-11-20 paper/pldi16-submission/introduction.tex
@@
 %
-The function returns @true@ in the base case (instead of @1@).
-and so \ocaml's responds with the error message:
+The function returns @true@ in the base case (instead of @1@),
+and so \ocaml responds with the error message:
 %
@@
 \end{verbatim}
-
-This message is makes perfect sense to an expert who is familiar
+%
+This message makes perfect sense to an expert who is familiar
 with the language and has a good mental model of how the type
@@
 %
-However, it may perplex a novice who is still to develop such a
+However, it may perplex a novice who has yet to develop such a
 mental model.
 %
-To make matters worse, unification based type inference algorithms
+To make matters worse, unification-based type inference algorithms
 often report errors far removed from their source.
@@
 %
-We have developed \toolname\ an interactive tool which uses
+We have developed \toolname, an interactive tool that uses
 the source of the ill-typed function to automatically synthesize
@@
 shows how the recursive calls reduce to a configuration where
-the program ``goes wrong'' \ie\ the @int@ value @1@ is to be
+the program ``goes wrong'' --- \ie\ the @int@ value @1@ is to be
 multiplied with the @bool@ value @true@.
@@
 We solve this problem by developing a novel operational semantics
-that corresponds to executing the program with \emph{holes}
+that corresponds to executing the program with \emph{holes} ---
 values whose type is unknown.
@@
 Our semantics \emph{conservatively} instantiates holes with concrete
-values; thereby dynamically inferring the type of the input type
-until the point the program goes wrong.
+values; thereby dynamically inferring the type of the input
+until the program goes wrong.
 %
@@
 which means, intuitively, that if a witness is found for a given
-ill-typed function, then \emph{for all} input types, there exist
+ill-typed function, then, \emph{for all} input types, there exist
 inhabitants that can make the function go wrong.
@@
 Our second contribution is a novel interactive visualization of the
-execution of \ocaml\ programs well-typed or not~(\S~\ref{sec:interactive}).
+execution of \ocaml\ programs, well-typed or not~(\S~\ref{sec:interactive}).
 %
@@
 The strategies include many of the standard debugging moves, \eg\
-stepping \emph{forward} or \emph{into} or \emph{over} calls as well
+stepping \emph{forward} or \emph{into} or \emph{over} calls, as well
 stepping or jumping \emph{backward} to understand how a particular

### 2015-11-20 paper/pldi16-submission/overview.tex
tweak overview

@@
 At this point we choose a random @int@ for the instantiation and
-concrete execution takes over entirely leading us to the expected crash
+concrete execution takes over entirely, leading us to the expected crash
 in the multiplication.
@@
 \paragraph{Witness Generality}
-We show in \S~\ref{sec:soundness} our lazy instantiation of holes
+We show in \S~\ref{sec:soundness} that our lazy instantiation of holes
 produces \emph{general witnesses}.
@@
 It is clear, syntactically, that @fac@ takes \emph{at least} one
-argument, but in a higher-order language with currying syntax can be
+argument, but in a higher-order language with currying, syntax can be
 deceiving.
@@
 %
-The \ocaml compiler deduces that @incAllByOne@ takes a single argument
+The \ocaml\ compiler deduces that @incAllByOne@ takes a single argument
 because @List.map@'s \emph{type} says it takes two arguments, and it is
@@
 We have shown how to reliably find witnesses to type errors in \ocaml,
-but this does not fully address our original goal -- to \emph{explain}
+but this does not fully address our original goal --- to \emph{explain}
 the errors.
@@
 %
-The existing debuggers and interpreters for \ocaml assume a type-correct
+The existing debuggers and interpreters for \ocaml\ assume a type-correct
 program, so unfortunately we cannot use them off-the-shelf.
@@
 applicable traversal strategies --- formalized in
-\S~\ref{sec:explaining} --- to insert another term into the
+\S~\ref{sec:interactive} --- to insert another term into the
 visualization.
@@
 %
-Finally, we take a ``step forward'' from @fac 0@, bringing us to @fac@'s
+Finally, we take a ``step forward'' from \hbox{@fac 0@,} bringing us to @fac@'s
 body. If we mouse over the body term we will see a popup with the

### 2015-11-20 paper/pldi16-submission/semantics.tex
revise proof of lemma 4

@@
 We solve this problem be equipping our term language with a notion
-of a \emph{hole}, written \ehole{}, which represent \emph{unconstrained}
+of a \emph{hole}, written \ehole{}, which represents \emph{unconstrained}
 values that may replaced with values from \emph{any} type.
@@
 Our semantics ensure the generality of witnesses by incrementally
-\emph{refining} holes filling in just as much information as is
+\emph{refining} holes, filling in just as much information as is
 needed locally to make progress (inspired by the manner in
-which SmallCheck uses lazy evaluation~\cite{smallcheck}).
+which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
 %
@@
 input an term \texttt{e} and refines it to have the type \texttt{t} by
-yielding a pair of either the same expression and a value substitution
-the refines the holes in \texttt{e} to ensure it has type \texttt{t}, or
-yields the stuck term if no such refinement is possible. In the case
-where \texttt{e} is a hole, it first checks in the given $\vsu$ to see
-if the hole has already been instantiated and, if so, returns the
-existing instantiation:
+yielding a pair of either the same expression and an empty value
+substitution, or yields the stuck term if no such refinement is
+possible. In the case where \texttt{e} is a hole, it first checks in the
+given $\vsu$ to see if the hole has already been instantiated and, if
+so, returns the existing instantiation:
 %
@@
 %
-We write \steps{e}{\vsu}{e'}{\vsu'} if \steps{e}{\vsu}{e'}{\vsu'}
+We write \steps{e}{\vsu}{e'}{\vsu'} if \steptr{\trace}{e}{\vsu}{e'}{\vsu'}
 for some trace $\trace$.
@@
 \paragraph{Primitive Reductions}
-Primitive reduction steps -- addition, if-elimination, and
-function application -- use \forcesym to ensure that values have the
+Primitive reduction steps --- addition, if-elimination, and
+function application --- use \forcesym to ensure that values have the
 appropriate type (and that holes are instantiated) before continuing the
@@
 If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
-$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is single
-step extension of $\trace$)
+$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is a single-step
+extension of $\trace$)
 %
@@
   %
-  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
+  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
 \end{proof}
@@
 the partial type upto trace $\trace$ will cause $f$ to get stuck
-in \emph{at most} $k+1$ steps where $k$ is the length of $\trace$.
+in \emph{at most} $k$ steps where $k$ is the length of $\trace$.
 %
@@
 \label{lem:k-stuck}
-  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
-     \tincompat{t}{\ptype{\trace}{f}}, and
-     \hastype{v}{t},
+  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu} and
+     \hastype{v}{t} and
+     \tincompat{t}{\ptype{\trace}{f}},
   then
@@
 \end{lem}
-\ES{self-todo: take a close look at the $\trace$'s vs k's}
+% \ES{self-todo: take a close look at the $\trace$'s vs k's}
 \begin{proof}
   By induction on $k$, the length of $\trace$.
-  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$.
-  Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
+  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$ and
+  $\trace' = \pair{\eapp{f}{\ehole{}}}{\emptysu},...,\pair{e'}{\vsu'}$ 
+  such that $\step{e'}{\vsu'}{e}{\vsu}$.
+  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
   will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
-  in at most $k+1$ steps:
+  in at most $k$ steps:
 %
@@
   % \item
-  [Case \tincompat{t}{s_{\trace}}:]
+  [Case \tincompat{t}{s_{\trace'}}:]
     The inductive hypothesis applies.
   %\item
-  [Case \tcompat{t}{s_{\trace}} but \tincompat{t}{s_{\trace'}}:]
+  [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
     By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
-    $k+1$.
+    $k$.
     %
-    A case analysis of the applicable rules shows that \forcesym cannot
-    have succeeded.
+    A case analysis of the rules shows that \forcesym cannot have
+    succeeded.
   %\end{description}
@@
 
-\paragraph{Proof of Theorem~\ref{thm:soundness}}
+\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
 %
 Suppose $\trace$ witnesses that $f$ gets stuck,
-and let $t$ be the partial type of $f$ upto $\trace$.
+and let $t = \ptype{\trace}{f}$.
 We show that \emph{all} types $s$ have stuck-inducing
 values by splitting cases on whether the type is
-compatible with the partial type upto $\trace$.
+compatible with $t$. %the partial type upto $\trace$.
 %
@@
 %\end{description}
+\end{proof}
 

### 2015-11-20 paper/pldi16-submission/semantics.tex
tweak semantics

@@
 %
-Next, we describe what it takes as input and output, and how it
+Next, we describe what its input and output, and how it
 addresses the above challenges to search the space of possible
@@
 Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
-denoting the list of \emph{top-level definitions}
+denoting the list of \emph{top-level definitions}:
 %
@@
 %
-We determine whether a given value is stuck by using the predicate
+We determine whether a given value is stuck by using the predicate:
 %
@@
 apply the parameters to the target and close the result under the top-level
-let-binders, prior to invoking @eval@.
+let-binders, prior to invoking \hbox{@eval@.}
 %
@@
 
-\begin{thm}{\textbf{[Witness Generation]}}
+\begin{cor}{\textbf{[Witness Generation]}}
 \label{thm:generation}
-  If @genWitness k@ $[(f_1, e_1), \ldots, (f_n, e_n)]$ is non-empty,
+  If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
   then for every type $t$ there exists $\hastype{v}{t}$ such that
   $\steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
-\end{thm}
+\end{cor}
 

### 2015-11-20 paper/pldi16-submission/interactive.tex
tweak intro
up to related

@@
 \section{Explaining Type Errors With Traces}
-\label{sec:explaining}
+\label{sec:interactive}
 
@@
 
-Next, we show how to build an visual interactive debugger
+Next, we show how to build a visual interactive debugger
 from the traced semantics, by describing the visualization
 \emph{state} \ie\ what the user sees at any given moment,
-the set of \emph{commands} available to user, and what
+the set of \emph{commands} available to user and what
 they do, and finally how we use a command to \emph{update}
@@
 successor. In other words, the visualization state looks
-like a set of linear lists of expressions as shown in Figure~\ref{fig:FIXTHIS}.
+like a set of linear lists of expressions as shown in Figure~\ref{fig:nanomaly-factorial}.
 %
@@
 and the current visualization state to compute the new
-expression that should be added to the visualization state
+expression that should be added to the visualization state,
 and @updState vs cmd e@ then updates the graph
@@
 @putRoot vs e e'@ returns the modified version of @vs@ extended
-with a new root vertex @e@ and its successor @e'@, and,
+with a new root vertex @e@ and its successor @e'@.
 %
@@
 % source vertex.
-%
+
 @getNext vs e@ returns the immediate successor of @e@ in $\tr$.
@@
 %
-Note that the sub-term edges $\searrow$ allow us to decompose an
-expression into a sub-expression and the surrounding context, thus
-enabling the \stepintosym\ and \stepoversym\ traversals.
+% Note that the sub-term edges $\searrow$ allow us to decompose an
+% expression into a sub-expression and the surrounding context, thus
+% enabling the \stepintosym\ and \stepoversym\ traversals.
 
-\RJ{EXAMPLE of interaction from overview goes here}
+% \RJ{EXAMPLE of interaction from overview goes here}
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.8\linewidth]{fac-steps.png}
+\caption{A sequence of interactions with the trace of
+  \texttt{fac 1}. The stuck term is red, in each node the redex is
+  highlighted. Thick arrows denote a multi-step transition, thin arrows
+  denote a single-step transition. We start in step 1. In step 2 we jump
+  forward from the witness to the next function call. In step 3 we step
+  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
+  of execution. In step 4 we take a single step forward from
+  \texttt{fac 0}.} % (hiding the context for space).}
+\label{fig:nanomaly-factorial}
+\end{figure}
 

### 2015-11-20 paper/pldi16-submission/overview.tex
up to related

@@
 
-A sample interaction with the trace of @fac 1@ can be seen in
-Figure~\ref{fig:nanomaly-factorial}.
-%
-\begin{figure*}[t]
-\centering
-\includegraphics[width=0.8\linewidth]{fac-steps.png}
-\caption{A sequence of interactions with the trace of
-  \texttt{fac 1}. The stuck term is red, in each node the redex is
-  highlighted. Thick arrows denote a multi-step transition, thin arrows
-  denote a single-step transition. We start in step 1. In step 2 we jump
-  forward from the witness to the next function call. In step 3 we step
-  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
-  of execution. In step 4 we take a single step forward from
-  \texttt{fac 0} (hiding the context for space).}
-\label{fig:nanomaly-factorial}
-\end{figure*}
-%
-The initial state of the visualization tells us that after some number
-of steps -- the thick arrow denotes a multi-step transition -- we try to
-multiply @1@ by @true@.
+% A sample interaction with the trace of @fac 1@ can be seen in
+% Figure~\ref{fig:nanomaly-factorial}.
+% %
+% % \begin{figure*}[t]
+% % \centering
+% % \includegraphics[width=0.8\linewidth]{fac-steps.png}
+% % \caption{A sequence of interactions with the trace of
+% %   \texttt{fac 1}. The stuck term is red, in each node the redex is
+% %   highlighted. Thick arrows denote a multi-step transition, thin arrows
+% %   denote a single-step transition. We start in step 1. In step 2 we jump
+% %   forward from the witness to the next function call. In step 3 we step
+% %   into the recursive \texttt{fac 0} call, which spawns a new ``thread''
+% %   of execution. In step 4 we take a single step forward from
+% %   \texttt{fac 0} (hiding the context for space).}
+% % \label{fig:nanomaly-factorial}
+% % \end{figure*}
+% %
+% The initial state of the visualization tells us that after some number
+% of steps -- the thick arrow denotes a multi-step transition -- we try to
+% multiply @1@ by @true@.
 
-Upon seeing the stuck term, we might wonder where @true@ came from.
-%
-To investigate we select the stuck term and click the ``jump backward''
-button to search backwards from the stuck term for the most recent
-function call, which brings us to @1 * fac 0@. Notice at this point that
-@fac 0@ is highlighted while @1 *@ is grayed out. This tells us that
-@fac 0@ is the redex in this term.
+% Upon seeing the stuck term, we might wonder where @true@ came from.
+% %
+% To investigate we select the stuck term and click the ``jump backward''
+% button to search backwards from the stuck term for the most recent
+% function call, which brings us to @1 * fac 0@. Notice at this point that
+% @fac 0@ is highlighted while @1 *@ is grayed out. This tells us that
+% @fac 0@ is the redex in this term.
 
-@fac 0@ seems like the right thing to do so we choose to ``step into''
-it, which inserts a new multi-step transition from @fac 0@ to @true@.
-%
-Finally, we take a ``step forward'' from \hbox{@fac 0@,} bringing us to @fac@'s
-body. If we mouse over the body term we will see a popup with the
-environment at this point, notably telling us that @n = 0@. At this point
-it is clear that @fac@ handles the @n <= 0@ case incorrectly and should
-instead return an @int@.
+% @fac 0@ seems like the right thing to do so we choose to ``step into''
+% it, which inserts a new multi-step transition from @fac 0@ to @true@.
+% %
+% Finally, we take a ``step forward'' from \hbox{@fac 0@,} bringing us to @fac@'s
+% body. If we mouse over the body term we will see a popup with the
+% environment at this point, notably telling us that @n = 0@. At this point
+% it is clear that @fac@ handles the @n <= 0@ case incorrectly and should
+% instead return an @int@.
 

### 2015-11-20 paper/pldi16-submission/evaluation.tex
up to related
11 pages exactly!

@@
 \item The number of single-step reductions along the path from the
-  witness to the stuck term. This can be thought of as a worst-case
-  complexity, \ie ``How big is the fully-expanded trace?''
+  witness to the stuck term.
+  % This can be thought of as a worst-case
+  % complexity, \ie ``How big is the fully-expanded trace?''
 \item The number forward (or backward) jumps that can be taken along the
@@
 there are some degenerate cases where it is unable to detect
-them. Consider, for example, the following buggy version of @replicate@,
-which is supposed to produce a list containing @n@ copies of @x@.
+them. Consider, for example, the following buggy version of @replicate@.
+% ,
+% which is supposed to produce a list containing @n@ copies of @x@.
 %
@@
   let rec replicate n x =
-    if n <= 0 then
-      []
-    else
-      replicate (n-1) [x]
+    if n <= 0 then [] else replicate (n-1) [x]
 \end{code}
@@
 This code produces a nested list (with @n@ levels of nesting) containing
-a single copy of @x@. \ocaml detects a cyclic \hbox{@'a = 'a list@}
-constraint in the recursive call and throws a type error, whereas
-\nanomaly happily recurses @n@ times to produce the nested list.
-Strictly speaking, this function itself cannot ``go wrong'', the
-program would not get stuck until a \emph{client} attempted to use the
-result expecting a flat list. But this is not very satisfying as
-@replicate@ is clearly to blame. Furthermore, in our experience,
-infinite-type errors are some of the more difficult ones to debug (and
-to explain to novices), so better support for this scenario would be
-useful.
+a single copy of @x@, instead of a list with @n@ copies of @x@. \ocaml
+detects a cyclic \hbox{@'a = 'a list@} constraint in the recursive call
+and throws a type error, whereas \nanomaly happily % recurses @n@ times to
+produces the nested list.  Strictly speaking, this function itself cannot
+``go wrong'', the program would not get stuck until a \emph{client}
+attempted to use the result expecting a flat list. But this is not very
+satisfying as @replicate@ is clearly to blame. Furthermore, in our
+experience, infinite-type errors are often difficult to %some of the more difficult ones to
+debug (and to explain to novices), so better support for this scenario
+would be useful.
 

### 2015-11-20 paper/pldi16-submission/related.tex
11 pages exactly!

@@
 Chen and Erwig~\cite{chen_counter-factual_2014} use a variational type
-system represent the possibility that the inferred type of an expression
-may be incorrect, because the expression may be the source of the
-error. It then attempts to deduce the error source by searching for an
-expression whose type can be changed such that type inference would
-succeed. In contrast to Seminal, which searches for changes at the
-value-level, by searching at the type level the search is complete due
+system to allow for the possibility of changing an expression's type,
+and search for an expression whose type can be changed such that type
+inference would succeed.
+%
+% It then attempts to deduce the error source by searching for an
+% expression whose type can be changed such that type inference would
+% succeed.
+%
+In contrast to Lerner, who searches for changes at the
+value-level, by searching at the type level Chen's search is complete due
 the finite universe of types in a program.
@@
 %
-Pavlinovic \etal~\cite{pavlinovic_finding_2014} translate the error
+Pavlinovic \etal~\cite{pavlinovic_finding_2014} translate the %error
 localization problem to a MaxSMT optimization problem, using
@@
 incrementally test refactorings, which often cause type errors in
-distant pieces of code. They replace any expression that fails to type
-check with a \emph{runtime} error, but do not perform any type checking
+distant functions. They replace any expression that fails to type
+check with a \emph{runtime} error, but do not check types
 at runtime.
@@
 %
-enables this by symbolic execution (to track which path a given input
-vector triggers) with concrete execution (to ensure test failures are
+combines symbolic execution (to track which path a given input
+triggers) with concrete execution (to ensure failures are
 not spurious). The system collects a path condition during execution,
@@
 satisfy the negated path condition, \ie inputs that will not trigger the
-same path. Thus, dynamic-symbolic execution can prune the search space
+same path. Thus, it can prune the search space
 much faster than techniques based on enumerating values, but is limited
@@
 by a call to the solver. We chose to use lazy, random generation instead
-because it does not incur the overhead of an external solver, and
+because it is efficient, and % does not incur the overhead of an external solver, and
 produces high coverage for our domain of novice programs.
@@
 of reductions that produced it. Notably, they allow the user to supply a
-\emph{partial value} -- a value containing holes -- and in turn present
-a partial slice, containing only those steps that affected the specified
-components of the partial value. This system is designed to answer
-questions of the form ``Where did this value come from?'' and thus is
-focused on backward exploration. In contrast, our trace visualization
-supports forward \emph{and} backward exploration, though our backward
-steps are more limited than Perera's. Specifically, we do not support
-selecting a value and inserting the intermediate terms that preceded it
-while ignoring unrelated computation steps; this would be interesting
-future work.
+\emph{partial value} -- containing holes -- and present a partial slice,
+containing only those steps that affected the the partial value. This
+system is designed to answer questions of the form ``Where did this
+value come from?'' and thus is focused on backward exploration. In
+contrast, our trace visualization supports forward \emph{and} backward
+exploration, though our backward steps are more limited than
+Perera's. Specifically, we do not support selecting a value and
+inserting the intermediate terms that preceded it while ignoring
+unrelated computation steps; this would be interesting future work.
 
-\ES{todo: more on program exploration}
+% \ES{todo: more on program exploration}
 

### 2015-11-20 paper/pldi16-submission/haskellListings.tex
update graph api/update figure
ugh colors..

@@
 \lstdefinelanguage{HaskellUlisses}{
-	basicstyle=\codesize\ttfamily,
+	basicstyle=\footnotesize\ttfamily,
 	sensitive=true,
-	morecomment=[l][\color{gray_ulisses}\ttfamily\codesize]{--},
-	morecomment=[s][\color{gray_ulisses}\ttfamily\codesize]{\{-}{-\}},
+	morecomment=[l][\color{gray_ulisses}\ttfamily\footnotesize]{--},
+	morecomment=[s][\color{gray_ulisses}\ttfamily\footnotesize]{\{-}{-\}},
 	morestring=[b]",
@@
 	showtabs=false,
+        escapeinside={(*}{*)},%
         % mathescape=true,
@@
                 padAverage,focusUp,execute,checkSMT,inputTypes,outputType,toReft,app,
-								genArgs, genWitness, close, witness, mkApps, mkLets, isStuck
+                genArgs,genWitness,close,witness,mkApps,mkLets,isStuck,findExpr,updState,putBefore,putAfter,
+                putRoot,getNext,getPrev,getSubterms,applyCtx,findApp,findVal,replicate,loop,fac,incAllByOne,map
                 %, inTypes, inputTypes, outputType, execute, smtFindModel, smtRefuteModel
@@
                 len,isBH,isBal,bh,isRB,keys,Sorted,RBT,Col,isBlack,OrdRBT,Set,sz,
-                StackSet,NoDuplicates,Data,RBTree,XMonad,Generic,true
+                StackSet,NoDuplicates,Data,RBTree,XMonad,Generic,
+                VState,Expr,Ctx,Cmd,int,list,string,bool
 	},
@@
 	{[5]
-		PS,Tip,Node,Black,Red,EQ,False,GT,Just,LT,Left,Nothing,Right,True,Show,Eq,Ord,Num,C,N,Leaf,Bin,CounterExample
+		PS,Tip,Node,Black,Red,EQ,False,GT,Just,LT,Left,Nothing,Right,True,Show,Eq,Ord,Num,C,N,Leaf,Bin,CounterExample,
+                StepForward,StepBack,JumpForward,JumpBack,StepOver,StepInto,true
 	},

### 2015-11-20 paper/pldi16-submission/introduction.tex
edit related

@@
 
-In this paper we propose a new approach to explaining type
-\emph{static} type errors: by \emph{dynamically} witnessing
+In this paper we propose a new approach that explains 
+static type errors by \emph{dynamically} witnessing
 how exactly an ill-typed program goes wrong.

### 2015-11-20 paper/pldi16-submission/interactive.tex
back to old traversal
edit related

@@
 
-% The trace takes the form of a \emph{reduction graph}, where the nodes
-% are terms and the edges represent either the single-step
-% $\hookrightarrow$ or a ``sub-term'' relation. For example, evaluating
-% the expression @1 + 2 + 3@ would produce the graph in
-% Figure~\ref{fig:simple-reduction}.
-% %
-% \begin{figure}[t]
-%   \centering
-%   \includegraphics[width=\linewidth]{simple.png}
-% \caption{The reduction graph for \texttt{1 + 2 + 3}.}
-% \label{fig:simple-reduction}
-% \end{figure}
-% %
-% We choose this graph representation instead of a simple, linear sequence
-% of expressions because it will allow us to express a variety of
-% traversals, such as ``step into'' and ``step over'' -- commonly found in
-% traditional debuggers.
-
 \subsection{Tracing Semantics}
@@
 \begin{figure*}[t]
-% \relDescription{Trace Syntax}
-% $$
-% \begin{array}{rrcl}
-  % & \tr & ::= & \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr
-% \end{array}
-% $$
-% \\
 \relDescription{Computing Sub-Terms}
@@
 
-% After evaluation the reduction graph can be constructed directly from
-% the trace $\tr$ as follows:
-% \[
-% G(\tr) = \pair{\{e \spmid e \in \tr\}}{\tr}
-% \]
-
-% \subsection{Traversing the Reduction Graph}
 \subsection{Interactive Debugging}
@@
 %
-The procedure @findExpr vs cmd e@ traverses $\tr$
+The procedure \findExpr{\vstate}{\cmd}{e} traverses $\tr$
 and the current visualization state to compute the new
-expression that should be added to the visualization state,
-and @updState vs cmd e@ then updates the graph
-by inserting the new expression appropriately, using one of
-the following functions.
+expression that should be added to the visualization state;
+and \updState{\vstate}{\cmd}{e} then updates the graph
+by inserting the new expression appropriately, using
+one of the following graph manipulating functions:
 %
-@putBefore vs e e'@ \hbox{(resp. @putAfter vs e e'@)}
-returns the modified version of @vs@ where @e'@ is the
-immediate predecessor of @e@ (resp. the immediate successor of @e@);
+\putBefore{\vstate}{e}{e'} (resp. \putAfter{\vstate}{e}{e'})
+returns the modified version of \vstate\ where $e'$ is the
+immediate predecessor of $e$ (resp.\ the immediate successor of $e$);
 %
-@putRoot vs e e'@ returns the modified version of @vs@ extended
-with a new root vertex @e@ and its successor @e'@.
+\putRoot{\vstate}{e}{e'} returns the modified version of \vstate
+extended with a new root vertex $e$ with successor $e'$.
 %

### 2015-11-20 paper/pldi16-submission/overview.tex
remove sub-term edges from the graph
tweak

@@
 The \ocaml\ compiler deduces that @incAllByOne@ takes a single argument
-because @List.map@'s \emph{type} says it takes two arguments, and it is
+because the \emph{type} of \hbox{@List.map@'s} says it takes two arguments, and it is
 partially applied to @(+ 1)@.
@@
   \centering
-  \includegraphics[width=\linewidth]{simple.pdf}
+  \includegraphics[height=2in]{simple.pdf}
   \caption{The reduction graph for \texttt{1+2+3}. The two edges

### 2015-11-20 paper/pldi16-submission/interactive.tex
remove sub-term edges from the graph
done interactive

@@
 \begin{figure*}[t]
-\relDescription{Computing Sub-Terms}
-\begin{gather*}
-\begin{array}{lcl}
-\subtermssym                 & \dcolon & e \to \tr \\
-\subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
-\subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
-\subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
-                                &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
-                                &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
-\subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
-                                &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
-\subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
-\subterms{e}                 & \defeq & \bullet
-\end{array}
-\end{gather*}
+%% \relDescription{Computing Sub-Terms}
+%% \begin{gather*}
+%% \begin{array}{lcl}
+%% \subtermssym                 & \dcolon & e \to \tr \\
+%% \subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
+%% \subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
+%% \subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
+                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
+                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
+%% \subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
+                                %% &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
+%% \subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
+%% \subterms{e}                 & \defeq & \bullet
+%% \end{array}
+%% \end{gather*}
 \judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
@@
 %
-A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
-intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
+%A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
+%intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
 %
-A \emph{reduction graph} is a set of edges:
-$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr \spmid \subterm{e}{e}; \tr$$
+A \emph{reduction graph} is a set of steps-to edges:
+$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr$$  %% \spmid \subterm{e}{e}; \tr$$
 
@@
 Next, we describe the general recipe for extending a transition
-rule to collect edges, and provide a selection of examples
+rule to collect edges.
+%
+The general recipe for collecting steps-to edges is to
+to recording the consequent of each original rule in the
+trace. That is, each original judgment \step{e}{\vsu}{e'}{\vsu'}
+becomes \stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \tr}.
+%
+As the translation is mechanical, we provide a selection of examples
 in Figure~\ref{fig:interactive}.
-%
-The steps-to edges are collected by recording the consequent of
-each original rule in the trace. That is, each original judgment
-\step{e}{\vsu}{e'}{\vsu'} becomes
-\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \tr}.
-%
-The sub-term edges are delegated to a helper function \subtermssym\
-which adds edges from an expression to each of its
-\emph{immediate} sub-expressions.
-%
-We collect \subtermssym edges after each transition,
-to get the following template for the small-step relation:
-\[
-\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \subterms{e'}; \tr}
-\]
+
+%%% The sub-term edges are delegated to a helper function \subtermssym\
+%%% which adds edges from an expression to each of its
+%%% \emph{immediate} sub-expressions.
+%%% %
+%%% We collect \subtermssym\ edges after each transition,
+%%% to get the following template for the small-step relation:
+%%% \[
+%%% \stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \subterms{e'}; \tr}
+%%% \]
 
@@
 exposition, we assume we have a (global) trace:
-$\stepg{e_0}{\emptysu}{\bullet}{e_n}{\_}{\tr}$, where
+$\stepg{e_0}{\emptysu}{\bullet}{e_n}{\vsu}{\tr}$, where
 $e_0$ and $e_n$ are the \emph{initial} and \emph{final}
@@
 %
-A \emph{visualization state} @VState@ is a \emph{directed graph}
+A \emph{visualization state} $\vstate$ is a \emph{directed graph}
 whose vertices are expressions and whose edges are such
@@
 successor. In other words, the visualization state looks
-like a set of linear lists of expressions as shown in Figure~\ref{fig:nanomaly-factorial}.
+like a set of linear lists of expressions as shown in
+Figure~\ref{fig:nanomaly-factorial}.
 %
@@
 edge linking the initial and final expressions.
+
+\paragraph{Visualization Context}
+%
+The \emph{visualization context} of each expression $e$
+in the visualization state \vstate is the (unique) linear chain
+in which it the expression $e$ belongs.
+%
+We write $\vroot{\vstate}{e}$ for the \emph{first} (or root)
+expression appearing in the visualization context of $e$ in
+\vstate.
 
@@
 
-\input{traversals.tex}
+\input{traversals_new.tex}
 
 \paragraph{Update}
-Figure~\ref{fig:traversing-graph} shows how we update the state
-for each command.
 %
-The procedure \findExpr{\vstate}{\cmd}{e} traverses $\tr$
-and the current visualization state to compute the new
-expression that should be added to the visualization state;
-and \updState{\vstate}{\cmd}{e} then updates the graph
-by inserting the new expression appropriately, using
-one of the following graph manipulating functions:
+Figure~\ref{fig:traversing-graph} shows
+how we compute the \emph{next} expression
+(to be added to the visualization state)
+given the current visualization state
+$\vstate$, command $\cmd$ and selected
+expression $e$.
 %
-\putBefore{\vstate}{e}{e'} (resp. \putAfter{\vstate}{e}{e'})
-returns the modified version of \vstate\ where $e'$ is the
-immediate predecessor of $e$ (resp.\ the immediate successor of $e$);
-%
-\putRoot{\vstate}{e}{e'} returns the modified version of \vstate
-extended with a new root vertex $e$ with successor $e'$.
-%
-% @getRoot vs e@ returns the vertex (expression) obtained
-% by transitively following the predecessors of @e@ until a
-% source vertex.
+It is straightforward to then \emph{update}
+the visualization graph by adding the new
+term before (resp.\ after) the selected
+expression $e$ if the command was a step
+or jump forward (resp.\ backward), or
+to create a new visualization context
+if the command was $\stepintosym$.
 
-@getNext vs e@ returns the immediate successor of @e@ in $\tr$.
-%
-@getPrev vs e@ computes the path @p@ between @e@ and its immediate
-predecessor in the current visualization, and then returns @e@'s immediate
-predecessor along @p@.
-%
-@getSubterms vs e@ traverses the sub-term edges to decompose an
-expression into a list of sub-expressions paired with their context.
-%
-@applyCtx vs e ctx@ applies @ctx@ to @e@, traversing the sub-term edges
-in reverse to find the super-term of @e@.
-%
-\hbox{@findApp vs e@} builds on top of @getSubterms@ to find the first
-application sub-term (if any), \ie the first sub-term that looks like
-$\eapp{v_1}{v_2}$.
-%
-@findVal vs e@ traverses the single-step edges to find the final value
-that @e@ reduces to.
-%
-% Note that the sub-term edges $\searrow$ allow us to decompose an
-% expression into a sub-expression and the surrounding context, thus
-% enabling the \stepintosym\ and \stepoversym\ traversals.
+%%% and \updState{\vstate}{\cmd}{e} then updates the graph
+%%% by inserting the new expression appropriately, using
+%%% one of the following graph manipulating functions:
+%%% %
+%%% \putBefore{\vstate}{e}{e'} (resp. \putAfter{\vstate}{e}{e'})
+%%% returns the modified version of \vstate\ where $e'$ is the
+%%% immediate predecessor of $e$ (resp.\ the immediate successor of $e$);
+%%% %
+%%% \putRoot{\vstate}{e}{e'} returns the modified version of \vstate
+%%% extended with a new root vertex $e$ with successor $e'$.
+%%% %
+%%% @getNext vs e@ returns the immediate successor of @e@ in $\tr$.
+%%% %
+%%% @getPrev vs e@ computes the path @p@ between @e@ and its immediate
+%%% predecessor in the current visualization, and then returns @e@'s immediate
+%%% predecessor along @p@.
+%%% %
+%%% @getSubterms vs e@ traverses the sub-term edges to decompose an
+%%% expression into a list of sub-expressions paired with their context.
+%%% %
+%%% @applyCtx vs e ctx@ applies @ctx@ to @e@, traversing the sub-term edges
+%%% in reverse to find the super-term of @e@.
+%%% %
+%%% \hbox{@findApp vs e@} builds on top of @getSubterms@ to find the first
+%%% application sub-term (if any), \ie the first sub-term that looks like
+%%% $\eapp{v_1}{v_2}$.
+%%% %
+%%% @findVal vs e@ traverses the single-step edges to find the final value
+%%% that @e@ reduces to.
 
-% \RJ{EXAMPLE of interaction from overview goes here}
 \begin{figure}[t]
@@
   of execution. In step 4 we take a single step forward from
-  \texttt{fac 0}.} % (hiding the context for space).}
+  \texttt{fac 0}.}
 \label{fig:nanomaly-factorial}
 \end{figure}
-
-%
-%
-%The initial path $p$ is required for the backward-steps as a node may
-%have multiple incoming $\leadsto$ edges, \eg
-%\singlestep{\eplus{1}{2}}{3} and \singlestep{\eplus{2}{1}}{3}.
-%

### 2015-11-20 paper/pldi16-submission/abstract.tex
up to related
done eval

@@
   Third, we evaluate our approach on two data sets
-  comprising more than 4500 ill-typed student programs.
+  comprising over 4,500 ill-typed student programs.
   Our technique is able to generate witnesses for
-  nearly 90\% of the programs, and our reduction graph
+  88\% of the programs, and our reduction graph
   is yields small counterexamples for 90\% of the witnesses.

### 2015-11-20 paper/pldi16-submission/evaluation.tex
done eval

@@
 
-%
-% The immediate question is ``What fraction of type errors admit
-% witnesses?''
-%
 We have implemented a prototype of our search procedure and trace
 visualization for the pure subset of \ocaml, in a tool called \nanomaly.
-% , \ie \lang extended with
-% algebraic datatypes and records.
-%
-In our implementation we instantiated \gensym with a simple random
-generation of values, which we will show is more than sufficient for the
+In our implementation we instantiated \gensym\ with a simple random
+generation of values, which we will show suffices for the
 majority of type errors.
 
+\paragraph{Evaluation Goals}
+%
 There are two questions we seek to answer with our evaluation:
+%
 \begin{enumerate}
-\item What fraction of type errors \emph{admit} witnesses?
-\item How \emph{complex} are the traces our witnesses produce?
+\item \emphbf{Witness Coverage}
+      How many ill-typed programs can we find witnesses for?
+\item \emphbf{Witness Complexity}
+      How \emph{complex} are the traces produced by the witnesses?
 \end{enumerate}
 
-We answer these questions on two sets of known-bad programs, \ie
-programs that were rejected by the \ocaml compiler because of a type
-error.
+\paragraph{Benchmarks}
+We answer these questions on two sets of ill-typed programs, \ie
+programs that were rejected by the \ocaml\ compiler because of a
+type error.
 %
@@
 %
-We recorded each interaction with the \ocaml top-level system over the
+We recorded each interaction with the \ocaml\ top-level system over the
 course of the first three assignments (IRB \# hidden for blind review), % \#140608),
@@
 %
-The second dataset -- widely used in the literature -- comes from a
-similar course at the University of
-Washington~\cite{lerner_seminal:_2006}, from which we extracted 284
-ill-typed programs.
+The second dataset --- widely used in the literature --- comes from a
+similar course at the University of Washington~\cite{lerner_seminal:_2006},
+from which we extracted 284 ill-typed programs.
 
@@
 We ran our search algorithm on each program for 1,000 iterations, with
-the entry point set to the function that \ocaml had identified as
+the entry point set to the function that \ocaml\ had identified as
 containing a type error.
@@
 per program.
+%
 % Due to the possibility of non-termination we set a limit on the number
@@
 \label{sec:trace-complexity}
-For each of the ill-typed programs for which we could find a witness, we
-measure the complexity of the generated trace according to the following
-metrics:
+
+For each of the ill-typed programs for which we could
+find a witness, we measure the complexity of the generated
+trace according to two metrics.
+
+\paragraph{Jump Compression}
+%
+A jump compressed trace is one whose edges are limited to forward
+or backward jumps. (The sizes of these traces is equal to the jump
+metric above.)
+%
+In our experience, jump compression abstracts many details of the
+computation that are often uninteresting or irrelevant to the
+explanation, or at any rate, easily summarized as call-return
+pairs.
+%
+Note that jump compressed traces are not quite the same as
+stack-traces as they show \emph{all} function calls, including
+those that returned successfully.
+%
+For example, the display on the right in Figure shows
+a jump-compressed trace, of size 4. The full trace for
+the same example has size \RJ{Eric please fix this...}
+
+\paragraph{Metrics} Thus, our two metrics are:
+% size of the full trace,
+% \ie the number of small-step reductions, and the size of the jump-compressed
+% version of the trace.
 %
 \begin{enumerate}
-\item The number of single-step reductions along the path from the
-  witness to the stuck term.
+\item \emphbf{Single-Step Metric}
+      The number of single-step reductions along the path from the
+      witness to the stuck term in the given witness execution, and
   % This can be thought of as a worst-case
   % complexity, \ie ``How big is the fully-expanded trace?''
-\item The number forward (or backward) jumps that can be taken along the
-  path from the witness to the stuck term\footnote{This is not quite the same as
-  a stack-trace as it shows \emph{all} function calls, including those
-  that returned successfully.}.
-  % These abstract many details of the computation that are often
-  % uninteresting or irrelevant to the explanation.
+\item \emphbf{Jump Metric}
+      The size of the jump-compressed trace for the given witness execution,
+      \ie\ the number forward (or backward) jumps that can be taken along the
+      path from the witness to the stuck term.
+\end{enumerate}
+
 % \item \ES{others?}
-\end{enumerate}
 %
@@
 \label{sec:discussion}
+
 To summarize, our experiments demonstrate that \nanomaly finds witnesses
@@
   let rec replicate n x =
-    if n <= 0 then [] else replicate (n-1) [x]
+    if n <= 0 then
+      []
+    else
+      replicate (n-1) [x]
 \end{code}
@@
 program. Enhancing our visualization to slice out program paths relevant
-to specific values, \eg in the style of~\cite{perera_functional_2012},
+to specific values, \eg\ in the style of~\cite{perera_functional_2012},
 would likely help alleviate this issue, allowing users to

### 2015-11-20 paper/pldi16-submission/interactive.tex
misc edits

@@
 The \emph{visualization context} of each expression $e$
-in the visualization state \vstate is the (unique) linear chain
-in which it the expression $e$ belongs.
+in the visualization state $\vstate$ is the (unique) linear chain
+in which the expression $e$ belongs.
 %
@@
 expression appearing in the visualization context of $e$ in
-\vstate.
+$\vstate$.
 
@@
 \item \stepforwardsym, \stepbackwardsym:
-      show result of a single step forward or backward respectively,
+      show the result of a single step forward or backward respectively,
 %
-\item \jumpforwardsym:
-      show result of taking multiple steps (a \emph{``big''} step)
+\item \jumpforwardsym, \jumpbackwardsym:
+      show the result of taking multiple steps (a \emph{``big''} step)
       upto the first beta-reduction forward or backward respectively,
@@
 \item \stepintosym:
-      show result of stepping into a function call in a sub-term,
+      show the result of stepping into a function call in a sub-term,
       isolating it from the context,
@@
 \item \stepoversym:
-      show result of skipping over a function call in a sub-term.
+      show the result of skipping over a function call in a sub-term.
 \end{itemize}
@@
 \centering
-\includegraphics[width=0.8\linewidth]{fac-steps.png}
+\includegraphics[width=\linewidth]{fac-steps.png}
 \caption{A sequence of interactions with the trace of

### 2015-11-20 paper/pldi16-submission/evaluation.tex
misc edits
fix RJ comment

@@
 or backward jumps. (The sizes of these traces is equal to the jump
-metric above.)
+metric below.)
 %
@@
 %
-For example, the display on the right in Figure shows
+For example, the display on the right in Figure~\ref{fig:factorial} shows
 a jump-compressed trace, of size 4. The full trace for
-the same example has size \RJ{Eric please fix this...}
+the same example has size 19.
 
@@
       The size of the jump-compressed trace for the given witness execution,
-      \ie\ the number forward (or backward) jumps that can be taken along the
+      \ie\ the number of forward (or backward) jumps that can be taken along the
       path from the witness to the stuck term.
@@
 %
-The average number of single-step reductions per trace is 26 for our
+The average number of single-step reductions per trace is 26 for the \ucsdbench\
 dataset (27 for the \uwbench\ dataset) with a maximum of 2,744 (659 for
@@
 % \end{itemize}
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2015-11-20 paper/pldi16-submission/introduction.tex
update fac-overview and unhack code font size
tweak fac-overview
done interactive
misc edits
Wes comments. minor wording/typo issues checked in directly, others left as "WRW" comments (search) that I will mention via email.

@@
 \begin{verbatim}
- File "fac.ml", line 5, characters 8-23:
+ File "fac.ml", line 3, characters 8-23:
  Error: This expression has type bool
@@
 \begin{minipage}{.49\linewidth}
+\centering
 \begin{code}
@@
 \end{minipage}
-\hspace{0.3in}
-\begin{minipage}{.3\linewidth}
-  \includegraphics[height=2.5in]{fac-overview.png}
+% \hspace{0.3in}
+\begin{minipage}{.49\linewidth}
+\centering
+  \includegraphics[height=1.5in]{fac-overview.png}
 \end{minipage}
-\caption{(L) An ill-typed \texttt{fac} function; (R) Dynamically witnessing the type error in \texttt{fac}}
+\caption{(L) An ill-typed \texttt{fac} function; (R) Dynamically witnessing the type error in \texttt{fac}.}
 \label{fig:factorial}

### 2015-11-20 paper/pldi16-submission/semantics.tex

@@
 so, returns the existing instantiation:
+% WRW thinks we should explain why \force(\ehole,t,sigma) 
+% is just sigma(i) and not \force(sigma(i),t, sigma) -- at this
+% point the reader may not understand, especially since just two paragraphs
+% above you made a big deal about sigma mapping holes to things that
+% contain other holes.
 %
@@
 %
+% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
+% how the let rule would work for something "let rec"-y, and there's not
+% function call rule). I only mention this because I can imagine a reviewer
+% wondering about your ability to generate good witnesses for function
+% types. This could likely be addressed in text, by a forward reference to
+% Section 3.4 where higher-order functions are handled, without changing
+% any of the formalisms at the last minute.
 Figure~\ref{fig:operational} describes the small-step contextual
@@
 
+% WRW is still minorly worried about this theorem as phrased (cf. every
+% inhabitable type). Perhaps jsut a footnote explaining why \alpha -> \beta
+% is not considered here?  
 \begin{thm}{\textbf{[Witness Generality]}}
@@
 
+% WRW: Just to be sure, the type of "lambda x. x + 1" would seem
+% to be box_1 -> box_2 here, while one might be tempted to get
+% int -> int. Some extra prose text contrasting typeof() with
+% "apply f to box to get a partial type" would likely help. 
 \paragraph{The Type of an Expression} The (dynamic) type of an
@@
     %
+    % WRW suggests one more sentence here: do the thinking for the reader.
+    % Why does t being incompatible with s_\tau mean that narrow
+    % cannot succeed? I think officially you need to refer to Preservation
+    % here in your induction, to argue that the type being narrowed *to*
+    % is the same as the types you introduced earlier, and *thus* the
+    % narrow must fail (because they're incompatible). Otherwise readers
+    % might imagine a weird case where t and s_\tau are incompatible, but
+    % the final narrow is to some other type t'' that is compatible. 
+    %
+    % This is a primary result, so more detail may not hurt.  
     A case analysis of the rules shows that \forcesym cannot have
@@
 %
-Next, we describe what its input and output, and how it
+Next, we describe its input and output, and how it
 addresses the above challenges to search the space of possible
@@
 %
-Obviously, the problem of generating inputs is undecidable in general.
+The problem of generating inputs is undecidable in general.
 %
-The witness generation procedure takes two inputs.
+Our witness generation procedure takes two inputs.
 %
@@
 %
-In the sequel, we refer to @fn@ as the \emph{target} of witness generation.
+In what follows, we refer to @fn@ as the \emph{target} of witness generation.
 
@@
 Due to the non-determinism introduced by \gensym, a call \hbox{|eval($e$, $\vsu$)|}
-returns the \emph{list} possible results of the form $\pair{e'}{\vsu'}$ such
+returns the \emph{list} of possible results of the form $\pair{e'}{\vsu'}$ such
 that $\steps{e}{\vsu}{e'}{\vsu'}$.
@@
 This is achieved via a simple @loop@ that keeps adding holes until
+% WRW: Where / to what does the loop add holes? Be specific for the reader. 
 @eval@uating the closed term yields a non-lambda value.

### 2015-11-20 paper/pldi16-submission/interactive.tex
compact fac-steps to save space
Wes minor changes to the second half of the paper. See WRW comments.

@@
 an novel interactive debugger that allows the user
-to effectively visualize \emph{how} the program goes (wrong.)
+to effectively visualize \emph{how} the program goes (wrong).
 
@@
   Figure~\ref{fig:operational}, extended to collect a full reduction
+  % WRW: \subterms is Used here in the figure before being Defined (indeed,
+  % it is never defined anywhere that I can see). A simple bit in text to
+  % explain what you mean would help. Why is subterms necessary? What does
+  % it collect up for you? 
   graph.}
@@
 A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
-intuitively indicates that $e_1$ steps, in a single step, to $e_2$.
+% WRW: This was "e_1 steps, in a single step" which was circular.
+intuitively indicates that $e_1$ reduces, in a single step, to $e_2$.
 %
@@
 they do, and finally how we use a command to \emph{update}
-the visualization state. In the sequel, for clarity of
+the visualization state. In what follows, for clarity of
 exposition, we assume we have a (global) trace:

### 2015-11-20 paper/pldi16-submission/semantics.tex
address some WRW comments

@@
 
-% WRW is still minorly worried about this theorem as phrased (cf. every
-% inhabitable type). Perhaps jsut a footnote explaining why \alpha -> \beta
-% is not considered here?  
 \begin{thm}{\textbf{[Witness Generality]}}
@@
   For any function $f$, if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
-  then for every type $t$ there exists $\hastype{v}{t}$ such that
+  then for every inhabitable type\footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b.
+\tfun{a}{b}$ to be uninhabitable.}  $t$ there exists $\hastype{v}{t}$ such that
   $\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
@@
 
-% WRW: Just to be sure, the type of "lambda x. x + 1" would seem
-% to be box_1 -> box_2 here, while one might be tempted to get
-% int -> int. Some extra prose text contrasting typeof() with
-% "apply f to box to get a partial type" would likely help. 
 \paragraph{The Type of an Expression} The (dynamic) type of an
@@
   \]
+%
+Note that we approximate the type of expressions rather than
+performing full type inference, for example
+$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun{\thole{}}{\thole{}}$,
+instead of the usual $\tfun{\tint}{\tint}$.
 

### 2015-11-20 paper/pldi16-submission/evaluation.tex

@@
 \caption{Results of our coverage testing and the distribution of test
+  % WRW notes that the pie chart does not print on standard department
+  % printer, and is emailing a photo as proof. Possible solution:
+  % convert PDF -> PNG and include PNG? 
   outcomes. Our random search successfully finds witnesses for 83\% of
@@
 %
-A jump compressed trace is one whose edges are limited to forward
+A \emph{jump compressed} trace is one whose edges are limited to forward
 or backward jumps. (The sizes of these traces is equal to the jump
@@
 pairs.
+% WRW: Say more here. This is critical to your evaluatino and to your
+% conclusion (e.g., where we say that jump-compressed traces hide boring
+% details). Why are these details boring and others relevant? Do the
+% boring ones contain less information somehow? 
 %

### 2015-11-20 paper/pldi16-submission/interactive.tex
kill subterm
more WRW comments

@@
   {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
-  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \subterms{C[e_1]}; \tr_1}}
+  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \tr_1}}
 \\ \\
@@
   {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
-         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
+         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \tr}}
 \\ \\
@@
   Figure~\ref{fig:operational}, extended to collect a full reduction
-  % WRW: \subterms is Used here in the figure before being Defined (indeed,
-  % it is never defined anywhere that I can see). A simple bit in text to
-  % explain what you mean would help. Why is subterms necessary? What does
-  % it collect up for you? 
   graph.}
@@
 A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
-% WRW: This was "e_1 steps, in a single step" which was circular.
 intuitively indicates that $e_1$ reduces, in a single step, to $e_2$.
@@
 \centering
-\includegraphics[width=\linewidth]{fac-steps.png}
+\includegraphics[width=\linewidth]{fac-steps-new.png}
 \caption{A sequence of interactions with the trace of

### 2015-11-20 paper/pldi16-submission/semantics.tex
more WRW comments

@@
 so, returns the existing instantiation:
-% WRW thinks we should explain why \force(\ehole,t,sigma) 
-% is just sigma(i) and not \force(sigma(i),t, sigma) -- at this
-% point the reader may not understand, especially since just two paragraphs
-% above you made a big deal about sigma mapping holes to things that
-% contain other holes.
 %
@@
 % \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
+\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
 \force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
@@
   % \end{array} \\
-\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
 \force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
@@
 $$
+%
+While a hole may map to a value that \emph{contains} another hole, \eg a
+lambda, it may not map \emph{directly} to another hole, so in the second
+case above we do not need to \forcesym\ the result of the substitution.
 
@@
 \end{lem}
-% \ES{self-todo: take a close look at the $\trace$'s vs k's}
 \begin{proof}
@@
     %
-    % WRW suggests one more sentence here: do the thinking for the reader.
-    % Why does t being incompatible with s_\tau mean that narrow
-    % cannot succeed? I think officially you need to refer to Preservation
-    % here in your induction, to argue that the type being narrowed *to*
-    % is the same as the types you introduced earlier, and *thus* the
-    % narrow must fail (because they're incompatible). Otherwise readers
-    % might imagine a weird case where t and s_\tau are incompatible, but
-    % the final narrow is to some other type t'' that is compatible. 
+    Lemma~\ref{lem:refine-partial} tells us
+    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
+    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
     %
-    % This is a primary result, so more detail may not hurt.  
     A case analysis of the rules shows that \forcesym cannot have
@@
 %
-This is achieved via a simple @loop@ that keeps adding holes until
-% WRW: Where / to what does the loop add holes? Be specific for the reader. 
-@eval@uating the closed term yields a non-lambda value.
+This is achieved via a simple @loop@ that keeps adding holes to the
+target application until @eval@uating the closed term yields a
+non-lambda value.
 %

### 2015-11-20 paper/pldi16-submission/evaluation.tex
more WRW comments
replace pdf figures with pngs

--- 2015-11-20 paper/pldi16-submission/evaluation.tex
@@
 \begin{minipage}{\linewidth}
-\includegraphics[width=\linewidth]{coverage.pdf}
+\includegraphics[width=\linewidth]{coverage.png}
 \end{minipage}
 \begin{minipage}{\linewidth}
-\includegraphics[width=\linewidth]{distrib.pdf}
+\includegraphics[width=\linewidth]{distrib.png}
 \end{minipage}
@@
 pairs.
+%
+Expanding the initial visualization to a jump-compressed trace is often
+an ideal first step;
+% expanding a jump-compressed traces ideal for the initial state of the
+they provide a concise overview of the execution and allow the user to
+step directly into an interesting function call, at which point the can
+proceed with more fine-grained steps.
 %
@@
 \centering
-% \includegraphics[width=\linewidth]{trace_size.pdf}
-% \begin{minipage}{0.49\linewidth}
-\includegraphics[width=\linewidth]{trace_size_step.pdf}
-% \end{minipage}
-% \begin{minipage}{0.49\linewidth}
-\includegraphics[width=\linewidth]{trace_size_jump.pdf}
-% \end{minipage}
+\includegraphics[width=\linewidth]{trace_size_step.png}
+\includegraphics[width=\linewidth]{trace_size_jump.png}
 \caption{Complexity of the generated traces. At least 90\% of the traces

### 2015-11-20 paper/pldi16-submission/introduction.tex
replace pdf figures with pngs

@@
 \begin{verbatim}
- File "fac.ml", line 3, characters 8-23:
+ File "fac.ml", line 5, characters 8-17:
  Error: This expression has type bool
@@
 static type errors by \emph{dynamically} witnessing
-how exactly an ill-typed program goes wrong.
+how an ill-typed program goes wrong.
 %

### 2015-11-20 paper/pldi16-submission/evaluation.tex
add para about recursion, and figure demonstrating jump compression

@@
 computation that are often uninteresting or irrelevant to the
-explanation, or at any rate, easily summarized as call-return
-pairs.
+explanation.
 % 
-Expanding the initial visualization to a jump-compressed trace is often
-an ideal first step;
-% expanding a jump-compressed traces ideal for the initial state of the
-they provide a concise overview of the execution and allow the user to
-step directly into an interesting function call, at which point the can
-proceed with more fine-grained steps.
+In particular, jump compressed traces hide low-level operations and
+summarize function calls as call-return pairs, see
+Figure~\ref{fig:fac-jump} for a variant of @fac@ that implements the
+subtraction as a function call instead of a primitive.
+%
+\begin{figure}[t]
+\centering
+\includegraphics[height=1.5in]{fac-minus.png}
+\caption{Jump-compressed trace of \texttt{fac 1} with subtraction
+  implemented as a function call.}
+\label{fig:fac-jump}
+\end{figure}
+%
+Once the user has identified an interesting call-return pair, they can
+step into the call and proceed with more fine-grained steps.
+%
+% Expanding the initial visualization to a jump-compressed trace is often
+% an ideal first step;
+% % expanding a jump-compressed traces ideal for the initial state of the
+% they provide a concise overview of the execution and allow the user to
+% step directly into an interesting function call, at which point the can
+% proceed with more fine-grained steps.
 %

### 2015-11-20 paper/pldi16-submission/related.tex
add para about recursion, and figure demonstrating jump compression

@@
 
-\subsection{Type Errors}
-\label{sec:type-error}
+% \paragraph{Type Errors}
+% \label{sec:type-error}
 
-\paragraph{Diagnosis and Repair}
+\paragraph{Diagnosing and Repairing Type Errors}
 \label{sec:diagnosis-repair}
@@
 
-\subsection{Testing}
+\paragraph{Testing}
 \label{sec:testing}
@@
 
-\subsection{Program Exploration}
+\paragraph{Program Exploration}
 Perera \etal~\cite{perera_functional_2012} present a tracing semantics

### 2015-11-20 paper/pldi16-submission/semantics.tex
add para about recursion, and figure demonstrating jump compression

@@
 lambda, it may not map \emph{directly} to another hole, so in the second
-case above we do not need to \forcesym\ the result of the substitution.
+case above we do not need to \forcesym\ the result of the substitution,
+the sub-hole will be narrowed when the context demands it.
 
@@
 computation. Importantly, beta-reduction \emph{does not} type-check its
-argument, it only ensures that the value being applied is a function.
+argument, it only ensures that $v_1$ is a function.
+
+\paragraph{Recursion}
+Fixed-point operators often cannot be typed in static type systems, but
+we are not concerned with \emph{assigning} types to terms, rather with
+showing that \emph{no type} can be assigned. We are simply executing the
+untyped $\lambda$-calculus, which has no issue handling recursion.
 

### 2015-11-20 paper/pldi16-submission/related.tex
slightly bigger figure

@@
 \emph{partial value} -- containing holes -- and present a partial slice,
-containing only those steps that affected the the partial value. This
-system is designed to answer questions of the form ``Where did this
-value come from?'' and thus is focused on backward exploration. In
-contrast, our trace visualization supports forward \emph{and} backward
+containing only those steps that affected the the partial value.
+% This
+% system is designed to answer questions of the form ``Where did this
+% value come from?'' and thus is focused on backward exploration. 
+Perera's system focuses on backward exploration; in
+contrast, our visualization supports forward \emph{and} backward
 exploration, though our backward steps are more limited than

### 2015-11-20 paper/pldi16-submission/plots.py
graph colors

@@ -6,7 +6,8 @@ import numpy as np
 UCSD = 'AUTHORS'
 
 BUCKETS = [0.1, 0.2, 1.0, 10.0, 60.0 ] # range(500, 3001, 500)
-COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
+COLORS=['#90B0D4', '#90D492', '#D4B490', '#D490D2']
+#COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
 
 SAFE = ['S', 'T']
 SAFE_L = ['Safe', 'Timeout']
@@ -15,8 +16,8 @@ UNSAFE_L = ['Unsafe', 'Unbound', 'Diverge'] #, 'Output']
 ALL = UNSAFE + SAFE
 ALL_L = UNSAFE_L + SAFE_L
 
-ALL_D = [ ['S', 'T'], ['U'], ['B'], ['D'] ]
-ALL_DL = [ 'No Witness', 'Witness', 'Unbound', 'Diverge' ] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
+ALL_D = [ ['U'], ['B'], ['D'], ['S', 'T']]
+ALL_DL = [ 'Witness', 'Unbound', 'Diverge', 'No Witness'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
 
 
 def read_csv(f):

### 2015-11-20 paper/pldi16-submission/introduction.tex
minor typos in abstract and intro

@@
 We show that for both data sets, our technique is able to generate
-witnesses for nearly 90\% of the programs, in under a second the
+witnesses for nearly 90\% of the programs, in under a second in the
 vast majority of cases.

### 2016-01-11 paper/pldi16-submission/semantics.tex
thread \vsu through \force per reviewer B suggestion.
cleaner presentation and fixes an annoying typo-bug

@@
 input an term \texttt{e} and refines it to have the type \texttt{t} by
-yielding a pair of either the same expression and an empty value
+yielding a pair of either the same expression and value
 substitution, or yields the stuck term if no such refinement is
@@
 \begin{array}{lcl}
-% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
-% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
-\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
+%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
+\forcesym            & ::     & (e, t, \vsu) \rightarrow \pair{e}{\vsu} \\
+\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\vsu} \\
 \force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
 \begin{cases}
-  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
-  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
+  \pair{\vsu(i)}{\vsu}        & \mbox{if}\ i \in dom(\vsu) \\
+  \pair{v}{\ehole{i} \mapsto v ; \vsu}   & v = \gen{t}\\
 \end{cases} \\
  % \begin{array}{l}
-    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
+    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\vsu}\ \mathtt{else} \\
          % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
   % \end{array} \\
-\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
-\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
-\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
-\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
-\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
-\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
+\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\vsu} \\
+\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\vsu} \\
+\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\vsu} \\
+\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\vsu} \\
+\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\vsu} \\
+\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\vsu} \\
 \end{array}

### 2016-01-13 paper/pldi16-submission/response.txt
add first draft of response

We thank the reviewers for their enthusiastic comments, and will briefly
respond to the main critiques.

Formalism
---------

While we have implemented (and evaluated) our technique on the pure
subset of OCaml -- including algebraic datatypes, polymorphism, and type
annotations -- we restricted the formalism to a simpler language because
the additional constructs do not any major complexity to the technique.
We simply treat constructor application, case analysis, type annotation,
etc. as primitive operations like `+` and `if`; they narrow the partial
types of values and force instantiation. However, in light of the
reviews it is clear that this is not obvious to the reader, so we will
include an extended formal semantics in the next iteration of the paper.

We also thank Reviewer B in particular for pointing out the two bugs in
our proof. We have refactored the substitutions in the formalism to be
properly threaded through `narrow` and `E-Plus-Good`, and have changed
our definition of type compatibility to the usual one. While partial
types do grow monotonically, in Lemma 4 specifically we are concerned
with *any* value (and type) so the example with incompatible pair types
applies.

Representivity of the datasets
------------------------------

Between our dataset and the UW dataset, we have numerous instances of
writing higher-order functions and instantiating their arguments, as
well as two different expression evaluators. So we believe the
benchmarks are representative of the range of programs an undergraduate
student may have to write.

Pedagogic motivation for witnesses and traces
---------------------------------------------

We are also convinced of the importance of witnesses and traces for
debugging and did not intend this paper to specifically present evidence
in their favor, rather to show that generating witnesses is feasible and
*practical* for type errors. (This is also why we did not discuss
producing witnesses for other runtime errors like divide-by-zero.) We
will update the motivation to be clearer about the scope of the paper
and results.

### 2016-01-13 paper/pldi16-submission/response.txt
Update response.txt

@@
 in their favor, rather to show that generating witnesses is feasible and
-*practical* for type errors. (This is also why we did not discuss
+practical for type errors. (This is also why we did not discuss
 producing witnesses for other runtime errors like divide-by-zero.) We

### 2016-01-20 paper/icfp16-submission/response.txt
copy to icfp16-submission, switch to icfp class

We thank the reviewers for their enthusiastic comments, and will briefly
respond to the main critiques.

Formalism
---------

While we have implemented (and evaluated) our technique on a pure subset
of OCaml -- including algebraic datatypes, polymorphism, and type
annotations -- we restricted the formalism to a simpler language because
the additional constructs do not add significant complexity to the
technique. We simply treat constructor application, case analysis, type
annotation, etc. as primitive operations like `+` and `if`; they narrow
the partial types of values and force instantiation. However, in light
of the reviews it is clear that this is not obvious to the reader, so we
will add text to Section 3 to list additional language features that can
be handled as primitive operations from the perspective of our
semantics.

We also thank Reviewer B in particular for pointing out two bugs in our
proof. We have refactored the substitutions in the formalism to be
properly threaded through `narrow` and `E-Plus-Good`. We have also
changed our definition of type compatibility to the usual one, fixing the

  (([]_1, []_2), []_3) ~ ([]_4, ([]_5, []_6))

example (while partial types do grow monotonically, in Lemma 4
specifically we are concerned with *any* value, so we must account for
this case).

Representivity of the datasets
------------------------------

Between our dataset and the UW dataset, we have numerous instances of
writing higher-order functions and instantiating their arguments, as
well as two different expression evaluators. We will expand the text in
Section 5 to describe the generality of the datasets, including a list
of language features and idioms that are present.

Pedagogic motivation for witnesses and traces
---------------------------------------------

We are also convinced of the importance of witnesses and traces for
debugging and did not intend this paper to specifically present evidence
in their favor. We will update the motivation and conclusion to clarify
that the scope of the paper is to present algorithmic and empirical
evidence that generating witnesses is practical for type errors.

### 2016-01-20 paper/icfp16-submission/abstract.tex
copy to icfp16-submission, switch to icfp class

--- 2015-11-20 paper/pldi16-submission/abstract.tex
@@
   %
-  First, given an ill-typed function, we first symbolically
+  First, given an ill-typed function, we symbolically
   execute the body to dynamically synthesize witness values
@@
   88\% of the programs, and our reduction graph
-  is yields small counterexamples for 90\% of the witnesses.
+  yields small counterexamples for 90\% of the witnesses.
 \end{abstract}

### 2016-01-20 paper/icfp16-submission/plots.py

@@ -1,0 +1,232 @@
+import csv
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+
+UCSD = 'AUTHORS'
+
+BUCKETS = [0.1, 0.2, 1.0, 10.0, 60.0 ] # range(500, 3001, 500)
+COLORS=['#90B0D4', '#90D492', '#D4B490', '#D490D2']
+#COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
+
+SAFE = ['S', 'T']
+SAFE_L = ['Safe', 'Timeout']
+UNSAFE = ['U', 'B', 'D'] #, 'O']
+UNSAFE_L = ['Unsafe', 'Unbound', 'Diverge'] #, 'Output']
+ALL = UNSAFE + SAFE
+ALL_L = UNSAFE_L + SAFE_L
+
+ALL_D = [ ['U'], ['B'], ['D'], ['S', 'T']]
+ALL_DL = [ 'Witness', 'Unbound', 'Diverge', 'No Witness'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
+
+
+def read_csv(f):
+    with open(f) as f:
+        return list(csv.reader(f))
+
+def autolabel(ax, rects):
+    # attach some text labels
+    for rect in rects:
+        height = rect.get_height()
+        ax.text(rect.get_x() + rect.get_width()/2., height,
+                '%d' % int(height),
+                ha='center', va='bottom')
+
+def cumulative_coverage(data):
+    headers = data[0]
+    data = data[1:]
+    return [(l, int(100 * len([r for r in data
+                               if float(r[2]) <= l
+                               and r[4] in UNSAFE])
+                          / float(len(data))))
+            for l in BUCKETS]
+
+def plot_coverage(seminal, ucsd):
+    xy_s = cumulative_coverage(seminal)
+    xy_u = cumulative_coverage(ucsd)
+
+
+    N = len(xy_s)
+    ind = np.arange(N)    # the x locations for the groups
+    width = 0.35       # the width of the bars: can also be len(x) sequence
+
+    fig = plt.figure()
+    p1 = plt.bar(ind, [r[1] for r in xy_s], width,
+                 color=COLORS[0])
+    p2 = plt.bar(ind + width, [r[1] for r in xy_u], width,
[…]

### 2016-02-27 paper/icfp16-submission/semantics.tex
trascribing the extended semantics

--- 2016-01-11 paper/pldi16-submission/semantics.tex
@@
 Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
-calculus with integers and booleans.
+calculus with integers, booleans, and binary trees.
 %
 As we are specifically interested in programs that \emph{do} go wrong,
-we include an explicit \stuck\ state in our syntax.
+we include an explicit \stuck\ configuration in our syntax.
 
@@
 We solve this problem be equipping our term language with a notion
-of a \emph{hole}, written \ehole{}, which represents \emph{unconstrained}
+of a \emph{hole}, written \ehole, which represents \emph{unconstrained}
 values that may replaced with values from \emph{any} type.
 %
-Dually, holes may also appear in types, where they may be thought
-of as type variables that we will not generalize over.
+Dually, holes may also appear in types, written \thole, where they may
+be thought of as type variables that we will not generalize over.
 %
@@
 \label{sec:semantics}
-\input{operational.tex}
 %
@@
 
-\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
-input an term \texttt{e} and refines it to have the type \texttt{t} by
-yielding a pair of either the same expression and value
-substitution, or yields the stuck term if no such refinement is
-possible. In the case where \texttt{e} is a hole, it first checks in the
-given $\vsu$ to see if the hole has already been instantiated and, if
-so, returns the existing instantiation:
+\paragraph{Narrowing Types} The procedure $\force{v}{t}{\vsu}{\tsu}$,
+defined in Figure~\ref{fig:narrow}, takes as input a value $v$ and
+refines it to have the type $t$ by yielding a pair of either the same
+value and value substitution, or yields the stuck term if no such
+refinement is possible. In the case where $v$ is a hole, it first checks
+in the given $\vsu$ to see if the hole has already been instantiated
+and, if so, returns the existing instantiation.
 %
+\begin{figure*}[t]
 $$
@@
 %% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
-\forcesym            & ::     & (e, t, \vsu) \rightarrow \pair{e}{\vsu} \\
-\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\vsu} \\
-\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
+\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
+\force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
 \begin{cases}
-  \pair{\vsu(i)}{\vsu}        & \mbox{if}\ i \in dom(\vsu) \\
-  \pair{v}{\ehole{i} \mapsto v ; \vsu}   & v = \gen{t}\\
+  \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
+  \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
 \end{cases} \\
+\force{\vhole{t_1}}{t_2}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
+\begin{cases}
+  \triple{v}{\vsu}{\tsu''}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
+                                         \tsu' = \unify{t_1}{t_2}{\tsu},
+                                         \tsu'' = \unify{\subst{\tsu'}{t_1}}
+                                                        {\typeof{v}}{\tsu'} \\
+  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
+  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu}, v = \gen{\subst{\tsu'}{t_1}}\\
+\end{cases} \\
+% \begin{cases}
+%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
+%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
+%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
+% \end{cases} \\
  % \begin{array}{l}
-    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\vsu}\ \mathtt{else} \\
-         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
+    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
+         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
   % \end{array} \\
-\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\vsu} \\
-\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\vsu} \\
-\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\vsu} \\
-\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\vsu} \\
-\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\vsu} \\
-\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\vsu} \\
+\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
+\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
+\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
+\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
+\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
+\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
 \end{array}
 $$
+\caption{Narrowing values}
+\label{fig:narrow}
+\end{figure*}
 %
 While a hole may map to a value that \emph{contains} another hole, \eg a
-lambda, it may not map \emph{directly} to another hole, so in the second
-case above we do not need to \forcesym\ the result of the substitution,
-the sub-hole will be narrowed when the context demands it.
+lambda or a tree, it may not map \emph{directly} to another hole, so in
+the second case above we do not need to \forcesym\ the result of the
+substitution, the sub-hole will be narrowed when the context demands it.
 
@@
 a lambda with a \emph{new} hole denoting the return value, and unconstrained
-types (denoted by $\thole{}$) yield fresh, unconstrained values
-(denoted by $\ehole{i}$).
+types (denoted by $\thole$) yield fresh, unconstrained values
+(denoted by $\ehole$).
 
@@
 \begin{array}{lcll}
-% \gensym       & ::      & t \rightarrow v \\
+\gensym       & ::      & t \rightarrow v \\
 \gen{\tint}   & \defeq  & n, & \quad \text{non-deterministic} \\
 \gen{\tbool}  & \defeq  & b, & \quad \text{non-deterministic} \\
-\gen{\tfun{t_1}{t_2}}   & \defeq & \efun{x}{\ehole{i}}, & \quad \text{$i$ is fresh} \\
-\gen{\thole{}} & \defeq & \ehole{i}, & \quad \text{$i$ is fresh} \\
+\gen{\ttree{t}}  & \defeq  & tr, & \quad \text{non-deterministic, well-typed} \\
+\gen{\tfun}   & \defeq & \efun{x}{\ehole}, & \quad \text{\ehole is fresh} \\
+\gen{\thole}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
 \end{array}
@@
 \paragraph{Steps and Traces}
+\input{operational.tex}
 %
@@
 %
-A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
-and a substitution $\vsu$.
+A \emph{closed expression} is a triple $\triple{e}{\vsu}{\tsu}$ of an expression $e$,
+a value substitution $\vsu$, and a type substitution $\tsu$.
 %
-We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
-transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
+We write $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ if the closed
+expression $\triple{e}{\vsu}{\tsu}$ transitions in a \emph{small step} to
+$\triple{e'}{\vsu'}{\tsu'}$.
 %
 A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
-$\pair{e_0}{\vsu_0}, \ldots, \pair{e_n}{\vsu_n}$ such that
-$\forall 0 \leq i < n$, we have \step{e_i}{\vsu_i}{e_{i+1}}{\vsu_{i+1}}.
+$\triple{e_0}{\vsu_0}{\tsu_0}, \ldots, \triple{e_n}{\vsu_n}{\tsu_n}$ such that
+$\forall 0 \leq i < n$, we have
+$\step{e_i}{\vsu_i}{\tsu_i}{e_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
 %
-We write \steptr{\trace}{e}{\vsu}{e'}{\vsu'} if $\trace$ is a trace of the form
-$\pair{e}{\vsu},\ldots,\pair{e'}{\vsu'}$.
+We write \steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if $\trace$ is
+a trace of the form $\triple{e}{\vsu}{\tsu},\ldots,\pair{e'}{\vsu'}{\tsu'}$.
 %
-We write \steps{e}{\vsu}{e'}{\vsu'} if \steptr{\trace}{e}{\vsu}{e'}{\vsu'}
-for some trace $\trace$.
+We write \steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if
+\steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} for some trace $\trace$.
 
 \paragraph{Primitive Reductions}
-Primitive reduction steps --- addition, if-elimination, and
-function application --- use \forcesym to ensure that values have the
-appropriate type (and that holes are instantiated) before continuing the
-computation. Importantly, beta-reduction \emph{does not} type-check its
-argument, it only ensures that $v_1$ is a function.
+Primitive reduction steps --- addition, if-elimination, function
+application, and data construction and case analysis --- use \forcesym
+to ensure that values have the appropriate type (and that holes are
+instantiated) before continuing the computation. Importantly,
+beta-reduction \emph{does not} type-check its argument, it only ensures
+that $v_1$ is a function.
 
@@
 \emph{general} manner. That is, given a function $f$, if we have
-$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$, then for every
-possible input type $t$, we can find a value $\hastype{v}{t}$
-such that $\eapp{f}{v}$ goes wrong.
+$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
+then for every possible input type $t$, we can find a value
+$\hastype{v}{t}$ such that $\eapp{f}{v}$ goes wrong.
 
@@
 \label{thm:soundness}
-  For any function $f$, if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
-  then for every inhabitable type\footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b.
-\tfun{a}{b}$ to be uninhabitable.}  $t$ there exists $\hastype{v}{t}$ such that
-  $\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
+  For any function $f$, if\\
+  \hbox{$\steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
+  then for every % inhabitable
+  type
+  % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
+  $t$ there exists $\hastype{v}{t}$ such that
+  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
 \end{thm}
@@
     \typeof{b}   & \defeq & \tbool & \\
-    \typeof{\efun{x}{e}} & \defeq & \tfun{\thole{i}}{\typeof{e}}, & \quad \text{$i$ is fresh} \\
-    \typeof{e} & \defeq & \thole{i}, & \quad \text{$i$ is fresh} \\
+    \typeof{\efun{x}{e}} & \defeq & \tfun \\
+    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
+    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
+    \typeof{\vhole{t}} & \defeq & t \\
+    \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
   \end{array}
@@
 performing full type inference, for example
-$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun{\thole{}}{\thole{}}$,
-instead of the usual $\tfun{\tint}{\tint}$.
+$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
+instead of the usual $\tint \rightarrow \tint$.
 
 \paragraph{Dynamic Type Inference}
-We can think of the evaluation of \eapp{f}{\ehole{}} as (dynamically)
+We can think of the evaluation of \eapp{f}{\ehole} as (dynamically)
 inferring a \emph{partial type} --- a type that may contain holes ---
@@
 %
-Formally, we say that if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su},
+Formally, we say that if
+$\steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$,
 then the \emph{partial type} of $f$ upto $\trace$, written \ptype{\trace}{f},
-is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
+is $\typeof{\subst{\su}{\ehole{}}} \rightarrow \typeof{e}$.
 %
@@
 \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
-$\exists \tsu.\ t = \subst{\tsu}{s} \lor s = \subst{\tsu}{t}$.
+$\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.
 
@@
 \label{lem:refine-partial}
-If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
-$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is a single-step
+If $\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots$ and
+$\trace' \defeq \trace, \triple{e'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
 extension of $\trace$)
@@
 If
-$\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu}, \ldots,\pair{e}{\vsu}$
+$\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
 and
-$\trace' \defeq \trace, \step{e}{\vsu}{e'}{\vsu'}$
+$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
 and
 $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
-then the final step $\step{e}{\vsu}{e'}{\vsu'}$ invokes \forcesym.
+then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ invokes \forcesym.
 \end{lem}
@@
 \label{lem:k-stuck}
-  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu} and
+  If \steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
      \hastype{v}{t} and
@@
   then
-     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
+     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}.
 \end{lem}
@@
   By induction on $k$, the length of $\trace$.
-  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$ and
-  $\trace' = \pair{\eapp{f}{\ehole{}}}{\emptysu},...,\pair{e'}{\vsu'}$ 
-  such that $\step{e'}{\vsu'}{e}{\vsu}$.
+  Let $s_{\trace} \rightarrow \_ = \ptype{\trace}{f}$ and
+  $\trace' = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
+  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
   Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
-  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
+  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}
   in at most $k$ steps:
@@
 %
+\ES{HERE}
 \begin{mcode}
-  eval :: ($e$, $\su$) -> [($v$, $\su$)]
+  eval :: ($e$, $\vsu$, $\tsu$) -> [($v$, $\vsu$, $\tsu$)]
 \end{mcode}
 %
-Due to the non-determinism introduced by \gensym, a call \hbox{|eval($e$, $\vsu$)|}
-returns the \emph{list} of possible results of the form $\pair{e'}{\vsu'}$ such
-that $\steps{e}{\vsu}{e'}{\vsu'}$.
+Due to the non-determinism introduced by \gensym, a call
+\hbox{|eval($e$, $\vsu$, $\tsu$)|} returns the \emph{list} of possible results
+of the form $\triple{e'}{\vsu'}{\tsu'}$ such that
+$\steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$.
 %
@@
 
-close :: [($x$, $e$)] -> [$v$] -> ($e$, $\vsu$)
-close bs vs = (e, $\emptysu$)
+close :: [($x$, $e$)] -> [$v$] -> ($e$, $\vsu$, $\tsu$)
+close bs vs = (e, $\emptysu$, $\emptysu$)
   where
@@
   then for every type $t$ there exists $\hastype{v}{t}$ such that
-  $\steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
+  $\steps{\eapp{f_n}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
 \end{cor}
@@
    vs  = genArgs defs []        -- (1)
-   $e$    = close defs vs        -- (2)
-   res = take n (eval e)      -- (3)
-   $\vsu$s   = [$\vsu$ | $\pair{\stuck}{\vsu}$ <- res]  -- (4)
+   cfg = close defs vs          -- (2)
+   res = take n (eval cfg)      -- (3)
+   $\vsu$s   = [$\vsu$ | $\triple{\stuck}{\vsu}{\tsu}$ <- res]  -- (4)
 \end{mcode}

### 2016-02-29 paper/icfp16-submission/semantics.tex
start adding appendix with proofs
abakst comments

@@
 of a \emph{hole}, written \ehole, which represents \emph{unconstrained}
-values that may replaced with values from \emph{any} type.
+values that may be replaced with values from \emph{any} type.
 %
@@
 We track how the holes are incrementally filled in, by using
-(value) \emph{substitutions} that map holes (denoted by their index)
+(value) \emph{substitutions} that map holes % (denoted by their index)
 to values which may, in turn, contain holes.
@@
 
-\paragraph{Narrowing Types} The procedure $\force{v}{t}{\vsu}{\tsu}$,
-defined in Figure~\ref{fig:narrow}, takes as input a value $v$ and
-refines it to have the type $t$ by yielding a pair of either the same
-value and value substitution, or yields the stuck term if no such
-refinement is possible. In the case where $v$ is a hole, it first checks
-in the given $\vsu$ to see if the hole has already been instantiated
-and, if so, returns the existing instantiation.
+\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
+$$
+\forcesym :: (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
+$$
+defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
+$t$, and the current value and type substitutions, and refines $v$ to
+have type $t$ by yielding a triple of either the same value and
+substitutions, or yields the stuck state if no such refinement is
+possible. In the case where $v$ is a hole, it first checks in the given
+$\vsu$ to see if the hole has already been instantiated and, if so,
+returns the existing instantiation.
+\ES{explain the subtlety of narrowing a value hole to an uninstantiated type}
 %
-\begin{figure*}[t]
+\begin{figure*}[ht]
 $$
@@
 lambda or a tree, it may not map \emph{directly} to another hole, so in
-the second case above we do not need to \forcesym\ the result of the
+the second case of \forcesym\ we do not need to \forcesym\ the result of the
 substitution, the sub-hole will be narrowed when the context demands it.
@@
 \paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
-takes as input a type \texttt{t} and returns a value of that type. For base
+takes as input a type $t$ and returns a value of that type. For base
 types, the procedure returns an arbitrary value. For functions, it returns
@@
 %
-A \emph{closed expression} is a triple $\triple{e}{\vsu}{\tsu}$ of an expression $e$,
-a value substitution $\vsu$, and a type substitution $\tsu$.
+A \emph{configuration} is a triple $\triple{e \cup \stuck}{\vsu}{\tsu}$
+of an expression $e$ or the stuck term $\stuck$, a value substitution
+$\vsu$, and a type substitution $\tsu$.
 %
-We write $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ if the closed
-expression $\triple{e}{\vsu}{\tsu}$ transitions in a \emph{small step} to
+We write $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ if the configuration
+$\triple{e}{\vsu}{\tsu}$ transitions in a \emph{small step} to
 $\triple{e'}{\vsu'}{\tsu'}$.
 %
-A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
+A (finite) \emph{trace} $\trace$ is a sequence of configurations
 $\triple{e_0}{\vsu_0}{\tsu_0}, \ldots, \triple{e_n}{\vsu_n}{\tsu_n}$ such that
@@
 We write \steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if $\trace$ is
-a trace of the form $\triple{e}{\vsu}{\tsu},\ldots,\pair{e'}{\vsu'}{\tsu'}$.
+a trace of the form $\triple{e}{\vsu}{\tsu},\ldots,\triple{e'}{\vsu'}{\tsu'}$.
 %
@@
 then for every possible input type $t$, we can find a value
-$\hastype{v}{t}$ such that $\eapp{f}{v}$ goes wrong.
+$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.
 
@@
     \typeof{\vhole{t}} & \defeq & t \\
+    \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
+    \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
     \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
@@
 \paragraph{Type Compatibility}
-Two types are compatible if one of them can be obtained by
-suitably filling the holes of the other (\ie by suitably ``instantiating''
-the type variables appearing in the other). Formally, a type $s$ is
+Two types are compatible if there exists a type subsitution that maps both
+types to the same type.
+% one of them can be obtained by
+% suitably filling the holes of the other (\ie by suitably ``instantiating''
+% the type variables appearing in the other).
+%
+Formally, a type $s$ is
 \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
@@
 \paragraph{Preservation}
-Given these two definitions, we show that each evaluation step
+Given these two definitions, we show that each successful evaluation step
 refines the partial type of $f$, \ie preserves type compatibility.
@@
 a general witness demonstrating that the program is ill-typed (\ie\ goes
-wrong for all inputs.)
+wrong for at least one input of every type.)
 In particular, we have shown how to non-deterministically find a witnesses

### 2016-03-01 paper/icfp16-submission/semantics.tex
proofs

@@
 inferring a \emph{partial type} --- a type that may contain holes ---
-for $f$.
+for $f$'s input.
 %
@@
 $\steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$,
-then the \emph{partial type} of $f$ upto $\trace$, written \ptype{\trace}{f},
-is $\typeof{\subst{\su}{\ehole{}}} \rightarrow \typeof{e}$.
+then the \emph{partial input type} of $f$ upto $\trace$, written
+\ptype{\trace}{f}, is
+$\typeof{\subst{\su}{\ehole{}}}$.
 %
@@
 \paragraph{Preservation}
-Given these two definitions, we show that each successful evaluation step
-refines the partial type of $f$, \ie preserves type compatibility.
+Given these two definitions, we show that each evaluation step
+refines the partial input type of $f$, \ie preserves type compatibility.
 %
@@
 \begin{proof}
-  By induction on the length of $\trace$ and case analysis on the evaluation rules.
+  By case analysis on the evaluation rules.
   %
@@
 %
-Furthermore, only a call to \forcesym can change the partial type of $f$.
+Furthermore, only a successful call to \forcesym can change the partial type of $f$.
 %
@@
   An examination of the rules shows that only place this happens is
-  in the first case of \forcesym.
+  in the second case of \forcesym.
 \end{proof}
@@
 the partial type upto trace $\trace$ will cause $f$ to get stuck
-in \emph{at most} $k$ steps where $k$ is the length of $\trace$.
+in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
 %
@@
   By induction on $k$, the length of $\trace$.
-  Let $s_{\trace} \rightarrow \_ = \ptype{\trace}{f}$ and
+  Let $s_{\trace} = \ptype{\trace}{f}$ and
   $\trace' = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
@@
 %
-  %\begin{description}
-  % \item
-  [Case \tincompat{t}{s_{\trace'}}:]
+  \begin{description}
+  \item [Case \tincompat{t}{s_{\trace'}}:]
     The inductive hypothesis applies.
-  %\item
-  [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
+  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
     By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
@@
     succeeded.
-  %\end{description}
+  \end{description}
 \end{proof}
@@
 %
-% \begin{description}
-  %\item
-  [Case \tcompat{s}{t}:]
-   The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
-   gets stuck.
-  %\item
-  [Case \tincompat{s}{t}:]
-   By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
+\begin{description}
+\item [Case \tcompat{s}{t}:]
+  Let $\trace = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{\stuck}{\vsu}{\tsu}$.
+  %
+  The value $v = \subst{\vsu}{\ehole}$ demonstrates that
   $\eapp{f}{v}$ gets stuck.
-%\end{description}
+\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
+  such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
+\end{description}
 \end{proof}

### 2016-03-01 paper/icfp16-submission/semantics.tex
much cleaning and simplification
- simplify unification calls
- use \alpha instead of \boxdot for type holes
- pass \tsu to \gen so we don't have to apply it recursively in the call
- write `leaf[t]` instead of `leaf @t`, looks better in tex
- add a bunch of ES notes
- extend proof appendix

@@
 As we are specifically interested in programs that \emph{do} go wrong,
-we include an explicit \stuck\ configuration in our syntax.
+we include an explicit \stuck\ state in our syntax.
 
@@
 We solve this problem be equipping our term language with a notion
-of a \emph{hole}, written \ehole, which represents \emph{unconstrained}
-values that may be replaced with values from \emph{any} type.
+of a \emph{hole}, written \vhole{\thole}, which represents an \emph{unconstrained}
+value that may be replaced with any value of an unspecified type \thole.
 %
@@
 \begin{cases}
-  \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
-  \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
+  \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
+  \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\
+
+  % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
+  % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
 \end{cases} \\
@@
 \begin{cases}
-  \triple{v}{\vsu}{\tsu''}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
-                                         \tsu' = \unify{t_1}{t_2}{\tsu},
-                                         \tsu'' = \unify{\subst{\tsu'}{t_1}}
-                                                        {\typeof{v}}{\tsu'} \\
+  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
+                                         \tsu' = \unify{\{t_1, t_2, \typeof{v}\}}{\tsu}\\
   \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
-  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu}, v = \gen{\subst{\tsu'}{t_1}}\\
+  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu}, v = \gen{t_1}{\tsu'}\\
 \end{cases} \\
@@
 \force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
-\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
-\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
+\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
+\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
 \force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
@@
 a lambda with a \emph{new} hole denoting the return value, and unconstrained
-types (denoted by $\thole$) yield fresh, unconstrained values
-(denoted by $\ehole$).
+types (denoted by $\thole$) yield fresh holes constrained to have type \thole
+(denoted by $\vhole{\thole}$). \ES{note that $\gen{\ttree{t}}{\tsu}$ will
+always place holes in the generated nodes, they will be filled in later,
+on demand.}
 
@@
 \begin{array}{lcll}
-\gensym       & ::      & t \rightarrow v \\
-\gen{\tint}   & \defeq  & n, & \quad \text{non-deterministic} \\
-\gen{\tbool}  & \defeq  & b, & \quad \text{non-deterministic} \\
-\gen{\ttree{t}}  & \defeq  & tr, & \quad \text{non-deterministic, well-typed} \\
-\gen{\tfun}   & \defeq & \efun{x}{\ehole}, & \quad \text{\ehole is fresh} \\
-\gen{\thole}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
+\gensym       & ::  & (t, \tsu) \rightarrow v \\
+\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
+\gen{\tint}{\tsu}   & \defeq  & n, & \quad \text{non-deterministic} \\
+\gen{\tbool}{\tsu}  & \defeq  & b, & \quad \text{non-deterministic} \\
+\gen{\ttree{t}}{\tsu}  & \defeq  & tr, & \quad \text{non-deterministic} \\
+\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, & \quad \text{\ehole and \thole are fresh} \\
+\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
 \end{array}
@@
 %
-A \emph{configuration} is a triple $\triple{e \cup \stuck}{\vsu}{\tsu}$
-of an expression $e$ or the stuck term $\stuck$, a value substitution
-$\vsu$, and a type substitution $\tsu$.
+A program state is a triple $\triple{e \cup \stuck}{\vsu}{\tsu}$ of an
+expression $e$ or the stuck term $\stuck$, a value substitution $\vsu$,
+and a type substitution $\tsu$. In the sequel we will write \estuck to
+denote either an expression $e$ or \stuck.
 %
-We write $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ if the configuration
-$\triple{e}{\vsu}{\tsu}$ transitions in a \emph{small step} to
-$\triple{e'}{\vsu'}{\tsu'}$.
+We write $\step{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ if the state
+$\triple{\estuck}{\vsu}{\tsu}$ transitions in a \emph{single step} to
+$\triple{\estuck'}{\vsu'}{\tsu'}$.
 %
 A (finite) \emph{trace} $\trace$ is a sequence of configurations
-$\triple{e_0}{\vsu_0}{\tsu_0}, \ldots, \triple{e_n}{\vsu_n}{\tsu_n}$ such that
+$\triple{\estuck_0}{\vsu_0}{\tsu_0}, \ldots, \triple{\estuck_n}{\vsu_n}{\tsu_n}$ such that
 $\forall 0 \leq i < n$, we have
-$\step{e_i}{\vsu_i}{\tsu_i}{e_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
+$\step{\estuck_i}{\vsu_i}{\tsu_i}{\estuck_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
 %
-We write \steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if $\trace$ is
-a trace of the form $\triple{e}{\vsu}{\tsu},\ldots,\triple{e'}{\vsu'}{\tsu'}$.
+We write \steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if $\trace$ is
+a trace of the form $\triple{\estuck}{\vsu}{\tsu},\ldots,\triple{\estuck'}{\vsu'}{\tsu'}$.
 %
-We write \steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if
-\steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} for some trace $\trace$.
+We write \steps{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if
+\steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} for some trace $\trace$.
 
@@
 \emph{general} manner. That is, given a function $f$, if we have
-$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
+$\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
 then for every possible input type $t$, we can find a value
@@
   For any function $f$, if\\
-  \hbox{$\steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
+  \hbox{$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
   then for every % inhabitable
@@
     \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
-    \typeof{\vhole{t}} & \defeq & t \\
+    \typeof{\vhole{\thole}} & \defeq & \thole \\
     \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
@@
 \paragraph{Dynamic Type Inference}
-We can think of the evaluation of \eapp{f}{\ehole} as (dynamically)
+We can think of the evaluation of \eapp{f}{\vhole{\thole}} as (dynamically)
 inferring a \emph{partial type} --- a type that may contain holes ---
-for $f$'s input.
+for $f$'s input, \ie a partial instantiation of \thole.
 %
-We can extract this type from evaluation traces.
+We can extract this type from an evaluation trace.
 %
 Formally, we say that if
-$\steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$,
+$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
 then the \emph{partial input type} of $f$ upto $\trace$, written
 \ptype{\trace}{f}, is
-$\typeof{\subst{\su}{\ehole{}}}$.
+$\typeof{\subst{\vsu}{\ehole}}$.
+\ES{should we say $\subst{\tsu}{\thole}$ instead? would need a helper function that does the repeated application of \tsu\ until the result has no $\thole \in dom(\tsu)$}
 %
@@
 \label{lem:refine-partial}
-If $\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots$ and
-$\trace' \defeq \trace, \triple{e'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
+If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
+$\trace' \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
 extension of $\trace$)
@@
   Note that all rules preserve partial types with the exception of when
-  \forcesym\ is called on a hole, in which case we instantiate the hole with
+  \forcesym\ is called on a hole, in which case we may instantiate the hole with
   a concrete value.
   %
-  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
+  But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
 \end{proof}
@@
 If
-$\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
+$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
 and
-$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
+$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$
 and
 $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
-then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ invokes \forcesym.
+then the final step $\step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ invokes \forcesym.
 \end{lem}
@@
 \label{lem:k-stuck}
-  If \steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
+  If \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
      \hastype{v}{t} and
@@
   Let $s_{\trace} = \ptype{\trace}{f}$ and
-  $\trace' = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
+  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
   such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
@@
 \item [Case \tcompat{s}{t}:]
-  Let $\trace = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{\stuck}{\vsu}{\tsu}$.
+  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},...,\triple{\stuck}{\vsu}{\tsu}$.
   %
   The value $v = \subst{\vsu}{\ehole}$ demonstrates that
-  $\eapp{f}{v}$ gets stuck.
+  $\eapp{f}{v}$ gets stuck. \ES{here we also need a repeated application of the substitution, as \ehole may map directly to $\vnode{\thole}{\vhole{\thole}}{\vleaf{\thole}}{\vleaf{\thole}}$}
 \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
@@
   loop vs  = case eval (close bs vs) of
-               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\ehole{}$:vs)
+               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\vhole{\thole}$ : vs)
                _        -> vs
@@
   If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
-  then for every type $t$ there exists $\hastype{v}{t}$ such that
+  then for every type $t$ there exists $v$ of type $t$ such that
   $\steps{\eapp{f_n}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.

### 2016-03-01 paper/icfp16-submission/semantics.tex
more simplificatino

@@
 \forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
-\force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
-\begin{cases}
-  \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
-  \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\
+% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
+% \begin{cases}
+%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
+%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\
 
-  % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
-  % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
-\end{cases} \\
-\force{\vhole{t_1}}{t_2}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
+%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
+%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
+% \end{cases} \\
+\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
 \begin{cases}
   \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
-                                         \tsu' = \unify{\{t_1, t_2, \typeof{v}\}}{\tsu}\\
+                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
   \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
-  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu}, v = \gen{t_1}{\tsu'}\\
+  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t_1}{\tsu'}\\
 \end{cases} \\
@@
 lambda or a tree, it may not map \emph{directly} to another hole, so in
-the second case of \forcesym\ we do not need to \forcesym\ the result of the
+the first case of \forcesym\ we do not need to \forcesym\ the result of the
 substitution, the sub-hole will be narrowed when the context demands it.
@@
   % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
-  $t$ there exists $\hastype{v}{t}$ such that
+  $t$ there exists a $v$ of type $t$ such that
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
@@
 
-\paragraph{The Type of an Expression} The (dynamic) type of an
-expression $e$ is defined as a function $\typeof{e}$ defined as:
+\paragraph{The Type of a a Value} The (dynamic) type of a
+value $v$ is defined as a function $\typeof{v}$ defined as:
   \[
@@
     \typeof{\vhole{\thole}} & \defeq & \thole \\
-    \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
-    \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
-    \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
+    % \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
+    % \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
+    % \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
   \end{array}
@@
 %
-Note that we approximate the type of expressions rather than
+Note that we approximate the type of values rather than
 performing full type inference, for example
@@
   The value $v = \subst{\vsu}{\ehole}$ demonstrates that
-  $\eapp{f}{v}$ gets stuck. \ES{here we also need a repeated application of the substitution, as \ehole may map directly to $\vnode{\thole}{\vhole{\thole}}{\vleaf{\thole}}{\vleaf{\thole}}$}
+  $\eapp{f}{v}$ gets stuck.
+\ES{here we also need a repeated application of the substitution, as \ehole may map directly to $\vnode{\thole}{\vhole{\thole}_1}{\vleaf{\thole}}{\vleaf{\thole}}$ and $\ehole_1$ may later map to \etrue.}
 \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$

### 2016-03-01 paper/icfp16-submission/semantics.tex
swap order of lemmas 2 and 3

@@
 Note that we approximate the type of values rather than
-performing full type inference, for example
+performing traditional type inference, for example
 $\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
@@
 %
-We can extract this type from an evaluation trace.
+We can extract this type from an evaluation trace, by repeatedly
+applying the final type substitution to \thole until it contains no
+holes in the domain of the substitution. We will call this process of
+repeated substition \emph{resolving} a hole, and will use
+\resolve{\thole}{\tsu} to denote the the resolution of \thole with
+respect to \tsu.
 %
@@
 \ptype{\trace}{f}, is
-$\typeof{\subst{\vsu}{\ehole}}$.
-\ES{should we say $\subst{\tsu}{\thole}$ instead? would need a helper function that does the repeated application of \tsu\ until the result has no $\thole \in dom(\tsu)$}
+$\resolve{\thole}{\tsu}$.
+%$\typeof{\subst{\vsu}{\ehole}}$.
+% \ES{should we say $\subst{\tsu}{\thole}$ instead? would need a helper function that does the repeated application of \tsu\ until the result has no $\thole \in dom(\tsu)$}
 %
@@
 %(or stuck.)
+
+% \begin{lem}
+% \label{lem:narrow-tsu}
+% If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
+% and $\trace' \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
+%     (\ie $\trace'$ is a single-step extension of $\trace$)
+% and $\tsu \neq \tsu'$
+% then the final step must have \emph{successfully} invoked \forcesym.
+% \end{lem}
+
+\paragraph{Narrowing}
+%
+Only a successful call to \forcesym can change the partial
+input type of $f$.
+%
+\begin{lem}
+\label{lem:force-inst}
+If
+$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
+and
+$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$
+(\ie $\trace'$ is a single-step extension of $\trace$)
+and
+$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
+then the final step $\step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ invokes \forcesym.
+\end{lem}
+
+\begin{proof}
+  By case analysis on the evaluation rules.
+  %
+  If $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$ then,
+  % one of the holes in $f$'s
+  % argument must have been instantiated with a concrete value at the last step.
+  by the definition of $\ptype{\trace}{f}$, $\tsu \neq \tsu'$, as \thole
+  does not change.
+  %
+  % An examination of the rules shows that only place this happens is
+  % in the second case of \forcesym.
+  An examination of the rules shows that only \forcesym can update \tsu,
+  and furthermore that only the successful cases of \forcesym do update
+  \tsu.
+\end{proof}
 
@@
 If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
-$\trace' \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
-extension of $\trace$)
+$\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
 %
@@
   %
-  Note that all rules preserve partial types with the exception of when
-  \forcesym\ is called on a hole, in which case we may instantiate the hole with
-  a concrete value.
-  %
-  But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
-\end{proof}
-
-\paragraph{Narrowing}
-%
-Furthermore, only a successful call to \forcesym can change the partial type of $f$.
-%
-\begin{lem}
-\label{lem:force-inst}
-If
-$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
-and
-$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$
-and
-$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
-then the final step $\step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ invokes \forcesym.
-\end{lem}
-
-\begin{proof}
-  By case analysis on the evaluation rules.
-  %
-  If $\ptype{\trace}{f} \neq \ptype{\trace}{f}$ then one of the holes in $f$'s
-  argument must have been instantiated with a concrete value at the last step.
-  %
-  An examination of the rules shows that only place this happens is
-  in the second case of \forcesym.
+  First note that by Lemma~\ref{lem:force-inst} we can immediately
+  discharge the \rulename{E-*-Bad} rules as they cannot change
+  \ptype{\trace}{f} at all, and are thus trivially
+  compatibility-preserving. For the \rulename{E-*-Good} rules we can
+  show that, by virtue of \forcesym succeeding, all must preseve
+  compatibility.
+  % Note that all rules preserve partial types with the exception of when
+  % \forcesym\ is called on a hole, in which case we may instantiate the hole with
+  % a concrete value.
+  % %
+  % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
 \end{proof}
@@
 \emph{Any} value that is \emph{incompatible} with
-the partial type upto trace $\trace$ will cause $f$ to get stuck
+the partial input type upto trace $\trace$ will cause $f$ to get stuck
 in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
@@
   then
-     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}.
+     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
+     in at most $k$ steps, where $k$ is the length of $\trace$.
 \end{lem}
 \begin{proof}
-  By induction on $k$, the length of $\trace$.
-  Let $s_{\trace} = \ptype{\trace}{f}$ and
-  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
+  By induction on $k$, the length of $\trace$. Let $s_{\trace} = \ptype{\trace}{f}$ and
+  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
   such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
   Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
-  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}
-  in at most $k$ steps:
+  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
+  in at most $k$ steps.
 %
@@
   \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
-    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
-    $k$.
+    Since $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$ we know
+    that $s_{\trace'} \neq s_{\trace}$, and
+    by Lemma~\ref{lem:force-inst} we know that we must have
+    successfully called \forcesym at step $k$.
     %
@@
     %
-    A case analysis of the rules shows that \forcesym cannot have
-    succeeded.
+    A case analysis of the evaluation rules shows that narrowing to
+    $t$ instead cannot succeed.
   \end{description}
@@
 \item [Case \tcompat{s}{t}:]
-  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},...,\triple{\stuck}{\vsu}{\tsu}$.
+  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{\stuck}{\vsu}{\tsu}$.
   %
-  The value $v = \subst{\vsu}{\ehole}$ demonstrates that
+  The value $v = \resolve{\ehole}{\vsu}$ demonstrates that
   $\eapp{f}{v}$ gets stuck.
-\ES{here we also need a repeated application of the substitution, as \ehole may map directly to $\vnode{\thole}{\vhole{\thole}_1}{\vleaf{\thole}}{\vleaf{\thole}}$ and $\ehole_1$ may later map to \etrue.}
 \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
   such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
+  % \ES{do we need to say anythign else?}
 \end{description}

### 2016-03-03 paper/icfp16-submission/semantics.tex
add base case
simplify 3.4 to just take a target expression
instead of top-level binder nonsense

@@
 returns the existing instantiation.
-\ES{explain the subtlety of narrowing a value hole to an uninstantiated type}
 %
@@
 
-\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
+\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}{\tsu}$
 takes as input a type $t$ and returns a value of that type. For base
@@
   in at most $k$ steps.
+
+  The base case where $k = 0$ (\ie empty traces) is trivial as no
+  evaluation has happened and thus there can be no values that are
+  incompatible with \ptype{0}{f}. So we proceed directly to the
+  inductive case and split cases on whether $t$ is compatible with
+  $s_{\trace'}$.
   %
@@
 %
-First, a search bound @n@ which is used to define the \emph{number} of
+First, a search bound $k$ which is used to define the \emph{number} of
 traces to explore. (We assume, without loss of generality, that all
@@
 %
-Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
-denoting the list of \emph{top-level definitions}:
-%
-\begin{code}
-  let f1 = e1 in
-  let f2 = e2 in
-  ...
-  let fn = en in
-\end{code}
+% Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
+% denoting the list of \emph{top-level definitions}:
+% %
+% \begin{code}
+%   let f1 = e1 in
+%   let f2 = e2 in
+%   ...
+%   let fn = en in
+% \end{code}
+% %
+% The witness generation procedure returns as output a list of (general)
+% witness expressions $e$ each of which is of the form @fn v1 ... vn@.
+Second, the target expression $e$ that contains the type error, and may be a curried function of multiple arguments.
 %
 The witness generation procedure returns as output a list of (general)
-witness expressions $e$ each of which is of the form @fn v1 ... vn@.
+witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
 %
 The \emph{empty} list is returned when no witness can be found after
-exploring $n$ traces.
+exploring $k$ traces.
 %
-In what follows, we refer to @fn@ as the \emph{target} of witness generation.
+% In what follows, we refer to @fn@ as the \emph{target} of witness generation.
 
@@
 %
-\ES{HERE}
 \begin{mcode}
-  eval :: ($e$, $\vsu$, $\tsu$) -> [($v$, $\vsu$, $\tsu$)]
+  eval :: $\triple{e}{\vsu}{\tsu}$ -> [$\triple{v \cup \stuck}{\vsu}{\tsu}$]
 \end{mcode}
@@
 Due to the non-determinism introduced by \gensym, a call
-\hbox{|eval($e$, $\vsu$, $\tsu$)|} returns the \emph{list} of possible results
-of the form $\triple{e'}{\vsu'}{\tsu'}$ such that
-$\steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$.
+\hbox{|eval$\triple{e}{\vsu}{\tsu}$|} returns the \emph{list} of possible results
+of the form $\triple{v \cup \stuck}{\vsu'}{\tsu'}$ such that
+$\steps{e}{\vsu}{\tsu}{v \cup \stuck}{\vsu'}{\tsu'}$.
 %
-We determine whether a given value is stuck by using the predicate:
-%
-\begin{mcode}
-  isStuck :: $v$ -> Bool
-\end{mcode}
+% We determine whether a given value is stuck by using the predicate:
+% %
+% \begin{mcode}
+%   isStuck :: $v$ -> Bool
+% \end{mcode}
 
@@
 %
-We address the issue of currying by defining a procedure @genArgs@ such that \\
-\hbox{@genArgs [(f1,e1),...,(fn,en)]@} yields a list of holes @[v1,...,vk]@
-such that \hbox{@fn v1 ... vk@} \emph{does not} evaluate to a lambda.
+We address the issue of currying by defining a procedure @genArgs@ such that
+\hbox{|genArgs $e$|} yields a list of holes |[$v_1$,...,$v_n$]|
+such that $\eapp{e}{v_1 \ldots v_n}$ \emph{does not} evaluate to a lambda.
 %
 This is achieved via a simple @loop@ that keeps adding holes to the
-target application until @eval@uating the closed term yields a
+target application until @eval@uating the term yields a
 non-lambda value.
 %
-The helper functions @witness@ and @close@ are used to respectively
-apply the parameters to the target and close the result under the top-level
-let-binders, prior to invoking \hbox{@eval@.}
+% The helper functions @witness@ and @close@ are used to respectively
+% apply the parameters to the target and close the result under the top-level
+% let-binders, prior to invoking \hbox{@eval@.}
 %
-That is, @mkApps@ creates a nested sequence of applications in
-the usual left-associative style, and @mkLets@ takes a list of
-binders and a body expression, and creates a sequence of nested
-let-binders that close the body expression.
+% That is, @mkApps@ creates a nested sequence of applications in
+% the usual left-associative style, and @mkLets@ takes a list of
+% binders and a body expression, and creates a sequence of nested
+% let-binders that close the body expression.
 
 \begin{figure}[t]
-\centering
+% $$
+% \begin{array}{lcl}
+% \genArgs{e} &\defeq& \begin{cases} \genArgs{\addArg{e}},& \mbox{if } \triple{\efun{x}{e'}}{\vsu}{\tsu},\ldots = \doeval{e} \\
+%  e,& \mbox{otherwise} \\
+% \end{cases}
+% \\
+% \addArg{e} &\defeq& \eapp{e}{\vhole{\thole}}, \qquad \mbox{where } \ehole, \thole \mbox{ are fresh }
+% \end{array}
+% $$
+
 \begin{mcode}
-genArgs :: [($x$, $e$)] -> [$v$]
-genArgs bs = loop []
+genArgs :: $e$ -> [$v$]
+genArgs $e$ = loop []
   where
-  loop vs  = case eval (close bs vs) of
-               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\vhole{\thole}$ : vs)
+  loop vs = case eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$ of
+             $\triple{\efun{x}{e}}{\vsu}{\tsu}$:_ -> loop ($\vhole{\thole}$ : vs)
+                           -- $\ehole$, $\thole$ are fresh
              _          -> vs
-
-close :: [($x$, $e$)] -> [$v$] -> ($e$, $\vsu$, $\tsu$)
-close bs vs = (e, $\emptysu$, $\emptysu$)
-  where
-    e       = mkLets bs (witness bs vs)
-
-witness :: [($x$, $e$)] -> [$v$] -> $e$
-witness bs vs = mkApps target vs
-  where
-    target    = snd (last bs)
 \end{mcode}
@@
 implementation of our search for witnesses, in the procedure @genWitness@,
-which takes as input a bound @n@ and a list of top level binders, and
-returns a list of witness expressions $e$ that demonstrate how the input
+which takes as input a bound @k@ and the target expression $e$, and
+returns a list of witness expressions $\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input
 program gets stuck.
@@
 @(1)@ we invoke @genArgs@ to determine the number of parameters required
-by the (curried) target function.
+by the (curried) target expression.
 %
-@(2)@ we obtain a closed expression $e$ by closing the target with the
-computed holes.
+% @(2)@ we obtain a closed expression $e$ by closing the target with the
+% computed holes.
 %
-@(3)@ we take the first @n@ traces returned by @eval@ on the closed $e$
+@(2)@ we take the first @k@ traces returned by @eval@ on the target $\eapp{e}{vs}$
 and
-@(4)@ we extract the substitutions corresponding to the $\stuck$ traces,
+@(3)@ we extract the substitutions corresponding to the $\stuck$ traces,
 and use them to return the list of witnesses.
@@
 \label{thm:generation}
-  If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
-  then for every type $t$ there exists $v$ of type $t$ such that
-  $\steps{\eapp{f_n}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
+  If |genWitness k $e$ = $\ $ $\triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$|,
+  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
+  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
 \end{cor}
@@
 \centering
+% $$
+% \begin{array}{lcll}
+% \genWitness{n}{e} &\defeq& \{ \resolve{e'}{\vsu} \spmid \triple{e'}{\vsu}{\tsu} \in \doeval{\genArgs{e}} \} & \\
+% \end{array}
+% $$
+
 \begin{mcode}
-genWitness :: Nat -> [($x$, $e$)] -> [$v$]
-genWitness n bs
-       = [witness bs ($\vsu$ vs) | $\vsu$ <- $\vsu$s]
+genWitness :: Nat -> $e$ -> [$e$]
+genWitness n $e$ = [$\resolve{\eapp{e}{vs}}{\vsu}$ | $\vsu$ <- $\vsu$s]
   where
-   vs  = genArgs defs []        -- (1)
-   cfg = close defs vs          -- (2)
-   res = take n (eval cfg)      -- (3)
-   $\vsu$s   = [$\vsu$ | $\triple{\stuck}{\vsu}{\tsu}$ <- res]  -- (4)
+   vs  = genArgs $e$                  -- (1)
+   res = take n (eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$) $\ $  -- (2)
+   $\vsu$s   = [$\vsu$ | $\triple{\stuck}{\vsu}{\tsu}$ <- res] $\ \ $   -- (3)
 \end{mcode}

### 2016-03-04 paper/icfp16-submission/interactive.tex
hide interactive semantics and exploration figures

--- 2015-11-20 paper/pldi16-submission/interactive.tex
@@
 %
-\begin{figure*}[t]
-%% \relDescription{Computing Sub-Terms}
-%% \begin{gather*}
-%% \begin{array}{lcl}
-%% \subtermssym                 & \dcolon & e \to \tr \\
-%% \subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
-%% \subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
-%% \subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
-                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
-                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
-%% \subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
-                                %% &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
-%% \subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
-%% \subterms{e}                 & \defeq & \bullet
-%% \end{array}
-%% \end{gather*}
-\judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
-\begin{gather*}
-\inference[\recontext]
-  {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
-  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \tr_1}}
-\\ \\
-\inference[\reappgood]
-  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
-  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
-         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \tr}}
-\\ \\
-\inference[\reappbad]
-  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
-  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
-\end{gather*}
-\caption{A selection of the operational semantics from
-  Figure~\ref{fig:operational}, extended to collect a full reduction
-  graph.}
-\label{fig:interactive}
-\end{figure*}
+% \begin{figure*}[t]
+% %% \relDescription{Computing Sub-Terms}
+% %% \begin{gather*}
+% %% \begin{array}{lcl}
+% %% \subtermssym                 & \dcolon & e \to \tr \\
+% %% \subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
+% %% \subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
+% %% \subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
+%                                 %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
+%                                 %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
+% %% \subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
+%                                 %% &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
+% %% \subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
+% %% \subterms{e}                 & \defeq & \bullet
+% %% \end{array}
+% %% \end{gather*}
+% \judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
+% \begin{gather*}
+% \inference[\recontext]
+%   {\stepg{e}{\su}{\tr}{e'}{\su'}{\tr'}}
+%   {\stepg{C[e]}{\su}{\tr}{C[e']}{\su'}{\singlestep{C[e]}{C[e']}; \tr'}}
+% \\ \\
+% \inference[\reappgood]
+%   {\pair{\efun{x}{e}}{\su'} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su}}
+%   {\stepg{\eapp{v_1}{v_2}}{\su}{\tr}
+%          {e\sub{x}{v_2}}{\su'}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \tr}}
+% \\ \\
+% \inference[\reappbad]
+%   {\pair{\stuck}{\su'} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su}}
+%   {\stepg{\eapp{v_1}{v_2}}{\su}{\tr}{\stuck}{\su'}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
+% \end{gather*}
+% \caption{A selection of the operational semantics from
+%   Figure~\ref{fig:operational}, extended to collect a full reduction
+%   graph.}
+% \label{fig:interactive}
+% \end{figure*}
 
@@
 Concretely, we extend the operational semantics to
-a relation of the form $\stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\tr'}$
+a relation of the form $\stepg{e}{\vsu}{\tsu}{\tr}{e'}{\vsu'}{\tsu'}{\tr'}$
 where $\tr'$ collects the edges of the transition.
@@
 to recording the consequent of each original rule in the
-trace. That is, each original judgment \step{e}{\vsu}{e'}{\vsu'}
-becomes \stepg{e}{\vsu}{\tr}{e'}{\vsu'}{\singlestep{e}{e'}; \tr}.
+trace. That is, each original judgment \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}
+becomes \stepg{e}{\vsu}{\tsu}{\tr}{e'}{\vsu'}{\tsu'}{\singlestep{e}{e'}; \tr}.
 %
-As the translation is mechanical, we provide a selection of examples
-in Figure~\ref{fig:interactive}.
+% As the translation is mechanical, we provide a selection of examples
+% in Figure~\ref{fig:interactive}.
+As the translation is mechanical, we will not discuss it further.
 
@@
 exposition, we assume we have a (global) trace:
-$\stepg{e_0}{\emptysu}{\bullet}{e_n}{\vsu}{\tr}$, where
+$\stepg{e_0}{\emptysu}{\emptysu}{\bullet}{e_n}{\vsu}{\tsu}{\tr}$, where
 $e_0$ and $e_n$ are the \emph{initial} and \emph{final}
@@
 
-\input{traversals_new.tex}
+% \input{traversals_new.tex}
 

### 2016-03-04 paper/icfp16-submission/gallery.tex
start gallery.tex

\section{The Advantage Of Traces}
\label{sec:advantage-traces}
In this section we present a qualitative evaluation of the witnesses and
traces that \toolname produces.

\begin{figure*}[ht]
\centering
\begin{minipage}{0.49\linewidth}
\centering
\begin{code}
let rec sqsum xs = match xs with
  | [] -> 0
  | h::t -> (sqsum t) @ (h * h)
\end{code}
\begin{verbatim}
File "prog3151.ml", line 7, characters 12-21:
Error: This expression has type
         int
       but an expression was expected of type
         'a list
\end{verbatim}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\centering
\includegraphics[height=200px]{sqsum.png}
\end{minipage}
\caption{caption}
\label{fig:traces}
\end{figure*}

\begin{figure*}[ht]
\centering
\begin{minipage}{0.49\linewidth}
\centering
\begin{code}
let rec sumList xs = match xs with
  | []    -> []
  | y::ys -> y + sumList ys
\end{code}
\begin{verbatim}
File "sumlist.ml", line 5, characters 17-27:
Error: This expression has type
         'a list
       but an expression was expected of type
         int
\end{verbatim}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\centering
\includegraphics[height=200px]{sumlist.png}
\end{minipage}
\caption{caption}
\label{fig:traces}
\end{figure*}

\begin{figure*}[ht]
\centering
\begin{minipage}{0.49\linewidth}
\centering
\begin{code}
let append xs ys = match xs with
  | []   -> ys
  | h::t -> h :: t :: ys
\end{code}
\begin{verbatim}
File "prog0164.ml", line 7, characters 17-18:
Error: This expression has type
         'a list
       but an expression was expected of type
         'a
       The type variable 'a occurs inside 'a list
\end{verbatim}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\centering
\includegraphics[height=100px]{append.png}
\end{minipage}
\caption{caption}
\label{fig:traces}
\end{figure*}

\begin{figure*}[ht]
\centering
\begin{minipage}{0.49\linewidth}
\centering
\begin{code}
let rec wwhile (f,b) =
  match f with
  | (z, false) -> z
  | (z, true)  -> wwhile (f, z)

let f x =
  let xx = x * x in
  (xx, (xx < 100))

let _ = wwhile (f, 2)
\end{code}
\begin{verbatim}
File "prog0358.ml", line 19, characters 16-17:
Error: This expression has type
         int -> int * bool
       but an expression was expected of type
         'a * bool
\end{verbatim}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\centering
\includegraphics[height=150px]{wwhile.png}
\end{minipage}
\caption{caption}
\label{fig:traces}
\end{figure*}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2016-03-04 paper/icfp16-submission/gallery.tex
clean up gallery figures

@@
 \centering
-\begin{code}
+\begin{ecode}
 let rec sqsum xs = match xs with
   | [] -> 0
-  | h::t -> (sqsum t) @ (h * h)
-\end{code}
+  | h::t -> __(sqsum t)__ @ (h * h)
+\end{ecode}
+% File "sqsum.ml", line 3, characters 12-21:
 \begin{verbatim}
-File "prog3151.ml", line 7, characters 12-21:
 Error: This expression has type
@@
 \centering
-\includegraphics[height=200px]{sqsum.png}
+\includegraphics[height=125px]{sqsum.png}
 \end{minipage}
@@
 \centering
-\begin{code}
+\begin{ecode}
 let rec sumList xs = match xs with
   | []    -> []
-  | y::ys -> y + sumList ys
-\end{code}
+  | y::ys -> y + __sumList ys__
+\end{ecode}
+% File "sumList.ml", line 3, characters 17-27:
 \begin{verbatim}
-File "sumlist.ml", line 5, characters 17-27:
 Error: This expression has type
@@
 \centering
-\includegraphics[height=200px]{sumlist.png}
+\includegraphics[height=125px]{sumlist.png}
 \end{minipage}
@@
 \centering
-\begin{code}
+\begin{ecode}
 let append xs ys = match xs with
   | []   -> ys
-  | h::t -> h :: t :: ys
-\end{code}
+  | h::t -> h :: __t__ :: ys
+\end{ecode}
+% File "append.ml", line 3, characters 17-18:
 \begin{verbatim}
-File "prog0164.ml", line 7, characters 17-18:
 Error: This expression has type
@@
 \centering
-\includegraphics[height=100px]{append.png}
+\includegraphics[height=75px]{append.png}
 \end{minipage}
@@
 \centering
-\begin{code}
+\begin{ecode}
 let rec wwhile (f,b) =
@@
 
-let _ = wwhile (f, 2)
-\end{code}
+let _ = wwhile (__f__, 2)
+\end{ecode}
+% File "wwhile.ml", line 10, characters 16-17:
 \begin{verbatim}
-File "prog0358.ml", line 19, characters 16-17:
 Error: This expression has type

### 2016-03-04 paper/icfp16-submission/plots.py

@@ -67,7 +67,7 @@ def plot_coverage(seminal, ucsd):
     autolabel(plt, p2)
 
     # plt.show()
-    fig.savefig('coverage.pdf')
+    fig.savefig('coverage.png')
     plt.close()
 
 def plot_trace_size(seminal, ucsd):
@@ -117,7 +117,7 @@ def plot_trace_size(seminal, ucsd):
 
     # autolabel(ax, p1)
 
-    plt.savefig('trace_size_step.pdf')
+    plt.savefig('trace_size_step.png')
     plt.close()
 
 
@@ -161,7 +161,7 @@ def plot_trace_size(seminal, ucsd):
     # autolabel(ax, p2)
 
     # plt.show()
-    plt.savefig('trace_size_jump.pdf')
+    plt.savefig('trace_size_jump.png')
     plt.close()
 
 def plot_distrib(seminal, ucsd):
@@ -215,7 +215,7 @@ def plot_distrib(seminal, ucsd):
 
 
     # plt.show()
-    plt.savefig('distrib.pdf')
+    plt.savefig('distrib.png')
     plt.close()
 
 

### 2016-03-05 paper/icfp16-submission/gallery.tex
captions!

@@
 \label{sec:advantage-traces}
-In this section we present a qualitative evaluation of the witnesses and
-traces that \toolname produces.
+Finally, having shown that \toolname produces small traces in an
+efficient manner, let us discuss the explanatory power of traces
+compared with typical type error messages. In this section we will
+present a qualitative evaluation of \toolname's jump-compressed traces
+on a series of examples drawn from real student programs in the
+\ucsdbench dataset. For each example we will present the code, a
+jump-compressed trace produced by \toolname, and the type error produced
+by the \ocaml compiler.
 
@@
 \end{minipage}
-\caption{caption}
+\caption{\texttt{sqsum xs} should square each element of \texttt{xs} and
+  then compute the sum of the result, but the student has used the
+  list-append operator |@| instead of \texttt{+} to compute the
+  sum. \ocaml blames the recursive call \texttt{sqsum t} (underlined
+  above) for being an \texttt{int} instead of a
+  \texttt{list}. \toolname's jump-compressed trace draws the eye to the
+  entire term \texttt{0 }|@|\texttt{ 1}.}
 \label{fig:traces}
@@
 \end{minipage}
-\caption{caption}
+\caption{\texttt{sumList xs} should sum the elements of \texttt{xs}, but
+  the student has returned \texttt{[]} in the base case instead of
+  \texttt{0}. \ocaml deduces from the base case that \texttt{sumList}
+  must return a \texttt{list} and thus blames the recursive call for
+  producing a \texttt{list} instead of an \texttt{int}. \toolname's
+  jump-compressed trace highlights the clash between the \texttt{[]} and
+  the \texttt{+}, but also demonstrates in the third step that
+  \texttt{sumList []} reduces to \texttt{[]}, which clarifies that the
+  conflict is between the base case and recursive case.}
 \label{fig:traces}
@@
 \end{minipage}
-\caption{caption}
+\caption{\texttt{append xs ys} should concatenate the two input lists,
+  but the student has tried to cons the tail of \texttt{xs} onto
+  \texttt{ys} instead of making a recursive call to \texttt{append}.
+  \ocaml blames \texttt{t} for being a \texttt{'a list} instead of a
+  \texttt{'a}, prompting an \emph{occurs-check} error. \toolname shows,
+  more concretely, that \texttt{[] :: [2]} gets stuck because lists are
+  homogeneous and \texttt{[]} is not an \texttt{int}.}
 \label{fig:traces}
@@
 \end{minipage}
-\caption{caption}
+\caption{\texttt{wwhile (f, b)} should repeatedly apply the function
+  \texttt{f} to the first element of its result pair, starting with the
+  initial input \texttt{b}, until the second element is \texttt{false}.
+  Unfortunately, the student has forgotten to apply \texttt{f} at all
+  and just matches it directly against a pair. This faulty definition of
+  \texttt{wwhile} still typechecks however, and \ocaml thus blames the
+  \emph{call-site} on line 10. \toolname makes no assumptions about the
+  input code and quickly draws the eye to the true error, the
+  \texttt{match} expression, and highlights the conflict in matching a
+  function against a pair pattern.}
 \label{fig:traces}

### 2016-03-05 paper/icfp16-submission/gallery.tex
misc

@@
   sum. \ocaml blames the recursive call \texttt{sqsum t} (underlined
-  above) for being an \texttt{int} instead of a
-  \texttt{list}. \toolname's jump-compressed trace draws the eye to the
+  above) for being an \texttt{int} instead of a \texttt{list}.
+  \toolname's jump-compressed trace draws the eye to the
   entire term \texttt{0 }|@|\texttt{ 1}.}
@@
   \texttt{0}. \ocaml deduces from the base case that \texttt{sumList}
-  must return a \texttt{list} and thus blames the recursive call for
-  producing a \texttt{list} instead of an \texttt{int}. \toolname's
-  jump-compressed trace highlights the clash between the \texttt{[]} and
-  the \texttt{+}, but also demonstrates in the third step that
-  \texttt{sumList []} reduces to \texttt{[]}, which clarifies that the
-  conflict is between the base case and recursive case.}
+  must return a \texttt{list} and thus blames the recursive call on line
+  3 for producing a \texttt{list} instead of an  \texttt{int}.
+  \toolname's jump-compressed trace highlights the clash
+  between the \texttt{[]} and the \texttt{+}, but also demonstrates in
+  the third step that \texttt{sumList []} reduces to \texttt{[]}, which
+  clarifies that the conflict is between the base case and recursive
+  case.}
 \label{fig:traces}
@@
   \texttt{ys} instead of making a recursive call to \texttt{append}.
-  \ocaml blames \texttt{t} for being a \texttt{'a list} instead of a
-  \texttt{'a}, prompting an \emph{occurs-check} error. \toolname shows,
-  more concretely, that \texttt{[] :: [2]} gets stuck because lists are
-  homogeneous and \texttt{[]} is not an \texttt{int}.}
+  \ocaml blames \texttt{t} on line 3 for being a \texttt{'a list}
+  instead of a \texttt{'a}, prompting an \emph{occurs-check}
+  error. \toolname shows, more concretely, that \texttt{[] :: [2]} gets
+  stuck because lists are homogeneous and \texttt{[]} is not an
+  \texttt{int}.}
 \label{fig:traces}
@@
   initial input \texttt{b}, until the second element is \texttt{false}.
-  Unfortunately, the student has forgotten to apply \texttt{f} at all
-  and just matches it directly against a pair. This faulty definition of
-  \texttt{wwhile} still typechecks however, and \ocaml thus blames the
-  \emph{call-site} on line 10. \toolname makes no assumptions about the
-  input code and quickly draws the eye to the true error, the
-  \texttt{match} expression, and highlights the conflict in matching a
-  function against a pair pattern.}
+  Unfortunately, the student has forgotten to apply \texttt{f} at all on
+  line 2, and just matches it directly against a pair. This faulty
+  definition of \texttt{wwhile} still typechecks however, and \ocaml
+  thus blames the \emph{call-site} on line 10. \toolname makes no
+  assumptions about the input code and quickly draws the eye to the true
+  error, the \texttt{match} expression, and highlights the conflict in
+  matching a function against a pair pattern.}
 \label{fig:traces}

### 2016-03-05 paper/icfp16-submission/gallery2.tex
add gallery2, with inline code instead of figures

\section{The Advantage Of Traces}
\label{sec:advantage-traces}
Having shown that \toolname produces small traces in an efficient
manner, let us discuss the explanatory power of traces compared with
typical type error messages. In this section we will present a
qualitative evaluation of \toolname's jump-compressed traces on a series
of examples drawn from real student programs in the \ucsdbench
dataset. We will show how \toolname's traces get to the heart of the
error by
%
\begin{enumerate}
\item highlighting the conflicting values that cause the program to get
  stuck, rather that blaming a single one,
\item showing the steps necessary to reach the stuck state, and
\item not assuming that a function is correct just because it type-checks.
\end{enumerate}
%
For each example we will present the code, the type error produced by
the \ocaml compiler, and a jump-compressed trace produced by \toolname.

% \begin{figure*}[ht]
% \centering
% \begin{minipage}{0.49\linewidth}
% \centering

First, we will examine a simple recursive function @sqsum@ that should
square each element of the input list and then compute the sum of the
result. In the following code snippets we will underline the source span
that \ocaml blames for the type error.
%
\begin{ecode}
let rec sqsum xs = match xs with
  | [] -> 0
  | h::t -> __(sqsum t)__ @ (h * h)
\end{ecode}
%
Unfortunately the student has used the list-append operator |@| instead
of \texttt{+} to compute the sum. \ocaml blames the recursive call
\texttt{sqsum t} on line 3 for being an \texttt{int} instead of a
\texttt{list}.
%
\begin{verbatim}
Error: This expression has type
         int
       but an expression was expected of type
         'a list
\end{verbatim}
%
In contrast, \toolname produces a trace showing how the evaluation of
@sqsum [1]@ eventually gets stuck when it tries to evaluate
\texttt{ 0 }|@|\texttt{ 1}.
%
\begin{center}
  \includegraphics[height=125px]{sqsum.png}
\end{center}
%
Note how the eye is drawn to the entire stuck term as opposed to just
the left sub-term. This emphasizes the \emph{conflict} between @int@ and
@list@ rather than assuming one or the other is correct.

Next we consider a function @sumList@ that just sums the elements of the
input list, but has a more unfortunate error.
\begin{ecode}
let rec sumList xs = match xs with
  | []    -> []
  | y::ys -> y + __sumList ys__
\end{ecode}
%
In this case the student has returned @[]@ in the base case instead of
@0@. Unfortunately, \ocaml deduces (incorrectly) from the base case that
@sumList@ must return a @list@, and then proceeds to blame the
\emph{recursive call} on line 3 for producing a @list@ instead of an
@int@.
%
\begin{verbatim}
Error: This expression has type
         'a list
       but an expression was expected of type
         int
\end{verbatim}
%
\toolname's jump compressed trace shows how the evaluation of
@sumList [1; 2]@ gets stuck when it tries to evaluate @2 + []@.
%
\begin{center}
  \includegraphics[height=125px]{sumlist.png}
\end{center}
%
This is essentially the same error that \ocaml produces, but the trace
clarifies immediately (via the third step) that the @[]@ is the result
of the recursive call @sumList []@, drawing attention to the incorrect
base case.

%% ES: append is actually a bit problematic as we don't find the nice
%% append [1] [2] witness. instead we could find something like
%% append [_] [], but it's not as clear IMO
% Our next example is the @append@ function, which should concatenate the
% two input lists.
% %
% \begin{ecode}
% let append xs ys = match xs with
%   | []   -> ys
%   | h::t -> h :: __t__ :: ys
% \end{ecode}
% %
% The student has forgotten to make a recursive call to @append@, and
% instead tries to cons the tail @t@ directly onto the second list @ys@.
% Consing @h@ back onto the result causes \ocaml to attempt to construct
% the infinite type @'a = 'a list@, triggering an \emph{occurs-check}
% error.
% %
% \begin{verbatim}
% Error: This expression has type
%          'a list
%        but an expression was expected of type
%          'a
%        The type variable 'a occurs inside 'a list
% \end{verbatim}
% %
% %
% \begin{center}
%   \includegraphics[height=75px]{append.png}
% \end{center}

Finally, consider the higher-order function @wwhile@ that emulates a
traditional while-loop. Concretely, @wwhile@ takes a function @f@ and
repeatedly calls @f@ on the first element of its output pair, starting
with the initial value @b@, until the second element is @false@.
%
\begin{ecode}
let rec wwhile (f,b) =
  match f with
  | (z, false) -> z
  | (z, true)  -> wwhile (f, z)

let f x =
  let xx = x * x in
  (xx, (xx < 100))

let _ = wwhile (__f__, 2)
\end{ecode}
%
Unfortunately, the student has forgotten to apply @f@ at all on line 2,
and just matches it directly against a pair. This faulty definition of
@wwhile@ still typechecks however, and \ocaml thus blames the
\emph{call-site} on line 10.
%
\begin{verbatim}
Error: This expression has type
         int -> int * bool
       but an expression was expected of type
         'a * bool
\end{verbatim}
%
\toolname makes no assumptions about the program and quickly draws the
eye to the true error, the @match@ expression on line 2, and highlights
the conflict in matching a function against a pair pattern.
%
\begin{center}
  \includegraphics[height=150px]{wwhile.png}
\end{center}
%

By highlighting conflicting values rather than blaming a single value
and not making any assumptions about the given program, \toolname avoids
misleading the user into focusing their attention on a piece of code
that is actually irrelevant to the error.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2016-03-07 paper/icfp16-submission/evaluation.tex
update numbers

--- 2015-11-20 paper/pldi16-submission/evaluation.tex
@@
 \centering
-\includegraphics[height=1.5in]{fac-minus.png}
+\includegraphics[height=2in]{fac-minus.png}
 \caption{Jump-compressed trace of \texttt{fac 1} with subtraction
@@
 \includegraphics[width=\linewidth]{trace_size_jump.png}
-\caption{Complexity of the generated traces. At least 90\% of the traces
-  have a jump complexity of less than 10, with an average of 3.}
+\caption{Complexity of the generated traces. 81\% of the combined traces
+  have a jump complexity of at most 10, with an average of 7 and a
+  median of 5.}
 \label{fig:results-complexity}
@@
 %
-The average number of single-step reductions per trace is 26 for the \ucsdbench\
-dataset (27 for the \uwbench\ dataset) with a maximum of 2,744 (659 for
-\uwbench). The average number of jumps per trace is 3 (also 3 for
-\uwbench) with a maximium of 177 (52 for \uwbench). In both datasets at
-least 90\% of the traces contain fewer than 10 jumps.
+The average number of single-step reductions per trace is 31 for the
+\ucsdbench\ dataset (35 for the \uwbench\ dataset) with a maximum of
+2,745 (660 for \uwbench) and a median of 16 (also 16 for \uwbench).
+%
+The average number of jumps per trace is 7 (also 7 for \uwbench) with a
+maximium of 353 (105 for \uwbench) and a median of 4 (also 4 for
+\uwbench).
+%
+In both datasets at least 80\% of the traces contain at most 10 jumps.
 %

### 2016-03-07 paper/icfp16-submission/introduction.tex
update numbers

--- 2015-11-20 paper/pldi16-submission/introduction.tex
@@
 Furthermore, we show that a simple interactive strategy yields
-compact counterexample traces with fewer than 5 steps for 60\%
-of the programs, and under 10 steps for 90\% of the programs.
+compact counterexample traces with at most 5 steps for 57\%
+of the programs, and at most 10 steps for 81\% of the programs.
 

### 2016-03-07 paper/icfp16-submission/plots.py

@@ -56,7 +56,7 @@ def plot_coverage(seminal, ucsd):
     p2 = plt.bar(ind + width, [r[1] for r in xy_u], width,
                  color=COLORS[1])
 
-    plt.xlabel('Witness found in <x seconds', fontsize=20)
+    plt.xlabel('Witness found in <= x seconds', fontsize=20)
     plt.ylabel('Witnesses Found (% of total programs)', fontsize=20)
     plt.title('Cumulative Coverage', fontsize=24)
     plt.xticks(ind + width, [r[0] for r in xy_s], fontsize='large')
@@ -69,6 +69,14 @@ def plot_coverage(seminal, ucsd):
     # plt.show()
     fig.savefig('coverage.png')
     plt.close()
+
+BINS = [5, 10, 20, 50, 100, 1000]
+
+def cumulative_trace_size(data):
+    return [len([r for r in data
+                 if r <= l])
+            / float(len(data))
+            for l in BINS]
 
 def plot_trace_size(seminal, ucsd):
     # xy = cumulative_coverage(data)
@@ -86,8 +94,7 @@ def plot_trace_size(seminal, ucsd):
     step_u = [int(r[5]) for r in ucsd[1:] if r[4] in UNSAFE and int(r[5]) > 0]
     jump_u = [int(r[6]) for r in ucsd[1:] if r[4] in UNSAFE and int(r[6]) > 0]
     
-    bins = [0, 5, 10, 20, 50, 100, 1000]
-    binlabels = ['[0,5)', '[5,10)', '[10,20)', '[20,50)', '[50,100)', '>=100']
+    binlabels = ['<= 5', '<= 10', '<= 20', '<= 50', '<= 100', '> 100']
     ind = np.arange(0, len(binlabels))
     width = 0.35
     # plt.figure(figsize=(100,50))
@@ -100,11 +107,18 @@ def plot_trace_size(seminal, ucsd):
     # ax.set(adjustable='box-forced', aspect=4)
 
     fig = plt.figure()
-    y,binEdges=np.histogram(step_s,bins=bins)
-    p1 = plt.bar(ind, y / float(len(step_s)), label='UW', width=width, color=COLORS[0])
+    # y,binEdges=np.histogram(step_s,bins=bins)
+    c_step_s = cumulative_trace_size(step_s)
+    print('step complexity')
[…]

### 2016-03-07 paper/icfp16-submission/interactive.tex
minor changes to overview

@@
 
-Next, we present a novel way to debug executions.
+In this section we present a novel way to debug executions.
 %
@@
 %
-The general recipe for collecting steps-to edges is to
-to recording the consequent of each original rule in the
+The general recipe for collecting steps-to edges is
+to record the consequent of each original rule in the
 trace. That is, each original judgment \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}

### 2016-03-07 paper/icfp16-submission/overview.tex
copy to icfp16-submission, switch to icfp class
much cleaning and simplification
- simplify unification calls
- use \alpha instead of \boxdot for type holes
- pass \tsu to \gen so we don't have to apply it recursively in the call
- write `leaf[t]` instead of `leaf @t`, looks better in tex
- add a bunch of ES notes
- extend proof appendix
minor changes to overview

--- 2015-11-20 paper/pldi16-submission/overview.tex
@@
 %
-A hole --- written \ehole{} --- is a \emph{placeholder} for a value; in
-our case we do not even know its type.
+A hole --- written \vhole{\thole} --- is a \emph{placeholder} for a
+value of some unknown type \thole.
 %
@@
 \begin{mcode}
-  fst : ($\thole{1}$ * $\thole{2}$) -> $\thole{1}$
+  fst : ($\thole_1$ * $\thole_2$) -> $\thole_1$
 \end{mcode}
@@
 The \ocaml\ compiler deduces that @incAllByOne@ takes a single argument
-because @List.map@'s \emph{type} says it takes two arguments, and it is
+because the \emph{type} of \hbox{@List.map@} says it takes two arguments, and it is
 partially applied to @(+ 1)@.
@@
 %
-These graphs have two types of edges:
+% These graphs have two types of edges:
+% %
+% (1) ``steps-to'' edges that capture the small-step transition between
+% two terms, and
+% %
+% (2) ``sub-term'' edges that capture the containment relation between two
+% terms.
 %
-(1) ``steps-to'' edges that capture the small-step transition between
-two terms, and
-%
-(2) ``sub-term'' edges that capture the containment relation between two
-terms.
-
 For example, evaluating the expression @1+2+3@ would produce the
@@
   \centering
-  \includegraphics[width=\linewidth]{simple.pdf}
+  \includegraphics[height=2in]{simple.png}
   \caption{The reduction graph for \texttt{1+2+3}. The two edges

### 2016-03-07 paper/icfp16-submission/gallery2.tex
10pt font for now

@@
 \begin{verbatim}
-Error: This expression has type
+  This expression has type
     int
@@
 \begin{verbatim}
-Error: This expression has type
+  This expression has type
     'a list
@@
 \begin{verbatim}
-Error: This expression has type
+  This expression has type
     int -> int * bool

### 2016-03-07 paper/icfp16-submission/introduction.tex
major cuts to sec:interactive in interactive-short.tex

@@
 \paragraph{2. Visualizing Witnesses}
-Our second contribution is a novel interactive visualization of the
+Our second contribution is an interactive visualization of the
 execution of \ocaml\ programs, well-typed or not~(\S~\ref{sec:interactive}).

### 2016-03-07 paper/icfp16-submission/overview.tex
major cuts to sec:interactive in interactive-short.tex

@@
 computation by selecting a visible term and choosing one of the
-applicable traversal strategies --- formalized in
+applicable traversal strategies --- described in
 \S~\ref{sec:interactive} --- to insert another term into the

### 2016-03-07 paper/icfp16-submission/overview.tex
aasdasd

--- 2016-03-07 paper/icfp16-submission/overview.tex
@@
 %
-A hole --- written \vhole{\thole} --- is a \emph{placeholder} for a
-value of some unknown type \thole.
+A hole --- written \ehole{} --- is a \emph{placeholder} for a value; in
+our case we do not even know its type.
 %
@@
 function to ``go wrong'', then there is
-\emph{no possible} type for the function.
+\emph{no possible} inhabitable type for the function.
 %
@@
 The \ocaml\ compiler deduces that @incAllByOne@ takes a single argument
-because the \emph{type} of \hbox{@List.map@} says it takes two arguments, and it is
+because the \emph{type} of \hbox{@List.map@'s} says it takes two arguments, and it is
 partially applied to @(+ 1)@.
@@
 
+\paragraph{Jump Compressed Witnesses}
+\RJ{add this para here with example}
+
 % A sample interaction with the trace of @fac 1@ can be seen in

### 2016-03-07 paper/icfp16-submission/semantics.tex
aasdasd

@@
 %
-We solve this problem be equipping our term language with a notion
-of a \emph{hole}, written \vhole{\thole}, which represents an \emph{unconstrained}
-value that may be replaced with any value of an unspecified type \thole.
+We solve this problem be equipping our term language with a
+notion of a \emph{hole}, written \vhole{\thole}, which represents
+an \emph{unconstrained} value $\ehole$ that may be replaced with
+\emph{any} value of type \thole.
 %
-Dually, holes may also appear in types, written \thole, where they may
-be thought of as type variables that we will not generalize over.
+Intuitively, the type holes \thole\ can be viewed as type variables
+that we will \emph{not} generalize over.
 %
+A \emph{normalized} value (resp.\ type) is one that is not a hole,
+but which may internally contain holes.
+%
+For example \RJ{add example here}
 
@@
 We track how the holes are incrementally filled in, by using
-(value) \emph{substitutions} that map holes % (denoted by their index)
-to values which may, in turn, contain holes.
+value (resp.\ type) \emph{substitutions} $\vsu$ (resp. $\tsu$)
+that map value (resp.\ type) holes to values (resp.\ types).
 %
-The substitutions let us ensure that we instantiate each hole
-with the same (partially defined) value, regardless of the multiple
-contexts in which the hole appears, and hence, ensures we can
-report a concrete (and general) witness for any (dynamically)
-discovered type errors.
+A \emph{normalized} substitution is one whose co-domain comprises
+of normalized values (resp.\ types).
+%
+In the sequel, we will assume and ensure that all substitutions
+are normalized.
+
+\paragraph{Resolving Holes}
+We \emph{resolve} a hole with respect to a substitution by
+transitively applying the substitution as long as it contains
+any holes that are defined in the substitution.
+%
+We write \resolve{\thole}{\tsu} to denote the the resolution
+of \thole\ with respect to \tsu.
+%
+Note that by definition, \resolve{\thole}{\tsu} does not contain
+any holes in the domain of \tsu.
 
@@
 %
+\RJ{Add intuition text -- "dynamic Hindley Milner"}
+%
+The substitutions let us ensure that we consistently instantiate
+each hole with the same (partially defined) value or type, regardless
+of the multiple contexts in which the hole appears. This ensures we
+can report a concrete (and general) witness for any (dynamically)
+discovered type errors.
 
@@
 %
-While a hole may map to a value that \emph{contains} another hole, \eg a
-lambda or a tree, it may not map \emph{directly} to another hole, so in
-the first case of \forcesym\ we do not need to \forcesym\ the result of the
-substitution, the sub-hole will be narrowed when the context demands it.
+%While a hole may map to a value that \emph{contains} another hole, \eg a
+%lambda or a tree, it may not map \emph{directly} to another hole,
+As the substitutions are normalized, in the first case of \forcesym\ we
+do not need to \forcesym\ the result of the substitution, the sub-hole
+will be narrowed when the context demands it.
 
-\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}{\tsu}$
-takes as input a type $t$ and returns a value of that type. For base
-types, the procedure returns an arbitrary value. For functions, it returns
-a lambda with a \emph{new} hole denoting the return value, and unconstrained
-types (denoted by $\thole$) yield fresh holes constrained to have type \thole
-(denoted by $\vhole{\thole}$). \ES{note that $\gen{\ttree{t}}{\tsu}$ will
-always place holes in the generated nodes, they will be filled in later,
-on demand.}
+\paragraph{Generating Values} The (non-deterministic)
+$\gen{t}{\tsu}$ in Figure~\ref{fig:gen}, takes
+as input a type $t$ and returns a value of that type.
+%
+For base types, the procedure returns an arbitrary value of
+that type.
+%
+For functions, it returns a lambda with a \emph{new} hole
+denoting the return value, and unconstrained types (denoted
+by $\thole$) yield fresh holes constrained to have type
+\thole (denoted by $\vhole{\thole}$).
+%
+Note that for an algebraic datatype, $\gen{\ttree{t}}{\tsu}$
+will always place holes in the generated nodes; they will be
+lazily filled in later, on demand.
 
+\begin{figure}[ht]
 $$
@@
 $$
+\caption{Generating values}
+\label{fig:gen}
+\end{figure}
 
@@
 \paragraph{Primitive Reductions}
-Primitive reduction steps --- addition, if-elimination, function
-application, and data construction and case analysis --- use \forcesym
-to ensure that values have the appropriate type (and that holes are
-instantiated) before continuing the computation. Importantly,
-beta-reduction \emph{does not} type-check its argument, it only ensures
-that $v_1$ is a function.
+%
+\RJ{Put high-level intuition about how "dynamic HM" is
+formalized in op-sem, to set up next few paragraphs}
+%
+Primitive reduction steps --- addition, if-elimination,
+function application, and data construction and case
+analysis --- use \forcesym to ensure that values have
+the appropriate type (and that holes are instantiated)
+before continuing the computation.
+%
+Importantly, beta-reduction \emph{does not} type-check its
+argument, it only ensures that ``the caller'' $v_1$ is indeed
+a function.
 
 \paragraph{Recursion}
-Fixed-point operators often cannot be typed in static type systems, but
-we are not concerned with \emph{assigning} types to terms, rather with
-showing that \emph{no type} can be assigned. We are simply executing the
-untyped $\lambda$-calculus, which has no issue handling recursion.
+\RJ{this para appears out of nowhere --- non-sequitur}
+Fixed-point operators often cannot be typed in static type
+systems, but we are not concerned with \emph{assigning}
+types to terms, rather with showing that \emph{no type}
+can be assigned.
+%
+We are simply executing the untyped $\lambda$-calculus,
+which has no issue handling recursion.
 
@@
 % \end{proof}
-%
+
 \subsection{Generality}\label{sec:soundness}
@@
 A key technical challenge in generating witnesses is
-that we have no (static) type information to rely upon. Thus, we
-must avoid the trap of generating \emph{spurious} witnesses that
-arise from picking irrelevant values, when instead there exist
-perfectly good values of a \emph{different} type, under which
-the program would not go wrong.
-
-\paragraph{Witness Generality}
-We now show that our evaluation relation instantiates holes in a
-\emph{general} manner. That is, given a function $f$, if we have
-$\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
-then for every possible input type $t$, we can find a value
+that we have no (static) type information to rely upon.
+%
+Thus, we must avoid the trap of generating \emph{spurious}
+witnesses that arise from picking irrelevant values, when
+instead there exist perfectly good values of a \emph{different}
+type, under which the program would not have gone wrong.
+%
+We now show that our evaluation relation instantiates holes
+in a \emph{general} manner. That is, given a function $f$,
+if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
+then \emph{for every} input type $t$, we can find a value
 $v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.
@@
 
-\paragraph{The Type of a a Value} The (dynamic) type of a
-value $v$ is defined as a function $\typeof{v}$ defined as:
-  \[
-  \begin{array}{lcll}
+\paragraph{The Type of a a Value} The \emph{dynamic type}
+of a value $v$ is defined as a function $\typeof{v}$ shown
+in Figure~\ref{fig:typeof}.
+%
+The types of primitive values are defined in the natural manner.
+%
+The types of functions are \emph{approximated}, which is all
+that is needed to ensure an application does not get stuck, \eg
+$$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
+instead of $\tint \rightarrow \tint$.
+%
+The types of (polymorphic) data types are obtained from the
+labels on their values.
+
+\begin{figure}[ht]
+\[ \begin{array}{lcll}
     \typeof{n}   & \defeq & \tint & \\
@@
     % \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
-  \end{array}
-  \]
-%
-Note that we approximate the type of values rather than
-performing traditional type inference, for example
-$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
-instead of the usual $\tint \rightarrow \tint$.
+  \end{array} \]
+\caption{The \emph{dynamic type} of a value.}
+\label{fig:typeof}
+\end{figure}
 
 \paragraph{Dynamic Type Inference}
-We can think of the evaluation of \eapp{f}{\vhole{\thole}} as (dynamically)
-inferring a \emph{partial type} --- a type that may contain holes ---
-for $f$'s input, \ie a partial instantiation of \thole.
+We can think of the evaluation of \eapp{f}{\vhole{\thole}}
+as synthesizing a partial instantiation of \thole, and hence,
+\emph{dynamically inferring} a (partial) type for $f$'s input.
 %
-We can extract this type from an evaluation trace, by repeatedly
-applying the final type substitution to \thole until it contains no
-holes in the domain of the substitution. We will call this process of
-repeated substition \emph{resolving} a hole, and will use
-\resolve{\thole}{\tsu} to denote the the resolution of \thole with
-respect to \tsu.
+We can extract this type from an evaluation trace, by
+\emph{resolving} the \thole\ with the final type
+substitution at the end of the trace.
 %
@@
 then the \emph{partial input type} of $f$ upto $\trace$, written
-\ptype{\trace}{f}, is
-$\resolve{\thole}{\tsu}$.
+\ptype{\trace}{f}, is $\resolve{\thole}{\tsu}$.
+
+%repeatedly
+%applying the final type substitution to \thole until it contains no
+%holes in the domain of the substitution. We will call this process of
+%repeated substition \emph{resolving} a hole, and will use
+%\resolve{\thole}{\tsu} to denote the the resolution of \thole with
+%respect to \tsu.
+
 %$\typeof{\subst{\vsu}{\ehole}}$.
@@
 
-\paragraph{Type Compatibility}
-Two types are compatible if there exists a type subsitution that maps both
-types to the same type.
-% one of them can be obtained by
-% suitably filling the holes of the other (\ie by suitably ``instantiating''
-% the type variables appearing in the other).
+\paragraph{Compatibility}
 %
-Formally, a type $s$ is
-\emph{compatible} with a type $t$, written \tcompat{s}{t}, if
-$\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.
+A \emph{type} $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t},
+if $\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.
+%
+That is, two types are compatible if there exists a type substitution
+that maps both types to the same type.
+%
+A \emph{value} $v$ is \emph{compatible} with a type $t$, written \vcompat{v}{t},
+if $\tcompat{\typeof{v}}{t}$, that is, if the dynamic type of $v$ is
+compatible with $t$.
 
 \paragraph{Preservation}
-Given these two definitions, we show that each evaluation step
-refines the partial input type of $f$, \ie preserves type compatibility.
+We prove that each evaluation step \emph{refines} the partial input type
+of $f$, \ie\ preserves type compatibility.
 %
@@
 \emph{Any} value that is \emph{incompatible} with
-the partial input type upto trace $\trace$ will cause $f$ to get stuck
-in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
+the partial input type upto trace $\trace$ will
+cause $f$ to get stuck in \emph{at most} $k$
+steps, where $k$ is the length of $\trace$.
 %
@@
 \label{lem:k-stuck}
-  If \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
-     \hastype{v}{t} and
-     \tincompat{t}{\ptype{\trace}{f}},
+  For all $v$,
+  if \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
+     \vincompat{v}{\ptype{\trace}{f}},
   then
@@
 \begin{proof}
-  By induction on $k$, the length of $\trace$. Let $s_{\trace} = \ptype{\trace}{f}$ and
+By induction on $k$, the length of $\trace$.
+%
+Suppose {\vincompat{v}{\ptype{\trace}{f}}}.
+%
+We show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
+in at most $k$ steps.
+%
+The base case, $k = 0$ is trivial, $\trace$ is empty
+and so $\ptype{\trace}{f}$ is a hole that is compatible
+with \emph{every} value $v$.
+%
+In the inductive case, let
 $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
-  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
-  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
-  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
-  in at most $k$ steps.
-
-  The base case where $k = 0$ (\ie empty traces) is trivial as no
-  evaluation has happened and thus there can be no values that are
-  incompatible with \ptype{0}{f}. So we proceed directly to the
-  inductive case and split cases on whether $t$ is compatible with
-  $s_{\trace'}$.
+be the prefix of $\trace$ of length $k-1$.
+%
+Furthermore, let $s_{\trace} = \ptype{\trace}{f}$ and $s_{\trace'} = \ptype{\trace'}{f}$.
+%
+Let us split cases on whether $v$ is compatible with $s_{\trace'}$.
 %
 \begin{description}
-  \item [Case \tincompat{t}{s_{\trace'}}:]
+\item [Case \vincompat{v}{s_{\trace'}}:]
   The inductive hypothesis applies.
-  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
-    Since $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$ we know
-    that $s_{\trace'} \neq s_{\trace}$, and
-    by Lemma~\ref{lem:force-inst} we know that we must have
-    successfully called \forcesym at step $k$.
+
+\item [Case $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$:]
+  Since $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$ we know
+  that $s_{\trace'} \neq s_{\trace}$.
+  By Lemma~\ref{lem:force-inst} we know that we must have
+  invoked \forcesym\ at step $k$.
   %
-    Lemma~\ref{lem:refine-partial} tells us
+  The Preservation Lemma~\ref{lem:refine-partial} implies that
   $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
-    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
+  specifically narrowed $s_{\trace'}$ to a type incompatible with $v$.
   %
-    A case analysis of the evaluation rules shows that narrowing to
-    $t$ instead cannot succeed.
+  A case analysis of the evaluation rules shows that such an
+  invocation of \forcesym\ at step $k$
+  cannot succeed, \ie\ yields \stuck.
+   \RJ{some intuition needed --- seems like key step.}
 \end{description}

### 2016-03-07 paper/icfp16-submission/evaluation.tex

@@
 a jump-compressed trace, of size 4. The full trace for
-the same example has size 19.
+the same example has size 20.
 
@@
 \begin{enumerate}
-\item \emphbf{Single-Step Metric}
-      The number of single-step reductions along the path from the
-      witness to the stuck term in the given witness execution, and
+\item \emphbf{Single-Step Metric} The size of the trace after expanding
+  all of the single-step edges from the witness to the stuck term, and
   % This can be thought of as a worst-case
   % complexity, \ie ``How big is the fully-expanded trace?''
-\item \emphbf{Jump Metric}
-      The size of the jump-compressed trace for the given witness execution,
-      \ie\ the number of forward (or backward) jumps that can be taken along the
-      path from the witness to the stuck term.
+\item \emphbf{Jump Metric} The size of the trace for after expanding
+  only the forward (or backward) jumps from the witness to the stuck
+  term.
 \end{enumerate}
@@
 %
-\begin{figure}[t]
+\begin{figure*}[t]
 \centering
-\includegraphics[width=\linewidth]{trace_size_step.png}
-\includegraphics[width=\linewidth]{trace_size_jump.png}
+\includegraphics[width=0.49\linewidth]{trace_size_step.png}
+\includegraphics[width=0.49\linewidth]{trace_size_jump.png}
 \caption{Complexity of the generated traces. 81\% of the combined traces
-  have a jump complexity of at most 10, with an average of 7 and a
-  median of 5.}
+  have a jump complexity of at most 10, with an average complexity of 7
+  and a median of 5.}
 \label{fig:results-complexity}
-\end{figure}
+\end{figure*}
 %
@@
 compile-time analysis, and (2) with traces that have a low average
-complexity of 3 function calls.
+complexity of 7 jumps.
 

### 2016-03-07 paper/icfp16-submission/interactive-short.tex

--- 2016-03-07 paper/icfp16-submission/interactive.tex
@@
 
-In this section we present a novel way to debug executions.
+In this section we present an interactive visualization of program
+executions.
 %
-First, we develop a notion of a \emph{reduction graphs}
-and extend our semantics with a form of \emph{tracing}
-so that they incrementally collect the edges
-in the graph (\S~\ref{sec:inter-semant}).
+First, we extend our semantics (\S~\ref{sec:inter-semant}) to record
+each reduction step in a \emph{trace}, producing a \emph{reduction
+  graph} alongside the witness.
 %
-Next, we express a set of common \emph{interactive debugging}
-steps as graph traversals (\S~\ref{sec:traversing-graph}), yielding
-an novel interactive debugger that allows the user
-to effectively visualize \emph{how} the program goes (wrong).
+Then we describe a set of common \emph{interactive debugging} steps that
+can be expressed as simple traversals over the reduction graph
+(\S~\ref{sec:traversing-graph}), yielding an interactive debugger that
+allows the user to effectively visualize \emph{how} the program goes
+(wrong).
 
@@
 %
-Next, we describe the general recipe for extending a transition
-rule to collect edges.
+% Next, we describe the general recipe for extending a transition
+% rule to collect edges.
 %
@@
 % in Figure~\ref{fig:interactive}.
-As the translation is mechanical, we will not discuss it further.
+% As the translation is mechanical, we will not discuss it further.
 
@@
 from the traced semantics, by describing the visualization
-\emph{state} \ie\ what the user sees at any given moment,
-the set of \emph{commands} available to user and what
-they do, and finally how we use a command to \emph{update}
-the visualization state. In what follows, for clarity of
-exposition, we assume we have a (global) trace:
-$\stepg{e_0}{\emptysu}{\emptysu}{\bullet}{e_n}{\vsu}{\tsu}{\tr}$, where
-$e_0$ and $e_n$ are the \emph{initial} and \emph{final}
-expressions respectively.
+\emph{state} --- \ie\ what the user sees at any given moment ---
+and the set of \emph{commands} available to user and what
+they do.
+% , and finally how we use a command to \emph{update}
+% the visualization state. In what follows, for clarity of
+% exposition, we assume we have a (global) trace:
+% $\stepg{e_0}{\emptysu}{\emptysu}{\bullet}{e_n}{\vsu}{\tsu}{\tr}$, where
+% $e_0$ and $e_n$ are the \emph{initial} and \emph{final}
+% expressions respectively.
 
@@
 %
-A \emph{visualization state} $\vstate$ is a \emph{directed graph}
+A \emph{visualization state} % $\vstate$
+is a \emph{directed graph}
 whose vertices are expressions and whose edges are such
@@
 
-\paragraph{Visualization Context}
-%
-The \emph{visualization context} of each expression $e$
-in the visualization state $\vstate$ is the (unique) linear chain
-in which the expression $e$ belongs.
-%
-We write $\vroot{\vstate}{e}$ for the \emph{first} (or root)
-expression appearing in the visualization context of $e$ in
-$\vstate$.
+\begin{figure}[t]
+\centering
+\includegraphics[width=\linewidth]{fac-steps-new.png}
+\caption{A sequence of interactions with the trace of
+  \texttt{fac 1}. The stuck term is red, in each node the redex is
+  highlighted. Thick arrows denote a multi-step transition, thin arrows
+  denote a single-step transition. We start in step 1. In step 2 we jump
+  forward from the witness to the next function call. In step 3 we step
+  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
+  of execution. In step 4 we take a single step forward from
+  \texttt{fac 0}.}
+\label{fig:nanomaly-factorial}
+\end{figure}
+
+% \paragraph{Visualization Context}
+% %
+% The \emph{visualization context} of each expression $e$
+% in the visualization state $\vstate$ is the (unique) linear chain
+% in which the expression $e$ belongs.
+% %
+% We write $\vroot{\vstate}{e}$ for the \emph{first} (or root)
+% expression appearing in the visualization context of $e$ in
+% $\vstate$.
 
@@
       show the result of taking multiple steps (a \emph{``big''} step)
-      upto the first beta-reduction forward or backward respectively,
+      up to the first function call, or return, forward or backward
+      respectively,
 %
@@
       show the result of stepping into a function call in a sub-term,
-      isolating it from the context,
-
+      isolating it in a new thread, and
+%
 \item \stepoversym:
@@
 
-\paragraph{Update}
-%
-Figure~\ref{fig:traversing-graph} shows
-how we compute the \emph{next} expression
-(to be added to the visualization state)
-given the current visualization state
-$\vstate$, command $\cmd$ and selected
-expression $e$.
-%
-It is straightforward to then \emph{update}
-the visualization graph by adding the new
-term before (resp.\ after) the selected
-expression $e$ if the command was a step
-or jump forward (resp.\ backward), or
-to create a new visualization context
-if the command was $\stepintosym$.
+% \paragraph{Update}
+% %
+% Figure~\ref{fig:traversing-graph} shows
+% how we compute the \emph{next} expression
+% (to be added to the visualization state)
+% given the current visualization state
+% $\vstate$, command $\cmd$ and selected
+% expression $e$.
+% %
+% It is straightforward to then \emph{update}
+% the visualization graph by adding the new
+% term before (resp.\ after) the selected
+% expression $e$ if the command was a step
+% or jump forward (resp.\ backward), or
+% to create a new visualization context
+% if the command was $\stepintosym$.
 
@@
 %%% that @e@ reduces to.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\linewidth]{fac-steps-new.png}
-\caption{A sequence of interactions with the trace of
-  \texttt{fac 1}. The stuck term is red, in each node the redex is
-  highlighted. Thick arrows denote a multi-step transition, thin arrows
-  denote a single-step transition. We start in step 1. In step 2 we jump
-  forward from the witness to the next function call. In step 3 we step
-  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
-  of execution. In step 4 we take a single step forward from
-  \texttt{fac 0}.}
-\label{fig:nanomaly-factorial}
-\end{figure}

### 2016-03-07 paper/icfp16-submission/semantics.tex
fixed genWitness

@@
 
-The witness generation procedure @genWitness@ is formalized in
+The witness generation procedure $\genWitnessN$ is formalized in
 Figure~\ref{fig:algo-gen-witness}.
@@
 %
-First, a search bound $k$ which is used to define the \emph{number} of
+First, a search bound $n$ which is used to define the \emph{number} of
 traces to explore. (We assume, without loss of generality, that all
@@
 %
-% Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
-% denoting the list of \emph{top-level definitions}:
-% %
-% \begin{code}
-%   let f1 = e1 in
-%   let f2 = e2 in
-%   ...
-%   let fn = en in
-% \end{code}
-% %
-% The witness generation procedure returns as output a list of (general)
-% witness expressions $e$ each of which is of the form @fn v1 ... vn@.
-Second, the target expression $e$ that contains the type error, and may be a curried function of multiple arguments.
+Second, the target expression $e$ that contains the type error,
+and may be a curried function of multiple arguments.
 %
@@
 The \emph{empty} list is returned when no witness can be found after
-exploring $k$ traces.
-%
-% In what follows, we refer to @fn@ as the \emph{target} of witness generation.
+exploring $n$ traces.
 
 \paragraph{Modeling Semantics}
-%: \texttt{eval} and \texttt{isStuck}}
 %
@@
 %
-\begin{mcode}
-  eval :: $\triple{e}{\vsu}{\tsu}$ -> [$\triple{v \cup \stuck}{\vsu}{\tsu}$]
-\end{mcode}
+$$
+\evalN :: \triple{e}{\vsu}{\tsu} \rightarrow [\triple{v \cup \stuck}{\vsu}{\tsu}]
+$$
 %
 Due to the non-determinism introduced by \gensym, a call
-\hbox{|eval$\triple{e}{\vsu}{\tsu}$|} returns the \emph{list} of possible results
-of the form $\triple{v \cup \stuck}{\vsu'}{\tsu'}$ such that
-$\steps{e}{\vsu}{\tsu}{v \cup \stuck}{\vsu'}{\tsu'}$.
-%
-% We determine whether a given value is stuck by using the predicate:
-% %
-% \begin{mcode}
-%   isStuck :: $v$ -> Bool
-% \end{mcode}
+$\evalfn{\triple{e}{\vsu}{\tsu}}$ returns the \emph{list}
+of possible results of the form $\triple{v \cup \stuck}{\vsu'}{\tsu'}$
+such that $\steps{e}{\vsu}{\tsu}{v \cup \stuck}{\vsu'}{\tsu'}$.
 
@@
 %
-We address the issue of currying by defining a procedure @genArgs@ such that
-\hbox{|genArgs $e$|} yields a list of holes |[$v_1$,...,$v_n$]|
-such that $\eapp{e}{v_1 \ldots v_n}$ \emph{does not} evaluate to a lambda.
+We address the issue of currying by defining a procedure
+$\genArgs{e}$ that yields a list of holes $[v_1, \ldots, v_k]$
+such that $\eapp{e}{v_1 \ldots v_k}$ \emph{does not} evaluate to
+a lambda.
 %
-This is achieved via a simple @loop@ that keeps adding holes to the
-target application until @eval@uating the term yields a
+This is achieved via a loop that keeps adding holes to the
+target application until evaluating the term yields a
 non-lambda value.
@@
 \paragraph{Generating Witnesses}
-% Procedure \texttt{genWitness}}
 %
 Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
-implementation of our search for witnesses, in the procedure @genWitness@,
-which takes as input a bound @k@ and the target expression $e$, and
-returns a list of witness expressions $\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input
-program gets stuck.
+implementation of our search for witnesses, in the procedure
+$\genWitness{k}{e}$, which takes as input a bound $k$ and the
+target expression $e$, and returns a list of witness expressions
+$\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input program
+gets stuck.
 %
@@
 %
-@(1)@ we invoke @genArgs@ to determine the number of parameters required
-by the (curried) target expression.
-%
-% @(2)@ we obtain a closed expression $e$ by closing the target with the
-% computed holes.
-%
-@(2)@ we take the first @k@ traces returned by @eval@ on the target $\eapp{e}{vs}$
-and
-@(3)@ we extract the substitutions corresponding to the $\stuck$ traces,
-and use them to return the list of witnesses.
+\begin{enumerate}
+  \item We invoke $\genArgsN$ to determine the number of
+        parameters required by the (curried) target expression.
+
+  \item We take the first $k$ traces returned by $\evalN$
+        on the target $\eapp{e}{vs}$, and
+
+  \item We extract the substitutions corresponding to the
+        $\stuck$ traces, and use them to return the list
+        of witnesses.
+\end{enumerate}
 %
@@
 \centering
-% $$
-% \begin{array}{lcll}
-% \genWitness{n}{e} &\defeq& \{ \resolve{e'}{\vsu} \spmid \triple{e'}{\vsu}{\tsu} \in \doeval{\genArgs{e}} \} & \\
-% \end{array}
-% $$
-
-\begin{mcode}
-genWitness :: Nat -> $e$ -> [$e$]
-genWitness n $e$ = [$\resolve{\eapp{e}{vs}}{\vsu}$ | $\vsu$ <- $\vsu$s]
-  where
-   vs  = genArgs $e$                  -- (1)
-   res = take n (eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$) $\ $  -- (2)
-   $\vsu$s   = [$\vsu$ | $\triple{\stuck}{\vsu}{\tsu}$ <- res] $\ \ $   -- (3)
-\end{mcode}
+$$
+\begin{array}{lclr}
+\genWitnessN       & :: & (\mathsf{Nat} \times e) \rightarrow 2^{e} & \\
+\genWitness{n}{e}  & =  & \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \} & \\
+\quad \mbox{\textbf{where}} &    & & \\
+\quad \quad vs     & =  & \genArgs{e} & (1) \\
+\quad \quad res    & =  & \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}} & (2) \\
+\quad \quad \Sigma & =  & \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\} & (3)
+\end{array}
+$$
+%% \begin{mcode}
+%% $\genWitnessN :: (\mathsf{Nat} \times e) \rightarrow 2^{e}$
+%% $\genWitness{n}{e} = \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \}$
+  %% where
+   %% $vs     = \genArgs{e}$               -- (1)
+   %% $res    = \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}}$  -- (2)
+   %% $\Sigma = \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\}$   -- (3)
+%% \end{mcode}
 \caption{Generating Witnesses}

### 2016-03-07 paper/icfp16-submission/semantics.tex
fixed genWitness

@@
 \end{mcode}
-\caption{Find number of arguments for target function}
+\caption{Find number of arguments for target function\RJ{FIX FORMAT}}
 \label{fig:algo-gen-args}
@@
 \label{thm:generation}
-  If |genWitness k $e$ = $\ $ $\triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$|,
+  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$$
   then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that

### 2016-03-07 paper/icfp16-submission/semantics.tex
make genArgs mathy

--- 2016-03-07 paper/icfp16-submission/semantics.tex
@@
 $$
-\forcesym :: (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
+\forcesym : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
 $$
@@
 %% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
-\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
+\forcesym                  & : & v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
 % \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
@@
 \begin{array}{lcll}
-\gensym       & ::  & (t, \tsu) \rightarrow v \\
+\gensym       & :   & t \times \tsu \rightarrow v \\
 \gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
@@
 We resolve the non-determinism in the operational semantics
-(\S~\ref{sec:semantics}) via the procedure:
+(\S~\ref{sec:semantics}) via the procedure
 %
 $$
-\evalN :: \triple{e}{\vsu}{\tsu} \rightarrow [\triple{v \cup \stuck}{\vsu}{\tsu}]
+\evalN : e \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}^{*}
 $$
@@
 Due to the non-determinism introduced by \gensym, a call
-$\evalfn{\triple{e}{\vsu}{\tsu}}$ returns the \emph{list}
-of possible results of the form $\triple{v \cup \stuck}{\vsu'}{\tsu'}$
-such that $\steps{e}{\vsu}{\tsu}{v \cup \stuck}{\vsu'}{\tsu'}$.
+$\evalfn{e}$ returns the \emph{list}
+of possible results of the form $\triple{v \cup \stuck}{\vsu}{\tsu}$
+such that $\steps{e}{\emptysu}{\emptysu}{v \cup \stuck}{\vsu}{\tsu}$.
 
-\paragraph{Generating Target Arguments}
-%
-We address the issue of currying by defining a procedure
-$\genArgs{e}$ that yields a list of holes $[v_1, \ldots, v_k]$
-such that $\eapp{e}{v_1 \ldots v_k}$ \emph{does not} evaluate to
-a lambda.
+We address the issue of currying by defining a procedure \genArgs{e},
+defined in Figure~\ref{fig:algo-gen-args},
+that yields a list of holes $v_1, \ldots, v_n$ such that
+$\eapp{e}{v_1 \ldots v_n}$ \emph{does not} evaluate to a lambda.
 %
@@
 % $$
+$$
+\begin{array}{lcl}
+\genArgsN   & : & e \rightarrow e^{*} \\
+\genArgs{e} & = & \loopfn{[]} \\
+\quad \mbox{\textbf{where}} &    & \\
+  \quad \quad \loopfn{vs} & = & \mbox{\textbf{case }} \evalfn{\eapp{e}{vs}} \mbox{\textbf{ of}} \\
+             &   & \quad \triple{\efun{x}{e}}{\vsu}{\tsu},\ldots \rightarrow \loopfn{\vhole{\thole} :: vs} \\
+             &   & \qquad (\ehole, \thole \mbox{ are fresh}) \\
+             &   & \quad \_ \rightarrow vs
+\end{array}
+$$
 
-\begin{mcode}
-genArgs :: $e$ -> [$v$]
-genArgs $e$ = loop []
-  where
-  loop vs = case eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$ of
-             $\triple{\efun{x}{e}}{\vsu}{\tsu}$:_ -> loop ($\vhole{\thole}$ : vs)
-                           -- $\ehole$, $\thole$ are fresh
-             _          -> vs
-\end{mcode}
-\caption{Find number of arguments for target function}
+% \begin{mcode}
+% genArgs :: $e$ -> [$v$]
+% genArgs $e$ = loop []
+%   where
+%   loop vs = case eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$ of
+%              $\triple{\efun{x}{e}}{\vsu}{\tsu}$:_ -> loop ($\vhole{\thole}$ : vs)
+%                            -- $\ehole$, $\thole$ are fresh
+%              _          -> vs
+% \end{mcode}
+\caption{Finding the number of arguments for the target expression.}
 \label{fig:algo-gen-args}
@@
 \label{thm:generation}
-  If |genWitness k $e$ = $\ $ $\triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$|,
+  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$,
   then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
@@
 \begin{array}{lclr}
-\genWitnessN       & :: & (\mathsf{Nat} \times e) \rightarrow 2^{e} & \\
+\genWitnessN       & : & \mathsf{Nat} \times e \rightarrow e^{*} & \\
 \genWitness{n}{e}  & = & \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \} & \\
@@
 \quad \quad vs     & = & \genArgs{e} & (1) \\
-\quad \quad res    & =  & \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}} & (2) \\
+\quad \quad res    & = & \takefn{n}{\evalfn{\eapp{e}{vs}}} & (2) \\
 \quad \quad \Sigma & = & \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\} & (3)
@@
 %% \end{mcode}
-\caption{Generating Witnesses}
+\caption{Generating witnesses.}
 \label{fig:algo-gen-witness}

### 2016-03-08 paper/icfp16-submission/gallery2.tex
fixed genWitness

@@
-\section{The Advantage Of Traces}
-\label{sec:advantage-traces}
-Having shown that \toolname produces small traces in an efficient
-manner, let us discuss the explanatory power of traces compared with
-typical type error messages. In this section we will present a
-qualitative evaluation of \toolname's jump-compressed traces on a series
-of examples drawn from real student programs in the \ucsdbench
-dataset. We will show how \toolname's traces get to the heart of the
-error by
+\section{The Advantage Of Traces}\label{sec:advantage-traces}
+
+Next, we present a \emph{qualitative} evaluation that compares
+the explanations provided by \toolname's dynamic witnesses with
+the static reports produced by the \ocaml\ compiler and \sherrloc,
+a state-of-the-art fault localization approach~\cite{ZhangMyers}.
 %
-\begin{enumerate}
-\item highlighting the conflicting values that cause the program to get
+In particular, we illustrate, using a series of examples drawn
+from student programs in the \ucsdbench\ dataset how \toolname's
+jump-compressed traces can get to the heart of the error by
+%
+highlighting the conflicting values that cause the program to get
 stuck, rather that blaming a single one,
-\item showing the steps necessary to reach the stuck state, and
-\item not assuming that a function is correct just because it type-checks.
-\end{enumerate}
 %
-For each example we will present the code, the type error produced by
-the \ocaml compiler, and a jump-compressed trace produced by \toolname.
+showing the steps necessary to reach the stuck state, and
+%
+not assuming that a function is correct just because it type-checks.
+%
+For each example we will present
+(1)~the code,
+(2)~the error message returned \ocaml,
+(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),
+(4)~the jump-compressed trace produced by \toolname.
 
@@
 
-First, we will examine a simple recursive function @sqsum@ that should
-square each element of the input list and then compute the sum of the
-result. In the following code snippets we will underline the source span
-that \ocaml blames for the type error.
+\paragraph{Example: Recursion with Bad Operator}
+The recursive function @sqsum@ should square each
+element of the input list and then compute the sum
+of the result.
 %
@@
     | [] -> 0
-    | h::t -> __(sqsum t)__ @ (h * h)
+    | h::t -> ==(sqsum t)== @ (h * h)
 \end{ecode}
 %
-Unfortunately the student has used the list-append operator |@| instead
-of \texttt{+} to compute the sum. \ocaml blames the recursive call
-\texttt{sqsum t} on line 3 for being an \texttt{int} instead of a
-\texttt{list}.
+Unfortunately the student has used the list-append
+operator |@| instead of \texttt{+} to compute the sum.
+%
+Both \ocaml\ and \sherrloc\ blame the \emph{wrong location},
+namely the recursive call @sqsum t@ with the message
 %
@@
 %
-In contrast, \toolname produces a trace showing how the evaluation of
-@sqsum [1]@ eventually gets stuck when it tries to evaluate
-\texttt{ 0 }|@|\texttt{ 1}.
+\toolname\ produces the following trace showing how the evaluation of
+@sqsum [1]@ gets stuck:
 %
@@
 %
-Note how the eye is drawn to the entire stuck term as opposed to just
-the left sub-term. This emphasizes the \emph{conflict} between @int@ and
-@list@ rather than assuming one or the other is correct.
+The figure highlights the entire stuck term
+(not just the recursive call), emphasizing
+the \emph{conflict} between @int@ and @list@
+rather than assuming one or the other is correct.
 
-Next we consider a function @sumList@ that just sums the elements of the
-input list, but has a more unfortunate error.
+\paragraph{Example: Recursion with Bad Base Case}
+%
+The function @sumList@ is supposed to add up
+the elements of its input list.
+%
 \begin{ecode}
   let rec sumList xs = match xs with
-    | []    -> []
+    | []    -> ==[]==
     | y::ys -> y + __sumList ys__
@@
 %
-In this case the student has returned @[]@ in the base case instead of
-@0@. Unfortunately, \ocaml deduces (incorrectly) from the base case that
-@sumList@ must return a @list@, and then proceeds to blame the
-\emph{recursive call} on line 3 for producing a @list@ instead of an
-@int@.
+Here, the student has incorrectly returned @[]@
+in the base case instead of @0@.
+%
+Unfortunately, \ocaml\ deduces (incorrectly)
+from the base case that @sumList@ must return
+a @list@, and then proceeds to blame the
+\emph{recursive call} on line 3 for
+producing a @list@ instead of an @int@.
 %

### 2016-03-08 paper/icfp16-submission/semantics.tex

@@
 %
-First, a search bound $n$ which is used to define the \emph{number} of
+First, a search bound $k$ which is used to define the \emph{number} of
 traces to explore. (We assume, without loss of generality, that all
@@
 Due to the non-determinism introduced by \gensym, a call
-$\evalfn{e}$ returns the \emph{list}
+$\evalfn{e}$ returns a \emph{list}
 of possible results of the form $\triple{v \cup \stuck}{\vsu}{\tsu}$
@@
 
+\paragraph{Currying}
 We address the issue of currying by defining a procedure \genArgs{e},
-defined in Figure~\ref{fig:algo-gen-args},
-that yields a list of holes $v_1, \ldots, v_n$ such that
-$\eapp{e}{v_1 \ldots v_n}$ \emph{does not} evaluate to a lambda.
+defined in Figure~\ref{fig:algo-gen-args}, that takes as input an
+expression $e$ and produces a \emph{saturated} expression of the form
+$\eapp{e}{\ehole_1[\thole_1] \ldots \ehole_n[\thole_n]}$ that
+\emph{does not} evaluate to a lambda.
 %
-This is achieved via a loop that keeps adding holes to the
-target application until evaluating the term yields a
-non-lambda value.
+This is achieved with a simple loop that keeps adding holes to the
+target application until evaluating the term yields a non-lambda value.
 %
@@
 $$
-\begin{array}{lcl}
-\genArgsN   & : & e \rightarrow e^{*} \\
-\genArgs{e} & = & \loopfn{[]} \\
-\quad \mbox{\textbf{where}} &    & \\
-  \quad \quad \loopfn{vs} & = & \mbox{\textbf{case }} \evalfn{\eapp{e}{vs}} \mbox{\textbf{ of}} \\
-             &   & \quad \triple{\efun{x}{e}}{\vsu}{\tsu},\ldots \rightarrow \loopfn{\vhole{\thole} :: vs} \\
-             &   & \qquad (\ehole, \thole \mbox{ are fresh}) \\
-             &   & \quad \_ \rightarrow vs
+\begin{array}{lclr}
+\genArgsN   & : & e \rightarrow e \\
+\genArgs{e} & = & \mbox{\textbf{case }} \evalfn{e} \mbox{\textbf{ of}} \\
+ \quad \triple{\efun{x}{e}}{\vsu}{\tsu},\ldots &\rightarrow& \genArgs{\eapp{e}{\vhole{\thole}}} & (\ehole, \thole \mbox{ are fresh}) \\
+ \quad \_ &\rightarrow& e \\
 \end{array}
 $$
+\caption{Generating a saturated application.}
+\label{fig:algo-gen-args}
 
-% \begin{mcode}
-% genArgs :: $e$ -> [$v$]
-% genArgs $e$ = loop []
-%   where
-%   loop vs = case eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$ of
-%              $\triple{\efun{x}{e}}{\vsu}{\tsu}$:_ -> loop ($\vhole{\thole}$ : vs)
-%                            -- $\ehole$, $\thole$ are fresh
-%              _          -> vs
-% \end{mcode}
-\caption{Finding the number of arguments for the target expression.}
-\label{fig:algo-gen-args}
 \end{figure}
@@
 \begin{enumerate}
-  \item We invoke $\genArgsN$ to determine the number of
-        parameters required by the (curried) target expression.
+  \item We invoke $\genArgs{e}$ to produce a \emph{saturated}
+        application $e_{sat}$.
 
   \item We take the first $k$ traces returned by $\evalN$
-        on the target $\eapp{e}{vs}$, and
+        on the target $e_{sat}$, and
 
@@
 \label{thm:generation}
-  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$,
+  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$$
   then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
@@
 \genWitnessN       & : & \mathsf{Nat} \times e \rightarrow e^{*} & \\
-\genWitness{n}{e}  & = & \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \} & \\
+\genWitness{n}{e}  & = & \{ \resolve{e_{sat}}{\vsu} \mid \vsu \in \Sigma \} & \\
 \quad \mbox{\textbf{where}} &    & & \\
-\quad \quad vs     & = & \genArgs{e} & (1) \\
-\quad \quad res    & = & \takefn{n}{\evalfn{\eapp{e}{vs}}} & (2) \\
+\quad \quad e_{sat} & = & \genArgs{e} & (1) \\
+\quad \quad res    & = & \takefn{n}{\evalfn{e_{sat}}} & (2) \\
 \quad \quad \Sigma & = & \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\} & (3)

### 2016-03-08 paper/icfp16-submission/semantics.tex
misc semantics

@@
 for ill-typed programs that precisely defines the notion
-of a \emph{witness}~\S~\ref{sec:semantics},
+of a \emph{witness}~\S~\ref{sec:semantics}.
 %
 Third, we formalize and prove a notion of \emph{generality} for
-witnesses, which states, intuitively that if we find a
+witnesses, which states, intuitively, that if we find a
 single witness then for \emph{every possible} type
-assignment, there exist inputs that are guaranteed to make
-the program ``go wrong''~\S~\ref{sec:soundness}, and finally,
+assignment there exist inputs that are guaranteed to make
+the program ``go wrong''~\S~\ref{sec:soundness}.
 %
-Fourth, we refine the operational semantics into a
+Finally, we refine the operational semantics into a
 \emph{search procedure} that returns concrete (general)
@@
 %
-For example \RJ{add example here}
+For example
+$\vnode{\thole}{\vhole{\thole}}{\vleaf{\thole}}{\vleaf{\thole}}$ is a
+normalized value, and its type $\ttree{\thole}$ is a normalized type.
 
@@
 %
-A \emph{normalized} substitution is one whose co-domain comprises
+A \emph{normalized} substitution is one whose co-domain is comprised
 of normalized values (resp.\ types).
@@
 %
-Note that for an algebraic datatype, $\gen{\ttree{t}}{\tsu}$
-will always place holes in the generated nodes; they will be
-lazily filled in later, on demand.
+When generating a $\ttree{t}$ we must take care to ensure
+the resulting tree is well-typed. For a polymorphic type
+$\ttree{\thole}$, eg in \recasegoodone, we will
+place holes in the generated nodes; they will be lazily
+filled in later, on demand.
 

### 2016-03-08 paper/icfp16-submission/gallery2.tex
fixed genWitness

@@
 (3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),
+\footnote{When the locations from \ocaml\ and \sherrloc\ overlap,
+we just underline the relevant code.}
 (4)~the jump-compressed trace produced by \toolname.
@@
     | [] -> 0
-    | h::t -> ==(sqsum t)== @ (h * h)
+    | h::t -> __(sqsum t)__ @ (h * h)
 \end{ecode}
@@
 %
-The function @sumList@ is supposed to add up
+The function @sumList@ should add up
 the elements of its input list.
@@
 %
-Here, the student has incorrectly returned @[]@
+Unfortunately, it incorrectly returns @[]@
 in the base case instead of @0@.
 %
-Unfortunately, \ocaml\ deduces (incorrectly)
-from the base case that @sumList@ must return
-a @list@, and then proceeds to blame the
-\emph{recursive call} on line 3 for
-producing a @list@ instead of an @int@.
+\sherrloc\ blames the base case, and \ocaml\
+assumes the base case is correct and blames
+the \emph{recursive call} on line 3:
 %
@@
 %
-\toolname's jump compressed trace shows how the evaluation of
-@sumList [1; 2]@ gets stuck when it tries to evaluate @2 + []@.
+Both of the above are parts of the full story, which
+is succinctly summarized by \toolname's trace showing
+how @sumList [1; 2]@ gets stuck at @2 + []@.
 %
@@
 %
-This is essentially the same error that \ocaml produces, but the trace
-clarifies immediately (via the third step) that the @[]@ is the result
-of the recursive call @sumList []@, drawing attention to the incorrect
-base case.
+The trace clarifies immediately (via the third step)
+that the @[]@ is the result of the recursive call
+@sumList []@, and shows how it is incompatible with
+the subsequent \texttt{+} operation.
 
@@
 
-Finally, consider the higher-order function @wwhile@ that emulates a
-traditional while-loop. Concretely, @wwhile@ takes a function @f@ and
-repeatedly calls @f@ on the first element of its output pair, starting
-with the initial value @b@, until the second element is @false@.
+\paragraph{Example: Higher-Order Functions}
+%
+The higher-order function @wwhile@ is supposed
+to emulate a traditional while-loop. It takes
+a function @f@ and repeatedly calls @f@ on the
+first element of its output pair, starting with
+the initial value @b@, until the second element
+is @false@.
 %
@@
 %
-Unfortunately, the student has forgotten to apply @f@ at all on line 2,
-and just matches it directly against a pair. This faulty definition of
-@wwhile@ still typechecks however, and \ocaml thus blames the
-\emph{call-site} on line 10.
+Unfortunately, the student has forgotten to \emph{apply}
+@f@ at all on line 2, and just matches it directly against
+a pair.
+This faulty definition of @wwhile@ still typechecks however,
+and hence, are ``assumed'' to be correct by and both \ocaml\
+and \sherrloc\ which blame the \emph{use-site} on line 10.
 %
@@
 %
-\toolname makes no assumptions about the program and quickly draws the
-eye to the true error, the @match@ expression on line 2, and highlights
+\toolname\ synthesizes a trace that draws the eye to the
+true error: the @match@ expression on line 2, and highlights
 the conflict in matching a function against a pair pattern.
@@
 %
-
-By highlighting conflicting values rather than blaming a single value
-and not making any assumptions about the given program, \toolname avoids
-misleading the user into focusing their attention on a piece of code
-that is actually irrelevant to the error.
+By highlighting conflicting values (\ie\ the source and sink
+of the problem) and not making any assumptions, \toolname\
+focusses the user's attention on the piece of code that is
+actually relevant to the error.
 

### 2016-03-08 paper/icfp16-submission/gallery2.tex
add digitsOfInt

@@
 the static reports produced by the \ocaml\ compiler and \sherrloc,
-a state-of-the-art fault localization approach~\cite{ZhangMyers}.
+a state-of-the-art fault localization approach~\cite{zhang_toward_2014}.
 %
@@
 
+\paragraph{Example: Bad Helper Function that Type-Checks}
+%
+The function @digitsOfInt@ is supposed to return a list of
+the digits of the input integer.
+%
+\begin{ecode}
+  let append x xs =
+    match xs with
+    | [] -> [x]
+    | _  -> x :: xs
+
+  let rec digitsOfInt n =
+    if n <= 0 then
+      []
+    else
+      append (==digitsOfInt (n / 10)==)
+             [__n mod 10__]
+\end{ecode}
+%
+Unfortunately, the student's @append@ function \emph{conses} an element
+onto a list instead of appending two lists.
+%
+Though incorrect, @append@ still type-checks and thus \ocaml and
+\sherrloc blame the \emph{use-site} on line 10.
+%
+\begin{verbatim}
+  This expression has type
+    int
+  but an expression was expected of type
+    'a list
+\end{verbatim}
+%
+\toolname, in contrast, makes no asummptions about @append@ and produces
+a trace that illustrates the true error on line 4, by
+highlighting the conflict in consing a list onto a list of integers.
+%
+\begin{center}
+  \includegraphics[height=175px]{digitsOfInt.png}
+\end{center}
+%
+
 \paragraph{Example: Higher-Order Functions}

### 2016-03-09 paper/icfp16-submission/semantics.tex
type substitutions are not normalized
consider `node v[a_1] leaf leaf`, we'll end up generating a fresh a_2
and a_3 for the leafs and extend the type substition with `a_1 = a_2 = a_3`

@@
 %
-A \emph{normalized} value (resp.\ type) is one that is not a hole,
+A \emph{normalized} value is one that is not a hole,
 but which may internally contain holes.
@@
 $\vnode{\thole}{\vhole{\thole}}{\vleaf{\thole}}{\vleaf{\thole}}$ is a
-normalized value, and its type $\ttree{\thole}$ is a normalized type.
+normalized value.
 
@@
 %
-A \emph{normalized} substitution is one whose co-domain is comprised
-of normalized values (resp.\ types).
+A \emph{normalized} value substitution is one whose co-domain is
+comprised of normalized values.
 %
-In the sequel, we will assume and ensure that all substitutions
+In the sequel, we will assume and ensure that all value substitutions
 are normalized.
@@
   \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
-  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t_1}{\tsu'}\\
+  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
 \end{cases} \\
@@
 %lambda or a tree, it may not map \emph{directly} to another hole,
-As the substitutions are normalized, in the first case of \forcesym\ we
+As the value substitution is normalized, in the first case of \forcesym\ we
 do not need to \forcesym\ the result of the substitution, the sub-hole

### 2016-03-10 paper/icfp16-submission/semantics.tex
rewrite lemma k-stuck

@@
 
-\paragraph{Incompatible Values Are Wrong}
+\begin{lem}
+\label{lem:vsu-ext}
+If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
+and $\trace'$ is a single-step extension of $\trace$
+and $\resolve{\vhole{\thole}}{\vsu} \neq \resolve{\vhole{\thole}}{\vsu'}$
+then $\vsu' = \vsu[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
+\end{lem}
+\begin{proof}
+  By case analysis on the evaluation rules.
   %
-\emph{Any} value that is \emph{incompatible} with
-the partial input type upto trace $\trace$ will
-cause $f$ to get stuck in \emph{at most} $k$
-steps, where $k$ is the length of $\trace$.
+  $\vhole{\thole}$ does not change, so if the resolved values differ then
+  $\vsu \neq \vsu'$.
   %
+  Note that only \forcesym can update \vsu\
+  and only the first case,
+  when we extend \vsu\ with $[\ehole \mapsto v]$,
+  does.
+\end{proof}
+
+\paragraph{Incompatible Types Are Wrong}
+\emph{For all} types that are \emph{incompatible} with the
+partial input type up to trace $\trace$, there exists a value
+that will cause $f$ to get stuck in \emph{at most} $k$ steps,
+where $k$ is the length of $\trace$.
+
 \begin{lem}
 \label{lem:k-stuck}
-  For all $v$,
-  if \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
-     \vincompat{v}{\ptype{\trace}{f}},
-  then
-     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
-     in at most $k$ steps, where $k$ is the length of $\trace$.
+For all $t$,
+if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
+   $\tincompat{t}{\ptype{\trace}{f}}$,
+   then there exists a $v$ such that $\hastype{v}{t}$ and
+   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$ in at most
+   $k$ steps, where $k$ is the length of $\trace$.
 \end{lem}
 \begin{proof}
-By induction on $k$, the length of $\trace$.
+  We can construct $v$ from $\trace$ by induction on $k$, the length of $\trace$.
   %
-Suppose {\vincompat{v}{\ptype{\trace}{f}}}.
+  The goal is to find the step at which the refinement of $\vhole{\thole}$,
+  $\resolve{\vhole{\thole}}{\vsu}$ becomes incompatible with $t$.
   %
-We show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
-in at most $k$ steps.
-%
-The base case, $k = 0$ is trivial, $\trace$ is empty
-and so $\ptype{\trace}{f}$ is a hole that is compatible
-with \emph{every} value $v$.
+  The base case, $k = 0$ is trivial, $\vsu_0 = \emptysu$ and
+  $\resolve{\vhole{\thole}}{\emptysu} = \vhole{\thole}$ which is compatible with
+  \emph{any} type $t$.
   %
   In the inductive case, let
-$\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
+  $$\trace_{k-1} = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e_{k-1}}{\vsu_{k-1}}{\tsu_{k-1}}$$
   be the prefix of $\trace$ of length $k-1$.
   %
-Furthermore, let $s_{\trace} = \ptype{\trace}{f}$ and $s_{\trace'} = \ptype{\trace'}{f}$.
-%
-Let us split cases on whether $v$ is compatible with $s_{\trace'}$.
+  Let us split cases on whether $\resolve{\vhole{\thole}}{\vsu_{k-1}}$ is compatible with $t$.
   %
   \begin{description}
-\item [Case \vincompat{v}{s_{\trace'}}:]
+  \item[Case $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k-1}}}{t}$:]
+    %
     The inductive hypothesis applies.
+    %
+  \item[Case $\vcompat{\resolve{\vhole{\thole}}{\vsu_{k-1}}}{t}$
+        but  $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$:]
+    %
+    In this case we will show that $\resolve{\vhole{\thole}}{\vsu_{k-1}}$
+    must contain some other hole $\ehole'$ which is instantiated with a
+    concrete value at step $k$.
+    %
+    Furthermore, $\ehole'$ is instantiated in such a way that
+    $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$.
+    %
+    Finally, we will show that if we had instantiated $\ehole'$ with a
+    value such that
+    $\vcompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$, the current
+    step would have gotten $\stuck$.
 
-\item [Case $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$:]
-  Since $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$ we know
-  that $s_{\trace'} \neq s_{\trace}$.
-  By Lemma~\ref{lem:force-inst} we know that we must have
-  invoked \forcesym\ at step $k$.
+    % By Lemma~\ref{lem:fixme} we know that
+    % $\vcompat{\resolve{\vhole{\thole}}{\vsu}}{\resolve{\thole}{\tsu}}$.
     %
-  The Preservation Lemma~\ref{lem:refine-partial} implies that
-  $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
-  specifically narrowed $s_{\trace'}$ to a type incompatible with $v$.
+    By Lemma~\ref{lem:vsu-ext} we know that
+    $\vsu_{k} = \vsu_{k-1}[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
     %
-  A case analysis of the evaluation rules shows that such an
-  invocation of \forcesym\ at step $k$
-  cannot succeed, \ie\ yields \stuck.
-   \RJ{some intuition needed --- seems like key step.}
+    We will assume, without loss of generality, that
+    $\vsu_{k} = \vsu_{k-1}[\ehole' \mapsto v']$.
+
+    Next, we introduce a notion of \emph{value contexts}, defined
+    analogously to evaluation contexts.
+    %
+    $$
+      V ::= \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V}
+    $$
+    %
+    Since $\vsu_{k-1}$ and $\vsu_{k}$ differ only in $\ehole'$ but the resolved
+    values differ,
+    $\resolve{\vhole{\thole}}{\vsu_{k-1}} = \vcompat{\invctx{\ehole'[\thole']}}{t}$
+    and
+    $\resolve{\vhole{\thole}}{\vsu_{k}} = \vincompat{\invctx{v'}}{t}$.
+    %
+    Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
+    we show by case analysis on the evaluation relation that
+    $\invctx{u}$ could not make progress at this step, and thus
+    $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
   \end{description}
 \end{proof}
+
+% \paragraph{Incompatible Values Are Wrong}
+%
+% \emph{Any} value that is \emph{incompatible} with
+% the partial input type upto trace $\trace$ will
+% cause $f$ to get stuck in \emph{at most} $k$
+% steps, where $k$ is the length of $\trace$.
+% %
+% \begin{lem}
+% \label{lem:k-stuck}
+%   For all $v$,
+%   if \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
+%      \vincompat{v}{\ptype{\trace}{f}},
+%   then
+%      \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
+%      in at most $k$ steps, where $k$ is the length of $\trace$.
+% \end{lem}
+% \begin{proof}
+% By induction on $k$, the length of $\trace$.
+% %
+% Suppose {\vincompat{v}{\ptype{\trace}{f}}}.
+% %
+% We show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
+% in at most $k$ steps.
+% %
+% The base case, $k = 0$ is trivial, $\trace$ is empty
+% and so $\ptype{\trace}{f}$ is a hole that is compatible
+% with \emph{every} value $v$.
+% %
+% In the inductive case, let
+% $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
+% be the prefix of $\trace$ of length $k-1$.
+% %
+% Furthermore, let $s_{\trace} = \ptype{\trace}{f}$ and $s_{\trace'} = \ptype{\trace'}{f}$.
+% %
+% Let us split cases on whether $v$ is compatible with $s_{\trace'}$.
+% %
+% \begin{description}
+% \item [Case \vincompat{v}{s_{\trace'}}:]
+%   The inductive hypothesis applies.
+
+% \item [Case $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$:]
+%   Since $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$ we know
+%   that $s_{\trace'} \neq s_{\trace}$.
+%   By Lemma~\ref{lem:force-inst} we know that we must have
+%   invoked \forcesym\ at step $k$.
+%   %
+%   The Preservation Lemma~\ref{lem:refine-partial} implies that
+%   $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
+%   specifically narrowed $s_{\trace'}$ to a type incompatible with $v$.
+%   %
+%   A case analysis of the evaluation rules shows that such an
+%   invocation of \forcesym\ at step $k$
+%   cannot succeed, \ie\ yields \stuck.
+%    \RJ{some intuition needed --- seems like key step.}
+% \end{description}
+% \end{proof}
 
@@
   $\eapp{f}{v}$ gets stuck.
-\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
-  such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
-  % \ES{do we need to say anythign else?}
+\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, we can derive
+  a $v$ from $\trace$ such that \hastype{v}{t} and $\eapp{f}{v}$ gets
+  stuck.
+% \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
+%   such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
+%   % \ES{do we need to say anythign else?}
 \end{description}

### 2016-03-10 paper/icfp16-submission/semantics.tex
no induction in lemma k-stuck

@@
 \begin{proof}
-  We can construct $v$ from $\trace$ by induction on $k$, the length of $\trace$.
+  We can construct $v$ from $\trace$ as follows.
   %
-  The goal is to find the step at which the refinement of $\vhole{\thole}$,
-  $\resolve{\vhole{\thole}}{\vsu}$ becomes incompatible with $t$.
+  Let
+  $$
+  \trace_i = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},
+             \ldots,
+             \triple{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}},
+             \triple{e_{i}}{\vsu_{i}}{\tsu_{i}}
+  $$
+  be the shortest prefix of $\trace$ such that
+  $\tincompat{\ptype{\trace_i}{f}}{t}$.
   %
-  The base case, $k = 0$ is trivial, $\vsu_0 = \emptysu$ and
-  $\resolve{\vhole{\thole}}{\emptysu} = \vhole{\thole}$ which is compatible with
-  \emph{any} type $t$.
-  %
-  In the inductive case, let
-  $$\trace_{k-1} = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e_{k-1}}{\vsu_{k-1}}{\tsu_{k-1}}$$
-  be the prefix of $\trace$ of length $k-1$.
-  %
-  Let us split cases on whether $\resolve{\vhole{\thole}}{\vsu_{k-1}}$ is compatible with $t$.
-  %
-  \begin{description}
-  \item[Case $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k-1}}}{t}$:]
-    %
-    The inductive hypothesis applies.
-    %
-  \item[Case $\vcompat{\resolve{\vhole{\thole}}{\vsu_{k-1}}}{t}$
-        but  $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$:]
-    %
-    In this case we will show that $\resolve{\vhole{\thole}}{\vsu_{k-1}}$
+  We will show that $\resolve{\vhole{\thole}}{\vsu_{i-1}}$
   must contain some other hole $\ehole'$ which is instantiated with a
-    concrete value at step $k$.
+  concrete value at step $i$.
   %
   Furthermore, $\ehole'$ is instantiated in such a way that
-    $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$.
+  $\vincompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$.
+  \ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
   %
@@
   value such that
-    $\vcompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$, the current
+  $\vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$, the current
   step would have gotten $\stuck$.
@@
   By Lemma~\ref{lem:vsu-ext} we know that
-    $\vsu_{k} = \vsu_{k-1}[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
+  $\vsu_{i} = \vsu_{i-1}[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
   %
   We will assume, without loss of generality, that
-    $\vsu_{k} = \vsu_{k-1}[\ehole' \mapsto v']$.
+  $\vsu_{i} = \vsu_{i-1}[\ehole' \mapsto v']$.
 
@@
   %
-    Since $\vsu_{k-1}$ and $\vsu_{k}$ differ only in $\ehole'$ but the resolved
+  Since $\vsu_{i-1}$ and $\vsu_{i}$ differ only in $\ehole'$ but the resolved
   values differ,
-    $\resolve{\vhole{\thole}}{\vsu_{k-1}} = \vcompat{\invctx{\ehole'[\thole']}}{t}$
+  $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \vcompat{\invctx{\ehole'[\thole']}}{t}$
   and
-    $\resolve{\vhole{\thole}}{\vsu_{k}} = \vincompat{\invctx{v'}}{t}$.
+  $\resolve{\vhole{\thole}}{\vsu_{i}} = \vincompat{\invctx{v'}}{t}$.
   %
@@
   $\invctx{u}$ could not make progress at this step, and thus
-    $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
-  \end{description}
+  $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
 \end{proof}

### 2016-03-10 paper/icfp16-submission/semantics.tex
add intuition text for 3.2 and clean up recursion para

@@
 %
-First, we define a core calculus within which we will work~\S~\ref{sec:syntax}.
+First, we define a core calculus within which we will work~(\S~\ref{sec:syntax}).
 %
@@
 for ill-typed programs that precisely defines the notion
-of a \emph{witness}~\S~\ref{sec:semantics}.
+of a \emph{witness}~(\S~\ref{sec:semantics}).
 %
@@
 assignment there exist inputs that are guaranteed to make
-the program ``go wrong''~\S~\ref{sec:soundness}.
+the program ``go wrong''~(\S~\ref{sec:soundness}).
 %
@@
 \emph{search procedure} that returns concrete (general)
-witnesses for ill-typed programs~\S~\ref{sec:search-algorithm}.
+witnesses for ill-typed programs~\S~(\ref{sec:search-algorithm}).
 
@@
 As we are specifically interested in programs that \emph{do} go wrong,
-we include an explicit \stuck\ state in our syntax.
+we include an explicit \stuck\ term in our syntax. We write $\estuck$ to
+denote terms that may be \stuck, and $e$ to denote terms that may not be
+stuck.
 
@@
 an \emph{unconstrained} value $\ehole$ that may be replaced with
-\emph{any} value of type \thole.
+\emph{any} value of an unknown type \thole.
 %
@@
 %
+The substitutions let us ensure that we consistently instantiate each
+hole with the same (partially defined) value or type, regardless of the
+multiple contexts in which the hole appears.
+%
+This ensures we can report a concrete (and general) witness for any
+(dynamically) discovered type errors.
+
 A \emph{normalized} value substitution is one whose co-domain is
@@
 %
-\RJ{Add intuition text -- "dynamic Hindley Milner"}
+Recall that our goal is to synthesize a value that demonstrates
+why (and how) a function goes wrong.
 %
-The substitutions let us ensure that we consistently instantiate
-each hole with the same (partially defined) value or type, regardless
-of the multiple contexts in which the hole appears. This ensures we
-can report a concrete (and general) witness for any (dynamically)
-discovered type errors.
+We accomplish this by combining evaluation with type inference,
+giving us a form of dynamic type inference.
+% starting with an unconstrained hole and gradually refining it
+% as evaluation constrains its type.
+%
+Each primitive evaluation step tells us more about the types of the
+program values. For example, addition tells us that the addends must be
+integers, and case-analysis tells us that the scrutinee must be a tree.
+%
+When a hole appears in such a context, we know what type it must have
+in order to make progress and can fill it in with a concrete value.
 
@@
 When generating a $\ttree{t}$ we must take care to ensure
-the resulting tree is well-typed. For a polymorphic type
-$\ttree{\thole}$, eg in \recasegoodone, we will
-place holes in the generated nodes; they will be lazily
-filled in later, on demand.
+the resulting tree is well-typed.
+%
+For a polymorphic type $\ttree{\thole}$ %, eg in \recasegoodone,
+we will place holes in the generated nodes; they will be lazily filled
+in later, on demand.
 
@@
 %
-A program state is a triple $\triple{e \cup \stuck}{\vsu}{\tsu}$ of an
+A configuration is a triple $\triple{\estuck}{\vsu}{\tsu}$ of an
 expression $e$ or the stuck term $\stuck$, a value substitution $\vsu$,
-and a type substitution $\tsu$. In the sequel we will write \estuck to
-denote either an expression $e$ or \stuck.
+and a type substitution $\tsu$.
+%
+% We write \estuck to
+% denote either an expression $e$ or \stuck.
 %
@@
 %
-\RJ{Put high-level intuition about how "dynamic HM" is
-formalized in op-sem, to set up next few paragraphs}
+% \RJ{Put high-level intuition about how "dynamic HM" is
+% formalized in op-sem, to set up next few paragraphs}
 %
@@
 \paragraph{Recursion}
-\RJ{this para appears out of nowhere --- non-sequitur}
+%
+% \RJ{this para appears out of nowhere --- non-sequitur}
+Our semantics lacks a built-in $\mathtt{fix}$ construct
+for defining recursive function, which may surprise
+the reader.
+%
 Fixed-point operators often cannot be typed in static type
-systems, but we are not concerned with \emph{assigning}
-types to terms, rather with showing that \emph{no type}
-can be assigned.
+systems, but our system would simply approximate its type
+as $\tfun$, apply it, and move along with evaluation.
 %
-We are simply executing the untyped $\lambda$-calculus,
-which has no issue handling recursion.
+Thus we do not need a built-in recursion construct, we
+can use any of the standard fixed-point operators.
+
+%  but we are not concerned with \emph{assigning}
+% types to terms, rather with showing that \emph{no type}
+% can be assigned.
+% %
+% We are simply executing the untyped $\lambda$-calculus,
+% which has no issue handling recursion.
 

### 2016-03-11 paper/icfp16-submission/semantics.tex
state lemma that resolve(_, s) ~ resolve(a, t)
sketch induction proof, need to fill in cases in inductive step

@@
 and
-$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$
+$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
 (\ie $\trace'$ is a single-step extension of $\trace$)
@@
 $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
-then the final step $\step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ invokes \forcesym.
+then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
+ successfully invokes \forcesym.
+%
+\ES{not clear we still need this}
 \end{lem}
@@
 then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
+%
+\ES{not clear we still need this}
 \end{lem}
@@
   % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
+\end{proof}
+
+\begin{lem}
+\label{lem:resolve-compat}
+For all traces
+$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
+$\vcompat{\resolve{\ehole}{\vsu}}{\resolve{\thole}{\tsu}}$.
+\end{lem}
+\begin{proof}
+  By induction on $\trace$.
+  %
+  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
+  and $\vhole{\thole}$ is trivially compatible with $\thole$.
+  %
+  In the inductive case, consider the single-step extension of $\trace$,
+  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
+  %
+  We show by case analysis on the evaluation rules that if
+  $\vcompat{\resolve{\ehole}{\vsu}}{\resolve{\thole}{\tsu}}$, then
+  $\vcompat{\resolve{\ehole}{\vsu'}}{\resolve{\thole}{\tsu'}}$.
+  %
+  \ES{fill in cases in the appendix}
 \end{proof}
@@
   does.
+  %
+  \ES{fill in cases in the appendix}
 \end{proof}
@@
   $\vincompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$.
-  \ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
   %
@@
   values differ,
-  $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \vcompat{\invctx{\ehole'[\thole']}}{t}$
+  $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{\ehole'[\thole']}$
   and
-  $\resolve{\vhole{\thole}}{\vsu_{i}} = \vincompat{\invctx{v'}}{t}$.
+  $\resolve{\vhole{\thole}}{\vsu_{i}} = \invctx{v'}$.
+  %
+  Furthermore, by Lemma~\ref{lem:resolve-compat} we know that
+  $$
+  \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
+  =
+  \tcompat{\ptype{i-1}{f}}{t}
+  $$
+  $$
+  \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
+  =
+  \tincompat{\ptype{i}{f}}{t}
+  $$
+  %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
   %

### 2016-03-11 paper/icfp16-submission/semantics.tex
new sketch of lemma k-stuck

@@
 $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
-$\vcompat{\resolve{\ehole}{\vsu}}{\resolve{\thole}{\tsu}}$.
+$\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$.
 \end{lem}
@@
   In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
-  and $\vhole{\thole}$ is trivially compatible with $\thole$.
+  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
   %
@@
   We show by case analysis on the evaluation rules that if
-  $\vcompat{\resolve{\ehole}{\vsu}}{\resolve{\thole}{\tsu}}$, then
-  $\vcompat{\resolve{\ehole}{\vsu'}}{\resolve{\thole}{\tsu'}}$.
+  $\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$, then
+  $\vsub{\resolve{\thole}{\tsu'}}{\resolve{\ehole}{\vsu'}}$.
   %
@@
 and $\trace'$ is a single-step extension of $\trace$
-and $\resolve{\vhole{\thole}}{\vsu} \neq \resolve{\vhole{\thole}}{\vsu'}$
-then $\vsu' = \vsu[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
+and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
+then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
 \end{lem}
@@
   %
-  $\vhole{\thole}$ does not change, so if the resolved values differ then
-  $\vsu \neq \vsu'$.
+  $\thole$ does not change, so if the resolved types differ then
+  $\tsu \neq \tsu'$.
   %
-  Note that only \forcesym can update \vsu\
-  and only the first case,
-  when we extend \vsu\ with $[\ehole \mapsto v]$,
-  does.
+  Note that only \forcesym can change \tsu, and it can only do so
+  via \unifysym, which can only extend \tsu.
+  % and only the first case,
+  % when we extend \vsu\ with $[\ehole \mapsto v]$,
+  % does.
   %
@@
   %
-  We will show that $\resolve{\vhole{\thole}}{\vsu_{i-1}}$
-  must contain some other hole $\ehole'$ which is instantiated with a
-  concrete value at step $i$.
+  We will show that $\resolve{\thole}{\tsu_{i-1}}$
+  must contain some other hole $\thole'$ that is
+  instantiated at step $i$.
   %
-  Furthermore, $\ehole'$ is instantiated in such a way that
-  $\vincompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$.
+  Furthermore, $\thole'$ is instantiated in such a way that
+  $\tincompat{\resolve{\thole}{\tsu_{i}}}{t}$.
   %
-  Finally, we will show that if we had instantiated $\ehole'$ with a
-  value such that
-  $\vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$, the current
+  Finally, we will show that if we had instantiated $\thole'$ such that
+  $\tcompat{\resolve{\thole}{\tsu_{i}}}{t}$, the current
   step would have gotten $\stuck$.
@@
   By Lemma~\ref{lem:vsu-ext} we know that
-  $\vsu_{i} = \vsu_{i-1}[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
+  $\tsu_{i} = \tsu_{i-1}[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
   %
   We will assume, without loss of generality, that
-  $\vsu_{i} = \vsu_{i-1}[\ehole' \mapsto v']$.
+  $\tsu_{i} = \tsu_{i-1}[\thole' \mapsto t']$.
 
-  Next, we introduce a notion of \emph{value contexts}, defined
+  Next, we introduce a notion of type % and value
+  contexts, defined
   analogously to evaluation contexts.
@@
   $$
-    V ::= \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V}
+  \begin{array}{lcl}
+    T &::=& \bullet \spmid \ttree{T} \\
+    % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
+  \end{array}
   $$
   %
-  Since $\vsu_{i-1}$ and $\vsu_{i}$ differ only in $\ehole'$ but the resolved
-  values differ,
-  $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{\ehole'[\thole']}$
+  Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
+  types differ,
+  $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
   and
-  $\resolve{\vhole{\thole}}{\vsu_{i}} = \invctx{v'}$.
+  $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
   %
-  Furthermore, by Lemma~\ref{lem:resolve-compat} we know that
-  $$
-  \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
-  =
-  \tcompat{\ptype{i-1}{f}}{t}
-  $$
-  $$
-  \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
-  =
-  \tincompat{\ptype{i}{f}}{t}
-  $$
+  Let $s$ be a concrete type such that $\intctx{s} = t$.
+  %
+  We show by case analysis on the evaluation rules that
+  %
+  $$\step{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}[\thole' \mapsto s]}{\stuck}{\vsu}{\tsu}$$
+  %
+  Finally, by Lemma~\ref{lem:resolve-compat} we know that
+  $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
+  thus $\vsub{\intctx{\thole'}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$.
+  %
+  Let $u = \gen{s}{\tsu}$
+  and $v = \resolve{\vhole{\thole}}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s}$,
+  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$ in $i$ steps.
+  %and
+  % $\vsub{\resolve{\thole}{\tsu_{i}}}{\resolve{\ehole}{\vsu_{i}}}$.
+  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
+  % =
+  % \tcompat{\ptype{i-1}{f}}{t}
+  % \\
+  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
+  % =
+  % \tincompat{\ptype{i}{f}}{t}
+  %
+  % Finally, by the definition of $\typeofsym$,
+  % we know that $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{v}$
+  % such that $\vsub{T'[\thole']}{v}$.
+  %such that $\resolve{\typeof{v}}{\tsu_{i-1}} = T'[\thole']$
+  %
   %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
   %
-  Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
-  we show by case analysis on the evaluation relation that
-  $\invctx{u}$ could not make progress at this step, and thus
-  $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
+  % Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
+  % we show by case analysis on the evaluation relation that
+  % $\invctx{u}$ could not make progress at this step, and thus
+  % $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
 \end{proof}

### 2016-03-12 paper/icfp16-submission/semantics.tex
add proof of lem:resolve-compat
clean up

@@
 We now show that our evaluation relation instantiates holes
-in a \emph{general} manner. That is, given a function $f$,
+in a \emph{general} manner. That is, given a lambda-term $f$,
 if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
@@
 \label{thm:soundness}
-  For any function $f$, if\\
+  For any lambda-term $f$, if
   \hbox{$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
@@
 
-\paragraph{The Type of a a Value} The \emph{dynamic type}
+\paragraph{The Type of a Value} The \emph{dynamic type}
 of a value $v$ is defined as a function $\typeof{v}$ shown
@@
 %
-A \emph{type} $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t},
+A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t},
 if $\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.
@@
 %
-A \emph{value} $v$ is \emph{compatible} with a type $t$, written \vcompat{v}{t},
+A value $v$ is \emph{compatible} with a type $t$, written \vcompat{v}{t},
 if $\tcompat{\typeof{v}}{t}$, that is, if the dynamic type of $v$ is
@@
 
+\paragraph{Type Refinement}
+A type $s$ is a \emph{refinement} of a type $t$, written $\tsub{s}{t}$,
+if $\exists \theta. s = \subst{\theta}{t}$.
+%
+In other words, $s$ is a refinement of $t$ if there exists a type
+substitution that maps $t$ directly to $s$.
+%
+A type $t$ is a \emph{refinement} of a value $v$, written $\tsub{t}{v}$,
+if $\tsub{t}{\typeof{v}}$, \ie if $t$ is a refinement of the
+dynamic type of $v$.
+%
+\ES{calling a type a refinement of a value sounds really weird..}
+
+We prove that the partial instantiation of a type hole $\thole$
+is always a refinement of the partial instantiation of the
+associated value hole $\ehole$.
+
 \begin{lem}
@@
   %
-  \ES{fill in cases in the appendix}
 \end{proof}
@@
 and $\trace'$ is a single-step extension of $\trace$
-and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
+and $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$
+% and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
 then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
@@
   %
-  $\thole$ does not change, so if the resolved types differ then
+  $\thole$ does not change, so if the partial input types differ then
   $\tsu \neq \tsu'$.
@@
 \label{lem:k-stuck}
-For all $t$,
+For all types $t$,
 if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
@@
   %
-  We will show that $\resolve{\thole}{\tsu_{i-1}}$
+  We will show that $\ptype{\trace_{i-1}}{f}$ % $\resolve{\thole}{\tsu_{i-1}}$
   must contain some other hole $\thole'$ that is
@@
   Furthermore, $\thole'$ is instantiated in such a way that
-  $\tincompat{\resolve{\thole}{\tsu_{i}}}{t}$.
+  $\tincompat{\ptype{\trace_i}{f}}{t}$.
+  % $\tincompat{\resolve{\thole}{\tsu_{i}}}{t}$.
   %
   Finally, we will show that if we had instantiated $\thole'$ such that
-  $\tcompat{\resolve{\thole}{\tsu_{i}}}{t}$, the current
-  step would have gotten $\stuck$.
+  $\tcompat{\ptype{\trace_i}{f}}{t}$,
+  % $\tcompat{\resolve{\thole}{\tsu_{i}}}{t}$,
+  the current step would have gotten $\stuck$.
 
@@
   $\tsu_{i} = \tsu_{i-1}[\thole' \mapsto t']$.
-
+  %
   Next, we introduce a notion of type % and value
@@
   types differ,
-  $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
+  $\ptype{\trace_{i-1}}{f} = \intctx{\thole'}$
   and
-  $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
+  $\ptype{\trace_i}{f} = \intctx{t'}$.
+  % $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
+  % and
+  % $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
   %
@@
   Finally, by Lemma~\ref{lem:resolve-compat} we know that
-  $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
-  thus $\vsub{\intctx{\thole'}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$.
+  % $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
+  $\vsub{\ptype{\trace_{i-1}}{f}}{\resolve{\ehole}{\vsu_{i-1}}}$,
+  thus $\vsub{\intctx{\thole'}}{\resolve{\ehole}{\vsu_{i-1}}}$.
   %
   Let $u = \gen{s}{\tsu}$
-  and $v = \resolve{\vhole{\thole}}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s}$,
+  and $v = \resolve{\ehole}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s}$,
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$ in $i$ steps.

### 2016-03-14 paper/icfp16-submission/semantics.tex
add context compatibility lemma

@@
 
+Next, we introduce a notion of type % and value
+contexts, defined analogously to evaluation contexts.
+%
+$$
+\begin{array}{lcl}
+  T &::=& \bullet \spmid \ttree{T} \\
+  % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
+\end{array}
+$$
+%
+
+\begin{lem}
+\label{lem:context-compat}
+For all types $s$ and $t$,
+if $\tcompat{\intctx{s}}{t}$
+then $t = \intctx{s'}$
+such that $\tcompat{s}{s'}$.
+\end{lem}
+\begin{proof}
+  By induction on $t$.
+\end{proof}
+
 \paragraph{Incompatible Types Are Wrong}
@@
   %
-  Next, we introduce a notion of type % and value
-  contexts, defined
-  analogously to evaluation contexts.
-  %
-  $$
-  \begin{array}{lcl}
-    T &::=& \bullet \spmid \ttree{T} \\
-    % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
-  \end{array}
-  $$
-  %
   Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
@@
   %
-  Let $s$ be a concrete type such that $\intctx{s} = t$.
+  By Lemma~\ref{lem:context-compat},
+  let $s$ be a concrete type such that $\intctx{s} = t$.
   %
@@
 Suppose $\trace$ witnesses that $f$ gets stuck,
-and let $t = \ptype{\trace}{f}$.
-We show that \emph{all} types $s$ have stuck-inducing
-values by splitting cases on whether the type is
-compatible with $t$. %the partial type upto $\trace$.
+and let $s = \ptype{\trace}{f}$.
+We show that \emph{all} types $t$ have stuck-inducing
+values by splitting cases on whether $t$ is
+compatible with $s$. %the partial type upto $\trace$.
 %
@@
 \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, we can derive
-  a $v$ from $\trace$ such that \hastype{v}{t} and $\eapp{f}{v}$ gets
+  a $v$ from $\trace$ such that $\hastype{v}{t}$ and $\eapp{f}{v}$ gets
   stuck.

### 2016-03-14 paper/icfp16-submission/semantics.tex
kill the resolve business

@@
 are normalized.
-
-\paragraph{Resolving Holes}
-We \emph{resolve} a hole with respect to a substitution by
-transitively applying the substitution as long as it contains
-any holes that are defined in the substitution.
-%
-We write \resolve{\thole}{\tsu} to denote the the resolution
-of \thole\ with respect to \tsu.
-%
-Note that by definition, \resolve{\thole}{\tsu} does not contain
-any holes in the domain of \tsu.
+%
+We ensure additionally that the co-domain of a substitution does not
+refer to any elements of its domain, \ie when we extend a substitution
+with a new binding we apply the substitution to itself.
+
+% \paragraph{Resolving Holes}
+% We \emph{resolve} a hole with respect to a substitution by
+% transitively applying the substitution as long as it contains
+% any holes that are defined in the substitution.
+% %
+% We write \resolve{\thole}{\tsu} to denote the the resolution
+% of \thole\ with respect to \tsu.
+% %
+% Note that by definition, \resolve{\thole}{\tsu} does not contain
+% any holes in the domain of \tsu.
 

### 2016-03-14 paper/icfp16-submission/semantics.tex
add pairs. need to update proofs

@@
 Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
-calculus with integers, booleans, and binary trees.
+calculus with integers, booleans, pairs, and binary trees.
 %
@@
 \force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
+\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq & \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''}, \mbox{if}\ \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu}, \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'}  \\
 \force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
@@
 \gensym       & :   & t \times \tsu \rightarrow v \\
-\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
-\gen{\tint}{\tsu}   & \defeq  & n, & \quad \text{non-deterministic} \\
-\gen{\tbool}{\tsu}  & \defeq  & b, & \quad \text{non-deterministic} \\
-\gen{\ttree{t}}{\tsu}  & \defeq  & tr, & \quad \text{non-deterministic} \\
-\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, & \quad \text{\ehole and \thole are fresh} \\
-\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
+\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, &  \text{if } \thole \in dom(\tsu) \\
+\gen{\tint}{\tsu}   & \defeq  & n, &  \text{non-deterministic} \\
+\gen{\tbool}{\tsu}  & \defeq  & b, &  \text{non-deterministic} \\
+\gen{\tprod{t_1}{t_2}}{\tsu}  & \defeq  & \epair{\gen{t_1}{\tsu}}{\gen{t_2}{\tsu}} & \\ % \begin{subarray}{l} v_1 = \gen{t_1}{\tsu}\\ v_2 = \gen{t_2}{\tsu}\end{subarray} \\
+\gen{\ttree{t}}{\tsu}  & \defeq  & tr, &  \text{non-deterministic} \\
+\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, &  \text{\ehole and \thole are fresh} \\
+\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \text{\ehole is fresh} \\
 \end{array}
@@
     \typeof{\efun{x}{e}} & \defeq & \tfun \\
+    \typeof{\epair{v_1}{v_2}} & \defeq & \tprod{\typeof{v_1}}{\typeof{v_2}} \\
     \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
@@
 \end{cor}
+\begin{proof}
+  For any function $f$ of multiple arguments, we can define $f'$ to be
+  the uncurried version of $f$ that takes all of its arguments as a
+  nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.
+\end{proof}
 

### 2016-03-14 paper/icfp16-submission/semantics.tex
update proofs
comment out unused lemmas

@@
 \begin{cases}
-  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
+  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\vhole{\thole}},
                                          \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
-  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
-  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
+  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\vhole{\thole}} \\
+  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
 \end{cases} \\
@@
 if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
-then \emph{for every} input type $t$, we can find a value
+then \emph{for every} concrete type $t$, we can find a value
 $v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.
@@
 %
-We can extract this type from an evaluation trace, by
-\emph{resolving} the \thole\ with the final type
-substitution at the end of the trace.
+We can extract this type from an evaluation trace by
+applying the final type substitution to \thole.
+% \emph{resolving} the \thole\ with the final type
+% substitution at the end of the trace.
 %
@@
 $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
-then the \emph{partial input type} of $f$ upto $\trace$, written
+then the \emph{partial input type} of $f$ up to $\trace$, written
 \ptype{\trace}{f}, is $\resolve{\thole}{\tsu}$.
@@
 
-\paragraph{Narrowing}
-%
-Only a successful call to \forcesym can change the partial
-input type of $f$.
-%
-\begin{lem}
-\label{lem:force-inst}
-If
-$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
-and
-$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
-(\ie $\trace'$ is a single-step extension of $\trace$)
-and
-$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
-then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
- successfully invokes \forcesym.
-%
-\ES{not clear we still need this}
-\end{lem}
-
-\begin{proof}
-  By case analysis on the evaluation rules.
-  %
-  If $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$ then,
-  % one of the holes in $f$'s
-  % argument must have been instantiated with a concrete value at the last step.
-  by the definition of $\ptype{\trace}{f}$, $\tsu \neq \tsu'$, as \thole
-  does not change.
-  %
-  % An examination of the rules shows that only place this happens is
-  % in the second case of \forcesym.
-  An examination of the rules shows that only \forcesym can update \tsu,
-  and furthermore that only the successful cases of \forcesym do update
-  \tsu.
-\end{proof}
+% \paragraph{Narrowing}
+% %
+% Only a successful call to \forcesym can change the partial
+% input type of $f$.
+% %
+% \begin{lem}
+% \label{lem:force-inst}
+% If
+% $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
+% and
+% $\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
+% (\ie $\trace'$ is a single-step extension of $\trace$)
+% and
+% $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
+% then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
+%  successfully invokes \forcesym.
+% %
+% \ES{not clear we still need this}
+% \end{lem}
+
+% \begin{proof}
+%   By case analysis on the evaluation rules.
+%   %
+%   If $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$ then,
+%   % one of the holes in $f$'s
+%   % argument must have been instantiated with a concrete value at the last step.
+%   by the definition of $\ptype{\trace}{f}$, $\tsu \neq \tsu'$, as \thole
+%   does not change.
+%   %
+%   % An examination of the rules shows that only place this happens is
+%   % in the second case of \forcesym.
+%   An examination of the rules shows that only \forcesym can update \tsu,
+%   and furthermore that only the successful cases of \forcesym do update
+%   \tsu.
+% \end{proof}
 
@@
 
-\paragraph{Preservation}
-We prove that each evaluation step \emph{refines} the partial input type
-of $f$, \ie\ preserves type compatibility.
-%
-\begin{lem}
-\label{lem:refine-partial}
-If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
-$\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
-%
-%The partial type of $f$ upto $\trace$ is compatible
-%with the partial type upto $\trace'$, \ie\
-%
-then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
-%
-\ES{not clear we still need this}
-\end{lem}
-\begin{proof}
-  By case analysis on the evaluation rules.
-  %
-  First note that by Lemma~\ref{lem:force-inst} we can immediately
-  discharge the \rulename{E-*-Bad} rules as they cannot change
-  \ptype{\trace}{f} at all, and are thus trivially
-  compatibility-preserving. For the \rulename{E-*-Good} rules we can
-  show that, by virtue of \forcesym succeeding, all must preseve
-  compatibility.
-  % Note that all rules preserve partial types with the exception of when
-  % \forcesym\ is called on a hole, in which case we may instantiate the hole with
-  % a concrete value.
-  % %
-  % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
-\end{proof}
+% \paragraph{Preservation}
+% We prove that each evaluation step \emph{refines} the partial input type
+% of $f$, \ie\ preserves type compatibility.
+% %
+% \begin{lem}
+% \label{lem:refine-partial}
+% If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
+% $\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
+% %
+% %The partial type of $f$ upto $\trace$ is compatible
+% %with the partial type upto $\trace'$, \ie\
+% %
+% then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
+% %
+% \ES{not clear we still need this}
+% \end{lem}
+% \begin{proof}
+%   By case analysis on the evaluation rules.
+%   %
+%   First note that by Lemma~\ref{lem:force-inst} we can immediately
+%   discharge the \rulename{E-*-Bad} rules as they cannot change
+%   \ptype{\trace}{f} at all, and are thus trivially
+%   compatibility-preserving. For the \rulename{E-*-Good} rules we can
+%   show that, by virtue of \forcesym succeeding, all must preseve
+%   compatibility.
+%   % Note that all rules preserve partial types with the exception of when
+%   % \forcesym\ is called on a hole, in which case we may instantiate the hole with
+%   % a concrete value.
+%   % %
+%   % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
+% \end{proof}
 
@@
 is always a refinement of the partial instantiation of the
-associated value hole $\ehole$.
-
-\begin{lem}
-\label{lem:resolve-compat}
-For all traces
-$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
-$\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$.
-\end{lem}
-\begin{proof}
-  By induction on $\trace$.
-  %
-  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
-  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
-  %
-  In the inductive case, consider the single-step extension of $\trace$,
-  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
-  %
-  We show by case analysis on the evaluation rules that if
-  $\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$, then
-  $\vsub{\resolve{\thole}{\tsu'}}{\resolve{\ehole}{\vsu'}}$.
-  %
-\end{proof}
-
+associated value hole $\vhole{\thole}$.
+
+\paragraph{Preservation}
+%
+We prove two preservation lemmas. First, we show that each evaluation
+step refines the partial input type of $f$, thus preserving type
+compatibility.
+%
 \begin{lem}
@@
   %
-  \ES{fill in cases in the appendix}
+  % \ES{fill in cases in the appendix}
 \end{proof}
-
+%
+Second, we show that at each step of evaluation, the partial input type of $f$
+is a refinement of the instantiation of $\vhole{\thole}$.
+%
+\begin{lem}
+\label{lem:resolve-compat}
+For all traces
+$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
+$\vsub{\ptype{\trace}{f}}{\resolve{\vhole{\thole}}{\vsu}}$.
+\end{lem}
+\begin{proof}
+  By induction on $\trace$.
+  %
+  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
+  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
+  %
+  In the inductive case, consider the single-step extension of $\trace$,
+  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
+  %
+  We show by case analysis on the evaluation rules that if
+  $\vsub{\ptype{\trace}{f}}{\resolve{\vhole{\thole}}{\vsu}}$, then
+  $\vsub{\ptype{\trace'}{f}}{\resolve{\vhole{\thole}}{\vsu'}}$.
+  %
+\end{proof}
+
+\ES{Where does this fit in??}
 Next, we introduce a notion of type % and value
@@
 \begin{array}{lcl}
-  T &::=& \bullet \spmid \ttree{T} \\
+  T &::=& \bullet \spmid \ttree{T} \spmid \tprod{T}{t} \spmid \tprod{t}{T} \\
   % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
@@
   $$\step{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}[\thole' \mapsto s]}{\stuck}{\vsu}{\tsu}$$
-  %
+
   Finally, by Lemma~\ref{lem:resolve-compat} we know that
   % $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
-  $\vsub{\ptype{\trace_{i-1}}{f}}{\resolve{\ehole}{\vsu_{i-1}}}$,
-  thus $\vsub{\intctx{\thole'}}{\resolve{\ehole}{\vsu_{i-1}}}$.
-  %
-  Let $u = \gen{s}{\tsu}$
-  and $v = \resolve{\ehole}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s}$,
+  $\vsub{\ptype{\trace_{i-1}}{f}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$
+  and thus $\vsub{\intctx{\thole'}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$.
+  %
+  Let
+  $$
+  \begin{array}{lcl}
+  u &=& \gen{s}{\tsu} \\
+  v &=& \resolve{\vhole{\thole}}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s} \\
+  \end{array}
+  $$
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$ in $i$ steps.
@@
   %
-  The value $v = \resolve{\ehole}{\vsu}$ demonstrates that
+  The value $v = \resolve{\vhole{\thole}}{\vsu}$ demonstrates that
   $\eapp{f}{v}$ gets stuck.

### 2016-03-14 paper/icfp16-submission/introduction.tex
emphasize jump-compression in intro

@@
 \end{code}
-\end{minipage}
-% \hspace{0.3in}
-\begin{minipage}{.49\linewidth}
+\vspace{3em}
+% \end{minipage}
+% % \hspace{0.3in}
+% \begin{minipage}{.39\linewidth}
 \centering
@@
 \end{minipage}
-\caption{(L) An ill-typed \texttt{fac} function; (R) Dynamically witnessing the type error in \texttt{fac}.}
+\begin{minipage}{.49\linewidth}
+\centering
+\includegraphics[height=3in]{fac-long.png}
+\end{minipage}
+\caption{(TL) An ill-typed \texttt{fac} function; (BL) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (R) The same trace, fully expanded.}
 \label{fig:factorial}
@@
 the source of the ill-typed function to automatically synthesize
-the result on the right in Figure~\ref{fig:factorial}, which
+the result on the bottom-left in Figure~\ref{fig:factorial}, which
 shows how the recursive calls reduce to a configuration where
@@
 We solve this problem by developing a novel operational semantics
-that corresponds to executing the program with \emph{holes} ---
-values whose type is unknown.
+that combines evaluation and type inference.
+%
+We execute the program with \emph{holes} --- values whose type is
+unknown --- as the inputs.
+% that corresponds to executing the program with \emph{holes} ---
+% values whose type is unknown.
+%
+A hole remains abstract until the evaluation context tells us what type
+it must have, for example the parameters to an addition operation must
+both be integers.
 %
@@
 steps that allow the user to keep track of a term's provenance.
+
+We introduce a notion of \emph{jump-compressed} traces to abstract away
+the irrelevant details of a computation.
+%
+A jump-compressed trace includes only big steps to the next function
+call or return, for example the trace in the bottom-left of
+Figure~\ref{fig:factorial} is a jump-compressed trace.
+%
+Jump-compressed traces are similar to stack traces in that both show a
+sequence of function calls that lead to a crash, but the jump-compressed
+trace also shows the return values of successful calls, which can be
+useful in understanding why a particular path was taken.
 

### 2016-03-14 paper/icfp16-submission/evaluation.tex
move "jump compression" to interactive-short
move lem:context-compat to the appendix

@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
 \centering
 \begin{minipage}{\linewidth}
-\includegraphics[width=\linewidth]{coverage.png}
-\end{minipage}
-\begin{minipage}{\linewidth}
-\includegraphics[width=\linewidth]{distrib.png}
+\centering
+\includegraphics[width=0.49\linewidth]{coverage.png}
+% \end{minipage}
+% \begin{minipage}{\linewidth}
+\includegraphics[width=0.49\linewidth]{distrib.png}
 \end{minipage}
@@
 \label{fig:results-witness}
-\end{figure}
+\end{figure*}
 
@@
 
-\paragraph{Jump Compression}
-%
-A \emph{jump compressed} trace is one whose edges are limited to forward
-or backward jumps. (The sizes of these traces is equal to the jump
-metric below.)
-%
-In our experience, jump compression abstracts many details of the
-computation that are often uninteresting or irrelevant to the
-explanation.
-% 
-In particular, jump compressed traces hide low-level operations and
-summarize function calls as call-return pairs, see
-Figure~\ref{fig:fac-jump} for a variant of @fac@ that implements the
-subtraction as a function call instead of a primitive.
-%
-\begin{figure}[t]
-\centering
-\includegraphics[height=2in]{fac-minus.png}
-\caption{Jump-compressed trace of \texttt{fac 1} with subtraction
-  implemented as a function call.}
-\label{fig:fac-jump}
-\end{figure}
-%
-Once the user has identified an interesting call-return pair, they can
-step into the call and proceed with more fine-grained steps.
-%
-% Expanding the initial visualization to a jump-compressed trace is often
-% an ideal first step;
-% % expanding a jump-compressed traces ideal for the initial state of the
-% they provide a concise overview of the execution and allow the user to
-% step directly into an interesting function call, at which point the can
-% proceed with more fine-grained steps.
-%
-Note that jump compressed traces are not quite the same as
-stack-traces as they show \emph{all} function calls, including
-those that returned successfully.
-%
-For example, the display on the right in Figure~\ref{fig:factorial} shows
-a jump-compressed trace, of size 4. The full trace for
-the same example has size 20.
-
-\paragraph{Metrics} Thus, our two metrics are:
+% \paragraph{Metrics} Thus, our two metrics are:
 % size of the full trace,
@@
 \begin{enumerate}
-\item \emphbf{Single-Step Metric} The size of the trace after expanding
+\item \emphbf{Single-step Metric} The size of the trace after expanding
   all of the single-step edges from the witness to the stuck term, and
@@
   % complexity, \ie ``How big is the fully-expanded trace?''
-\item \emphbf{Jump Metric} The size of the trace for after expanding
-  only the forward (or backward) jumps from the witness to the stuck
-  term.
+\item \emphbf{Jump-compressed Metric} The size of the jump-compressed trace.
 \end{enumerate}

### 2016-03-14 paper/icfp16-submission/gallery2.tex
fixed clobbering
move "jump compression" to interactive-short
move lem:context-compat to the appendix

@@
 (2)~the error message returned \ocaml,
-(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),
+(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold)
 \footnote{When the locations from \ocaml\ and \sherrloc\ overlap,
-we just underline the relevant code.}
+we just underline the relevant code.},
 (4)~the jump-compressed trace produced by \toolname.
@@
 %
-Unfortunately, it incorrectly returns @[]@
-in the base case instead of @0@.
+Unfortunately, in the base case, it returns @[]@
+instead of @0@.
 %

### 2016-03-14 paper/icfp16-submission/interactive-short.tex
fixed genWitness
move "jump compression" to interactive-short
move lem:context-compat to the appendix

@@
 
+\paragraph{Jump Compression}
+%
+A \emph{jump compressed} trace is one whose edges are limited to forward
+or backward jumps.
+% (The sizes of these traces is equal to the jump
+% metric below.)
+%
+In our experience, jump compression abstracts many details of the
+computation that are often uninteresting or irrelevant to the
+explanation.
+%
+In particular, jump compressed traces hide low-level operations and
+summarize function calls as call-return pairs, see
+Figure~\ref{fig:fac-jump} for a variant of @fac@ that implements the
+subtraction as a function call instead of a primitive.
+%
+\begin{figure}[t]
+\centering
+\includegraphics[height=2in]{fac-minus.png}
+\caption{Jump-compressed trace of \texttt{fac 1} with subtraction
+  implemented as a function call.}
+\label{fig:fac-jump}
+\end{figure}
+%
+Once the user has identified an interesting call-return pair, they can
+step into the call and proceed with more fine-grained steps.
+%
+% Expanding the initial visualization to a jump-compressed trace is often
+% an ideal first step;
+% % expanding a jump-compressed traces ideal for the initial state of the
+% they provide a concise overview of the execution and allow the user to
+% step directly into an interesting function call, at which point the can
+% proceed with more fine-grained steps.
+%
+Note that jump compressed traces are not quite the same as
+stack-traces as they show \emph{all} function calls, including
+those that returned successfully.
+%
+For example, the display on the bottom-left in Figure~\ref{fig:factorial} shows
+a jump-compressed trace, of size 4. The full trace for
+the same example (on the right of the Figure~\ref{fig:factorial}) has size 19.
+
+% \RJ{add para about jump-compressed traces + example. include full trace and compressed trace.
+% could merge with example in overview.}
+
 % \input{traversals_new.tex}

### 2016-03-14 paper/icfp16-submission/overview.tex
move "jump compression" to interactive-short
move lem:context-compat to the appendix

--- 2016-03-07 paper/icfp16-submission/overview.tex
@@
 A hole --- written \vhole{\thole} --- is a \emph{placeholder} for a
-value of some unknown type \thole.
+value \ehole of some unknown type \thole.
 %
@@
 %
-Having identified an input vector that triggers a bug, a common next
+Having identified an input vector that triggers a crash, a common next
 step is to step through the program with a \emph{debugger} to observe
@@
 visualization.
+
+\paragraph{Jump-compressed Witnesses}
+It is rare for the initial state of the visualization to be
+informative enough to diagnose the error.
+%
+We provide a short-cut to expand the witness to a \emph{jump-compressed}
+trace, which contains every function call and return step.
+%
+The jump-compressed trace abstracts the computation as a sequence of
+call-response pairs, providing a high-level overview of steps taken
+to reach the crash, and a high level of compression compared to the
+full trace.
+%
+For example, the jump-compressed trace in Figure~\ref{fig:factorial}
+contains 4 nodes compared to the 19 in the fully expanded trace.
+%
+Our benchmark suite of student programs shows that jump-compression is
+practical, with an average trace size of 7 nodes and a median of 5.
 

### 2016-03-14 paper/icfp16-submission/semantics.tex
move "jump compression" to interactive-short
move lem:context-compat to the appendix

@@
 witnesses for ill-typed programs~\S~(\ref{sec:search-algorithm}).
+%
+We have formalized and tested our semantics and generality theorem
+in \textsc{PLT-Redex}~\cite{felleisen2009semantics}.
 
@@
 
-\ES{Where does this fit in??}
-Next, we introduce a notion of type % and value
-contexts, defined analogously to evaluation contexts.
-%
-$$
-\begin{array}{lcl}
-  T &::=& \bullet \spmid \ttree{T} \spmid \tprod{T}{t} \spmid \tprod{t}{T} \\
-  % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
-\end{array}
-$$
-%
-
-\begin{lem}
-\label{lem:context-compat}
-For all types $s$ and $t$,
-if $\tcompat{\intctx{s}}{t}$
-then $t = \intctx{s'}$
-such that $\tcompat{s}{s'}$.
-\end{lem}
-\begin{proof}
-  By induction on $t$.
-\end{proof}
-
 \paragraph{Incompatible Types Are Wrong}
@@
   the current step would have gotten $\stuck$.
+
+  % Next, we introduce a notion of type % and value
+  % contexts, defined analogously to evaluation contexts.
+  % %
+  % $$
+  % \begin{array}{lcl}
+  %   T &::=& \bullet \spmid \ttree{T} \spmid \tprod{T}{t} \spmid \tprod{t}{T} \\
+  %   % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
+  % \end{array}
+  % $$
 

### 2016-03-14 paper/icfp16-submission/gallery2.tex
misc edits everywhere

@@
 This faulty definition of @wwhile@ still typechecks however,
-and hence, are ``assumed'' to be correct by and both \ocaml\
-and \sherrloc\ which blame the \emph{use-site} on line 10.
+and is assumed to be correct by and both \ocaml\
+and \sherrloc\ which blame the use-site on line 10.
 %

### 2016-03-14 paper/icfp16-submission/overview.tex
misc edits everywhere

@@
 %
-We provide a short-cut to expand the witness to a \emph{jump-compressed}
-trace, which contains every function call and return step.
+Rather than abandon the user, we provide a short-cut to expand the witness
+to a \emph{jump-compressed} trace, which contains every function call
+and return step.
 %
@@
 Our benchmark suite of student programs shows that jump-compression is
-practical, with an average trace size of 7 nodes and a median of 5.
+practical, with an average jump-compressed trace size of 7 nodes and a
+median of 5.
 

### 2016-03-14 paper/icfp16-submission/semantics.tex
misc edits everywhere

@@
 program values. For example, addition tells us that the addends must be
-integers, and case-analysis tells us that the scrutinee must be a tree.
+integers, and % case-analysis tells us that the scrutinee must be a tree.
+an if-expression tells us the condition must be a boolean.
 %
@@
 \paragraph{Generating Values} The (non-deterministic)
-$\gen{t}{\tsu}$ in Figure~\ref{fig:gen}, takes
+$\gen{t}{\tsu}$ in Figure~\ref{fig:gen} takes
 as input a type $t$ and returns a value of that type.
 %
-For base types, the procedure returns an arbitrary value of
+For base types the procedure returns an arbitrary value of
 that type.
 %
-For functions, it returns a lambda with a \emph{new} hole
-denoting the return value, and unconstrained types (denoted
-by $\thole$) yield fresh holes constrained to have type
+For functions it returns a lambda with a \emph{new} hole
+denoting the return value.
+%
+For unconstrained types (denoted
+by $\thole$) it yields a fresh hole constrained to have type
 \thole (denoted by $\vhole{\thole}$).
@@
 For a polymorphic type $\ttree{\thole}$ %, eg in \recasegoodone,
-we will place holes in the generated nodes; they will be lazily filled
+or $\tprod{\thole_1}{\thole_2}$
+we will place holes in the generated value; they will be lazily filled
 in later, on demand.
@@
 instead there exist perfectly good values of a \emph{different}
-type, under which the program would not have gone wrong.
+type under which the program would not have gone wrong.
 %
@@
   % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
-  $t$ there exists a $v$ of type $t$ such that
-  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
+  $t$ there exists a value $v$ of type $t$ such that
+  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
 \end{thm}
@@
 First, we show how our evaluation rules encode a dynamic form of
-type inference, and next, we show that the types inferred via
+type inference, and then we show that the witnesses found by
 evaluation are indeed maximally general.
@@
 The types of functions are \emph{approximated}, which is all
-that is needed to ensure an application does not get stuck, \eg
+that is needed to ensure an application does not get stuck.
+%
+For example,
 $$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
@@
 %
-The types of (polymorphic) data types are obtained from the
-labels on their values.
+The types of (polymorphic) trees are obtained from the labels on their
+values, and the types of tuples directly from their values.
 
@@
 We can think of the evaluation of \eapp{f}{\vhole{\thole}}
-as synthesizing a partial instantiation of \thole, and hence,
+as synthesizing a partial instantiation of \thole, and thus
 \emph{dynamically inferring} a (partial) type for $f$'s input.
@@
 
-We prove that the partial instantiation of a type hole $\thole$
-is always a refinement of the partial instantiation of the
-associated value hole $\vhole{\thole}$.
+% We prove that the partial instantiation of a type hole $\thole$
+% is always a refinement of the partial instantiation of the
+% associated value hole $\vhole{\thole}$.
 
@@
 \emph{For all} types that are \emph{incompatible} with the
-partial input type up to trace $\trace$, there exists a value
+partial input type up to $\trace$, there exists a value
 that will cause $f$ to get stuck in \emph{at most} $k$ steps,
@@
 a general witness demonstrating that the program is ill-typed (\ie\ goes
-wrong for at least one input of every type.)
+wrong for at least one input of every type).
 In particular, we have shown how to non-deterministically find a witnesses
@@
 The \emph{empty} list is returned when no witness can be found after
-exploring $n$ traces.
+exploring $k$ traces.
 
@@
 Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
-implementation of our search for witnesses, in the procedure
+implementation of our search for witnesses with the procedure
 $\genWitness{k}{e}$, which takes as input a bound $k$ and the
@@
 \label{thm:generation}
-  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$$
+  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$$
   then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
@@
   For any function $f$ of multiple arguments, we can define $f'$ to be
-  the uncurried version of $f$ that takes all of its arguments as a
+  the uncurried version of $f$ that takes all of its arguments as a single
   nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.

### 2016-03-15 paper/icfp16-submission/evaluation.tex
misc edits everywhere
misc edits

@@
 We have implemented a prototype of our search procedure and trace
-visualization for the pure subset of \ocaml, in a tool called \nanomaly.
+visualization for the purely functional subset of \ocaml, in a tool called \nanomaly.
 In our implementation we instantiated \gensym\ with a simple random
@@
 %
-In both datasets at least 80\% of the traces contain at most 10 jumps.
+In both datasets 80\% or more traces have at most 10 jumps.
 %
@@
 there are some degenerate cases where it is unable to detect
-them. Consider, for example, the following buggy version of @replicate@.
-% ,
-% which is supposed to produce a list containing @n@ copies of @x@.
+them. Consider, the following buggy @replicate@
 %
@@
   let rec replicate n x =
-    if n <= 0 then
-      []
-    else
-      replicate (n-1) [x]
+    if n <= 0 then []
+              else replicate (n-1) [x]
 \end{code}
@@
 program. Enhancing our visualization to slice out program paths relevant
-to specific values, \eg\ in the style of~\cite{perera_functional_2012},
+to specific values~\cite{perera_functional_2012},
 would likely help alleviate this issue, allowing users to

### 2016-03-15 paper/icfp16-submission/semantics.tex
misc edits

@@
   Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
-  types differ,
+  types differ, we have, for some type context $T$,
   $\ptype{\trace_{i-1}}{f} = \intctx{\thole'}$
@@
   $\ptype{\trace_i}{f} = \intctx{t'}$.
-  % $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
-  % and
-  % $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
-  %
-  By Lemma~\ref{lem:context-compat},
-  let $s$ be a concrete type such that $\intctx{s} = t$.
+  %
+  We prove, in Lemma~\ref{lem:context-compat},
+  that \RJ{rephrase lamma here}.
+  Thus, by Lemma~\ref{lem:context-compat}, let $s$ be a
+  concrete type such that $\intctx{s} = t$.
   %

### 2016-03-15 paper/icfp16-submission/semantics.tex
remove type contexts

--- 2016-03-14 paper/icfp16-submission/semantics.tex
@@
   Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
-  types differ,
-  $\ptype{\trace_{i-1}}{f} = \intctx{\thole'}$
+  types differ, we have
+  $\thole' \in \ptype{\trace_{i-1}}{f}$
   and
-  $\ptype{\trace_i}{f} = \intctx{t'}$.
-  % $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
-  % and
-  % $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
-  %
-  By Lemma~\ref{lem:context-compat},
-  let $s$ be a concrete type such that $\intctx{s} = t$.
+  $\ptype{\trace_i}{f} = \ptype{\trace_{i-1}}{f}\sub{\thole'}{t'}$.
+  %
+  % We prove\includeTechReport{, in Lemma~\ref{lem:context-compat},}
+  % that for all types $s$ and $t$, if.
+  Let $s$ be a
+  concrete type such that $\ptype{\trace_{i-1}}{f}\sub{\thole'}{s} = t$.
   %
@@
   $\vsub{\ptype{\trace_{i-1}}{f}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$
-  and thus $\vsub{\intctx{\thole'}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$.
+  and thus $\thole' \in \resolve{\vhole{\thole}}{\vsu_{i-1}}$.
+  %and thus $\vsub{\intctx{\thole'}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$.
   %

### 2016-03-15 paper/icfp16-submission/evaluation.tex
address various minor reviewer comments

@@
 from which we extracted 284 ill-typed programs.
+%
+Both datasets include programs that demonstrate a variety of functional
+programming idioms, including (tail) recursive functions, higher-order
+functions, polymorphic data types, and expression evaluators.
 
@@
 Interestingly, while the vast majority of witnesses corresponded to a
-type-error, as expected, 3-4\% triggered an unbound variable error and
-2-3\% triggered an infinite recursion error.
+type-error, as expected, 3-4\% triggered an unbound variable error (even
+though \ocaml reported a type error) and 2-3\% triggered an infinite
+recursion error.
 %

### 2016-03-15 paper/icfp16-submission/introduction.tex
address various minor reviewer comments

@@
 Our second contribution is an interactive visualization of the
-execution of \ocaml\ programs, well-typed or not~(\S~\ref{sec:interactive}).
+execution of purely functional \ocaml\ programs, well-typed or not~(\S~\ref{sec:interactive}).
 %

### 2016-03-15 paper/icfp16-submission/overview.tex
address various minor reviewer comments

@@
 %
-At this point we choose a random @int@ for the instantiation and
+At this point we choose a
+random @int@\footnote{With standard heuristics to favor small values.}
+for the instantiation and
 concrete execution takes over entirely, leading us to the expected crash
@@
 %
-There is another wrinkle though, how did we know
+There is another wrinkle, though; how did we know
 that @fac@ takes a single argument instead of two

### 2016-03-16 paper/icfp16-submission/evaluation.tex
Made a pass over the complete paper, clarifying and breaking up some convoluted sentences, and removing references to "blind review" for ICFP.

@@
 We recorded each interaction with the \ocaml\ top-level system over the
-course of the first three assignments (IRB \# hidden for blind review), % \#140608),
+course of the first three assignments (IRB  
+% \# hidden for blind review), 
+\#140608),
 from which we extracted \ucsdsize\ distinct, ill-typed \ocaml programs.

### 2016-03-16 paper/icfp16-submission/gallery2.tex
misc edits
Made a pass over the complete paper, clarifying and breaking up some convoluted sentences, and removing references to "blind review" for ICFP.

@@
-\section{The Advantage Of Traces}\label{sec:advantage-traces}
+\subsection{The Advantage Of Traces}\label{sec:advantage-traces}
 
@@
 In particular, we illustrate, using a series of examples drawn
-from student programs in the \ucsdbench\ dataset how \toolname's
-jump-compressed traces can get to the heart of the error by
+from student programs in the \ucsdbench\ dataset, how \toolname's
+jump-compressed traces can get to the heart of the error. Our approach
 %
-highlighting the conflicting values that cause the program to get
+highlights the conflicting values that cause the program to get
 stuck, rather that blaming a single one,
 %
-showing the steps necessary to reach the stuck state, and
+shows the steps necessary to reach the stuck state, and
 %
-not assuming that a function is correct just because it type-checks.
+does not assume that a function is correct just because it type-checks.
 %
@@
 (2)~the error message returned \ocaml,
-(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold)
-\footnote{When the locations from \ocaml\ and \sherrloc\ overlap,
-we just underline the relevant code.},
-(4)~the jump-compressed trace produced by \toolname.
+(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),\footnote{When the locations from \ocaml\ and \sherrloc\ overlap,
+we just underline the relevant code.}
+and (4)~the jump-compressed trace produced by \toolname.
 
@@
 By highlighting conflicting values (\ie\ the source and sink
-of the problem) and not making any assumptions, \toolname\
+of the problem) and not making assumption about function correctness, \toolname\
 focusses the user's attention on the piece of code that is

### 2016-03-16 paper/icfp16-submission/introduction.tex
Made a pass over the complete paper, clarifying and breaking up some convoluted sentences, and removing references to "blind review" for ICFP.

@@
 %
-The graph lets us build a visualization that shows a sequence of
-steps from the source witness to the stuck term and allows the user
-to interactively expand the computation to expose intermediate steps,
+The graph lets us visualize the sequence of
+steps from the source witness to the stuck term. The user can 
+interactively expand the computation to expose intermediate steps
 by selecting an expression and choosing a traversal strategy.

### 2016-05-03 paper/icfp16-submission/rebuttal.txt
add ICFP reviews and 1st draft of rebuttal

We thank the reviewers for their thoughtful remarks, and will respond
briefly to a few categories of concerns.

1. Related Work

We thank reviewer C for the pointers to extra related work. We will
greatly expand the section in the next revision of the paper.

TODO: list a few additional citations we would talk about

2. Bias in Type Variable Instantiation

In the `fac` example, asserting that `n` must be an int without looking
at the rest of the code is indeed biased, but any inference scheme will
be biased in one direction or another. Our bias is towards the
evaluation semantics of the language, other techniques are biased by
their traversal strategy (eg top-down vs bottom-up) or by various
heuristics (whether provided by the compiler or through generic
reasoning).

We would argue, however, (as in Section 5.4) that our bias is more
natural as it is based on a concrete instance of the program crashing.

3. Overloading / Type Classes

Overloaded operators can be problematic for our approach. In reviewer
C's variant of `fac`, our approach would get stuck at the `if n <= m`
test, because we cannot know how to instantiate `n` or `m` at that point
in the program. This does not signify a type error of course, but our
approach would fail to discover a witness. We suspect this is relatively
uncommon in novice OCaml programs (TODO: there were no instances of
this scenario in either of our datasets), but the issue would surely be
exacerbated if we ported the approach to Haskell due to the prevalence
of type classes.

We will add a paragraph describing this issue to the Discussion section,
and defer a treatment of overloading to future work.

4. Scaling to Real-World Code

Our benchmarks are small programs as they were drawn from homework
submissions. The UW dataset had programs generally in the 100-200 LoC
range, while the programs from our dataset were generally <50 LoC (we
had already sliced out a minimal program for each error in our dataset).
It is quite possible that our heuristic for compressing the trace based
on function calls would not suffice for larger programs, though there is
a broad literature on program slicing to draw from there.

5. Language Features

By "pure subset" of OCaml we mean polymorphic datatypes and records, and
we will clarify this in the text. We do not support references, modules,
or the object system.

It is true that we implicitly assume all types are inhabited in the
formalism, and that the generality theorem is weaker in the presence of
uninhabited types. We will clarify this in the text.

### 2016-05-03 paper/icfp16-submission/rebuttal.txt
no instances of generic comparison in benchmarks

@@
 C's variant of `fac`, our approach would get stuck at the `if n <= m`
-test, because we cannot know how to instantiate `n` or `m` at that point
-in the program. This does not signify a type error of course, but our
-approach would fail to discover a witness. We suspect this is relatively
-uncommon in novice OCaml programs (TODO: there were no instances of
-this scenario in either of our datasets), but the issue would surely be
-exacerbated if we ported the approach to Haskell due to the prevalence
-of type classes.
+test, because `n` and `m` are both unconstrained at this point (apart
+from having the same type), so we do not know how to instantiate
+them. This does not signify a type error of course, but our approach
+would fail to discover a witness. We suspect this is relatively uncommon
+in novice OCaml programs (there were no instances of this scenario in
+either of our datasets), but the issue would surely be exacerbated if we
+ported the approach to Haskell due to the prevalence of type classes.
 

### 2016-05-04 paper/icfp16-submission/rebuttal.txt
add list of extra papers to cite and discuss

@@
 We thank the reviewers for their thoughtful remarks, and will respond
-briefly to a few categories of concerns.
+to a few categories of concerns.
 
@@
 
-We thank reviewer C for the pointers to extra related work. We will
-greatly expand the section in the next revision of the paper.
+We thank reviewers B and C for the pointers to extra related work. We
+will greatly expand the section in the next revision of the paper to
+include many of the following references.
 
-TODO: list a few additional citations we would talk about
+- Haack and Wells, ESOP 2003
+- Rahli et al., Tech Report 2010
+- Schilling, TFP 2011
+- Gast, IFL 2004
+- Heeren et al., ICFP 2003
+- Hage and Heeren, IFL 2006
+- Christiansen, TFP 2014
+- Serrano and Hage, ESOP 2016
+- Chitil, ICFP 2001
+- Stuckey et al., Haskell 2003
+- Bayne et al., ICSE 2011
+- Zhang et al., PLDI 2015
 

### 2016-05-04 paper/icfp16-submission/rebuttal.txt
revise related work rebuttal

@@
 
-We thank reviewers B and C for the pointers to extra related work. We
-will greatly expand the section in the next revision of the paper to
-include many of the following references.
+We thank reviewers B and C for the pointers to extra related work.
 
-- Haack and Wells, ESOP 2003
-- Rahli et al., Tech Report 2010
-- Schilling, TFP 2011
-- Gast, IFL 2004
-- Heeren et al., ICFP 2003
-- Hage and Heeren, IFL 2006
-- Christiansen, TFP 2014
-- Serrano and Hage, ESOP 2016
-- Chitil, ICFP 2001
-- Stuckey et al., Haskell 2003
-- Bayne et al., ICSE 2011
-- Zhang et al., PLDI 2015
+Type error slicing shares a common goal with us of providing
+more diagnostic information to the programmer, though the techniques
+differ significantly. Error slicing is based on a static analysis of
+the typing data, usually as a constraint system, and presents
+multiple candidate error locations, possibly ranked according to
+some heuristic. In contrast, our approach executes the program and
+presents a concrete execution trace demonstrating the error. Gast's
+IFL 2004 paper is perhaps the most similar to ours as he produces
+a data flow that would violate the typing rules.
+
+Customizable type error messages along the lines of Heeren 2003,
+Christiansen 2015, and Serrano Mena 2016 also addresses the issue
+of poor type errors. The difference with our work is more pronounced,
+as these techniques often rely on a library author (or instructor)
+with domain-specific knowledge to craft high-quality error messages,
+whereas our approach is fully automatic but ignorant of the application
+domain.
+
+We might also add the algorithmic type debugging work of Chitil 2001
+and Stuckey 2003, which presents the user with a series of queries
+about the types of intermediate expressions to help isolate and
+explain the error.
+
+We will extend the Related Work section to reference all of these
+techniques and papers.
 

### 2016-05-04 paper/icfp16-submission/rebuttal.txt
Add in a few sentences to the rebuttal.
499 words

--- 2016-05-04 paper/icfp16-submission/rebuttal.txt
@@
 
-We thank reviewers B and C for the pointers to extra related work. We
-will greatly expand the section in the next revision of the paper to
-include many of the following references.
+We thank reviewers B and C for the pointers to extra related work.
 
-- Haack and Wells, ESOP 2003
-- Rahli et al., Tech Report 2010
-- Schilling, TFP 2011
-- Gast, IFL 2004
-- Heeren et al., ICFP 2003
-- Hage and Heeren, IFL 2006
-- Christiansen, TFP 2014
-- Serrano and Hage, ESOP 2016
-- Chitil, ICFP 2001
-- Stuckey et al., Haskell 2003
-- Bayne et al., ICSE 2011
-- Zhang et al., PLDI 2015
+We share a common goal with type error slicing, though the techniques
+differ significantly. Error slicing involves a static analysis of the
+typing data and presents multiple candidate error locations, possibly
+ranked according to some heuristic. In contrast, we execute the program
+and present an execution trace demonstrating the error. Gast's IFL 2004
+paper is perhaps the most similar to ours as he produces a data flow
+that would violate the typing rules.
 
-2. Bias in Type Variable Instantiation
+Customizable type error messages along the lines of Heeren 2003,
+Christiansen 2015, and Serrano Mena 2016 also addresses the issue of
+poor type errors. The difference with our work is more pronounced, as
+these techniques often rely on domain-specific knowledge to craft
+high-quality error messages, whereas our approach is fully automatic but
+ignorant of the application domain.
 
-In the `fac` example, asserting that `n` must be an int without looking
-at the rest of the code is indeed biased, but any inference scheme will
-be biased in one direction or another. Our bias is towards the
-evaluation semantics of the language, other techniques are biased by
-their traversal strategy (eg top-down vs bottom-up) or by various
-heuristics (whether provided by the compiler or through generic
-reasoning).
+We will extend the Related Work section to reference all of these
+techniques and papers.
+
+2. Bias in Type Instantiation
+
+In the `fac` example, asserting that `n` must be an int at the branch is
+indeed biased. Our bias is towards the evaluation semantics of the
+language, other techniques are biased by their traversal strategy or by
+various heuristics.
 
@@
 natural as it is based on a concrete instance of the program crashing.
+We will expand the relevant text to clarify this issue.
 
@@
 
-Overloaded operators can be problematic for our approach. In reviewer
-C's variant of `fac`, our approach would get stuck at the `if n <= m`
-test, because `n` and `m` are both unconstrained at this point (apart
-from having the same type), so we do not know how to instantiate
-them. This does not signify a type error of course, but our approach
-would fail to discover a witness. We suspect this is relatively uncommon
-in novice OCaml programs (there were no instances of this scenario in
-either of our datasets), but the issue would surely be exacerbated if we
-ported the approach to Haskell due to the prevalence of type classes.
+Overloaded operators are problematic for our approach. In reviewer C's
+variant of `fac`, we would get stuck at the `if n <= m` test because `n`
+and `m` are constrained to the same unknown type. Thus we do not know
+how to instantiate them, and would fail to discover a witness. Our
+benchmarks suggest this is rare in novice programs, but the issue would
+be exacerbated if we ported the approach to Haskell, due to the
+prevalence of type classes.
 
-We will add a paragraph describing this issue to the Discussion section,
-and defer a treatment of overloading to future work.
+We will describe this issue in the Discussion section, and defer a
+treatment of overloading to future work.
 
@@
 Our benchmarks are small programs as they were drawn from homework
-submissions. The UW dataset had programs generally in the 100-200 LoC
-range, while the programs from our dataset were generally <50 LoC (we
-had already sliced out a minimal program for each error in our dataset).
-It is quite possible that our heuristic for compressing the trace based
-on function calls would not suffice for larger programs, though there is
-a broad literature on program slicing to draw from there.
+submissions, the largest being around 200 LoC. It is quite possible
+that our heuristic for compressing the trace based on function calls
+would not suffice for larger programs, though there is a broad
+literature on program slicing to draw from there. We will augment our
+discussion of this threat to generality in the text.
 

### 2016-05-04 paper/icfp16-submission/rebuttal.txt
Update rebuttal.txt

@@
-We thank the reviewers for their thoughtful remarks, and will respond
-to a few categories of concerns.
+Thank you for your thoughtful feedback!
 
@@
 
-We thank reviewers B and C for the pointers to extra related work.
+We thank reviewers B and C for the pointers to extra related 
+work -- we will extend the Related Work section to reference 
+all of these techniques and papers.
 
@@
 
-We will extend the Related Work section to reference all of these
-techniques and papers.
-
-2. Bias in Type Instantiation
+2. Bias in Type Instantiation (?CUT?)
 
@@
 
-Overloaded operators are problematic for our approach. In reviewer C's
-variant of `fac`, we would get stuck at the `if n <= m` test because `n`
-and `m` are constrained to the same unknown type. Thus we do not know
-how to instantiate them, and would fail to discover a witness. Our
-benchmarks suggest this is rare in novice programs, but the issue would
-be exacerbated if we ported the approach to Haskell, due to the
-prevalence of type classes.
+Currently our approach does not support overloading.
+
+With an overloaded language, reviewer C's variant of `fac`, 
+we would get stuck at the `if n <= m` test because `n` and `m` 
+are constrained to the same unknown type, but as we do not know 
+how to instantiate such types, we  would fail to discover a witness.
+
+Thus, our formalization (theory) assumes there are no overloaded
+(or polymorphic) operators. Furthermore, in our implementation, 
+i.e. over *all* our benchmarks,  all the operands are concretized 
+*before* they are applied the any (overloaded) operator.
 
@@
 Our benchmarks are small programs as they were drawn from homework
-submissions, the largest being around 200 LoC. It is quite possible
-that our heuristic for compressing the trace based on function calls
-would not suffice for larger programs, though there is a broad
-literature on program slicing to draw from there. We will augment our
-discussion of this threat to generality in the text.
+submissions, the largest being around 200 LoC. While our jump compression 
+heuristic may not suffice for larger programs, we could extend it with 
+ideas from the slicing literature.
 
@@
 
-By "pure subset" of OCaml we mean polymorphic datatypes and records, and
-we will clarify this in the text. We do not support references, modules,
-or the object system.
+By "pure subset" of OCaml we mean polymorphic datatypes and records, 
+and we will clarify this in the text. Currently, we do not support 
+references, modules, or the object system.
 
-It is true that we implicitly assume all types are inhabited in the
-formalism, and that the generality theorem is weaker in the presence of
-uninhabited types. We will clarify this in the text.
+Indeed, we implicitly assume all types are inhabited in the formalism, 
+and that the generality theorem is weaker in the presence of uninhabited 
+types. We will clarify this in the text.

### 2016-05-04 paper/icfp16-submission/rebuttal.txt
fixes

@@
 
-1. Related Work
+1. Related Work [B, C]
 
@@
 
-2. Bias in Type Instantiation (?CUT?)
+2. Bias in Type Instantiation [C]
 
-In the `fac` example, asserting that `n` must be an int at the branch is
-indeed biased. Our bias is towards the evaluation semantics of the
-language, other techniques are biased by their traversal strategy or by
-various heuristics.
+Our goal is to be as lenient as possible in type instantation. In the
+`fac` example, we assert that `n` must be an int at the branch condition
+because that is the only way to make progress. This is indeed biased, as
+the `n <= 0` expression may be the source of the error, but it is
+necessary to continue evaluation.
 
-We would argue, however, (as in Section 5.4) that our bias is more
-natural as it is based on a concrete instance of the program crashing.
 We will expand the relevant text to clarify this issue.
 
-3. Overloading / Type Classes
+3. Overloading / Type Classes [A,B,C]
 
@@
 
-4. Scaling to Real-World Code
+4. Scaling to Real-World Code [C]
 
@@
 
-5. Language Features
+5. Language Features [A]
 

### 2016-05-24 paper/icfp16-cameraready/evaluation.tex
misc

--- 2016-03-16 paper/icfp16-submission/evaluation.tex
@@
 course of the first three assignments (IRB  
-% \# hidden for blind review), 
-\#140608),
+\# hidden for blind review), 
+% \#140608),
 from which we extracted \ucsdsize\ distinct, ill-typed \ocaml programs.
@@
 The second dataset --- widely used in the literature --- comes from a
-similar course at the University of Washington~\cite{lerner_seminal:_2006},
+graduate-level course at the University of Washington~\cite{Lerner2006-pj},
 from which we extracted 284 ill-typed programs.
@@
   10 seconds. In both datasets we detect actual type errors about 82\%
-  of the time, unbound variables or constructors 3-4\% of the time, and
-  diverging loops 2-3\% of the time. For the remaining 11-12\% of the
+  of the time, unbound variables or constructors 3--4\% of the time, and
+  diverging loops 2--3\% of the time. For the remaining 11--12\% of the
   programs we are unable to provide any useful feedback.  }
@@
 In both datasets our tool was able to find a witness for 83\% of the
-programs in under one second, \ie fast enough to be integrated as a
+programs in under one second, \ie\ fast enough to be integrated as a
 compile-time check. If we extend our tolerance to a 10 second timeout,
@@
 Interestingly, while the vast majority of witnesses corresponded to a
-type-error, as expected, 3-4\% triggered an unbound variable error (even
-though \ocaml reported a type error) and 2-3\% triggered an infinite
+type-error, as expected, 3--4\% triggered an unbound variable error (even
+though \ocaml\ reported a type error) and 2--3\% triggered an infinite
 recursion error.
@@
 %
-While a more advanced search procedure, \eg dynamic-symbolic execution,
+While a more advanced search procedure, \eg\ dynamic-symbolic execution,
 could likely trigger more of the type errors, our experiments show that
@@
 program. Enhancing our visualization to slice out program paths relevant
-to specific values~\cite{perera_functional_2012},
+to specific values~\cite{Perera2012-dy},
 would likely help alleviate this issue, allowing users to

### 2016-05-24 paper/icfp16-cameraready/gallery2.tex
misc

--- 2016-03-16 paper/icfp16-submission/gallery2.tex
@@
-\subsection{The Advantage Of Traces}\label{sec:advantage-traces}
+\subsection{The Advantage of Traces}\label{sec:advantage-traces}
 
@@
 the static reports produced by the \ocaml\ compiler and \sherrloc,
-a state-of-the-art fault localization approach~\cite{zhang_toward_2014}.
+a state-of-the-art fault localization approach~\cite{Zhang2014-lv}.
 %
@@
 %
-\toolname, in contrast, makes no asummptions about @append@ and produces
+\toolname, in contrast, makes no assumptions about @append@ and produces
 a trace that illustrates the true error on line 4, by

### 2016-05-24 paper/icfp16-cameraready/introduction.tex
misc

--- 2016-03-16 paper/icfp16-submission/introduction.tex
@@
 Much recent work has focused on analyzing unification constraints
-to properly \emph{localize} a type error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
+to properly \emph{localize} a type error~\cite{Lerner2007-dt,Chen2014-gd,Zhang2014-lv,Pavlinovic2014-mr},
 but an accurate source location still does not explain \emph{why} the
@@
 \end{minipage}
-\caption{(TL) An ill-typed \texttt{fac} function; (BL) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (R) The same trace, fully expanded.}
+\caption{(top-left) An ill-typed \texttt{fac} function; (bottom-left) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (right) The same trace, fully expanded.}
 \label{fig:factorial}
@@
 %
-The first is the \uwbench\ data set~\cite{lerner_searching_2007}
+The first is the \uwbench\ data set~\cite{Lerner2007-dt}
 comprising \uwsize\ ill-typed programs.

### 2016-05-24 paper/icfp16-cameraready/overview.tex
misc

--- 2016-03-15 paper/icfp16-submission/overview.tex
@@
 To defer input generation, we borrow the notion of a ``hole'' from
-SmallCheck~\cite{runciman_smallcheck_2008}.
+SmallCheck~\cite{Runciman2008-ka}.
 %
@@
 \label{sec:visual-witness}
-We have shown how to reliably find witnesses to type errors in \ocaml,
+We have described how to reliably find witnesses to type errors in \ocaml,
 but this does not fully address our original goal --- to \emph{explain}

### 2016-05-24 paper/icfp16-cameraready/related.tex
misc

--- 2015-11-20 paper/pldi16-submission/related.tex
@@
 
-Lerner \etal~\cite{lerner_searching_2007} attempt to localize the type
+Lerner \etal~\cite{Lerner2007-dt} attempt to localize the type
 error and suggest a fix by replacing expressions (or removing them
@@
 %
-Chen and Erwig~\cite{chen_counter-factual_2014} use a variational type
+Chen and Erwig~\cite{Chen2014-gd} use a variational type
 system to allow for the possibility of changing an expression's type,
@@
 %
-Neubauer and Thiemann~\cite{neubauer_discriminative_2003} present a
+Neubauer and Thiemann~\cite{Neubauer2003-xv} present a
 decidable type system based on discriminative sum types, in which all
@@
 %
-Zhang and Myers~\cite{zhang_toward_2014} present an algorithm for
+Zhang and Myers~\cite{Zhang2014-lv} present an algorithm for
 identifying the most likely culprit in a system of unsatisfiable
@@
 %
-Pavlinovic \etal~\cite{pavlinovic_finding_2014} translate the %error
+Pavlinovic \etal~\cite{Pavlinovic2014-mr} translate the %error
 localization problem to a MaxSMT optimization problem, using
@@
 \label{sec:running-ill-typed}
-Vytiniotis \etal~\cite{vytiniotis_equality_2012} extend the Haskell
+Vytiniotis \etal~\cite{Vytiniotis2012-gh} extend the Haskell
 compiler GHC to support compiling ill-typed programs, but their intent
@@
 Our use of holes to represent unknown values is inspired by the work of
-Runciman, Naylor, and Lindblad~\cite{runciman_smallcheck_2008,naylor_finding_2007,lindblad_property_2007},
+Runciman, Naylor, and Lindblad~\cite{Runciman2008-ka,Naylor2007-mi,Lindblad2007-oy},
 %
@@
 %
-Random test generation~\cite{claessen_quickcheck:_2000,csallner_jcrasher:_2004,pacheco_feedback-directed_2007}
+Random test generation~\cite{Claessen2000-lj,Csallner2004-bf,Pacheco2007-at}
 %
@@
 %
-Dynamic-symbolic execution~\cite{godefroid_dart:_2005,cadar_klee:_2008,tillmann_pex_2008}
+Dynamic-symbolic execution~\cite{Godefroid2005-am,Cadar2008-kg,Tillmann2008-qc}
 %
@@
 % \begin{itemize}
-% \item \cite{claessen_quickcheck:_2000}
+% \item \cite{Claessen2000-lj}
 % \item 
-% \item \cite{godefroid_dart:_2005}
-% \item \cite{cadar_klee:_2008}
-% \item \cite{tillmann_pex_2008}
+% \item \cite{Godefroid2005-am}
+% \item \cite{Cadar2008-kg}
+% \item \cite{Tillmann2008-qc}
 % \end{itemize}
@@
 \paragraph{Program Exploration}
-Perera \etal~\cite{perera_functional_2012} present a tracing semantics
+Perera \etal~\cite{Perera2012-dy} present a tracing semantics
 for functional programs that tags values with their provenance, enabling

### 2016-05-24 paper/icfp16-cameraready/semantics.tex
init paper/icfp16-cameraready
misc

--- 2016-03-15 paper/icfp16-submission/semantics.tex
@@
 We have formalized and tested our semantics and generality theorem
-in \textsc{PLT-Redex}~\cite{felleisen2009semantics}.
+in \textsc{PLT-Redex}~\cite{Felleisen2009-ya}.
 
@@
 %
-We solve this problem be equipping our term language with a
+We solve this problem by equipping our term language with a
 notion of a \emph{hole}, written \vhole{\thole}, which represents
@@
 needed locally to make progress (inspired by the manner in
-which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
+which SmallCheck uses lazy evaluation~\cite{Runciman2008-ka}).
 %
@@
 \gensym       & :   & t \times \tsu \rightarrow v \\
-\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, &  \text{if } \thole \in dom(\tsu) \\
-\gen{\tint}{\tsu}   & \defeq  & n, &  \text{non-deterministic} \\
-\gen{\tbool}{\tsu}  & \defeq  & b, &  \text{non-deterministic} \\
+\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu} &  \text{if } \thole \in dom(\tsu) \\
+\gen{\tint}{\tsu}   & \defeq  & n &  \text{non-det.} \\
+\gen{\tbool}{\tsu}  & \defeq  & b &  \text{non-det.} \\
 \gen{\tprod{t_1}{t_2}}{\tsu}  & \defeq  & \epair{\gen{t_1}{\tsu}}{\gen{t_2}{\tsu}} & \\ % \begin{subarray}{l} v_1 = \gen{t_1}{\tsu}\\ v_2 = \gen{t_2}{\tsu}\end{subarray} \\
-\gen{\ttree{t}}{\tsu}  & \defeq  & tr, &  \text{non-deterministic} \\
-\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, &  \text{\ehole and \thole are fresh} \\
-\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \text{\ehole is fresh} \\
+\gen{\ttree{t}}{\tsu}  & \defeq  & tr &  \text{non-det.} \\
+\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}} &  \text{\ehole, \thole are fresh} \\
+\gen{\thole}{\tsu}  & \defeq & \vhole{\thole} & \text{\ehole is fresh} \\
 \end{array}
@@
 Our semantics lacks a built-in $\mathtt{fix}$ construct
-for defining recursive function, which may surprise
+for defining recursive functions, which may surprise
 the reader.
@@
 %
-Thus we do not need a built-in recursion construct, we
-can use any of the standard fixed-point operators.
+Thus we can use any of the standard fixed-point operators
+and do not need a built-in recursion construct.
 
@@
 %
-\ES{calling a type a refinement of a value sounds really weird..}
+%\ES{calling a type a refinement of a value sounds really weird..}
 
@@
 \label{lem:vsu-ext}
-If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
-and $\trace'$ is a single-step extension of $\trace$
+If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
+and $\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
+  (\ie\ $\trace'$ is a single-step extension of $\trace$)
+% and $\trace'$ is a single-step extension of $\trace$
 and $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$
@@
   %
-  Note that only \forcesym can change \tsu, and it can only do so
+  Note that only \forcesym\ can change \tsu,
+  %, and it can only do so via \unifysym
   via \unifysym, which can only extend \tsu.
@@
 %   % \ES{do we need to say anythign else?}
+\qedhere
 \end{description}

### 2016-05-24 paper/icfp16-cameraready/overview.tex
cite heuristics for favoring small values

@@
 At this point we choose a
-random @int@\footnote{With standard heuristics to favor small values.}
+random @int@\footnote{With standard heuristics~\cite{Claessen2000-lj} to favor small values.}
 for the instantiation and

### 2016-05-24 paper/icfp16-cameraready/evaluation.tex
address ad-hoc polymorphism

@@
 
-There are, of course, drawbacks to our approach. Three that stand out
+There are, of course, drawbacks to our approach. Four that stand out
 are: (1) coverage limits due to random generation, (2) the inability to
-handle certain instances of infinite types, and (3) dealing with an
-explosion in the size of generated traces.
+handle certain instances of infinite types, (3) dealing with an
+explosion in the size of generated traces, and (4) handling ad-hoc
+polymorphism.
 
@@
 
+\paragraph{Ad-hoc Polymorphism}
+Our approach can only support ad-hoc polymorphism (\eg\ type-classes in
+\haskell\ or polymorphic comparison functions in \ocaml) in limited cases
+where we have enough typing information at the call-site to resolve the
+overloading. For example, consider the @n <= 0@ test in our @fac@ example.
+@<=@ is polymorphic in \ocaml, but in this case we can make progress because
+the literal @0@ is not. If we parameterized @fac@ by a lower bound, \eg
+%
+\begin{code}
+  let rec fac n m =
+    if n <= m then
+      1
+    else
+      n * fac (n - 1) m
+\end{code}
+%
+and called @fac@ with two holes, we would get stuck at the @n <= m@
+test; not because of a type error, but because all we know about
+@n@ and @m@ at that point is that they must have the same (unknown)
+type.
+
+This issue is uncommon in \ocaml\ (we did not detect a single instance
+of it across all of our benchmarks), but it would surely be exacerbated
+by a language like \haskell, which makes heavy use of overloading. We
+suspect that dynamic-symbolic execution would allow us to handle ad-hoc
+polymorphism, but defer a proper treatment to future work.
+
 % \begin{itemize}

### 2016-05-25 paper/icfp16-cameraready/evaluation.tex
fix formatting of plots

@@
 The first dataset comes from the Spring 2014 undergraduate Programming
-Languages course at our institution.
+Languages course at UC San Diego.
 %
@@
 course of the first three assignments (IRB  
-\# hidden for blind review), 
-% \#140608),
+% \# hidden for blind review),
+\#140608),
 from which we extracted \ucsdsize\ distinct, ill-typed \ocaml programs.

### 2016-05-25 paper/icfp16-cameraready/plots.py

@@ -1,0 +1,258 @@
+import csv
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+
+UCSD = 'UCSD'
+
+BUCKETS = [0.1, 0.2, 1.0, 10.0, 60.0 ] # range(500, 3001, 500)
+COLORS=['#90B0D4', '#90D492', '#D4B490', '#D490D2']
+#COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
+
+SAFE = ['S', 'T']
+SAFE_L = ['Safe', 'Timeout']
+UNSAFE = ['U', 'B', 'D'] #, 'O']
+UNSAFE_L = ['Unsafe', 'Unbound', 'Diverge'] #, 'Output']
+ALL = UNSAFE + SAFE
+ALL_L = UNSAFE_L + SAFE_L
+
+ALL_D = [ ['U'], ['B'], ['D'], ['S', 'T']]
+ALL_DL = [ 'Witness', 'Unbound', 'Diverge', 'No Witness'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
+
+
+def read_csv(f):
+    with open(f) as f:
+        return list(csv.reader(f))
+
+def autolabel(ax, rects):
+    # attach some text labels
+    for rect in rects:
+        height = rect.get_height()
+        ax.text(rect.get_x() + rect.get_width()/2., height,
+                '%d' % int(height),
+                ha='center', va='bottom')
+
+def cumulative_coverage(data):
+    headers = data[0]
+    data = data[1:]
+    return [(l, int(100 * len([r for r in data
+                               if float(r[2]) <= l
+                               and r[4] in UNSAFE])
+                          / float(len(data))))
+            for l in BUCKETS]
+
+def plot_coverage(seminal, ucsd):
+    xy_s = cumulative_coverage(seminal)
+    xy_u = cumulative_coverage(ucsd)
+
+
+    N = len(xy_s)
+    ind = np.arange(N)    # the x locations for the groups
+    width = 0.35       # the width of the bars: can also be len(x) sequence
+
+    fig = plt.figure()
+    p1 = plt.bar(ind, [r[1] for r in xy_s], width,
+                 color=COLORS[0])
+    p2 = plt.bar(ind + width, [r[1] for r in xy_u], width,
+                 color=COLORS[1])
+
[…]

### 2016-05-25 paper/icfp16-cameraready/evaluation.tex
clarify size of benchmarks

@@
 \paragraph{Benchmarks}
-We answer these questions on two sets of ill-typed programs, \ie
+We answer these questions on two sets of ill-typed programs, \ie\
 programs that were rejected by the \ocaml\ compiler because of a
@@
 \#140608),
-from which we extracted \ucsdsize\ distinct, ill-typed \ocaml programs.
+from which we extracted \ucsdsize\ distinct, ill-typed \ocaml\ programs.
 %
@@
 %
-Both datasets include programs that demonstrate a variety of functional
-programming idioms, including (tail) recursive functions, higher-order
-functions, polymorphic data types, and expression evaluators.
+Both datasets contain relatively small programs, the largest being 348
+SLoC; however, they demonstrate a variety of functional programming
+idioms including (tail) recursive functions, higher-order functions,
+polymorphic data types, and expression evaluators.
 
@@
 %
-We also added a na\"ive check for infinite recursion; at each recursive
+We also added a na{\"\i}ve check for infinite recursion; at each recursive
 function call we check whether the new arguments are identical to the
@@
 can address this short-coming by performing an exhaustive search for
-inputs (\emph{resp}. paths through the program). As our experiments
+inputs (\emph{resp}.\ paths through the program). As our experiments
 show, however, novice programs do not appear to require more advanced
@@
 Though the average complexity of our generated traces is low in terms of
-jumps, there are some extreme outliers. We cannot reasonably expect a
-novice user to explore a trace containing 50+ terms and draw a
-conclusion about which pieces contributed to the bug in their
-program. Enhancing our visualization to slice out program paths relevant
-to specific values~\cite{Perera2012-dy},
-would likely help alleviate this issue, allowing users to
-highlight a confusing value and ask: ``Where did this come from?''
+jumps, there are some extreme outliers.
+%
+We cannot reasonably expect a novice user to explore a trace containing
+50+ terms and draw a conclusion about which pieces contributed to the
+bug in their program.
+%
+Enhancing our visualization to slice out program paths relevant to
+specific values~\cite{Perera2012-dy}, would likely help alleviate this
+issue, allowing users to highlight a confusing value and ask: ``Where
+did this come from?''
 

### 2016-06-07 paper/icfp16-cameraready/abstract.tex
init paper/icfp16-cameraready
clarify various things relating to language features and inhabitants

--- 2016-01-20 paper/icfp16-submission/abstract.tex
@@
   \emph{general witnesses} in that if a witness is
-  found, then \emph{for all} input types, there exist
-  inhabitants that can make the function go wrong.
+  found, then \emph{for all} inhabited input types, there exist
+  values that can make the function go wrong.
   %
@@
   88\% of the programs, and our reduction graph
-  yields small counterexamples for 90\% of the witnesses.
+  yields small counterexamples for 81\% of the witnesses.
 \end{abstract}

### 2016-06-07 paper/icfp16-cameraready/evaluation.tex
clarify various things relating to language features and inhabitants

@@
 We have implemented a prototype of our search procedure and trace
-visualization for the purely functional subset of \ocaml, in a tool called \nanomaly.
+visualization for a purely functional subset of \ocaml\
+--- with polymorphic types and records, but no modules, objects, or polymorphic variants ---
+in a tool called \nanomaly.
+%
+We treat explicit type signatures, \eg @(x : int)@, as
+primitive operations that narrow the type of the wrapped value.
+%
 In our implementation we instantiated \gensym\ with a simple random
-generation of values, which we will show suffices for the
-majority of type errors.
+generation of values, which we will show suffices for the majority of
+type errors.
 

### 2016-06-07 paper/icfp16-cameraready/introduction.tex
clarify various things relating to language features and inhabitants

@@
 %
-We prove that our procedure synthesizes \emph{general witnesses},
-which means, intuitively, that if a witness is found for a given
-ill-typed function, then, \emph{for all} input types, there exist
-inhabitants that can make the function go wrong.
+We prove that our procedure synthesizes \emph{general witnesses}, which
+means, intuitively, that if a witness is found for a given ill-typed
+function, then, \emph{for all} (inhabited) input types, there exist
+values that can make the function go wrong.
 

### 2016-06-07 paper/icfp16-cameraready/evaluation.tex
misc related work

@@
 %
-Due to the possibility of non-termination we set a timeout of 10 minutes
-per program.
+Due to the possibility of non-termination we set a timeout of one minute
+total per program.
 %
@@
 feedback as they either completed 1,000 tests successfully, or timed out
-after 10 minutes.
+after one minute.
 %

### 2016-06-07 paper/icfp16-cameraready/related.tex
misc related work

@@
 value-level, by searching at the type level Chen's search is complete due
-the finite universe of types in a program.
+the finite universe of types used in the program.
 %
@@
 %
-Zhang and Myers~\cite{Zhang2014-lv} present an algorithm for
+Zhang and Myers~\cite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
 identifying the most likely culprit in a system of unsatisfiable
@@
 \label{sec:running-ill-typed}
-Vytiniotis \etal~\cite{Vytiniotis2012-gh} extend the Haskell
+Vytiniotis~\etal~\cite{Vytiniotis2012-gh} extend the \haskell
 compiler GHC to support compiling ill-typed programs, but their intent
@@
 at runtime.
+%
+Bayne~\etal~\cite{Bayne2011-cn} also provide a semantics for running
+ill-typed (\java) programs, but in constrast transform the program to
+perform nearly all type checking at run-time. The key difference between
+Bayne~\etal and our work is that we use the dynamic semantics to
+automatically search for a witness to the type error, while their focus
+is on incremental, programmer-driven testing.
 
@@
 of reductions that produced it. Notably, they allow the user to supply a
-\emph{partial value} -- containing holes -- and present a partial slice,
+\emph{partial value} --- containing holes --- and present a partial slice,
 containing only those steps that affected the the partial value.
@@
 inserting the intermediate terms that preceded it while ignoring
-unrelated computation steps; this would be interesting future work.
+unrelated computation steps. %; this would be interesting future work.
 

### 2016-06-07 paper/icfp16-cameraready/related.tex
update related work

@@
 
-\paragraph{Diagnosing and Repairing Type Errors}
+\paragraph{Localizing and Repairing Type Errors}
 \label{sec:diagnosis-repair}
@@
 
-Lerner \etal~\cite{Lerner2007-dt} attempt to localize the type
-error and suggest a fix by replacing expressions (or removing them
-entirely) with alternatives based on the surrounding program context.
+% \subparagraph{Localization}
+The traditional Damas-Milner type inference
+algorithm~\cite{Damas1982-uw} reports the first program location where a
+type mismatch is discovered (subject to the traversal
+strategy~\cite{Lee1998-ys}).
 %
-Chen and Erwig~\cite{Chen2014-gd} use a variational type
-system to allow for the possibility of changing an expression's type,
-and search for an expression whose type can be changed such that type
-inference would succeed.
+As a result the error can be reported far away from its
+source~\cite{McAdam1998-ub} without enough information to guide the
+user.
+%
+Type-error slicing~\cite{Haack2003-vc,Schilling2011-yf,Rahli2015-tt,Sagonas2013-bf,Gast2004-zd,Neubauer2003-xv}
+%treats type inference as a constraint-satisfaction problem and
+recognizes this flaw and instead produces a slice of the program
+containing \emph{all} program locations that are connected to the type
+error.
+%
+%% \citet{Neubauer2003-xv} present a decidable type system based on
+%% discriminative sum types, in which all terms are typeable and type
+%% derivations contain all type errors in a program. They then use the
+%% typing derivation to slice out the parts of the expression related to
+%% each error.
+%
+Though the program slice must contain the source of the error, it can
+suffer from the opposite problem of providing \emph{too much}
+information, motivating recent work in ranking the program locations.
+%
+Zhang~\etal~\cite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
+identifying the most likely culprit using Bayesian reasoning.
+%
+\citet{Pavlinovic2014-mr,Pavlinovic2015-kh} translate the %error
+localization problem to a MaxSMT optimization problem, using
+compiler-provided weights to rank the possible sources.
+
+% \subparagraph{Repairing}
+In addition to localizing the error, \citet{Lerner2007-dt} attempt to
+suggest a fix by replacing expressions (or removing them entirely) with
+alternatives based on the surrounding program context.
+%
+\citet{Chen2014-gd} use a variational type system to allow for the
+possibility of changing an expression's type, and search for an
+expression whose type can be changed such that type inference would
+succeed.
 %
@@
 %
-Neubauer and Thiemann~\cite{Neubauer2003-xv} present a
-decidable type system based on discriminative sum types, in which all
-terms are typeable and type derivations contain all type errors in a
-program. They then use the typing derivation to slice out the parts of
-the expression related to each error.
-%
-Zhang and Myers~\cite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
-identifying the most likely culprit in a system of unsatisfiable
-constraints (\eg type equalities), based on Bayesian reasoning.
-%
-Pavlinovic \etal~\cite{Pavlinovic2014-mr} translate the %error
-localization problem to a MaxSMT optimization problem, using
-compiler-provided weights to rank the possible sources.
 % \item \cite{chen_error-tolerant_2012}
@@
 % \item \cite{thatte_type_1988}
+%
 
@@
 
+\paragraph{Improving Error Messages}
+The content and quality of the error messages themselves has also been
+studied extensively.
+%
+\citet{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
+messages in novice environments and present suggestions for improving
+their quality and consistency.
+%
+\citet{Hage2006-hc} identify a variety of general heuristics to improve
+the quality of type error messages, based on their teaching experience.
+%
+\citet{Heeren2003-db}, \citet{Christiansen2014-qc}, and
+\citet{Serrano2016-oo} provide methods for library authors to specialize
+type errors with domain-specific knowledge.
+%
+The difference with our work is more pronounced here as we do not
+attempt to improve the quality of the error message, instead we search
+for a witness to the error and explain it with the resulting execution
+trace.
+
 \paragraph{Running Ill-Typed Programs}
 \label{sec:running-ill-typed}
-Vytiniotis~\etal~\cite{Vytiniotis2012-gh} extend the \haskell
+\citet{Vytiniotis2012-gh} extend the \haskell
 compiler GHC to support compiling ill-typed programs, but their intent
@@
 %
-Bayne~\etal~\cite{Bayne2011-cn} also provide a semantics for running
+\citet{Bayne2011-cn} also provide a semantics for running
 ill-typed (\java) programs, but in constrast transform the program to
 perform nearly all type checking at run-time. The key difference between
-Bayne~\etal and our work is that we use the dynamic semantics to
+Bayne~\etal\ and our work is that we use the dynamic semantics to
 automatically search for a witness to the type error, while their focus
@@
 %
-combines symbolic execution (to track which path a given input
-triggers) with concrete execution (to ensure failures are
-not spurious). The system collects a path condition during execution,
-which tracks symbolically what conditions must be met to trigger the
-current path. Upon successfully completing a test run, it negates the
-path condition and queries a solver for another set of inputs that
-satisfy the negated path condition, \ie inputs that will not trigger the
-same path. Thus, it can prune the search space
-much faster than techniques based on enumerating values, but is limited
-by the expressiveness of the underlying solver. Our operational
-semantics is amenable to dynamic-symbolic execution, one would just need
-to collect the path condition and replace our implementation of \gensym
-by a call to the solver. We chose to use lazy, random generation instead
-because it is efficient, and % does not incur the overhead of an external solver, and
-produces high coverage for our domain of novice programs.
+combines symbolic execution (to track which path a given input triggers)
+with concrete execution (to ensure failures are not spurious). The
+system collects a path condition during execution, which tracks
+symbolically what conditions must be met to trigger the current
+path. Upon successfully completing a test run, it negates the path
+condition and queries a solver for another set of inputs that satisfy
+the negated path condition, \ie inputs that will not trigger the same
+path. Thus, it can prune the search space much faster than techniques
+based on enumerating values, but is limited by the expressiveness of the
+underlying solver.
+
+Our operational semantics is amenable to dynamic-symbolic execution, one
+would just need to collect the path condition and replace our
+implementation of \gensym by a call to the solver. We chose to use lazy,
+random generation instead because it is efficient, and % does not incur
+the overhead of an external solver, and produces high coverage for our
+domain of novice programs.
 

### 2016-06-07 paper/icfp16-cameraready/related.tex
misc

@@
 %
-In contrast to Lerner, who searches for changes at the
-value-level, by searching at the type level Chen's search is complete due
-the finite universe of types used in the program.
+In contrast to Lerner~\etal, who search for changes at the value-level,
+Chen~\etal\ search at the type-level and are thus complete due the finite
+universe of types used in the program.
 %
@@
 % value come from?'' and thus is focused on backward exploration. 
-Perera's system focuses on backward exploration; in
-contrast, our visualization supports forward \emph{and} backward
-exploration, though our backward steps are more limited than
-Perera's. Specifically, we do not support selecting a value and
-inserting the intermediate terms that preceded it while ignoring
-unrelated computation steps. %; this would be interesting future work.
+Perera~\etal\ focus on backward exploration; in contrast, our
+visualization supports forward \emph{and} backward exploration, though
+our backward steps are more limited.
+%
+Specifically, we do not support selecting a value and inserting the
+intermediate terms that preceded it while ignoring unrelated computation
+steps. %; this would be interesting future work.
 

### 2016-06-07 paper/icfp16-cameraready/semantics.tex
clarify various things relating to language features and inhabitants
misc

@@
   \hbox{$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
-  then for every % inhabitable
+  then for every
+  (inhabited\footnote{All types in our formalism are inhabited, but in a larger language like \ocaml this may not be the case.})
   type
@@
 %
-Our witness generation procedure takes two inputs.
-%
-First, a search bound $k$ which is used to define the \emph{number} of
-traces to explore. (We assume, without loss of generality, that all
-traces are finite.)
-%
-Second, the target expression $e$ that contains the type error,
-and may be a curried function of multiple arguments.
+Our witness generation procedure takes two inputs:
+%
+(1) a search bound $k$ which is used to define the \emph{number} of
+traces to explore\footnote{We assume, without loss of generality, that all
+traces are finite.} and
+%
+(2) the target expression $e$ that contains the type error
+(which may be a curried function of multiple arguments).
 %

### 2016-06-08 paper/icfp16-cameraready/introduction.tex
misc

@@
 \emph{witnesses} to type errors, \ie\ inputs that cause a
-program to go wrong~(\S~\ref{sec:semantics}).
+program to go wrong~(\S~\ref{sec:searching-witness}).
 %
@@
 Of course, the problem of finding witnesses is
-undecidable. In fact, due to the necessarily
+undecidable in general. In fact, due to the necessarily
 conservative nature of static typing, there

### 2016-06-10 paper/icfp16-cameraready/study.R
add data from user study

@@ -1,0 +1,93 @@
+library(irr)
+
+a0 = read.csv(file="study-data/A-0-scores.csv")
+a1 = read.csv(file="study-data/A-1-scores.csv")
+a2 = read.csv(file="study-data/A-2-scores.csv")
+a3 = read.csv(file="study-data/A-Eric-scores.csv")
+b0 = read.csv(file="study-data/B-0-scores.csv")
+b1 = read.csv(file="study-data/B-1-scores.csv")
+b2 = read.csv(file="study-data/B-2-scores.csv")
+b3 = read.csv(file="study-data/B-Eric-scores.csv")
+
+reasons0 = c(a0$X1..append.reason,
+             a0$X3..digitsofint.reason,
+             a0$X5..sumlist.reason,
+             a0$X7..wwhile.reason,
+             b0$X1..sumlist.reason,
+             b0$X3..wwhile.reason,
+             b0$X5..append.reason,
+             b0$X7..digitsofint.reason)
+reasons1 = c(a1$X1..append.reason,
+             a1$X3..digitsofint.reason,
+             a1$X5..sumlist.reason,
+             a1$X7..wwhile.reason,
+             b1$X1..sumlist.reason,
+             b1$X3..wwhile.reason,
+             b1$X5..append.reason,
+             b1$X7..digitsofint.reason)
+reasons2 = c(a2$X1..append.reason,
+             a2$X3..digitsofint.reason,
+             a2$X5..sumlist.reason,
+             a2$X7..wwhile.reason,
+             b2$X1..sumlist.reason,
+             b2$X3..wwhile.reason,
+             b2$X5..append.reason,
+             b2$X7..digitsofint.reason)
+reasons3 = c(a3$X1..append.reason,
+             a3$X3..digitsofint.reason,
+             a3$X5..sumlist.reason,
+             a3$X7..wwhile.reason,
+             b3$X1..sumlist.reason,
+             b3$X3..wwhile.reason,
+             b3$X5..append.reason,
+             b3$X7..digitsofint.reason)
+reasons = data.frame(reasons0, reasons1, reasons2, reasons3)
+
+fixes0 = c(a0$X2..append.fix,
+           a0$X4..digitsofint.fix,
+           a0$X6..sumlist.fix,
+           a0$X8..wwhile.fix,
+           b0$X2..sumlist.fix,
[…]

### 2016-06-11 paper/icfp16-cameraready/evaluation.tex
start draft of user study section

@@
 
+\input{gallery2}
+
+\subsection{User Study}
+\label{sec:user-study}
+Finally, to test the explanatory power of our jump-compressed traces, we
+ran a user study (IRB \#XXXX) at the University of Virginia (UVA).
+%
+We added four questions to an exam in the Spring session of UVA's
+undergraduate Programming Languages course (CS XXX), in which presented
+the students with ill-typed \ocaml\ programs and asked them to
+%
+(1) \emph{explain} the type error, and
+%
+(2) \emph{fix} the type error.
+%
+For each problem the student would be given the ill-typed program and
+either \ocaml's error message or \toolname's jump-compressed trace.
+%
+The four problems assigned to the students were the
+%
+@sumList@, @digitsOfInt@, and @wwhile@ programs from \S~\ref{sec:advantage-traces}
+as well as the following @append@ program....\ES{HERE}
+
+We then instructed four annotators (one of whom is an author, the other
+three were teaching assistants at UCSD) to classify the answers as
+correct or incorrect.
+
+We performed an inter-rater reliability (IRR) analysis to determine the
+degree to which the annotators consistently graded the exams.
+%
+As we had more than two annotators assigning nominal ratings we used
+Fleiss' kappa~\cite{XXX} to measure IRR.
+%
+The resulting kappa values, $\kappa = 0.72$ for the explanations and
+$\kappa = 0.83$ for the fixes, indicate substantial agreement among
+the annotators~\cite{XXX}.
+
+The results of one of the annotators are shown in
+Figure~\ref{fig:results-user-study}.
+%
+\begin{figure*}[t]
+\centering
+\includegraphics[width=0.49\linewidth]{user-study-reason.png}
+\includegraphics[width=0.49\linewidth]{user-study-fix.png}
+\caption{Results of the user study. The students given \toolname's
+  jump-compressed trace consistently scored better than those
+  given \ocaml's type error.}
+\label{fig:results-user-study}
+\end{figure*}
+%
+The students given \toolname's jump-compressed trace were consistently
+more likely to correctly explain and fix the type error than those
+given \ocaml's error message.
+
 \subsection{Discussion}

### 2016-06-11 paper/icfp16-cameraready/plots.py

@@ -23,6 +23,9 @@ ALL_DL = [ 'Witness', 'Unbound', 'Diverge', 'No Witness'] #   ['S', 'T'], ['U'],
 def read_csv(f):
     with open(f) as f:
         return list(csv.reader(f))
+def read_csv_dict(f):
+    with open(f) as f:
+        return list(csv.DictReader(f))
 
 def autolabel(ax, rects):
     # attach some text labels
@@ -68,6 +71,93 @@ def plot_coverage(seminal, ucsd):
 
     # plt.show()
     fig.savefig('coverage.png')
+    plt.close()
+
+
+def plot_user_study():
+    a = read_csv_dict('study-data/A-Eric-scores.csv')
+    b = read_csv_dict('study-data/B-Eric-scores.csv')
+
+    def f(xs):
+        return [int(x) for x in xs if int(x) >= 0]
+
+    def err(xs):
+        p = np.average(xs)
+        return 100 * np.sqrt(p * (1-p) / len(xs))
+
+    ## REASON
+    sumlist_a = f([r['5: sumlist reason'] for r in a])
+    append_a = f([r['1: append reason'] for r in a])
+    digitsofint_a = f([r['3: digitsofint reason'] for r in a])
+    wwhile_a = f([r['7: wwhile reason'] for r in a])
+
+    sumlist_b = f([r['1: sumlist reason'] for r in b])
+    append_b = f([r['5: append reason'] for r in b])
+    digitsofint_b = f([r['7: digitsofint reason'] for r in b])
+    wwhile_b = f([r['3: wwhile reason'] for r in b])
+
+    ind = np.arange(4)
+    width = 0.35
+
+    fig = plt.figure()
+    p_o = plt.bar(ind,
+                  [100*np.average(sumlist_b), 100*np.average(append_a), 100*np.average(digitsofint_a), 100*np.average(wwhile_b)],
+                  width,
+                  color=COLORS[0],
+                  yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b]))
+    p_n = plt.bar(ind + width,
+                  [100*np.average(sumlist_a), 100*np.average(append_b), 100*np.average(digitsofint_b), 100*np.average(wwhile_a)],
+                  width,
+                  color=COLORS[1],
+                  yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a]))
[…]

### 2016-06-11 paper/icfp16-cameraready/study.R

@@ -42,6 +42,7 @@ reasons3 = c(a3$X1..append.reason,
              b3$X5..append.reason,
              b3$X7..digitsofint.reason)
 reasons = data.frame(reasons0, reasons1, reasons2, reasons3)
+reasons = reasons[reasons$reasons0 != -1 & reasons$reasons1 != -1 & reasons$reasons2 != -1 & reasons$reasons3 != -1,]
 
 fixes0 = c(a0$X2..append.fix,
            a0$X4..digitsofint.fix,
@@ -76,7 +77,7 @@ fixes3 = c(a3$X2..append.fix,
            b3$X6..append.fix,
            b3$X8..digitsofint.fix)
 fixes = data.frame(fixes0, fixes1, fixes2, fixes3)
-
+fixes = fixes[fixes$fixes0 != -1 & fixes$fixes1 != -1 & fixes$fixes2 != -1 & fixes$fixes3 != -1,]
 
 print("REASONS")
 print(table(reasons[,1]))
@@ -84,6 +85,7 @@ print(table(reasons[,2]))
 print(table(reasons[,3]))
 print(table(reasons[,4]))
 print(kappam.fleiss(reasons,detail=TRUE))
+#print(kappam.fleiss(reasons[1:3],detail=TRUE))
 
 print("FIXES")
 print(table(fixes[,1]))
@@ -91,3 +93,4 @@ print(table(fixes[,2]))
 print(table(fixes[,3]))
 print(table(fixes[,4]))
 print(kappam.fleiss(fixes,detail=TRUE))
+#print(kappam.fleiss(fixes[1:3],detail=TRUE))

### 2016-06-13 paper/icfp16-cameraready/evaluation.tex
update benchmark numbers

@@
 \caption{Results of our coverage testing and the distribution of test
-  outcomes. Our random search successfully finds witnesses for 83\% of
+  outcomes. Our random search successfully finds witnesses for 79--85\% of
   the programs in under one second, improving slightly to 87\% in under
@@
 \ucsdbench\ dataset (35 for the \uwbench\ dataset) with a maximum of
-2,745 (660 for \uwbench) and a median of 16 (also 16 for \uwbench).
+2,745 (986 for \uwbench) and a median of 17 (also 17 for \uwbench).
 %
 The average number of jumps per trace is 7 (also 7 for \uwbench) with a
-maximium of 353 (105 for \uwbench) and a median of 4 (also 4 for
+maximium of 353 (185 for \uwbench) and a median of 4 (also 4 for
 \uwbench).
@@
 Random test generation has difficulty generating highly constrained
-values, \eg red-black trees or a pair of equal integers. If the type
-error is hidden behind a complex branch condition \nanomaly may not be
+values, \eg\ red-black trees or a pair of equal integers. If the type
+error is hidden behind a complex branch condition \nanomaly\ may not be
 able to trigger it. Exhaustive testing and dynamic-symbolic execution
@@
 This code produces a nested list (with @n@ levels of nesting) containing
-a single copy of @x@, instead of a list with @n@ copies of @x@. \ocaml
+a single copy of @x@, instead of a list with @n@ copies of @x@. \ocaml\
 detects a cyclic \hbox{@'a = 'a list@} constraint in the recursive call
-and throws a type error, whereas \nanomaly happily % recurses @n@ times to
+and throws a type error, whereas \nanomaly\ happily % recurses @n@ times to
 produces the nested list.  Strictly speaking, this function itself cannot
@@
 %%% End:
+%!TEX root = main.tex

### 2016-06-13 paper/icfp16-cameraready/related.tex
update benchmark numbers

@@
 suffer from the opposite problem of providing \emph{too much}
-information, motivating recent work in ranking the program locations.
+information, motivating recent work in ranking the candidate locations.
 %

### 2016-06-15 paper/icfp16-cameraready/evaluation.tex
finish draft of user study

@@
 We have implemented a prototype of our search procedure and trace
-visualization for a purely functional subset of \ocaml\
---- with polymorphic types and records, but no modules, objects, or polymorphic variants ---
-in a tool called \nanomaly.
+visualization for a purely functional subset of \ocaml\ --- with
+polymorphic types and records, but no modules, objects, or polymorphic
+variants --- in a tool called \nanomaly.
 %
@@
 %
-There are two questions we seek to answer with our evaluation:
+There are three questions we seek to answer with our evaluation:
 %
@@
       How \emph{complex} are the traces produced by the witnesses?
+\item \emphbf{Witness Utility}
+      How \emph{helpful} are the witnesses and traces in debugging type errors?
 \end{enumerate}
@@
 \paragraph{Benchmarks}
-We answer these questions on two sets of ill-typed programs, \ie\
-programs that were rejected by the \ocaml\ compiler because of a
+We answer the first two questions on two sets of ill-typed programs,
+\ie\ programs that were rejected by the \ocaml\ compiler because of a
 type error.
@@
 
-\begin{figure*}[t]
+\begin{figure*}[ht]
 \centering
@@
 
-\subsection{Trace Complexity}
+\subsection{Witness Complexity}
 \label{sec:trace-complexity}
@@
 %
-\begin{figure*}[t]
+\begin{figure*}[ht]
 \centering
@@
 
-\subsection{User Study}
+\subsection{Measuring Witness Utility}
 \label{sec:user-study}
@@
 %
-We added four questions to an exam in the Spring session of UVA's
-undergraduate Programming Languages course (CS XXX), in which presented
-the students with ill-typed \ocaml\ programs and asked them to
+We included four problems in an exam in the Spring session of UVA's
+undergraduate Programming Languages course (CS XXX).
+%
+The problems presented the students with ill-typed \ocaml\ programs and
+asked them to
 %
@@
 %
-The four problems assigned to the students were the
+The four problems assigned to the students were the @sumList@,
+@digitsOfInt@, and @wwhile@ programs from \S~\ref{sec:advantage-traces},
+as well as the following @append@ program
 %
-@sumList@, @digitsOfInt@, and @wwhile@ programs from \S~\ref{sec:advantage-traces}
-as well as the following @append@ program....\ES{HERE}
-
+\begin{ecode}
+  let append x l =
+    match x with
+    | []   -> l
+    | h::t -> h :: t :: l
+\end{ecode}
+%
+which triggers an occurs-check error in the ``cons'' case.
+%
 We then instructed four annotators (one of whom is an author, the other
-three were teaching assistants at UCSD) to classify the answers as
+three are teaching assistants at UCSD) to classify the answers as
 correct or incorrect.
@@
 %
-As we had more than two annotators assigning nominal ratings we used
-Fleiss' kappa~\cite{XXX} to measure IRR.
+As we had more than two annotators assigning nominal (``correct'' or
+``incorrect'') ratings we used Fleiss' kappa~\cite{Fleiss1971-du} to
+measure IRR.\@
 %
-The resulting kappa values, $\kappa = 0.72$ for the explanations and
-$\kappa = 0.83$ for the fixes, indicate substantial agreement among
-the annotators~\cite{XXX}.
+Fleiss' kappa is measured on a scale from $1$, indicating total
+agreement, to $-1$, indicating total disagreement, with $0$ indicating
+random agreement.
 
-The results of one of the annotators are shown in
-Figure~\ref{fig:results-user-study}.
+\paragraph{Results}
 %
-\begin{figure*}[t]
+Figure~\ref{fig:results-user-study} summarizes a single annotator's
+results, which show that students given \toolname's jump-compressed
+trace were consistently more likely to correctly explain and fix the
+type error than those given \ocaml's error message.
+%
+The measured kappa values were $\kappa = 0.72$ for the explanations and
+$\kappa = 0.83$ for the fixes; while there is no formal notion for what
+consititutes strong agreement~\cite{Krippendorff2012-wd}, kappa values
+above $0.60$ are often called ``substantial''
+agreement~\cite{Landis1977-ey}.
+%
+\begin{figure*}[ht]
 \centering
@@
 \includegraphics[width=0.49\linewidth]{user-study-fix.png}
-\caption{Results of the user study. The students given \toolname's
-  jump-compressed trace consistently scored better than those
-  given \ocaml's type error.}
+\caption{A classification of students' explanations and fixes for type
+  errors, given either \ocaml's error % message
+  or \toolname's
+  jump-compressed trace. The students given \toolname's jump-compressed
+  trace consistently scored better than those given \ocaml's type
+  error.}
 \label{fig:results-user-study}
@@
 %
-The students given \toolname's jump-compressed trace were consistently
-more likely to correctly explain and fix the type error than those
-given \ocaml's error message.
 
@@
 to type errors: (1) with high coverage in a timespan amenable to
-compile-time analysis, and (2) with traces that have a low average
-complexity of 7 jumps.
+compile-time analysis, (2) with traces that have a low average
+complexity of 7 jumps, and (3) that are more helpful to novice
+programmers than traditional type error messages.
 
@@
   let rec replicate n x =
-    if n <= 0 then []
-              else replicate (n-1) [x]
+    if n <= 0 then
+      []
+    else
+      replicate (n-1) [x]
 \end{code}

### 2016-06-15 paper/icfp16-cameraready/gallery2.tex
finish draft of user study

@@
-\subsection{The Advantage of Traces}\label{sec:advantage-traces}
+\subsection{Witness Utility}\label{sec:advantage-traces}
 
@@
 Both of the above are parts of the full story, which
-is succinctly summarized by \toolname's trace showing
+is summarized by \toolname's trace showing
 how @sumList [1; 2]@ gets stuck at @2 + []@.
@@
 
+\pagebreak
 \paragraph{Example: Bad Helper Function that Type-Checks}

### 2016-06-15 paper/icfp16-cameraready/plots.py

@@ -104,12 +104,14 @@ def plot_user_study():
                   [100*np.average(sumlist_b), 100*np.average(append_a), 100*np.average(digitsofint_a), 100*np.average(wwhile_b)],
                   width,
                   color=COLORS[0],
-                  yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b]))
+                  # yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b])
+    )
     p_n = plt.bar(ind + width,
                   [100*np.average(sumlist_a), 100*np.average(append_b), 100*np.average(digitsofint_b), 100*np.average(wwhile_a)],
                   width,
                   color=COLORS[1],
-                  yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a]))
+                  # yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a])
+    )
 
     plt.title('Explanation',fontsize=24)
     # plt.xlabel('Problem', fontsize=20)
@@ -142,12 +144,14 @@ def plot_user_study():
                   [100*np.average(sumlist_b), 100*np.average(append_a), 100*np.average(digitsofint_a), 100*np.average(wwhile_b)],
                   width,
                   color=COLORS[0],
-                  yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b]))
+                  # yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b])
+    )
     p_n = plt.bar(ind + width,
                   [100*np.average(sumlist_a), 100*np.average(append_b), 100*np.average(digitsofint_b), 100*np.average(wwhile_a)],
                   width,
                   color=COLORS[1],
-                  yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a]))
+                  # yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a])
+    )
 
     plt.title('Fix',fontsize=24)
     # plt.xlabel('Problem', fontsize=20)

### 2016-06-18 paper/icfp16-cameraready/evaluation.tex
number of exams
add threats to validity

@@
   outcomes. Our random search successfully finds witnesses for 79--85\% of
-  the programs in under one second, improving slightly to 87\% in under
+  the programs in under one second, improving to 87\% in under
   10 seconds. In both datasets we detect actual type errors about 82\%
@@
 While a more advanced search procedure, \eg\ dynamic-symbolic execution,
-could likely trigger more of the type errors, our experiments show that
+could likely trigger more of the type errors, our experiments suggest that
 type errors are coarse enough (or that novice programs are \emph{simple}
@@
 %
-The problems presented the students with ill-typed \ocaml\ programs and
+We presented the students with ill-typed \ocaml\ programs and
 asked them to
@@
 %
-For each problem the student would be given the ill-typed program and
+For each problem the student was given the ill-typed program and
 either \ocaml's error message or \toolname's jump-compressed trace.
@@
 %
-which triggers an occurs-check error in the ``cons'' case.
+which triggers an occurs-check error on line 4.
 %
@@
 
-\paragraph{Results}
+\paragraph{Threats to Validity}
+\subparagraph{Construct}
+%
+Measuring understanding is a difficult task; we used the correctness of
+the student's explanation of, and fix for, the type error as a proxy for
+her understanding, but it is possible that other metrics would produce
+different results.
+
+\subparagraph{Internal}
+%
+We assigned students randomly to two groups. The first group was given
+\ocaml's errors for @append@ and \hbox{@digitsOfInt@,} and \toolname's trace
+for @sumList@ and @wwhile@. The second group was given the opposite
+assignment of errors and traces. This assignment ensured that (1) each
+student was given \ocaml and \toolname problems, and (2) each student
+was given an ``easy'' and ``hard'' problem for both \ocaml and
+\toolname. Students without sufficient knowledge of \ocaml could affect
+the results, as could the time-constrained nature of an exam. For these
+reasons we excluded any answers left blank from our analysis.
+
+\subparagraph{External}
+%
+Our experiment is based on students in the process of learning \ocaml,
+and thus may not generalize to all developers. The four
+programs we used were chosen manually, via a random selection and
+filtering of the programs in the \ucsdbench dataset. In some cases we made
+minor simplifying edits (\eg alpha-renaming, dead-code removal) to the
+programs to make them more understandable in the short timeframe of an
+exam; however, we never altered the resulting type-error. A different
+selection of programs may lead to different results.
+
+\subparagraph{Conclusion}
+%
+We collected exams from 60 students, though due to the nature of the
+study not every student completed every problem. The number of
+incomplete submissions ranges from 2 (for the \ocaml version of
+@sumList@) to 17 (for the \toolname version of @wwhile@). Collecting
+more responses per test pair was not possible, as it would require
+having students answer the same problem twice (once with \ocaml and once
+with \toolname).
+
+\subsubsection{Results}
 %

### 2016-06-18 paper/icfp16-cameraready/study.R

@@ -42,6 +42,7 @@ reasons3 = c(a3$X1..append.reason,
              b3$X5..append.reason,
              b3$X7..digitsofint.reason)
 reasons = data.frame(reasons0, reasons1, reasons2, reasons3)
+## filter out answers that any annotator marked as INCOMPLETE
 reasons = reasons[reasons$reasons0 != -1 & reasons$reasons1 != -1 & reasons$reasons2 != -1 & reasons$reasons3 != -1,]
 
 fixes0 = c(a0$X2..append.fix,
@@ -77,6 +78,7 @@ fixes3 = c(a3$X2..append.fix,
            b3$X6..append.fix,
            b3$X8..digitsofint.fix)
 fixes = data.frame(fixes0, fixes1, fixes2, fixes3)
+## filter out answers that any annotator marked as INCOMPLETE
 fixes = fixes[fixes$fixes0 != -1 & fixes$fixes1 != -1 & fixes$fixes2 != -1 & fixes$fixes3 != -1,]
 
 print("REASONS")

### 2016-06-18 paper/icfp16-cameraready/abstract.tex
mention user study in abstract and intro

@@
   %
-  Third, we evaluate our approach on two data sets
+  Third, we evaluate the coverage of our approach on two data sets
   comprising over 4,500 ill-typed student programs.
@@
   yields small counterexamples for 81\% of the witnesses.
+  %
+  Finally, we evaluate the utility of our witnesses in
+  helping students understand and fix type errors, and
+  find that students presented with our witnesses
+  consistently show a greater understanding of type errors
+  than those presented with a standard error message.
 \end{abstract}

### 2016-06-18 paper/icfp16-cameraready/introduction.tex
mention user study in abstract and intro

@@
  File "fac.ml", line 5, characters 8-17:
- Error: This expression has type bool
-        but an expression was expected
-        of type int
+ Error: This expression has type bool but an
+        expression was expected of type int.
 \end{verbatim}
@@
 to properly \emph{localize} a type error~\cite{Lerner2007-dt,Chen2014-gd,Zhang2014-lv,Pavlinovic2014-mr},
-but an accurate source location still does not explain \emph{why} the
+but an accurate source location does not explain \emph{why} the
 program is wrong.
@@
 Our semantics \emph{conservatively} instantiates holes with concrete
-values; thereby dynamically inferring the type of the input
+values, dynamically inferring the type of the input
 until the program goes wrong.
@@
 %
-A jump-compressed trace includes only big steps to the next function
-call or return, for example the trace in the bottom-left of
-Figure~\ref{fig:factorial} is a jump-compressed trace.
+A jump-compressed trace includes only function
+calls and returns, for example the trace in the bottom-left of
+Figure~\ref{fig:factorial} is jump-compressed.
 %
-Jump-compressed traces are similar to stack traces in that both show a
+Jump-compressed traces are similar to stack traces, both show a
 sequence of function calls that lead to a crash, but the jump-compressed
@@
 
+\subparagraph{Witness Utility}
+%
+Even if we can find small witnesses for the majority of type errors, it
+may be that the witnesses do not actually help developers
+\emph{understand} the errors.
+%
+In other words, perhaps the static error message is sufficient to
+diagnose and fix the error, or perhaps the witness simply does not add
+enough information to make a difference.
+%
+Thus, our final contribution is a user study that compares the utility
+of our witnesses with that of the error messages provided by the \ocaml
+compiler~(\S~\ref{sec:user-study}).
+%
+We show that students given one of our witnesses are consistently more
+likely to correctly explain and fix a type error than those given
+the standard error message produced by the \ocaml compiler.
+
 \smallskip
 Our results show that in the vast majority of
-cases, (novices') ill-typed programs \emph{do} go wrong.
+cases, (novices') ill-typed programs \emph{do} go wrong,
+and that the witnesses to these errors can be helpful in
+understanding the source of the error.
+%
 This, in turn, opens the door to a novel dynamic way to

### 2016-06-18 paper/icfp16-cameraready/related.tex
mention user study in abstract and intro

@@
 In contrast to these approaches, we do not attempt to localize or fix
-the type error. Instead we try to explain it to the novice user using a
-dynamic witness that demonstrates how their program is not just
+the type error. Instead we try to explain it to the user using a
+dynamic witness that demonstrates how the program is not just
 ill-typed but truly wrong. In addition, allowing users to run their
-program (even knowing that it is wrong) enables experimentation and the
+program % (even knowing that it is wrong)
+enables experimentation and the
 use of debuggers to step through the program and investigate its

### 2016-06-18 paper/icfp16-cameraready/introduction.tex
add NSF acks

@@
 
-\subparagraph{Witness Utility}
+The ultimate purpose of an error report is to assist the programmer 
+in \emph{understanding} and \emph{fixing} problematic code.
+%
+Thus, our final contribution is a user study that compares \ocaml's 
+errors and our dynamic witnesses against \ocaml's type errors 
+along this dimension~(\S~\ref{sec:user-study}).
+%
+We show that students given one of our witnesses are consistently more
+likely to correctly explain and fix a type error than those given
+the standard error message produced by the \ocaml compiler.
+
+% \subparagraph{Witness Utility}
 %
@@
 %
+
 Thus, our final contribution is a user study that compares the utility

### 2016-06-21 paper/icfp16-cameraready/abstract.tex
tweaks based on wes comments

--- 2016-06-07 paper/icfp16-cameraready/abstract.tex
@@
   yields small counterexamples for 81\% of the witnesses.
+  %
+  Finally, we evaluate the utility of our witnesses in
+  helping students understand and fix type errors, and
+  find that students presented with our witnesses
+  show a greater understanding of type errors
+  than those presented with a standard error message.
 \end{abstract}

### 2016-06-21 paper/icfp16-cameraready/evaluation.tex
tweaks based on wes comments

@@
 \item \emphbf{Witness Utility}
-      How \emph{helpful} are the witnesses and traces in debugging type errors?
+      How \emph{helpful} (qualitatively and quantitatively)
+      are the witnesses and traces in debugging type errors?
 \end{enumerate}
@@
 The first dataset comes from the Spring 2014 undergraduate Programming
-Languages course at UC San Diego.
+Languages (CSE 130) course at UC San Diego.
 %
@@
 polymorphic data types, and expression evaluators.
+
+We answer the third question in two steps.
+%
+First, we present a qualitative evaluation of \toolname's traces on a
+selection of programs drawn from the UCSD dataset.
+%
+Second, we present a quantitative user study of students in the
+University of Virginia's Spring 2016 undergraduate Programming Languages
+(CS 4501) course.
+%
+As part of an exam, we presented the students with ill-typed \ocaml\
+programs and asked them to
+%
+(1) \emph{explain} the type error, and
+%
+(2) \emph{fix} the type error (IRB \#2014009900).
+%
+For each problem the students were given the ill-typed program and
+either \ocaml's error message or \toolname's jump-compressed trace.
 
@@
 
+\subsection{Witness Utility}
 \input{gallery2}
 
-\subsection{Measuring Witness Utility}
+\subsubsection{Quantitative}
 \label{sec:user-study}
-Finally, to test the explanatory power of our jump-compressed traces, we
-ran a user study (IRB \#XXXX) at the University of Virginia (UVA).
+% Finally, to test the explanatory power of our jump-compressed traces, we
+% ran a user study (IRB \#2014009900) at the University of Virginia (UVA).
+% %
+% We included four problems in an exam in the Spring session of UVA's
+% undergraduate Programming Languages course (CS 4501).
+% %
+% We presented the 60 students in the course with ill-typed \ocaml\
+% programs and asked them to
+% %
+% (1) \emph{explain} the type error, and
+% %
+% (2) \emph{fix} the type error.
+% %
+% For each problem the student was given the ill-typed program and
+% either \ocaml's error message or \toolname's jump-compressed trace.
 %
-We included four problems in an exam in the Spring session of UVA's
-undergraduate Programming Languages course (CS XXX).
-%
-We presented the students with ill-typed \ocaml\ programs and
-asked them to
-%
-(1) \emph{explain} the type error, and
-%
-(2) \emph{fix} the type error.
-%
-For each problem the student was given the ill-typed program and
-either \ocaml's error message or \toolname's jump-compressed trace.
-%
-The four problems assigned to the students were the @sumList@,
-@digitsOfInt@, and @wwhile@ programs from \S~\ref{sec:advantage-traces},
-as well as the following @append@ program
+We assigned four problems to the ($n=60$) students in the course: the
+@sumList@, @digitsOfInt@, and @wwhile@ programs from
+\S~\ref{sec:advantage-traces}, as well as the following @append@ program
 %
@@
 %
+For each problem the students were given the ill-typed program and
+either \ocaml's error message or \toolname's jump-compressed trace.
+%
+Due to the nature of an in-class exam, not every student answered every
+question; we received between 13 and 28 (out of a possible 30) responses
+for each problem-tool pair.
+
 We then instructed four annotators (one of whom is an author, the other
@@
 correct or incorrect.
-
+%
 We performed an inter-rater reliability (IRR) analysis to determine the
@@
 We collected exams from 60 students, though due to the nature of the
-study not every student completed every problem. The number of
-incomplete submissions ranges from 2 (for the \ocaml version of
-@sumList@) to 17 (for the \toolname version of @wwhile@). Collecting
-more responses per test pair was not possible, as it would require
-having students answer the same problem twice (once with \ocaml and once
-with \toolname).
+study not every student completed every problem.
+%
+The number of complete submissions ranges from 13 (for the \toolname
+version of @wwhile@) to 28 (for the \ocaml version of @sumList@), out of
+a maximum of 30 per program-tool pair.
+%
+Collecting more responses per test pair was not possible, as it
+would require having students answer the same problem twice (once with
+\ocaml and once with \toolname).
 
-\subsubsection{Results}
+\paragraph{Results}
 %
@@
 results, which show that students given \toolname's jump-compressed
-trace were consistently more likely to correctly explain and fix the
-type error than those given \ocaml's error message.
+trace were consistently more likely to correctly explain
+and fix the type error than those given \ocaml's error message.
+%
+Across each problem the \toolname responses were marked correct
+$10-30\%$ more often than the \ocaml responses, which suggests that
+the students who had access to \toolname's traces had a better
+understanding of the type errors.
 %
@@
   jump-compressed trace. The students given \toolname's jump-compressed
-  trace consistently scored better than those given \ocaml's type
+  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
   error.}

### 2016-06-21 paper/icfp16-cameraready/plots.py

@@ -99,6 +99,11 @@ def plot_user_study():
     ind = np.arange(4)
     width = 0.35
 
+    print([np.average(sumlist_a) - np.average(sumlist_b),
+           np.average(append_b) - np.average(append_a),
+           np.average(digitsofint_b) - np.average(digitsofint_a),
+           np.average(wwhile_a) - np.average(wwhile_b)])
+
     fig = plt.figure()
     p_o = plt.bar(ind,
                   [100*np.average(sumlist_b), 100*np.average(append_a), 100*np.average(digitsofint_a), 100*np.average(wwhile_b)],

### 2016-06-21 paper/icfp16-cameraready/introduction.tex

--- 2016-06-18 paper/icfp16-cameraready/introduction.tex
@@
 
-\subparagraph{Witness Utility}
+The ultimate purpose of an error report is to help the programmer 
+\emph{comprehend} and \emph{fix} problematic code.
 %
-Even if we can find small witnesses for the majority of type errors, it
-may be that the witnesses do not actually help developers
-\emph{understand} the errors.
+Thus, our final contribution is a user study that compares \toolname's 
+dynamic witnesses against \ocaml's type errors along the dimension of
+comprehensibility~(\S~\ref{sec:user-study}).
 %
-In other words, perhaps the static error message is sufficient to
-diagnose and fix the error, or perhaps the witness simply does not add
-enough information to make a difference.
+Our study finds that students given one of our witnesses are 
+consistently more likely to correctly explain and fix a type
+error than those given the standard error message produced by 
+the \ocaml compiler.
+
 %
-Thus, our final contribution is a user study that compares the utility
-of our witnesses with that of the error messages provided by the \ocaml
-compiler~(\S~\ref{sec:user-study}).
+% \subparagraph{Witness Utility}
 %
-We show that students given one of our witnesses are consistently more
-likely to correctly explain and fix a type error than those given
-the standard error message produced by the \ocaml compiler.
+% Even if we can find small witnesses for the majority of type errors, it
+% may be that the witnesses do not actually help developers
+% \emph{understand} the errors.
+%
+% In other words, perhaps the static error message is sufficient to
+% diagnose and fix the error, or perhaps the witness simply does not add
+% enough information to make a difference.
+%
+%
+% Thus, our final contribution is a user study that compares the utility
+% of our witnesses with that of the error messages provided by the \ocaml
+% compiler~(\S~\ref{sec:user-study}).
+%
 
 \smallskip
-Our results show that in the vast majority of
-cases, (novices') ill-typed programs \emph{do} go wrong,
-and that the witnesses to these errors can be helpful in
-understanding the source of the error.
-%
-This, in turn, opens the door to a novel dynamic way to
-explain, understand, and appreciate the benefits of static typing.
+All together, our results show that in the vast majority of cases, (novices') ill-typed 
+programs \emph{do} go wrong, and that the witnesses to these errors can be 
+helpful in understanding the source of the error. This, in turn, opens the 
+door to a novel dynamic way to explain, understand, and appreciate the 
+benefits of static typing.
 

### 2016-06-22 paper/icfp16-cameraready/gallery2.tex
tweaks based on wes comments
try placing all traces in sec:advantage-traces in a single figure

@@
-\subsection{Witness Utility}\label{sec:advantage-traces}
+\subsubsection{Qualitative}\label{sec:advantage-traces}
 
@@
 we just underline the relevant code.}
-and (4)~the jump-compressed trace produced by \toolname.
+and (4)~the jump-compressed trace produced by \toolname (in Figure~\ref{fig:traces}).
 
@@
 % \centering
+
+\begin{figure*}[htp]
+  \centering
+  \includegraphics[height=125px]{sqsum.png}
+  \includegraphics[height=125px]{sumlist.png}
+  \includegraphics[height=150px]{digitsOfInt.png}
+  \includegraphics[height=125px]{wwhile.png}
+  \caption{(Left to right) Jump-compressed traces showing how
+    \texttt{sqsum}, \texttt{sumList}, \texttt{digitsOfInt}, and
+    \texttt{wwhile} go wrong in \S~\ref{sec:advantage-traces}.}
+  \label{fig:traces}
+\end{figure*}
 
@@
 %
-\toolname\ produces the following trace showing how the evaluation of
-@sqsum [1]@ gets stuck:
+\toolname\ produces a trace (see Figure~\ref{fig:traces}) showing how the evaluation of
+@sqsum [1]@ gets stuck.
 %
-\begin{center}
-  \includegraphics[height=125px]{sqsum.png}
-\end{center}
+% \begin{center}
+%   \includegraphics[height=125px]{sqsum.png}
+% \end{center}
 %
-The figure highlights the entire stuck term
+The trace highlights the entire stuck term
 (not just the recursive call), emphasizing
@@
 is summarized by \toolname's trace showing
-how @sumList [1; 2]@ gets stuck at @2 + []@.
+how @sumList [1; 2]@ gets stuck at @2 + []@ (see Figure~\ref{fig:traces}).
 %
-\begin{center}
-  \includegraphics[height=125px]{sumlist.png}
-\end{center}
+% \begin{center}
+%   \includegraphics[height=125px]{sumlist.png}
+% \end{center}
 %
-The trace clarifies immediately (via the third step)
+The trace clarifies (via the third step)
 that the @[]@ is the result of the recursive call
@@
 %
+%\pagebreak
 Unfortunately, the student's @append@ function \emph{conses} an element
@@
 \toolname, in contrast, makes no assumptions about @append@ and produces
-a trace that illustrates the true error on line 4, by
+a trace (see Figure~\ref{fig:traces}) that illustrates the true error on line 4, by
 highlighting the conflict in consing a list onto a list of integers.
 %
-\begin{center}
-  \includegraphics[height=175px]{digitsOfInt.png}
-\end{center}
+% \begin{center}
+%   \includegraphics[height=175px]{digitsOfInt.png}
+% \end{center}
 %
@@
 %
-\toolname\ synthesizes a trace that draws the eye to the
+\toolname\ synthesizes a trace (see Figure~\ref{fig:traces}) that draws the eye to the
 true error: the @match@ expression on line 2, and highlights
@@
 %
-\begin{center}
-  \includegraphics[height=150px]{wwhile.png}
-\end{center}
+% \begin{center}
+%   \includegraphics[height=150px]{wwhile.png}
+% \end{center}
 %

### 2016-06-22 paper/icfp16-cameraready/gallery2.tex
actually, let's group all of the graphs
leave the traces inline

--- 2016-06-15 paper/icfp16-cameraready/gallery2.tex
@@
 
+% \begin{figure*}[htp]
+%   \centering
+%   \includegraphics[height=125px]{sqsum.png}
+%   \includegraphics[height=125px]{sumlist.png}
+%   \includegraphics[height=150px]{digitsOfInt.png}
+%   \includegraphics[height=125px]{wwhile.png}
+%   \caption{(Left to right) Jump-compressed traces showing how
+%     \texttt{sqsum}, \texttt{sumList}, \texttt{digitsOfInt}, and
+%     \texttt{wwhile} go wrong in \S~\ref{sec:advantage-traces}.}
+%   \label{fig:traces}
+% \end{figure*}
+
 \paragraph{Example: Recursion with Bad Operator}
@@
 %
-\toolname\ produces the following trace showing how the evaluation of
-@sqsum [1]@ gets stuck:
+\toolname\ produces a trace showing how the evaluation of
+@sqsum [1]@ gets stuck.
 %
@@
 %
-The figure highlights the entire stuck term
+The trace highlights the entire stuck term
 (not just the recursive call), emphasizing
@@
 %
-The trace clarifies immediately (via the third step)
+The trace clarifies (via the third step)
 that the @[]@ is the result of the recursive call
@@
 
-\pagebreak
+%\pagebreak
 \paragraph{Example: Bad Helper Function that Type-Checks}

### 2016-06-23 paper/icfp16-cameraready/evaluation.tex
try placing all traces in sec:advantage-traces in a single figure
actually, let's group all of the graphs
leave the traces inline
done

@@
 idioms including (tail) recursive functions, higher-order functions,
-polymorphic data types, and expression evaluators.
+polymorphic and algebraic data types. % and expression evaluators.
 
@@
 
-\begin{figure*}[ht]
+\begin{figure*}[htp]
 \centering
@@
 \centering
-\includegraphics[width=0.49\linewidth]{coverage.png}
+\includegraphics[width=0.45\linewidth]{coverage.png}
 % \end{minipage}
 % \begin{minipage}{\linewidth}
-\includegraphics[width=0.49\linewidth]{distrib.png}
+\includegraphics[width=0.45\linewidth]{distrib.png}
 \end{minipage}
@@
 find a witness, we measure the complexity of the generated
-trace according to two metrics.
+trace using two metrics.
 
@@
 \begin{enumerate}
-\item \emphbf{Single-step Metric} The size of the trace after expanding
+\item \emphbf{Single-step:} The size of the trace after expanding
   all of the single-step edges from the witness to the stuck term, and
@@
   % complexity, \ie ``How big is the fully-expanded trace?''
-\item \emphbf{Jump-compressed Metric} The size of the jump-compressed trace.
+\item \emphbf{Jump-compressed:} The size of the jump-compressed trace.
 \end{enumerate}
@@
 %
-\begin{figure*}[ht]
+\begin{figure*}[htp]
 \centering
-\includegraphics[width=0.49\linewidth]{trace_size_step.png}
-\includegraphics[width=0.49\linewidth]{trace_size_jump.png}
+\includegraphics[width=0.45\linewidth]{trace_size_step.png}
+\includegraphics[width=0.45\linewidth]{trace_size_jump.png}
 \caption{Complexity of the generated traces. 81\% of the combined traces
@@
 %
+\begin{figure*}[htp]
+\centering
+\includegraphics[width=0.45\linewidth]{user-study-reason.png}
+\includegraphics[width=0.45\linewidth]{user-study-fix.png}
+\caption{A classification of students' explanations and fixes for type
+  errors, given either \ocaml's error % message
+  or \toolname's
+  jump-compressed trace. The students given \toolname's jump-compressed
+  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
+  error.}
+\label{fig:results-user-study}
+\end{figure*}
 
@@
 In both datasets 80\% or more traces have at most 10 jumps.
+
+\subsection{Qualitative Evaluation of Witness Utility}\label{sec:advantage-traces}
+
+Next, we present a \emph{qualitative} evaluation that compares
+the explanations provided by \toolname's dynamic witnesses with
+the static reports produced by the \ocaml\ compiler and \sherrloc,
+a state-of-the-art fault localization approach~\cite{Zhang2014-lv}.
+%
+In particular, we illustrate, using a series of examples drawn
+from student programs in the \ucsdbench\ dataset, how \toolname's
+jump-compressed traces can get to the heart of the error. Our approach
+%
+highlights the conflicting values that cause the program to get
+stuck, rather that blaming a single one,
+%
+shows the steps necessary to reach the stuck state, and
+%
+does not assume that a function is correct just because it type-checks.
+%
+For each example we will present
+(1)~the code,
+(2)~the error message returned \ocaml,
+(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),\footnote{When the locations from \ocaml\ and \sherrloc\ overlap,
+we just underline the relevant code.}
+and (4)~the jump-compressed trace produced by \toolname.
+
+% \begin{figure*}[ht]
+% \centering
+% \begin{minipage}{0.49\linewidth}
+% \centering
+
+% \begin{figure*}[htp]
+%   \centering
+%   \includegraphics[height=125px]{sqsum.png}
+%   \includegraphics[height=125px]{sumlist.png}
+%   \includegraphics[height=150px]{digitsOfInt.png}
+%   \includegraphics[height=125px]{wwhile.png}
+%   \caption{(Left to right) Jump-compressed traces showing how
+%     \texttt{sqsum}, \texttt{sumList}, \texttt{digitsOfInt}, and
+%     \texttt{wwhile} go wrong in \S~\ref{sec:advantage-traces}.}
+%   \label{fig:traces}
+% \end{figure*}
+
+\paragraph{Example: Recursion with Bad Operator}
+The recursive function @sqsum@ should square each
+element of the input list and then compute the sum
+of the result.
+%
+\begin{ecode}
+  let rec sqsum xs = match xs with
+    | [] -> 0
+    | h::t -> __(sqsum t)__ @ (h * h)
+\end{ecode}
+%
+Unfortunately the student has used the list-append
+operator |@| instead of \texttt{+} to compute the sum.
+%
+Both \ocaml\ and \sherrloc\ blame the \emph{wrong location},
+the recursive call @sqsum t@, with the message
+%
+\begin{verbatim}
+  This expression has type
+    int
+  but an expression was expected of type
+    'a list
+\end{verbatim}
+%
+\toolname\ produces a trace showing how the evaluation of
+@sqsum [1]@ gets stuck.
+%
+\begin{center}
+  \includegraphics[height=125px]{sqsum.png}
+\end{center}
+%
+The trace highlights the entire stuck term
+(not just the recursive call), emphasizing
+the \emph{conflict} between @int@ and @list@
+rather than assuming one or the other is correct.
+
+\paragraph{Example: Recursion with Bad Base Case}
+%
+The function @sumList@ should add up
+the elements of its input list.
+%
+\begin{ecode}
+  let rec sumList xs = match xs with
+    | []    -> ==[]==
+    | y::ys -> y + __sumList ys__
+\end{ecode}
+%
+Unfortunately, in the base case, it returns @[]@
+instead of @0@.
+%
+\sherrloc\ blames the base case, and \ocaml\
+assumes the base case is correct and blames
+the \emph{recursive call} on line 3:
+%
+\begin{verbatim}
+  This expression has type
+    'a list
+  but an expression was expected of type
+    int
+\end{verbatim}
+%
+Both of the above are parts of the full story, which
+is summarized by \toolname's trace showing
+how @sumList [1; 2]@ gets stuck at @2 + []@.
+%
+\begin{center}
+  \includegraphics[height=125px]{sumlist.png}
+\end{center}
+%
+The trace clarifies (via the third step)
+that the @[]@ is the result of the recursive call
+@sumList []@, and shows how it is incompatible with
+the subsequent \texttt{+} operation.
+
+%% ES: append is actually a bit problematic as we don't find the nice
+%% append [1] [2] witness. instead we could find something like
+%% append [_] [], but it's not as clear IMO
+% Our next example is the @append@ function, which should concatenate the
+% two input lists.
+% %
+% \begin{ecode}
+% let append xs ys = match xs with
+%   | []   -> ys
+%   | h::t -> h :: __t__ :: ys
+% \end{ecode}
+% %
+% The student has forgotten to make a recursive call to @append@, and
+% instead tries to cons the tail @t@ directly onto the second list @ys@.
+% Consing @h@ back onto the result causes \ocaml to attempt to construct
+% the infinite type @'a = 'a list@, triggering an \emph{occurs-check}
+% error.
+% %
+% \begin{verbatim}
+% Error: This expression has type
+%          'a list
+%        but an expression was expected of type
+%          'a
+%        The type variable 'a occurs inside 'a list
+% \end{verbatim}
+% %
+% %
+% \begin{center}
+%   \includegraphics[height=75px]{append.png}
+% \end{center}
+
+%\pagebreak
+\paragraph{Example: Bad Helper Function that Type-Checks}
+%
+The function @digitsOfInt@ should return a list of
+the digits of the input integer.
+%
+\begin{ecode}
+  let append x xs =
+    match xs with
+    | [] -> [x]
+    | _  -> x :: xs
+
+  let rec digitsOfInt n =
+    if n <= 0 then
+      []
+    else
+      append (==digitsOfInt (n / 10)==)
+             [__n mod 10__]
+\end{ecode}
+%
+%\pagebreak
+Unfortunately, the student's @append@ function \emph{conses} an element
+onto a list instead of appending two lists.
+%
+Though incorrect, @append@ still type-checks and thus \ocaml and
+\sherrloc blame the \emph{use-site} on line 10.
+%
+\begin{verbatim}
+  This expression has type
+    int
+  but an expression was expected of type
+    'a list
+\end{verbatim}
+%
+In contrast, \toolname makes no assumptions about @append@
+yielding a trace that illustrates the true error on line 4, by
+highlighting the conflict in consing a list onto a list of integers.
+%
+\begin{center}
+  \includegraphics[height=160px]{digitsOfInt.png}
+\end{center}
 %
 
-\subsection{Witness Utility}
-\input{gallery2}
+\paragraph{Example: Higher-Order Functions}
+%
+The higher-order function @wwhile@ is supposed
+to emulate a traditional while-loop. It takes
+a function @f@ and repeatedly calls @f@ on the
+first element of its output pair, starting with
+the initial @b@, till the second element is @false@.
+%
+\begin{ecode}
+  let rec wwhile (f,b) =
+    match f with
+    | (z, false) -> z
+    | (z, true)  -> wwhile (f, z)
 
-\subsubsection{Quantitative}
+  let f x =
+    let xx = x * x in
+    (xx, (xx < 100))
+
+  let _ = wwhile (__f__, 2)
+\end{ecode}
+%
+The student has forgotten to \emph{apply} @f@ at all on line 2,
+and just matches it directly against a pair.
+This faulty @wwhile@ definition nevertheless typechecks,
+and is assumed to be correct by both \ocaml\ and \sherrloc\
+which blame the use-site on line 10.
+%
+\begin{verbatim}
+  This expression has type
+    int -> int * bool
+  but an expression was expected of type
+    'a * bool
+\end{verbatim}
+%
+\toolname\ synthesizes a trace that draws the eye to the
+true error: the @match@ expression on line 2, and highlights
+the conflict in matching a function against a pair pattern.
+%
+\begin{center}
+  \includegraphics[height=135px]{wwhile.png}
+\end{center}
+%
+By highlighting conflicting values, \ie\ the source and sink
+of the problem, and not making assumption about function correctness, \toolname\
+focusses the user's attention on the piece of code that is
+actually relevant to the error.
+
+\subsection{Quantitative Evaluation of Witness Utility}
 \label{sec:user-study}
@@
 \paragraph{Threats to Validity}
+Measuring understanding is a difficult task; the following summarize
+the threats to the validity of our results.
+
 \subparagraph{Construct}
 %
-Measuring understanding is a difficult task; we used the correctness of
-the student's explanation of, and fix for, the type error as a proxy for
-her understanding, but it is possible that other metrics would produce
-different results.
+We used the correctness of the student's explanation of, and fix for,
+the type error as a proxy for her understanding, but it is possible
+that other metrics would produce different results.
 
@@
 %
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=0.49\linewidth]{user-study-reason.png}
-\includegraphics[width=0.49\linewidth]{user-study-fix.png}
-\caption{A classification of students' explanations and fixes for type
-  errors, given either \ocaml's error % message
-  or \toolname's
-  jump-compressed trace. The students given \toolname's jump-compressed
-  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
-  error.}
-\label{fig:results-user-study}
-\end{figure*}
 %

### 2016-06-23 paper/icfp16-cameraready/introduction.tex
done

@@
 %
-This problem is surprisingly tricky when we cannot rely on
-static type information.
+This problem is tricky when we cannot rely on
+static type information, as we must avoid the
+trap of \emph{spurious} inputs that cause
+irrelevant problems that would be avoided
+by picking values of a different, relevant type.
 %
-In particular, we must avoid the trap of \emph{spurious} inputs
-that cause irrelevant problems that would be avoided by picking
-values of a different, relevant type.
-%
-We solve this problem by developing a novel operational semantics
-that combines evaluation and type inference.
+We solve this problem by developing a novel
+operational semantics that combines evaluation
+and type inference.
 %
@@
 unknown --- as the inputs.
-% that corresponds to executing the program with \emph{holes} ---
-% values whose type is unknown.
 %
-A hole remains abstract until the evaluation context tells us what type
-it must have, for example the parameters to an addition operation must
-both be integers.
+A hole remains abstract until the evaluation
+context tells us what type it must have, for
+example the parameters to an addition operation
+must both be integers.
 %
-Our semantics \emph{conservatively} instantiates holes with concrete
-values, dynamically inferring the type of the input
-until the program goes wrong.
+Our semantics conservatively instantiates holes
+with concrete values, dynamically inferring the
+type of the input until the program goes wrong.
 %
-We prove that our procedure synthesizes \emph{general witnesses}, which
-means, intuitively, that if a witness is found for a given ill-typed
-function, then, \emph{for all} (inhabited) input types, there exist
-values that can make the function go wrong.
+We prove that our procedure synthesizes \emph{general}
+witnesses, which means, intuitively, that if a witness
+is found for a given ill-typed function, then, \emph{for all}
+(inhabited) input types, there exist values that can make
+the function go wrong.
 

### 2016-06-23 paper/icfp16-cameraready/related.tex
shrink
done

@@
 \label{sec:related-work}
-In this section we connect our work to related efforts on type errors,
-testing, and program exploration.
+% In this section we connect our work to related efforts on type errors,
+% testing, and program exploration.
 
@@
 \label{sec:diagnosis-repair}
-It is well known that unification-based type inference procedures can
-produce poor error messages, and in particular, can misidentify the
-\emph{source} of the type error.
+% It is well known that
+% unification-based type inference procedures can
+% produce poor error messages, and in particular, can misidentify the
+% \emph{source} of the type error.
 %
-Thus, many groups have explored techniques to pinpoint the true source
-of the error and recommend fixes.
-
+Many groups have explored techniques to pinpoint the true source
+of errors reported by static type checkers.
+% unification-based type inference procedures.
+%
 % \subparagraph{Localization}
-The traditional Damas-Milner type inference
-algorithm~\cite{Damas1982-uw} reports the first program location where a
-type mismatch is discovered (subject to the traversal
-strategy~\cite{Lee1998-ys}).
+The traditional Damas-Milner type inference algorithm~\cite{Damas1982-uw}
+reports the first program location where a type mismatch is discovered
+(subject to the traversal strategy~\cite{Lee1998-ys}).
 %
@@
 \paragraph{Improving Error Messages}
+%
 The content and quality of the error messages themselves has also been
@@
 trace.
+%
 
@@
 
-\paragraph{Testing}
-\label{sec:testing}
-\nanomaly is at its heart a test generator, and builds on a rich line of
-work.
+\paragraph{Testing}\label{sec:testing}
+%
+\nanomaly is at its heart a test generator, and as such,
+builds on a rich line of work.
 %
@@
 
+A function's type is a theorem about the function's behavior.
+Thus, \toolname's witnesses can be viewed as \emph{counterexamples},
+thereby connecting it to work on using test generation to find
+counterexamples prior to starting a proof proof~\cite{ACL2Testing, Seidel15}.
+
 % \begin{itemize}
@@
 \paragraph{Program Exploration}
+
+\cite{Flanagan96} describes a static debugger for Scheme, which helps
+the programmer interactively visualize problematic source-sink flows
+corresponding soft-typing errors. The debugger allows the user to explore
+an abstract reduction graph computed from a static value set analysis of
+the program. In contrast, \toolname generates witnesses and allows the user
+to explore the resulting dynamic execution.
+%
 Perera \etal~\cite{Perera2012-dy} present a tracing semantics

### 2016-09-08 paper/icfp16-cameraready/abstract.tex

--- 2016-06-07 paper/icfp16-cameraready/abstract.tex
@@
   %
-  \emph{Static} type errors are a common stumbling block
+  {Static} type errors are a common stumbling block
   for newcomers to typed functional languages.
   %
-  We present a \emph{dynamic} approach to explaining type
+  We present a {dynamic} approach to explaining type
   errors by generating counterexample witness inputs that
-  illustrate \emph{how} an ill-typed program goes wrong.
+  illustrate how an ill-typed program goes wrong.
   %
   First, given an ill-typed function, we symbolically
-  execute the body to dynamically synthesize witness values
-  that can make the program go wrong.
+  execute the body to synthesize witness values that
+  make the program go wrong.
   We prove that our procedure synthesizes
-  \emph{general witnesses} in that if a witness is
-  found, then \emph{for all} inhabited input types, there exist
-  values that can make the function go wrong.
+  {general} witnesses in that if a witness is
+  found, then for all inhabited input types,
+  there exist values that can make the function go wrong.
   %
   Second, we show how to extend the above procedure to
-  produce a \emph{reduction graph} that can be used to
+  produce a reduction graph that can be used to
   interactively visualize and debug witness executions.
   %
-  Third, we evaluate our approach on two data sets
-  comprising over 4,500 ill-typed student programs.
+  Third, we evaluate the coverage of our approach
+  on two data sets comprising over 4,500 ill-typed
+  student programs.
   Our technique is able to generate witnesses for
@@
   yields small counterexamples for 81\% of the witnesses.
+  %
+  Finally, we evaluate whether our witnesses help
+  students understand and fix type errors, and
+  find that students presented with our witnesses
+  show a greater understanding of type errors
+  than those presented with a standard error message.
 \end{abstract}

### 2016-09-08 paper/icfp16-cameraready/introduction.tex

@@
 \begin{verbatim}
- File "fac.ml", line 5, characters 8-17:
- Error: This expression has type bool but an
-        expression was expected of type int.
+  This expression has type
+    bool
+  but an expression was expected of type
+    int.
 \end{verbatim}
@@
 \centering
-\begin{code}
+\begin{ecode}
 let rec fac n =
@@
   else
-      n * fac (n-1)
-\end{code}
+    n * __fac (n-1)__
+\end{ecode}
 \vspace{3em}
@@
 \end{minipage}
-\caption{(top-left) An ill-typed \texttt{fac} function; (bottom-left) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (right) The same trace, fully expanded.}
+\caption{(top-left) An ill-typed \texttt{fac} function, the error location reported by \ocaml is underlined; (bottom-left) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (right) The same trace, fully expanded.}
 \label{fig:factorial}

### 2016-09-08 paper/icfp16-cameraready/related.tex

@@
 A function's type is a theorem about the function's behavior.
-Thus, \toolname's witnesses can be viewed as \emph{counterexamples},
+Thus, \toolname's witnesses can be viewed as \emph{counter-examples},
 thereby connecting it to work on using test generation to find
-counterexamples prior to starting a proof proof~\cite{ACL2Testing, Seidel15}.
+counter-examples prior to starting a proof~\cite{ACL2Testing, Seidel15}.
 
@@
 
-\cite{Flanagan96} describes a static debugger for Scheme, which helps
+Flanagan \etal~\cite{Flanagan96} describe a static debugger for Scheme, which helps
 the programmer interactively visualize problematic source-sink flows
-corresponding soft-typing errors. The debugger allows the user to explore
+corresponding to soft-typing errors. The debugger allows the user to explore
 an abstract reduction graph computed from a static value set analysis of

### 2017-03-05 paper/jfp-submission/conclusion.tex
let's see what the JFP format looks like..

\section{Conclusion}
\label{sec:conclusion}
In this paper we have demonstrated how to find \emph{dynamic} witnesses
to \emph{static} type errors.
%
Dynamic witnesses provide the user with more information than static
error messages, and enable the use of standard debugging techniques
to explore \emph{how} the program ``goes wrong''.
%
Our witness search produces general witnesses, precluding the
possiblity of a valid typing for the program.
%
We have shown a novel interactive visualization of the execution of
functional programs, which can be used to \emph{show} a novice how a
witness makes their program crash.
%
Our evaluation shows that our search is effective at finding witnesses
with a 87\% success rate, and that our jump-compressed traces
effectively hide boring details of the execution, with 90\% of the
benchmarks containing less than 10 jumps.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-03-05 paper/jfp-submission/evaluation-witness.tex
let's see what the JFP format looks like..

--- 2015-11-18 paper/pldi16-submission/evaluation-witness.tex
@@
 similar course at the University of
-Washington~\cite{lerner_seminal:_2006}, from which we extracted 284
+Washington~\cite{Lerner2006-pj}, from which we extracted 284
 ill-typed programs.
@@
 \end{minipage}
-\vspace{-8ex}
+% \vspace{-8ex}
 \caption{Experimental results. Our random search successfully finds
@@
   time. For the remaining 11-12\% of the programs we are unable to
-  provide any useful feedback. \ES{TODO: the timeout number for ucsd is
-    a bit misleading due to a mistake in the benchmarking code. re-running to fix..}
+  provide any useful feedback. 
 }

### 2017-03-05 paper/jfp-submission/interactive-short.tex
let's see what the JFP format looks like..

--- 2016-03-14 paper/icfp16-submission/interactive-short.tex
@@
 %
-Once the user has identified an interesting call-return pair, they can
-step into the call and proceed with more fine-grained steps.
+Once users have identified interesting call-return pairs, they can
+step into those calls and proceed with more fine-grained steps.
 %
@@
 %
-For example, the display on the bottom-left in Figure~\ref{fig:factorial} shows
-a jump-compressed trace, of size 4. The full trace for
-the same example (on the right of the Figure~\ref{fig:factorial}) has size 19.
+% For example, the display on the bottom-left in Figure~\ref{fig:factorial} shows
+% a jump-compressed trace, of size 4. The full trace for
+% the same example (on the right of the Figure~\ref{fig:factorial}) has size 19.
 

### 2017-03-05 paper/jfp-submission/related.tex
let's see what the JFP format looks like..

--- 2016-09-08 paper/icfp16-cameraready/related.tex
@@
 %
-%% \citet{Neubauer2003-xv} present a decidable type system based on
+%% \cite{Neubauer2003-xv} present a decidable type system based on
 %% discriminative sum types, in which all terms are typeable and type
@@
 %
-\citet{Pavlinovic2014-mr,Pavlinovic2015-kh} translate the %error
+\cite{Pavlinovic2014-mr,Pavlinovic2015-kh} translate the %error
 localization problem to a MaxSMT optimization problem, using
@@
 % \subparagraph{Repairing}
-In addition to localizing the error, \citet{Lerner2007-dt} attempt to
+In addition to localizing the error, \cite{Lerner2007-dt} attempt to
 suggest a fix by replacing expressions (or removing them entirely) with
@@
 %
-\citet{Chen2014-gd} use a variational type system to allow for the
+\cite{Chen2014-gd} use a variational type system to allow for the
 possibility of changing an expression's type, and search for an
@@
 %
-\citet{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
+\cite{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
 messages in novice environments and present suggestions for improving
@@
 %
-\citet{Hage2006-hc} identify a variety of general heuristics to improve
+\cite{Hage2006-hc} identify a variety of general heuristics to improve
 the quality of type error messages, based on their teaching experience.
 %
-\citet{Heeren2003-db}, \citet{Christiansen2014-qc}, and
-\citet{Serrano2016-oo} provide methods for library authors to specialize
+\cite{Heeren2003-db}, \cite{Christiansen2014-qc}, and
+\cite{Serrano2016-oo} provide methods for library authors to specialize
 type errors with domain-specific knowledge.
@@
 \label{sec:running-ill-typed}
-\citet{Vytiniotis2012-gh} extend the \haskell
+\cite{Vytiniotis2012-gh} extend the \haskell
 compiler GHC to support compiling ill-typed programs, but their intent
@@
 %
-\citet{Bayne2011-cn} also provide a semantics for running
+\cite{Bayne2011-cn} also provide a semantics for running
 ill-typed (\java) programs, but in constrast transform the program to

### 2017-03-05 paper/jfp-submission/plots.py

@@ -1,0 +1,367 @@
+import csv
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+
+UCSD = 'UCSD'
+
+BUCKETS = [0.1, 0.2, 1.0, 10.0, 60.0 ] # range(500, 3001, 500)
+COLORS=['#90B0D4', '#90D492', '#D4B490', '#D490D2']
+#COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
+
+SAFE = ['S', 'T']
+SAFE_L = ['Safe', 'Timeout']
+UNSAFE = ['U', 'B', 'D'] #, 'O']
+UNSAFE_L = ['Unsafe', 'Unbound', 'Diverge'] #, 'Output']
+ALL = UNSAFE + SAFE
+ALL_L = UNSAFE_L + SAFE_L
+
+ALL_D = [ ['U'], ['B'], ['D'], ['S', 'T']]
+ALL_DL = [ 'Witness', 'Unbound', 'Diverge', 'No Witness'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
+
+
+def read_csv(f):
+    with open(f) as f:
+        return list(csv.reader(f))
+def read_csv_dict(f):
+    with open(f) as f:
+        return list(csv.DictReader(f))
+
+def autolabel(ax, rects):
+    # attach some text labels
+    for rect in rects:
+        height = rect.get_height()
+        ax.text(rect.get_x() + rect.get_width()/2., height,
+                '%d' % int(height),
+                ha='center', va='bottom')
+
+def cumulative_coverage(data):
+    headers = data[0]
+    data = data[1:]
+    return [(l, int(100 * len([r for r in data
+                               if float(r[2]) <= l
+                               and r[4] in UNSAFE])
+                          / float(len(data))))
+            for l in BUCKETS]
+
+def plot_coverage(seminal, ucsd):
+    xy_s = cumulative_coverage(seminal)
+    xy_u = cumulative_coverage(ucsd)
+
+
+    N = len(xy_s)
+    ind = np.arange(N)    # the x locations for the groups
+    width = 0.35       # the width of the bars: can also be len(x) sequence
+
+    fig = plt.figure()
+    p1 = plt.bar(ind, [r[1] for r in xy_s], width,
+                 color=COLORS[0])
+    p2 = plt.bar(ind + width, [r[1] for r in xy_u], width,
[…]

### 2017-03-05 paper/jfp-submission/study.R

@@ -1,0 +1,204 @@
+library(irr)
+
+a0 = read.csv(file="study-data/A-0-scores.csv")
+a1 = read.csv(file="study-data/A-1-scores.csv")
+a2 = read.csv(file="study-data/A-2-scores.csv")
+a3 = read.csv(file="study-data/A-Eric-scores.csv")
+b0 = read.csv(file="study-data/B-0-scores.csv")
+b1 = read.csv(file="study-data/B-1-scores.csv")
+b2 = read.csv(file="study-data/B-2-scores.csv")
+b3 = read.csv(file="study-data/B-Eric-scores.csv")
+
+reasons0 = c(a0$X1..append.reason,
+             a0$X3..digitsofint.reason,
+             a0$X5..sumlist.reason,
+             a0$X7..wwhile.reason,
+             b0$X1..sumlist.reason,
+             b0$X3..wwhile.reason,
+             b0$X5..append.reason,
+             b0$X7..digitsofint.reason)
+reasons1 = c(a1$X1..append.reason,
+             a1$X3..digitsofint.reason,
+             a1$X5..sumlist.reason,
+             a1$X7..wwhile.reason,
+             b1$X1..sumlist.reason,
+             b1$X3..wwhile.reason,
+             b1$X5..append.reason,
+             b1$X7..digitsofint.reason)
+reasons2 = c(a2$X1..append.reason,
+             a2$X3..digitsofint.reason,
+             a2$X5..sumlist.reason,
+             a2$X7..wwhile.reason,
+             b2$X1..sumlist.reason,
+             b2$X3..wwhile.reason,
+             b2$X5..append.reason,
+             b2$X7..digitsofint.reason)
+reasons3 = c(a3$X1..append.reason,
+             a3$X3..digitsofint.reason,
+             a3$X5..sumlist.reason,
+             a3$X7..wwhile.reason,
+             b3$X1..sumlist.reason,
+             b3$X3..wwhile.reason,
+             b3$X5..append.reason,
+             b3$X7..digitsofint.reason)
+reasons = data.frame(reasons0, reasons1, reasons2, reasons3)
+## filter out answers that any annotator marked as INCOMPLETE
[…]

### 2017-04-20 paper/jfp-submission/semantics.tex
let's see what the JFP format looks like..

--- 2016-06-07 paper/icfp16-cameraready/semantics.tex
@@
 in \textsc{PLT-Redex}~\cite{Felleisen2009-ya}.
+%
+Detailed proofs for the theorems in this section can be found in
+%
+\ifthenelse{\equal{\isTechReport}{true}}
+{Appendix~\ref{sec:proofs}.}
+{Appendix A of the accompanying tech report~\cite{Seidel2016Dynamic-TechRep}.}
 
@@
 \paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
-$$
+\[
 \forcesym : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
-$$
+\]
 defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
@@
 %
-\begin{figure*}[ht]
-$$
-\begin{array}{lcl}
+\begin{figure}[t]
+\[
+\begin{array}{lll}
 %% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
-\forcesym                  & : & v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
+\forcesym                  & : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
 % \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
@@
 % \end{cases} \\
-\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
+\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq % \hspace{-1ex}
 \begin{cases}
-  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\vhole{\thole}},
-                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
-  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\vhole{\thole}} \\
-  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
+  \triple{v}{\vsu}{\tsu'}    & \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}, \\
+                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\end{array} \\
+  \triple{\stuck}{\vsu}{\tsu} & \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}\end{array} \\
+  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} & \textbf{if} \begin{array}{l} \tsu' = \unify{\{\thole, t\}}{\tsu}, \\ v = \gen{t}{\tsu'} \end{array}\\
 \end{cases}  \\
@@
   % \end{array} \\
-\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
-\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
-\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
-\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq & \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''}, \mbox{if}\ \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu}, \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'}  \\
-\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
-\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
-\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
+\force{n}{\tint}{\vsu}{\tsu}     & \defeq \triple{n}{\vsu}{\tsu} \\
+\force{b}{\tbool}{\vsu}{\tsu}    & \defeq \triple{b}{\vsu}{\tsu} \\
+\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq \triple{\efun{x}{e}}{\vsu}{\tsu} \\
+\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu},\\ \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'} \end{array} \\
+\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq \triple{\vleaf{t_1}}{\vsu}{\tsu'} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array}\\
+\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} \hspace{-3mm} & \defeq \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array} \\
+\force{v}{t}{\vsu}{\tsu} & \defeq \triple{\stuck}{\vsu}{\tsu}
 \end{array}
-$$
+\]
 \caption{Narrowing values}
 \label{fig:narrow}
-\end{figure*}
+\end{figure}
 %
@@
 
-\begin{figure}[ht]
-$$
+\begin{figure}[t]
+\[
 \begin{array}{lcll}
@@
 \end{array}
-$$
+\]
 \caption{Generating values}
@@
 %
-We write \steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if $\trace$ is
-a trace of the form $\triple{\estuck}{\vsu}{\tsu},\ldots,\triple{\estuck'}{\vsu'}{\tsu'}$.
+We write $\steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ if $\trace$ is
+a trace of the form $\triple{\estuck}{\vsu}{\tsu},\ldots,$ $\triple{\estuck'}{\vsu'}{\tsu'}$.
 %
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
 \[ \begin{array}{lcll}
@@
 %   % \ES{do we need to say anythign else?}
-\qedhere
+% \qedhere
 \end{description}

### 2017-04-21 paper/icfp16-cameraready/plots.py
bench

@@ -48,6 +48,8 @@ def plot_coverage(seminal, ucsd):
     xy_s = cumulative_coverage(seminal)
     xy_u = cumulative_coverage(ucsd)
 
+    print ('xy_s', xy_s)
+    print ('xy_u', xy_u)
 
     N = len(xy_s)
     ind = np.arange(N)    # the x locations for the groups
@@ -295,6 +297,9 @@ def plot_distrib(seminal, ucsd):
             for o in ALL_D]
     rs_u = [len([r for r in ucsd[1:] if r[4] in o])
             for o in ALL_D]
+
+    print ('rs_s', rs_s)
+    print ('rs_u', rs_u)
 
     # N = len(xy)
     # ind = np.arange(N)    # the x locations for the groups

### 2017-04-21 paper/icfp16-cameraready/plots.py
plot

@@ -82,8 +82,11 @@ def plot_user_study():
         return [int(x) for x in xs if int(x) >= 0]
 
     def err(xs):
-        p = np.average(xs)
-        return 100 * np.sqrt(p * (1-p) / len(xs))
+        #p = np.average(xs)
+        #return 100 * np.sqrt(p * (1-p) / len(xs))
+        s = np.std(xs)
+        n = len(xs)
+        return 100 * (s / np.sqrt(n))
 
     ## REASON
     sumlist_a = f([r['5: sumlist reason'] for r in a])
@@ -109,20 +112,22 @@ def plot_user_study():
                   [100*np.average(sumlist_b), 100*np.average(append_a), 100*np.average(digitsofint_a), 100*np.average(wwhile_b)],
                   width,
                   color=COLORS[0],
-                  # yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b])
+                  yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b]),
+                  error_kw={'linewidth': 5}
     )
     p_n = plt.bar(ind + width,
                   [100*np.average(sumlist_a), 100*np.average(append_b), 100*np.average(digitsofint_b), 100*np.average(wwhile_a)],
                   width,
                   color=COLORS[1],
-                  # yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a])
+                  yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a]),
+                  error_kw={'linewidth': 5}
     )
 
     plt.title('Explanation',fontsize=24)
     # plt.xlabel('Problem', fontsize=20)
     plt.ylabel('% Correct', fontsize=20)
     plt.xticks(ind + width, ['sumList', 'append', 'digitsOfInt', 'wwhile'], fontsize='large')
-    plt.legend((p_o[0], p_n[1]), ('OCaml', 'NanoMaLy'), 'lower right', fontsize=16)
+    plt.legend(('OCaml', 'NanoMaLy'), loc='lower right', fontsize=16)
     # autolabel(plt, p_o)
     # autolabel(plt, p_n)
 
@@ -149,20 +154,22 @@ def plot_user_study():
                   [100*np.average(sumlist_b), 100*np.average(append_a), 100*np.average(digitsofint_a), 100*np.average(wwhile_b)],
                   width,
                   color=COLORS[0],
[…]

### 2017-04-21 paper/jfp-submission/semantics.tex
fiddling with figures..

@@
 \begin{cases}
-  \triple{v}{\vsu}{\tsu'}    & \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}, \\
+  \triple{v}{\vsu}{\tsu'}    &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}, \\
                                          \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\end{array} \\
-  \triple{\stuck}{\vsu}{\tsu} & \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}\end{array} \\
-  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} & \textbf{if} \begin{array}{l} \tsu' = \unify{\{\thole, t\}}{\tsu}, \\ v = \gen{t}{\tsu'} \end{array}\\
+  \triple{\stuck}{\vsu}{\tsu} &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}\end{array} \\
+  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} &\ \textbf{if} \begin{array}{l} \tsu' = \unify{\{\thole, t\}}{\tsu}, \\ v = \gen{t}{\tsu'} \end{array}\\
 \end{cases}  \\
@@
 \force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq \triple{\efun{x}{e}}{\vsu}{\tsu} \\
-\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu},\\ \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'} \end{array} \\
-\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq \triple{\vleaf{t_1}}{\vsu}{\tsu'} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array}\\
-\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} \hspace{-3mm} & \defeq \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array} \\
+\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu},\\ \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'} \end{array} \\
+\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq \triple{\vleaf{t_1}}{\vsu}{\tsu'} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array}\\
+\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} \hspace{-3mm} & \defeq \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array} \\
 \force{v}{t}{\vsu}{\tsu} & \defeq \triple{\stuck}{\vsu}{\tsu}
@@
 \label{thm:generation}
-  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$$
+  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$
   then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that

### 2017-04-21 paper/jfp-submission/plots.py

@@ -48,6 +48,8 @@ def plot_coverage(seminal, ucsd):
     xy_s = cumulative_coverage(seminal)
     xy_u = cumulative_coverage(ucsd)
 
+    print ('xy_s', xy_s)
+    print ('xy_u', xy_u)
 
     N = len(xy_s)
     ind = np.arange(N)    # the x locations for the groups
@@ -64,7 +66,7 @@ def plot_coverage(seminal, ucsd):
     plt.title('Cumulative Coverage', fontsize=24)
     plt.xticks(ind + width, [r[0] for r in xy_s], fontsize='large')
     plt.yticks(np.arange(0, 101, 10), fontsize='large')
-    plt.legend((p1[0], p2[0]), ('UW', UCSD), 'lower right', fontsize=16)
+    plt.legend((p1[0], p2[0]), ('UW', UCSD), loc='lower right', fontsize=16)
     # plt.legend((p1[0], p2[0]), ('Men', 'Women'))
     autolabel(plt, p1)
     autolabel(plt, p2)
@@ -226,7 +228,7 @@ def plot_trace_size(seminal, ucsd):
     print('ucsd:\t\t{}\t{}\t{}'.format(c_step_u[0], c_step_u[1], len(step_u)))
     print('avg/med/max:\t{}\t{}\t{}'.format(np.mean(step_u), np.median(step_s), np.max(step_u)))
     p2 = plt.bar(ind + width, c_step_u, label=UCSD, width=width, color=COLORS[1])
-    plt.legend((p1[0],p2[0]), ('UW',UCSD), fontsize=16)
+    plt.legend((p1[0],p2[0]), ('UW',UCSD), loc='lower right', fontsize=16)
     plt.title('Trace Complexity', fontsize=24)
     plt.xlabel('Total Steps', fontsize=20)
     plt.ylabel('Traces (%)', fontsize=20)
@@ -262,7 +264,7 @@ def plot_trace_size(seminal, ucsd):
     print('ucsd:\t\t{}\t{}\t{}'.format(c_jump_u[0], c_jump_u[1], len(jump_u)))
     print('avg/med/max:\t{}\t{}\t{}'.format(np.mean(jump_u), np.median(jump_s), np.max(jump_u)))
     p2 = plt.bar(ind + width, c_jump_u, label=UCSD, width=width, color=COLORS[1])
-    plt.legend((p1[0],p2[0]), ('UW',UCSD), fontsize=16)
[…]

### 2017-04-22 paper/jfp-submission/evaluation.tex
fiddling with figures..
start typing up investigation of "safe" programs

--- 2016-06-23 paper/icfp16-cameraready/evaluation.tex
@@
 
-\begin{figure*}[htp]
-\centering
-\begin{minipage}{\linewidth}
-\centering
-\includegraphics[width=0.45\linewidth]{coverage.png}
+\begin{figure}[htp]
+\centerline{
+\begin{minipage}{1.2\textwidth}
+\includegraphics[width=0.5\linewidth]{coverage.png}
 % \end{minipage}
 % \begin{minipage}{\linewidth}
-\includegraphics[width=0.45\linewidth]{distrib.png}
+\includegraphics[width=0.5\linewidth]{distrib.png}
 \end{minipage}
-% \vspace{-8ex}
+}
+\vspace{3ex}
 \caption{Results of our coverage testing and the distribution of test
@@
 \label{fig:results-witness}
-\end{figure*}
+\end{figure}
 
@@
 
+\input{safe}
+
 \subsection{Witness Complexity}
@@
 %
-\begin{figure*}[htp]
-\centering
-\includegraphics[width=0.45\linewidth]{trace_size_step.png}
-\includegraphics[width=0.45\linewidth]{trace_size_jump.png}
+\begin{figure}[htp]
+\centerline{
+\begin{minipage}{1.2\textwidth}
+\includegraphics[width=0.5\linewidth]{trace_size_step.png}
+\includegraphics[width=0.5\linewidth]{trace_size_jump.png}
+\end{minipage}
+}
+\vspace{3ex}
 \caption{Complexity of the generated traces. 81\% of the combined traces
@@
 \label{fig:results-complexity}
-\end{figure*}
+\end{figure}
 %
-\begin{figure*}[htp]
-\centering
-\includegraphics[width=0.45\linewidth]{user-study-reason.png}
-\includegraphics[width=0.45\linewidth]{user-study-fix.png}
+\begin{figure}[htp]
+\centerline{
+\begin{minipage}{1.2\textwidth}
+\includegraphics[width=0.5\linewidth]{user-study-reason.png}
+\includegraphics[width=0.5\linewidth]{user-study-fix.png}
+\end{minipage}
+}
+\vspace{3ex}
 \caption{A classification of students' explanations and fixes for type
@@
 \label{fig:results-user-study}
-\end{figure*}
+\end{figure}
 
@@
 
-This issue is uncommon in \ocaml\ (we did not detect a single instance
-of it across all of our benchmarks), but it would surely be exacerbated
+This issue is uncommon in \ocaml\
+(we detected it in around 5\% of our benchmarks),
+but it would surely be exacerbated
 by a language like \haskell, which makes heavy use of overloading. We

### 2017-04-22 paper/jfp-submission/safe.tex
start typing up investigation of "safe" programs

\subsubsection{How Safe are the ``Safe'' Programs?}
\label{sec:how-safe}

An immediate question arises regarding that 13--16\% of programs for
which we could not synthesize a witness:
%
are they \emph{actually} safe (\ie is the type system being too conservative),
%
or did \toolname simply fail to find the witness?

To answer this question, we investigated the 720 \ucsdbench programs for
which we failed to find a witness.
%
We found that for 247 programs (roughly a third of the failures, or 5\%
of all programs), \toolname got stuck when it tried to compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While extremely convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@ as input, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.
%
Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |x| and |y|.

The remaining 473 programs for which we could not produce a witness did
not admit such an automatic diagnosis, so we selected a random sample of
50 programs and manually searched for a witness.
%
We found only 8 cases (or 16\%) where a witness existed, but \toolname
was unable to discover it.
%
Five involved synthesizing a \emph{pair} of specially-crafted inputs
that would result in the function returning values of incompatible
types.
%
The other three required synthesizing an input that would trigger a
particular path through the program, and would likely have been caught
by symbolic execution.
%
Overall, it appears that our random instantiation of holes is well-suited
to finding type errors in student programs.

We categorized the remaining 42 failures into four groups:
%
\begin{enumerate}
\item 7 infinite type. program doesn't go wrong, but could search for
  context that would go wrong
\item 15 monomorphic functions. recall we abstract a function's type as
  $\tfun$ rather than $\alpha \to \beta$. (may also admit witness by
  searching for context?)
\item 6 dead code. type error occurs in unreachable code, ``deficiency''
  of the type system.
\item 14 bad call. function is well-typed, student calls makes ill-typed
  call, but no error triggered. nothing to be done, as we do not control
  call-site.
\end{enumerate}

### 2017-04-22 paper/jfp-submission/safe.tex
some text

@@
 
-We categorized the remaining 42 failures into four groups:
+We categorized the remaining 42 failures into four groups.
+
+\paragraph{Dead Code}
 %
-\begin{enumerate}
-\item 7 infinite type. program doesn't go wrong, but could search for
-  context that would go wrong
-\item 15 monomorphic functions. recall we abstract a function's type as
-  $\tfun$ rather than $\alpha \to \beta$. (may also admit witness by
-  searching for context?)
-\item 6 dead code. type error occurs in unreachable code, ``deficiency''
-  of the type system.
-\item 14 bad call. function is well-typed, student calls makes ill-typed
-  call, but no error triggered. nothing to be done, as we do not control
-  call-site.
-\end{enumerate}
+Six programs (or 12\% the samples) contained type errors that were
+unreachable by any inputs.
+%
+While technically safe, dead code is generally considered a ``code
+smell''; students would likely benefit from a warning in this case.
+
+\paragraph{Bad Function Call}
+Fourteen programs contained a function call where the student had
+supplied incorrect inputs, yet no error was triggered.
+%
+Consider, for example the following |assoc| function, which
+looks up a key in an \emph{association list} and returns a default if
+the key cannot be found.
+%
+\begin{code}
+  let rec assoc (d, k, l) = match l with
+    | (ki, vi)::tl ->
+       if ki = k then
+         vi
+       else
+         assoc (d, k, tl)
+    | _ -> d
+
+  let _ = assoc ([], 123, [(123, "sad"); (321, "happy")])
+\end{code}
+%
+The student's definition of |assoc| is correct, but \ocaml rejects their
+subsequent call because the default value |[]| is incompatible with the
+|string| values in the list.
+%
+In this particular call, the key |123| is in the list, so the default
+will not be used (and even if it were, no error would be triggered) and
+\ocaml's complaint is moot.
+%
+Of course, \ocaml cannot be expected to know that this particular call
+is safe, its type system is not sophisticated enough to express the
+necessary conditions.
+
+We note that all of the programs in this category appear to come from
+interactions with the \ocaml top-level, \ie the student defined a
+function and is now experimenting with it.
+%
+These calls are thus particularly harmless, indeed they may even be
+\emph{beneficial}, as the student may next try looking up a key that
+\emph{does not} exist, and realize that the returned values are
+incompatible.
+
+% \begin{enumerate}
+% \item 7 infinite type. program doesn't go wrong, but could search for
+%   context that would go wrong
+% \item 15 monomorphic functions. recall we abstract a function's type as
+%   $\tfun$ rather than $\alpha \to \beta$. (may also admit witness by
+%   searching for context?)
+% \item 6 dead code. type error occurs in unreachable code, ``deficiency''
+%   of the type system.
+% \item 14 bad call. function is well-typed, student calls makes ill-typed
+%   call, but no error triggered. nothing to be done, as we do not control
+%   call-site.
+% \end{enumerate}

### 2017-04-22 paper/jfp-submission/plots.py

@@ -9,14 +9,14 @@ BUCKETS = [0.1, 0.2, 1.0, 10.0, 60.0 ] # range(500, 3001, 500)
 COLORS=['#90B0D4', '#90D492', '#D4B490', '#D490D2']
 #COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
 
-SAFE = ['S', 'T']
-SAFE_L = ['Safe', 'Timeout']
+SAFE = ['S', 'T', 'H']
+SAFE_L = ['Safe', 'Timeout', 'Hole']
 UNSAFE = ['U', 'B', 'D'] #, 'O']
 UNSAFE_L = ['Unsafe', 'Unbound', 'Diverge'] #, 'Output']
 ALL = UNSAFE + SAFE
 ALL_L = UNSAFE_L + SAFE_L
 
-ALL_D = [ ['U'], ['B'], ['D'], ['S', 'T']]
+ALL_D = [ ['U', 'O'], ['B'], ['D'], ['S', 'T', 'H']]
 ALL_DL = [ 'Witness', 'Unbound', 'Diverge', 'No Witness'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
 
 

### 2017-04-23 paper/jfp-submission/safe.tex
more text

@@
 Six programs (or 12\% the samples) contained type errors that were
-unreachable by any inputs.
+unreachable by any inputs, often due to overlapping patterns in a
+@match@ expression.
 %
@@
 
-\paragraph{Bad Function Call}
+\paragraph{Bad Function Calls}
 Fourteen programs contained a function call where the student had
@@
 
+\paragraph{Non-Parametric Function Type}
+%
+Twenty-two programs (or 44\% of the samples) lack a witness in our
+semantics due to our non-parametric $\tfun$ type for functions.
+%
+Recall that our goal is to expose the runtime errors that would have
+been prevented by the type systems.
+%
+At runtime, it is always safe to call a function, thus we give functions
+a simple type $\tfun$ that says they may be applied, but says nothing
+about their inputs or outputs.
+%
+But consider the following @clone@ function, which is supposed to
+produce a list containing @n@ copies of the input @x@.
+%
+% Seven programs violated the \emph{occurs check} with cyclic typing
+% constraints like @'a = 'a list@, for example the following @clone@
+% function.
+%
+\begin{code}
+  let rec clone x n =
+    if n > 0 then
+      clone [x] (n - 1)
+    else
+      []
+\end{code}
+%
+Unfortunately, the student instead constructs an @n@-level nested list
+containing a single @x@.
+%
+The \ocaml compiler rejects this program because the recursive call to
+@clone@ induces a cyclic typing constraint @'a = 'a list@, capturing the
+fact that each call increases the nesting of the list.
+%
+\toolname fails to catch this because we do not track the types of the
+inputs to @clone@.
+
+We note, however, that @clone@ cannot go wrong; it is perfectly safe to
+repeatedly enclose a list inside another (disregarding the fact that the
+nested list is never returned).
+%
+Still, such a function would be very difficult to \emph{call} safely, as
+the programmer would have to reason about the dependency between the
+input @n@ and the nesting of the output list, which cannot be expressed
+in \ocaml's type system.
+%
+Thus, is it not particularly satisfying that \toolname fails to produce
+a witness here; a possible solution could be to track the types of the
+inputs, and demonstrate to the user how they change between recursive
+calls.
+
 % \begin{enumerate}

### 2017-04-23 paper/jfp-submission/safe.tex
more text

@@
 %
-We found only 8 cases (or 16\%) where a witness existed, but \toolname
-was unable to discover it.
+We categorized the programs into four groups.
+
+\paragraph{Witness Exists}
+%
+We found only eight cases (or 16\% of the samples) where a witness existed,
+but \toolname was unable to discover it.
 %
@@
 %
-Overall, it appears that our random instantiation of holes is well-suited
-to finding type errors in student programs.
-
-We categorized the remaining 42 failures into four groups.
+% Overall, it appears that our random instantiation of holes is well-suited
+% to finding type errors in student programs.
+%
+%We categorized the remaining 42 failures into three groups.
 
@@
 \paragraph{Bad Function Calls}
-Fourteen programs contained a function call where the student had
-supplied incorrect inputs, yet no error was triggered.
+Fourteen programs (or 28\% of the samples) contained a function call
+where the student had supplied incorrect inputs, yet no error was
+triggered.
 %
@@
 
+To summarize, our investigation suggests that the majority of programs
+(88\%) for which we fail to find a witness do not, in fact, admit a
+witness.
+%
+Overall, it appears that our random instantiation of holes is well-suited
+to finding type errors in student programs.
+%
+The remaining programs were all cases where \ocaml's type system was
+overly conservative.
+%
+Of course, the conservatism is somewhat justified as each case pointed
+to a ``code smell'' or difficult-to-use API; it would be interesting to
+investigate how demonstrate these issues in a similar manner.
+
 % \begin{enumerate}

### 2017-04-24 paper/jfp-submission/semantics.tex

@@
   then for every
-  (inhabited\footnote{All types in our formalism are inhabited, but in a larger language like \ocaml this may not be the case.})
+  (inhabited\footnote{All types in \lang are inhabited, but in a larger language like \ocaml this may not be true.})
   type

### 2017-04-24 paper/jfp-submission/safe.tex
discuss speculative insantiation

@@
 or did \toolname simply fail to find the witness?
+
+\ES{Need better organization for the groupings of programs, maybe a pie chart to guide us?}
 
@@
 and |y| such that |x < y| is true.
-%
+
 Extending \toolname with support for symbolic execution would alleviate
@@
 until we learn how to instantiate |x| and |y|.
+%
+Alternatively, we could \emph{speculatively} instantiate both |x| and
+|y| with some arbitrary type, and proceed with execution until we
+discover a type error.
+%
+This speculative instantiation is of course unsound, we would have to
+take care to avoid reporting frivolous type errors that were caused by
+such instantiations.
+%
+We would need to track which holes were instantiated speculatively in
+order to distinguish type errors that would have happened regardless, as
+in the above @bad@ function, from type errors that were the fault of our
+instantiation.
+
+Further, suppose that our speculative instantiation induces a frivolous
+type error.
+%
+For example, suppose we are given
+%
+\begin{code}
+  let bad x y =
+    if x < y then
+      x *. y
+    else
+      0.
+\end{code}
+%
+and choose to (speculatively) instantiate @x@ and @y@ as @int@s and proceed
+down the ``true'' branch.
+%
+We will quickly discover this was the wrong choice, as they are immediately
+narrowed to @float@s.
+%
+We must now backtrack and try a different instantiation, but we no
+longer need to choose one at random.
+%
+Since our instantiation was speculative, we can still view @x@ and @y@
+\emph{morally} as holes, and thus treat the @*.@ operator as a normal
+narrowing point with two holes.
+%
+This tells us that the \emph{correct} instantiation was in fact @float@,
+and we can then proceed as normal from the backtracking point with a
+concrete choice of @float@s.
+
+Thus, it appears that speculative instantiation of holes may be a
+useful, lightweight alternative to symbolic execution for our purposes.
 
@@
 Of course, the conservatism is somewhat justified as each case pointed
-to a ``code smell'' or difficult-to-use API; it would be interesting to
+to a ``code smell'' or difficult-to-use API;\ it would be interesting to
 investigate how demonstrate these issues in a similar manner.

### 2017-04-24 paper/jfp-submission/safe.tex
revise safe.tex

@@
 which we failed to find a witness.
+%
+
+\paragraph{Overloaded Functions.}
 %
@@
     else
-      0.
+      0.0
 \end{code}
@@
 concrete choice of @float@s.
-
+%
 Thus, it appears that speculative instantiation of holes may be a
@@
 
+\paragraph{} \hspace{0ex} \\
 The remaining 473 programs for which we could not produce a witness did
@@
 %
-We categorized the programs into four groups.
+We categorized the programs into three groups.
 
-\paragraph{Witness Exists}
+\paragraph{Non-Parametric Function Type.}
 %
-We found only eight cases (or 16\% of the samples) where a witness existed,
-but \toolname was unable to discover it.
-%
-Five involved synthesizing a \emph{pair} of specially-crafted inputs
-that would result in the function returning values of incompatible
-types.
-%
-The other three required synthesizing an input that would trigger a
-particular path through the program, and would likely have been caught
-by symbolic execution.
-%
-% Overall, it appears that our random instantiation of holes is well-suited
-% to finding type errors in student programs.
-%
-%We categorized the remaining 42 failures into three groups.
-
-\paragraph{Dead Code}
-%
-Six programs (or 12\% the samples) contained type errors that were
-unreachable by any inputs, often due to overlapping patterns in a
-@match@ expression.
-%
-While technically safe, dead code is generally considered a ``code
-smell''; students would likely benefit from a warning in this case.
-
-\paragraph{Bad Function Calls}
-Fourteen programs (or 28\% of the samples) contained a function call
-where the student had supplied incorrect inputs, yet no error was
-triggered.
-%
-Consider, for example the following |assoc| function, which
-looks up a key in an \emph{association list} and returns a default if
-the key cannot be found.
-%
-\begin{code}
-  let rec assoc (d, k, l) = match l with
-    | (ki, vi)::tl ->
-       if ki = k then
-         vi
-       else
-         assoc (d, k, tl)
-    | _ -> d
-
-  let _ = assoc ([], 123, [(123, "sad"); (321, "happy")])
-\end{code}
-%
-The student's definition of |assoc| is correct, but \ocaml rejects their
-subsequent call because the default value |[]| is incompatible with the
-|string| values in the list.
-%
-In this particular call, the key |123| is in the list, so the default
-will not be used (and even if it were, no error would be triggered) and
-\ocaml's complaint is moot.
-%
-Of course, \ocaml cannot be expected to know that this particular call
-is safe, its type system is not sophisticated enough to express the
-necessary conditions.
-
-We note that all of the programs in this category appear to come from
-interactions with the \ocaml top-level, \ie the student defined a
-function and is now experimenting with it.
-%
-These calls are thus particularly harmless, indeed they may even be
-\emph{beneficial}, as the student may next try looking up a key that
-\emph{does not} exist, and realize that the returned values are
-incompatible.
-
-\paragraph{Non-Parametric Function Type}
-%
-Twenty-two programs (or 44\% of the samples) lack a witness in our
+44\% of the samples lack a witness in our
 semantics due to our non-parametric $\tfun$ type for functions.
@@
 in \ocaml's type system.
-%
-Thus, is it not particularly satisfying that \toolname fails to produce
+
+Thus, it is not particularly satisfying that \toolname fails to produce
 a witness here; a possible solution could be to track the types of the
@@
 calls.
+%
+This would require maintaining a typing environment of variables in
+addition to the environments we maintain for holes.
+%
+We would have to modify the rule $\reappgood$ from
+Figure~\ref{fig:operational} to additionally $\forcesym$ the function's
+type against the concrete inputs.
+%
+However, we would want to ensure that this $\forcesym$ cannot fail ---
+it is preferable to report a stuck term as that provides a fuller view
+of the error.
+%
+Rather, we would note which evaluation steps induced incompatible
+type refinements, and if a traditional witness cannot be found, we could
+then report a trace expanded to show precisely these steps.
+%
+This seems like a modest extension to our semantics, and would be
+interesting to explore further.
 
-To summarize, our investigation suggests that the majority of programs
-(88\%) for which we fail to find a witness do not, in fact, admit a
+\paragraph{Dead Code.}
+%
+40\% of the samples contained type errors that were unreachable, either
+because they were dead code, or because the student called the function
+with inputs that could not trigger the error.
+
+12\% of the samples contained type errors that were
+unreachable by any inputs, often due to overlapping patterns in a
+@match@ expression.
+%
+% While technically safe, dead code is generally considered a ``code
+% smell''; students would likely benefit from a warning in this case.
+% \ES{Wes says this is too informal, source the claim}
+%
+While technically safe, dead code is generally considered a maintenance
+risk, as the programmer may not realize that it is dead~\cite{Wheeler2014-fg}
+or may accidentally bring it back to life~\cite{Seven2014-gf}.
+%
+Thus, a warning like that provided by \ocaml's pattern-exhaustiveness
+checker would be helpful.
+
+%\paragraph{Bad Function Calls}
+A further 28\% of the samples included a function call where the
+student had supplied ill-typed inputs, yet the path induced
+by the call contained no errors.
+%
+Consider, for example the following |assoc| function, which
+looks up a key in an \emph{association list} and returns a default if
+the key cannot be found.
+%
+\begin{code}
+  let rec assoc (d, k, l) = match l with
+    | (ki, vi)::tl ->
+       if ki = k then
+         vi
+       else
+         assoc (d, k, tl)
+    | _ -> d
+
+  let _ = assoc ([], 123, [(123, "sad"); (321, "happy")])
+\end{code}
+%
+The student's definition of |assoc| is correct, but \ocaml rejects their
+subsequent call because the default value |[]| is incompatible with the
+|string| values in the list.
+%
+In this particular call, the key |123| is in the list, so the default
+will not be used (and even if it were, no error would be triggered) and
+\ocaml's complaint is moot.
+
+Of course, \ocaml cannot be expected to know that this particular call
+is safe, its type system is not sophisticated enough to express the
+necessary conditions.
+%
+\ES{TODO: add some comment about using dependent types, though this particular example would be quite bizarre to encode in DT..}
+
+% We note that many of the programs in this category appear to come from
+% interactions with the \ocaml top-level, \ie the student defined a
+% function and is now experimenting with it.
+% %
+% These calls are thus particularly harmless, indeed they may even be
+% \emph{beneficial}, as the student may next try looking up a key that
+% \emph{does not} exist, and realize that the returned values are
+% incompatible.
+% \ES{Wes doesn't like this, sounds like punting on the issue}
+
+\paragraph{Witness Exists.}
+%
+We found that 16\% of the samples admitted a witness that \toolname
+was unable to discover.
+%
+Slightly over half involved synthesizing a \emph{pair} of
+specially-crafted inputs that would result in the function returning
+values of incompatible types.
+%
+The rest required synthesizing an input that would trigger a
+particular path through the program, and would likely have been caught
+by symbolic execution.
+
+\paragraph{Summary.}
+%
+Our investigation suggests that the majority of programs
+(84\%) for which we fail to find a witness do not, in fact, admit a
 witness.
@@
 Of course, the conservatism is somewhat justified as each case pointed
-to a ``code smell'' or difficult-to-use API;\ it would be interesting to
-investigate how demonstrate these issues in a similar manner.
+to code that would be difficult to use or maintain;
+%
+it would be interesting to investigate how demonstrate these issues in a
+similar manner.
 
@@
 % \end{enumerate}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-04-25 paper/jfp-submission/safe.tex
reorganize safe.tex

@@
 
-An immediate question arises regarding that 13--16\% of programs for
+An immediate question arises regarding the 13--16\% of programs for
 which we could not synthesize a witness:
@@
 or did \toolname simply fail to find the witness?
-
-\ES{Need better organization for the groupings of programs, maybe a pie chart to guide us?}
+%
+% \ES{Need better organization for the groupings of programs, maybe a pie chart to guide us?}
 
@@
 %
+We used a combination of automatic and manual coding to categorize these
+programs into four classes.
+%
+The first class is easily detected by \toolname itself, and thus admits
+a precise count.
+%
+This left us with 473 programs requiring manual coding; we selected
+a random sample of 50 programs to investigate, and will report results
+based on that sample.
+%
+Figure~\ref{fig:no-witness} summarizes the results of our investigation,
+we note the classes that were based on a random sample with a ``*''.
+
+\begin{figure}[t]
+%\centering
+\includegraphics[width=0.8\linewidth]{distrib_ext.png}
+\vspace{-0.75cm}
+\caption{Results of our investigation into programs where \toolname
+  did not produce a witness. A ``*'' denotes that the percentage is an
+  estimate based on a random sampling of 50 programs.}
+\label{fig:no-witness}
+\end{figure}
 
@@
 %
-We found that for 247 programs (roughly a third of the failures, or 5\%
-of all programs), \toolname got stuck when it tried to compare two holes.
+We found that for 5\% programs \toolname got stuck when it tried to
+compare two holes.
 %
@@
 
-\paragraph{} \hspace{0ex} \\
-The remaining 473 programs for which we could not produce a witness did
-not admit such an automatic diagnosis, so we selected a random sample of
-50 programs and manually searched for a witness.
+% \paragraph{} \hspace{0ex} \\
+% The remaining 473 programs for which we could not produce a witness did
+% not admit such an automatic diagnosis, so we selected a random sample of
+% 50 programs and manually searched for a witness.
+% %
+% We categorized the programs into three groups.
+
+\paragraph{Non-Parametric Function Type *.}
 %
-We categorized the programs into three groups.
-
-\paragraph{Non-Parametric Function Type.}
-%
-44\% of the samples lack a witness in our
-semantics due to our non-parametric $\tfun$ type for functions.
+5\% of programs lack a witness in our semantics due to our
+non-parametric $\tfun$ type for functions.
 %
@@
 
-\paragraph{Dead Code.}
+\paragraph{Dead Code and ``Safe'' Function Calls *.}
 %
-40\% of the samples contained type errors that were unreachable, either
+4\% of programs contained type errors that were unreachable, either
 because they were dead code, or because the student called the function
@@
 
-12\% of the samples contained type errors that were
-unreachable by any inputs, often due to overlapping patterns in a
-@match@ expression.
+1\% contained type errors that were unreachable by any inputs, often due
+to overlapping patterns in a @match@ expression.
 %
@@
 %\paragraph{Bad Function Calls}
-A further 28\% of the samples included a function call where the
+A further 3\% included a function call where the
 student had supplied ill-typed inputs, yet the path induced
@@
 
-\paragraph{Witness Exists.}
+\paragraph{Witness Exists *.}
 %
-We found that 16\% of the samples admitted a witness that \toolname
+We found that only 2\% of programs admit a witness that \toolname
 was unable to discover.

### 2017-04-25 paper/jfp-submission/plots.py

@@ -6,8 +6,9 @@ import numpy as np
 UCSD = 'UCSD'
 
 BUCKETS = [0.1, 0.2, 1.0, 10.0, 60.0 ] # range(500, 3001, 500)
-COLORS=['#90B0D4', '#90D492', '#D4B490', '#D490D2']
-#COLORS=['#348ABD', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
+#COLORS=['#90B0D4', '#90D492', '#D4B490', '#D490D2']
+COLORS=['#8dd3c7','#bebada','#ffffb3','#fb8072','#80b1d3','#fdb462']
+COLORS_E=['#8dd3c7','#bebada','#80b1d3','#ffffb3','#fdb462','#fb8072']
 
 SAFE = ['S', 'T', 'H']
 SAFE_L = ['Safe', 'Timeout', 'Hole']
@@ -18,6 +19,9 @@ ALL_L = UNSAFE_L + SAFE_L
 
 ALL_D = [ ['U', 'O'], ['B'], ['D'], ['S', 'T', 'H']]
 ALL_DL = [ 'Witness', 'Unbound', 'Diverge', 'No Witness'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
+
+ALL_D_E = [ ['U', 'O', 'B', 'D'], ['H'], ['S', 'T']]
+ALL_DL_E = [ 'Witness Found', 'Overloaded Function', 'Non-Parametric Function *', 'Dead Code *', 'Safe Call *', 'Witness Exists *'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
 
 
 def read_csv(f):
@@ -229,7 +233,7 @@ def plot_trace_size(seminal, ucsd):
     print('avg/med/max:\t{}\t{}\t{}'.format(np.mean(step_u), np.median(step_s), np.max(step_u)))
     p2 = plt.bar(ind + width, c_step_u, label=UCSD, width=width, color=COLORS[1])
     plt.legend((p1[0],p2[0]), ('UW',UCSD), loc='lower right', fontsize=16)
-    plt.title('Trace Complexity', fontsize=24)
+    # plt.title('Trace Complexity', fontsize=24)
     plt.xlabel('Total Steps', fontsize=20)
     plt.ylabel('Traces (%)', fontsize=20)
     # ax.set_xlim(0,6)
[…]

### 2017-04-25 paper/jfp-submission/plots.py
minor tweak

@@ -377,7 +377,7 @@ def plot_distrib_extended(ucsd):
     plt.axes(aspect=1)
     p1 = plt.pie(rs_u,
                  labels=ALL_DL_E,
-                 explode=[0, 0.2, 0.2, 0.3, 0.3, 0.2],
+                 explode=[0, 0.2, 0.2, 0.2, 0.2, 0.2],
                  autopct='%.0f%%',
                  pctdistance=1.3,
                  labeldistance=10,

### 2017-04-25 paper/jfp-submission/plots.py
fix error bars

@@ -119,14 +119,14 @@ def plot_user_study():
                   width,
                   color=COLORS[0],
                   yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b]),
-                  error_kw={'linewidth': 5}
+                  error_kw={'linewidth': 3, 'ecolor': 'gray', 'capsize': 6, 'capthick': 3}
     )
     p_n = plt.bar(ind + width,
                   [100*np.average(sumlist_a), 100*np.average(append_b), 100*np.average(digitsofint_b), 100*np.average(wwhile_a)],
                   width,
                   color=COLORS[1],
                   yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a]),
-                  error_kw={'linewidth': 5}
+                  error_kw={'linewidth': 3, 'ecolor': 'gray', 'capsize': 6, 'capthick': 3}
     )
 
     plt.title('Explanation',fontsize=24)
@@ -161,14 +161,14 @@ def plot_user_study():
                   width,
                   color=COLORS[0],
                   yerr=map(err, [sumlist_b, append_a, digitsofint_a, wwhile_b]),
-                  error_kw={'linewidth': 5}
+                  error_kw={'linewidth': 3, 'ecolor': 'gray', 'capsize': 6, 'capthick': 3}
     )
     p_n = plt.bar(ind + width,
                   [100*np.average(sumlist_a), 100*np.average(append_b), 100*np.average(digitsofint_b), 100*np.average(wwhile_a)],
                   width,
                   color=COLORS[1],
                   yerr=map(err, [sumlist_a, append_b, digitsofint_b, wwhile_a]),
-                  error_kw={'linewidth': 5}
+                  error_kw={'linewidth': 3, 'ecolor': 'gray', 'capsize': 6, 'capthick': 3}
     )
 
     plt.title('Fix',fontsize=24)

### 2017-04-27 paper/jfp-submission/evaluation.tex
first draft of error location

@@
 %
-\begin{figure}[htp]
+\begin{figure}[t]
 \centerline{
@@
 %
-\begin{figure}[htp]
-\centerline{
-\begin{minipage}{1.2\textwidth}
-\includegraphics[width=0.5\linewidth]{user-study-reason.png}
-\includegraphics[width=0.5\linewidth]{user-study-fix.png}
-\end{minipage}
-}
-\vspace{3ex}
-\caption{A classification of students' explanations and fixes for type
-  errors, given either \ocaml's error % message
-  or \toolname's
-  jump-compressed trace. The students given \toolname's jump-compressed
-  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
-  error.}
-\label{fig:results-user-study}
-\end{figure}
-
 \paragraph{Results}
@@
 %
-We assigned four problems to the ($n=60$) students in the course: the
-@sumList@, @digitsOfInt@, and @wwhile@ programs from
+We assigned four problems to the 60 students in the course: the
+@sumList@, \hbox{@digitsOfInt@,} and @wwhile@ programs from
 \S~\ref{sec:advantage-traces}, as well as the following @append@ program
@@
 For each problem the students were given the ill-typed program and
-either \ocaml's error message or \toolname's jump-compressed trace.
+either \ocaml's error or \toolname's jump-compressed trace.
 %
@@
 
+\begin{figure}[t]
+\centerline{
+\begin{minipage}{1.2\textwidth}
+\includegraphics[width=0.5\linewidth]{user-study-reason.png}
+\includegraphics[width=0.5\linewidth]{user-study-fix.png}
+\end{minipage}
+}
+\vspace{3ex}
+\caption{A classification of students' explanations and fixes for type
+  errors, given either \ocaml's error % message
+  or \toolname's
+  jump-compressed trace. The students given \toolname's jump-compressed
+  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
+  error.}
+\label{fig:results-user-study}
+\end{figure}
+
 \paragraph{Results}
@@
 %
+
+\input{locating}
 

### 2017-04-27 paper/jfp-submission/locating.tex
first draft of error location

\subsection{Locating Errors with Witnesses}
\label{sec:locating}

We have seen that \toolname can effectively synthesize witnesses to
explain the majority of (novice) type errors, but a good error report
should also help \emph{locate} the source of the error.
%
Thus, our final experiment seeks to use \toolname's witnesses as
localizations.

As discussed at the beginning of \S~\ref{sec:evaluation}, we recorded
each interaction of our students with the \ocaml top-level system.
%
In addition to collecting ill-typed programs, we collected
subsequent, fixed versions of the programs.
%
We identify fixed programs by searching for the first type-correct
program that follows the ill-typed program in the interaction trace.
%
We then use an expression-level \emph{diff}~\cite{Lempsink2009-xf} to
determine which sub-expressions changed between the ill-typed program
and the student's fix, and treat those expressions as the source of the
type error.

Not all ill-typed programs will have an associated fix; furthermore,
at some point a ``fix'' becomes a ``rewrite''.
%
We do not wish to consider the ``rewrites'', so we discard outliers
where the fraction of expressions that have changed is more than one
standard deviation above the mean, establishing a diff threshold of
45\%.
%
This accounts for roughly 14\% of programs pairs we discovered, leaving
us with 2,425 program pairs.

For each pair of an ill-typed program and its fix, we run \toolname and
collect two sets of source locations:
%
(1) the source location corresponding to the stuck term; and
%
(2) the source locations that \emph{produced} the values inside the
stuck term.
%
For example, in the @sqsum@ program from \S~\ref{sec:advantage-traces}
the stuck term is \verb!0 @ 1!.
%
This corresponds to the call to \verb!@! on line 3, and contains
the literal @0@ from line 2 and the value @1@ produced by the
@*@ on line 3.

We compare \toolname's witness-based predictions against a baseline of
the \ocaml compiler as well as the state-of-the-art tools \sherrloc
and \mycroft.
%
\sherrloc attempts to predict the most likely source of a type error
by searching the typing constraint graph for constraints that participate
in many unsatisfiable paths and few satisfiable paths.
%
\mycroft reduces the localization problem to MaxSAT by searching for a
minimal subset of constraints that can be removed, such that the
resulting system is satisfiable.
%
Both tools produce a \emph{set} of equally-likely expressions to blame
for the error (in practice the set contains only a few expressions),
similar to \toolname's witness-based predictions.

We evaluate each tool based on whether \emph{any} of its predictions
identifies a changed expression.
%
There were a number of student programs where \mycroft or \sherrloc
encountered an unsupported language feature or timed out after one
minute, or where \toolname failed to produce a witness.
%
We discard all such programs in our evaluation to level the playing
field, leaving us with a benchmark set of 1,612 programs.

\begin{figure}[t]
\includegraphics[width=0.8\linewidth]{blame.png}
\caption{Accuracy of type error localization. \toolname's witness-based
  predictions outperform \ocaml by 20 points, and are competitive
  with the state-of-the-art tools \mycroft and \sherrloc.}
\label{fig:results-blame}
\end{figure}

\paragraph{Results}
Figure~\ref{fig:results-blame} summarizes our results, which show that
\toolname's witnesses are competitive with \mycroft and \sherrloc in
automatically locating the source of a type error.
%
\toolname, \mycroft, and \sherrloc all outperform the \ocaml compiler,
which is not surprising given that they can produce multiple possible
error locations, while the \ocaml compiler is limited to one predicted
error location.
%
Interestingly, while all tools have a median of 2 predicted error
locations per program, \mycroft and \sherrloc have a long tail with a
maximum of 22 (\resp 12) locations, while \toolname's maximum is 5
locations.
%
We also note that while \mycroft and \sherrloc were designed
specifically to \emph{localize} type errors, \toolname's foremost
purpose is to \emph{explain} them, we consider its ability to localize
type errors an added benefit.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-04-27 paper/jfp-submission/plots.py

@@ -36,7 +36,7 @@ def autolabel(ax, rects):
     for rect in rects:
         height = rect.get_height()
         ax.text(rect.get_x() + rect.get_width()/2., height,
-                '%d' % int(height),
+                '%d' % round(height),
                 ha='center', va='bottom')
 
 def cumulative_coverage(data):
@@ -395,6 +395,37 @@ def plot_distrib_extended(ucsd):
     plt.savefig('distrib_ext.png')
     plt.close()
 
+def plot_blame():
+    # xy_s = cumulative_coverage(seminal)
+    # xy_u = cumulative_coverage(ucsd)
+
+    # print ('xy_s', xy_s)
+    # print ('xy_u', xy_u)
+
+    # FIXME: load these numbers from csv...
+    tools = ['OCaml', 'NanoMaLy', 'Mycroft', 'SHErrLoc']
+    accs  = [45.7,    67.9,       71.5,      75.4      ]
+
+    N = len(tools)
+    ind = np.arange(N)    # the x locations for the groups
+    width = 0.5       # the width of the bars: can also be len(x) sequence
+
+    fig = plt.figure()
+    p1 = plt.bar(ind, accs, width,
+                 align='center',
+                 color=COLORS[0])
+
+    #plt.xlabel('Witness found in <= x seconds', fontsize=20)
+    plt.ylabel('Accuracy (%)', fontsize=20)
+    plt.title('Accuracy of Type Error Localization', fontsize=24)
+    plt.xticks(ind, tools, fontsize=20)
+    plt.yticks(np.arange(0, 101, 10), fontsize='large')
+    autolabel(plt, p1)
+    #autolabel(plt, p2)
+
+    # plt.show()
+    fig.savefig('blame.png')
+    plt.close()
 
 
 if __name__ == '__main__':
@@ -411,3 +442,5 @@ if __name__ == '__main__':
     plot_coverage(seminal, ucsd)
 
     plot_user_study()
+
+    plot_blame()

### 2017-04-27 paper/jfp-submission/study.R

@@ -99,7 +99,7 @@ print(kappam.fleiss(fixes,exact=TRUE,detail=TRUE))
 
 
 significance <- function(x,y) {
-  wilcox.test(x, y)
+  wilcox.test(x, y, alternative="t")
 }
 
 print("append reason")

### 2017-04-27 paper/jfp-submission/evaluation.tex
prune orphans

@@
 %
-There are three questions we seek to answer with our evaluation:
+There are four questions we seek to answer with our evaluation:
 %
@@
 \item \emphbf{Witness Utility}
-      How \emph{helpful} (qualitatively and quantitatively)
+      How \emph{helpful} %(qualitatively and quantitatively)
       are the witnesses and traces in debugging type errors?
+\item \emphbf{Witness-based Blame}
+      Can witnesses be used to \emph{locate} the source
+      of a type error?
 \end{enumerate}
@@
 either \ocaml's error message or \toolname's jump-compressed trace.
+
+We answer the last question on a subset of the \ucsdbench dataset.
+%
+For each ill-typed program we identify the student's \emph{fix}
+by searching for the first type-correct program that follows it in
+the interaction trace.
+%
+We then use an expression-level \emph{diff}~\cite{Lempsink2009-xf} to
+determine which sub-expressions changed between the ill-typed program
+and the student's fix, and treat those expressions as the source of the
+type error.
 
@@
 \ucsdbench\ dataset (35 for the \uwbench\ dataset) with a maximum of
-2,745 (986 for \uwbench) and a median of 17 (also 17 for \uwbench).
+2,745 (\resp 986) and a median of 17 (\resp 17).
 %
-The average number of jumps per trace is 7 (also 7 for \uwbench) with a
-maximium of 353 (185 for \uwbench) and a median of 4 (also 4 for
-\uwbench).
+The average number of jumps per trace is 7 (\resp 7) with a
+maximium of 353 (\resp 185) and a median of 4 (\resp 4).
 %
@@
 
-\subparagraph{Construct}
+\subparagraph{Construct.}
 %
@@
 
-\subparagraph{Internal}
+\subparagraph{Internal.}
 %
@@
 
-\subparagraph{External}
+\subparagraph{External.}
 %
-Our experiment is based on students in the process of learning \ocaml,
+Our experiment used students in the process of learning \ocaml,
 and thus may not generalize to all developers. The four
-programs we used were chosen manually, via a random selection and
+programs were chosen manually, via a random selection and
 filtering of the programs in the \ucsdbench dataset. In some cases we made
@@
 
-\subparagraph{Conclusion}
+\subparagraph{Conclusion.}
 %
@@
 to type errors: (1) with high coverage in a timespan amenable to
-compile-time analysis, (2) with traces that have a low average
-complexity of 7 jumps, and (3) that are more helpful to novice
-programmers than traditional type error messages.
+compile-time analysis; (2) with traces that have a low average
+complexity of 7 jumps; (3) that are more helpful to novice
+programmers than traditional type error messages; and (4) that
+can be used to automatically locate the source of a type error.
 

### 2017-04-27 paper/jfp-submission/interactive-short.tex
fiddling with figures..
prune orphans

@@
 (\S~\ref{sec:traversing-graph}), yielding an interactive debugger that
-allows the user to effectively visualize \emph{how} the program goes
+allows the user to visualize \emph{how} the program goes
 (wrong).
@@
 A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
-intuitively indicates that $e_1$ reduces, in a single step, to $e_2$.
+indicates that $e_1$ reduces, in a single step, to $e_2$.
 %
@@
 a relation of the form $\stepg{e}{\vsu}{\tsu}{\tr}{e'}{\vsu'}{\tsu'}{\tr'}$
-where $\tr'$ collects the edges of the transition.
+where $\tr'$ collects the transitions.
 
@@
 \centering
-\includegraphics[width=\linewidth]{fac-steps-new.png}
+\includegraphics[height=4in]{fac-steps-new.png}
 \caption{A sequence of interactions with the trace of
@@
 \item \stepforwardsym, \stepbackwardsym:
-      show the result of a single step forward or backward respectively,
+      show the result of a single step forward or backward;
 %
@@
       up to the first function call, or return, forward or backward
-      respectively,
+      respectively;
 %
@@
       show the result of stepping into a function call in a sub-term,
-      isolating it in a new thread, and
+      isolating it in a new reduction thread; and
 %

### 2017-04-27 paper/jfp-submission/safe.tex
prune orphans

@@
 %
-While extremely convenient to use, they pose a challenge for \toolname's
+While convenient to use, they pose a challenge for \toolname's
 combination of execution and inference.
@@
 %
-When given @bad@ as input, \toolname will generate two fresh holes
+When given @bad@, \toolname will generate two fresh holes
 $\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@@
 order to distinguish type errors that would have happened regardless, as
-in the above @bad@ function, from type errors that were the fault of our
-instantiation.
+in @bad@, from type errors that were caused by our instantiation.
 
@@
 %
-\ES{TODO: add some comment about using dependent types, though this particular example would be quite bizarre to encode in DT..}
+% \ES{TODO: add some comment about using dependent types, though this particular example would be quite bizarre to encode in DT..}
 

### 2017-04-27 paper/jfp-submission/semantics.tex
prune orphans

@@
 
-The evaluation relation is parameterized by a pair of functions:
-called \emph{narrow} (\forcesym) and \emph{generate} (\gensym),
+The evaluation relation is parameterized by a pair of functions,
+\emph{narrow} (\forcesym) and \emph{generate} (\gensym),
 that ``dynamically'' perform type-checking and hole-filling
@@
   %
-  Note that only \forcesym\ can change \tsu,
+  Only \forcesym\ can change \tsu,
   %, and it can only do so via \unifysym
@@
   In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
-  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
+  and $\thole$ trivially refines $\vhole{\thole}$.
   %
@@
 %
-The witness generation procedure returns as output a list of (general)
+The witness generation procedure returns a list of (general)
 witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
@@
 \begin{proof}
-  For any function $f$ of multiple arguments, we can define $f'$ to be
-  the uncurried version of $f$ that takes all of its arguments as a single
+  For any function $f$ of multiple arguments, we can define $f'$ as the
+  uncurried version of $f$ that takes all of its arguments as a single
   nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.

### 2017-04-27 paper/jfp-submission/locating.tex
prune orphans
update location numbers

@@
 %
+Intuitively, these two classes of locations correspond to \emph{sinks}
+and \emph{sources} for typing constraints.
+%
 For example, in the @sqsum@ program from \S~\ref{sec:advantage-traces}
@@
 We discard all such programs in our evaluation to level the playing
-field, leaving us with a benchmark set of 1,612 programs.
+field, leaving us with a benchmark set of 1,569 programs.
 
@@
 \caption{Accuracy of type error localization. \toolname's witness-based
-  predictions outperform \ocaml by 20 points, and are competitive
+  predictions outperform \ocaml by 22 points, and are competitive
   with the state-of-the-art tools \mycroft and \sherrloc.}

### 2017-04-27 paper/jfp-submission/plots.py

@@ -404,7 +404,7 @@ def plot_blame():
 
     # FIXME: load these numbers from csv...
     tools = ['OCaml', 'NanoMaLy', 'Mycroft', 'SHErrLoc']
-    accs  = [45.7,    67.9,       71.5,      75.4      ]
+    accs  = [49.0,    70.7,       71.0,      78.9      ]
 
     N = len(tools)
     ind = np.arange(N)    # the x locations for the groups

### 2017-04-27 paper/jfp-submission/evaluation.tex
tweak citations

@@
 \caption{Results of our coverage testing and the distribution of test
-  outcomes. Our random search successfully finds witnesses for 79--85\% of
-  the programs in under one second, improving to 87\% in under
-  10 seconds. In both datasets we detect actual type errors about 82\%
-  of the time, unbound variables or constructors 3--4\% of the time, and
-  diverging loops 2--3\% of the time. For the remaining 11--12\% of the
+  outcomes. Our random search successfully finds witnesses for 75--82\% of
+  the programs in under one second, improving to 84\% in under
+  10 seconds. In both datasets we detect actual type errors at least 75\%
+  of the time, unbound variables or constructors 4--5\% of the time, and
+  diverging loops 3\% of the time. For the remaining 13--16\% of the
   programs we are unable to provide any useful feedback.  }
@@
 %
-In both datasets our tool was able to find a witness for 83\% of the
+In both datasets our tool was able to find a witness for at least 75\% of the
 programs in under one second, \ie\ fast enough to be integrated as a
 compile-time check. If we extend our tolerance to a 10 second timeout,
-we hit a maximum of 87\% coverage.
+we hit a maximum of 83--86\% coverage.
 %
 Interestingly, while the vast majority of witnesses corresponded to a
-type-error, as expected, 3--4\% triggered an unbound variable error (even
-though \ocaml\ reported a type error) and 2--3\% triggered an infinite
+type-error, as expected, 4--5\% triggered an unbound variable error (even
+though \ocaml\ reported a type error) and 3\% triggered an infinite
 recursion error.
 %
-For the remaining 12\% of programs we were unable to provide any useful
+For the remaining 13--16\% of programs we were unable to provide any useful
 feedback as they either completed 1,000 tests successfully, or timed out
@@
 While a more advanced search procedure, \eg\ dynamic-symbolic execution,
-could likely trigger more of the type errors, our experiments suggest that
+could likely uncover more errors, our experiments suggest that
 type errors are coarse enough (or that novice programs are \emph{simple}
@@
 The trace clarifies (via the third step)
-that the @[]@ is the result of the recursive call
-@sumList []@, and shows how it is incompatible with
+that the @[]@ results from the recursive call
+\hbox{@sumList []@,} and shows how it is incompatible with
 the subsequent \texttt{+} operation.
@@
 %
-In contrast, \toolname makes no assumptions about @append@
-yielding a trace that illustrates the true error on line 4, by
+In contrast, \toolname makes no assumptions about @append@,
+yielding a trace that illustrates the error on line 4, by
 highlighting the conflict in consing a list onto a list of integers.
@@
 To summarize, our experiments demonstrate that \nanomaly finds witnesses
-to type errors: (1) with high coverage in a timespan amenable to
-compile-time analysis; (2) with traces that have a low average
-complexity of 7 jumps; (3) that are more helpful to novice
-programmers than traditional type error messages; and (4) that
-can be used to automatically locate the source of a type error.
+to type errors:
+%
+(1) with high coverage in a timespan amenable to compile-time analysis;
+%
+(2) with traces that have a low average complexity of 7 jumps;
+%
+(3) that are more helpful to novice programmers than traditional type
+error messages; and
+%
+(4) that can be used to automatically locate the source of a type error.
 
 There are, of course, drawbacks to our approach. Four that stand out
-are: (1) coverage limits due to random generation, (2) the inability to
-handle certain instances of infinite types, (3) dealing with an
-explosion in the size of generated traces, and (4) handling ad-hoc
-polymorphism.
+are:
+%
+(1) coverage limits due to random generation;
+%
+(2) the inability to handle certain instances of infinite types;
+%
+(3) dealing with an explosion in the size of generated traces; and
+%
+(4) handling ad-hoc polymorphism.
 
@@
 show, however, novice programs do not appear to require more advanced
-search techniques, likely because the novice programs tend to be simple.
+search techniques, likely because they tend to be simple.
 

### 2017-04-27 paper/jfp-submission/related.tex
tweak citations

@@
 %
-Zhang~\etal~\cite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
+Zhang~\etal~\shortcite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
 identifying the most likely culprit using Bayesian reasoning.
 %
-\cite{Pavlinovic2014-mr,Pavlinovic2015-kh} translate the %error
+Pavlinovic~\etal~\shortcite{Pavlinovic2014-mr,Pavlinovic2015-kh}
+translate the %error
 localization problem to a MaxSMT optimization problem, using
 compiler-provided weights to rank the possible sources.
+%
+Loncaric~\etal~\shortcite{Loncaric2016-uk} improve the scalability of
+Pavlinovic~\etal by reusing the existing type checker as
+a theory solver in the Nelson-Oppen~\shortcite{Nelson1979-td}
+style, thus requiring only a MaxSAT solver.
 
 % \subparagraph{Repairing}
-In addition to localizing the error, \cite{Lerner2007-dt} attempt to
+In addition to localizing the error, Lerner~\etal~\shortcite{Lerner2007-dt} attempt to
 suggest a fix by replacing expressions (or removing them entirely) with
@@
 %
-\cite{Chen2014-gd} use a variational type system to allow for the
+Chen~\&~Erwig~\shortcite{Chen2014-gd} use a variational type system to allow for the
 possibility of changing an expression's type, and search for an
@@
 In contrast to Lerner~\etal, who search for changes at the value-level,
-Chen~\etal\ search at the type-level and are thus complete due the finite
+Chen~\&~Erwig search at the type-level and are thus complete due the finite
 universe of types used in the program.
@@
 ill-typed but truly wrong. In addition, allowing users to run their
-program % (even knowing that it is wrong)
+program (even knowing that it is wrong)
 enables experimentation and the
@@
 %
-\cite{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
+Marceau~\etal~\shortcite{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
 messages in novice environments and present suggestions for improving
@@
 %
-\cite{Hage2006-hc} identify a variety of general heuristics to improve
+Hage~\&~Heeren~\shortcite{Hage2006-hc} identify a variety of general heuristics to improve
 the quality of type error messages, based on their teaching experience.
 %
-\cite{Heeren2003-db}, \cite{Christiansen2014-qc}, and
-\cite{Serrano2016-oo} provide methods for library authors to specialize
+Heeren~\etal~\shortcite{Heeren2003-db},
+Christiansen~\shortcite{Christiansen2014-qc}, and
+Serrano~\&~Hage~\shortcite{Serrano2016-oo}
+provide methods for library authors to specialize
 type errors with domain-specific knowledge.
@@
 \label{sec:running-ill-typed}
-\cite{Vytiniotis2012-gh} extend the \haskell
+Vytiniotis~\etal~\shortcite{Vytiniotis2012-gh} extend the \haskell
 compiler GHC to support compiling ill-typed programs, but their intent
@@
 %
-\cite{Bayne2011-cn} also provide a semantics for running
+Bayne~\etal~\shortcite{Bayne2011-cn} also provide a semantics for running
 ill-typed (\java) programs, but in constrast transform the program to
@@
 implementation of \gensym by a call to the solver. We chose to use lazy,
-random generation instead because it is efficient, and % does not incur
+random generation instead because it is efficient, does not incur
 the overhead of an external solver, and produces high coverage for our
@@
 
-A function's type is a theorem about the function's behavior.
+A function's type is a theorem about the its behavior.
 Thus, \toolname's witnesses can be viewed as \emph{counter-examples},
-thereby connecting it to work on using test generation to find
+thereby connecting it to work on using test cases to find
 counter-examples prior to starting a proof~\cite{ACL2Testing, Seidel15}.
@@
 
-Flanagan \etal~\cite{Flanagan96} describe a static debugger for Scheme, which helps
+Flanagan~\etal~\shortcite{Flanagan96} describe a static debugger for Scheme, which helps
 the programmer interactively visualize problematic source-sink flows
@@
 %
-Perera \etal~\cite{Perera2012-dy} present a tracing semantics
+Perera~\etal~\shortcite{Perera2012-dy} present a tracing semantics
 for functional programs that tags values with their provenance, enabling

### 2017-04-27 paper/jfp-submission/safe.tex
tweak citations

@@
 %
-This left us with 473 programs requiring manual coding; we selected
+This left us with 473 programs that required manual coding; we selected
 a random sample of 50 programs to investigate, and will report results
@@
 %
-Figure~\ref{fig:no-witness} summarizes the results of our investigation,
-we note the classes that were based on a random sample with a ``*''.
+Figure~\ref{fig:no-witness} summarizes the results of our investigation ---
+we note the classes that were based on the random sample with a ``*''.
 
@@
 
-\paragraph{Overloaded Functions.}
+\paragraph{Overloaded Functions}
 %
@@
 %
-This speculative instantiation is of course unsound, we would have to
+This speculative instantiation is of course unsound; we would have to
 take care to avoid reporting frivolous type errors that were caused by
@@
 %
-For example, suppose we are given
+For example, suppose we are given \pagebreak
 %
@@
 
-\paragraph{Non-Parametric Function Type *.}
+\paragraph{Non-Parametric Function Type *}
 %
@@
 
-\paragraph{Dead Code and ``Safe'' Function Calls *.}
+\paragraph{Dead Code and ``Safe'' Function Calls *}
 %
@@
 %
-Thus, a warning like that provided by \ocaml's pattern-exhaustiveness
+Thus, a warning like that provided by \ocaml's pattern exhaustiveness
 checker would be helpful.
@@
 A further 3\% included a function call where the
-student had supplied ill-typed inputs, yet the path induced
-by the call contained no errors.
+student supplied ill-typed inputs, but the path induced
+by the call did not contain an error.
 %
-Consider, for example the following |assoc| function, which
-looks up a key in an \emph{association list} and returns a default if
-the key cannot be found.
+Consider the following |assoc| function, which
+looks up a key in an \emph{association list}, returning a default if
+it cannot be found.
 %
@@
 %
-In this particular call, the key |123| is in the list, so the default
-will not be used (and even if it were, no error would be triggered) and
+In this particular call the key |123| is in the list, so the default
+will not be used (even if it were, there would not be an error) and
 \ocaml's complaint is moot.
-
+%
 Of course, \ocaml cannot be expected to know that this particular call
@@
 
-\paragraph{Witness Exists *.}
+\paragraph{Witness Exists *}
 %
@@
 
-\paragraph{Summary.}
+\paragraph{Summary}
 %
-Our investigation suggests that the majority of programs
-(84\%) for which we fail to find a witness do not, in fact, admit a
+Our investigation suggests that the vast majority of programs
+for which we fail to find a witness do not, in fact, admit a
 witness.
 %
-Overall, it appears that our random instantiation of holes is well-suited
-to finding type errors in student programs.
-%
-The remaining programs were all cases where \ocaml's type system was
+% Overall, it appears that our random instantiation of holes is well-suited
+% to finding type errors in student programs.
+These programs were generally cases where \ocaml's type system was
 overly conservative.
@@
 %
-it would be interesting to investigate how demonstrate these issues in a
-similar manner.
+it would be interesting to investigate how demonstrate these issues in an
+intuitive manner.
 

### 2017-04-27 paper/jfp-submission/semantics.tex
tweak citations

@@
 
-\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
+\paragraph{Narrowing Types}
+The procedure % $\force{v}{t}{\vsu}{\tsu}$,
 \[
 \forcesym : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
-\]
+\]%
 defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
@@
 
-\begin{thm}{\textbf{[Witness Generality]}}
+\begin{thm}[Witness Generality]
 \label{thm:soundness}
@@
 
-\begin{cor}{\textbf{[Witness Generation]}}
+\begin{cor}[Witness Generation]
 \label{thm:generation}

### 2017-04-27 paper/jfp-submission/evaluation.tex
use nate-style color boxes

@@
 %
-For each example we will present
-(1)~the code,
-(2)~the error message returned \ocaml,
-(3)~the error locations returned by \ocaml\ (underlined) and \sherrloc\ (in bold),\footnote{When the locations from \ocaml\ and \sherrloc\ overlap,
-we just underline the relevant code.}
-and (4)~the jump-compressed trace produced by \toolname.
+For each example we will present:
+(1)~the code;
+(2)~the error message returned \ocaml;
+(3)~the error locations returned by \hlOcaml{\ocaml} and \hlSherrloc{\sherrloc};
+and (4)~\toolname's jump-compressed trace.
 
@@
     | [] -> 0
-    | h::t -> __(sqsum t)__ @ (h * h)
+    | h::t -> (*@\hlOcaml{\hlSherrloc{sqsum t}}@*) @ (h * h)
 \end{ecode}
@@
   let rec sumList xs = match xs with
-    | []    -> ==[]==
-    | y::ys -> y + __sumList ys__
+    | []    -> (*@\hlSherrloc{hello}@*)
+    | y::ys -> y + (*@\hlOcaml{sumList ys}@*)
 \end{ecode}
@@
     else
-      append (==digitsOfInt (n / 10)==)
-             [__n mod 10__]
+      append ((*@\hlSherrloc{digitsOfInt (n / 10)}@*))
+             [(*@\hlOcaml{n mod 10}@*)]
 \end{ecode}
@@
 
-  let _ = wwhile (__f__, 2)
+  let _ = wwhile ((*@\hlOcaml{\hlSherrloc{f}}@*), 2)
 \end{ecode}

### 2017-04-27 paper/jfp-submission/introduction.tex
tweak citations
use nate-style color boxes

--- 2016-09-08 paper/icfp16-cameraready/introduction.tex
@@
     else
-    n * __fac (n-1)__
+      n * (*@\hlOcaml{fac (n-1)}@*)
 \end{ecode}
-\vspace{3em}
-% \end{minipage}
-% % \hspace{0.3in}
-% \begin{minipage}{.39\linewidth}
-\centering
+\vspace{2em}
 \includegraphics[height=1.5in]{fac-overview.png}
@@
 \end{minipage}
-\caption{(top-left) An ill-typed \texttt{fac} function, the error location reported by \ocaml is underlined; (bottom-left) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (right) The same trace, fully expanded.}
+\vspace{1em}
+\caption{(top-left) An ill-typed \texttt{fac} function \hlOcaml{highlighting} the error location reported by \ocaml.
+(bottom-left) Dynamically witnessing the type error in \texttt{fac}, showing only function call-return pairs.
+(right) The same trace, fully expanded to show each small-step reduction in the computation.}
 \label{fig:factorial}

### 2017-04-27 paper/jfp-submission/abstract.tex
tweak numbers...

--- 2016-09-08 paper/icfp16-cameraready/abstract.tex
@@
   Our technique is able to generate witnesses for
-  88\% of the programs, and our reduction graph
-  yields small counterexamples for 81\% of the witnesses.
+  around 85\% of the programs, and our reduction graph
+  yields small counterexamples for over 80\% of the witnesses.
   %

### 2017-04-27 paper/jfp-submission/evaluation.tex
tweak numbers...

@@
 \caption{Results of our coverage testing and the distribution of test
-  outcomes. Our random search successfully finds witnesses for 75--82\% of
-  the programs in under one second, improving to 84\% in under
+  outcomes. Our random search successfully finds witnesses for 76--82\% of
+  the programs in under one second, improving to 84--85\% in under
   10 seconds. In both datasets we detect actual type errors at least 75\%
@@
 %
-In both datasets our tool was able to find a witness for at least 75\% of the
+In both datasets our tool was able to find a witness for over 75\% of the
 programs in under one second, \ie\ fast enough to be integrated as a
 compile-time check. If we extend our tolerance to a 10 second timeout,
-we hit a maximum of 83--86\% coverage.
+we reach 84--85\% coverage, and if we allow a 60 second search,
+we hit a maximum of 84--87\% coverage.
 %
@@
 \vspace{3ex}
-\caption{Complexity of the generated traces. 81\% of the combined traces
+\caption{Complexity of the generated traces. Over 80\% of the combined traces
   have a jump complexity of at most 10, with an average complexity of 7
@@
 %
-The average number of single-step reductions per trace is 31 for the
-\ucsdbench\ dataset (35 for the \uwbench\ dataset) with a maximum of
-2,745 (\resp 986) and a median of 17 (\resp 17).
+The average number of single-step reductions per trace is 30 for the
+\ucsdbench\ dataset (45 for the \uwbench\ dataset) with a maximum of
+2,745 (\resp 982) and a median of 14.5 (\resp 14.5).
 %
-The average number of jumps per trace is 7 (\resp 7) with a
-maximium of 353 (\resp 185) and a median of 4 (\resp 4).
+The average number of jumps per trace is 7 (\resp 10) with a
+maximium of 353 (\resp 221) and a median of 4 (\resp 4).
 %

### 2017-04-27 paper/jfp-submission/introduction.tex
tweak numbers...

@@
 We show that for both data sets, our technique is able to generate
-witnesses for nearly 90\% of the programs, in under a second in the
+witnesses for around 85\% of the programs, in under a second in the
 vast majority of cases.
@@
 Furthermore, we show that a simple interactive strategy yields
-compact counterexample traces with at most 5 steps for 57\%
-of the programs, and at most 10 steps for 81\% of the programs.
+compact counterexample traces with at most 5 steps for 60\%
+of the programs, and at most 10 steps for over 80\% of the programs.
 

### 2017-04-27 paper/jfp-submission/plots.py

@@ -42,7 +42,7 @@ def autolabel(ax, rects):
 def cumulative_coverage(data):
     headers = data[0]
     data = data[1:]
-    return [(l, int(100 * len([r for r in data
+    return [(l, round(100 * len([r for r in data
                                if float(r[2]) <= l
                                and r[4] in UNSAFE])
                           / float(len(data))))
@@ -367,10 +367,10 @@ def plot_distrib_extended(ucsd):
     rs_u = [len([r for r in ucsd[1:] if r[4] in o])
             for o in ALL_D_E] + [0,0,0]
     missed = rs_u[2]
-    rs_u[2] = int(missed * 0.44) # Non-parametric fun
-    rs_u[3] = int(missed * 0.12) # Dead code
-    rs_u[4] = int(missed * 0.28) # Safe call
-    rs_u[5] = int(missed * 0.16) # True miss
+    rs_u[2] = round(missed * 0.44) # Non-parametric fun
+    rs_u[3] = round(missed * 0.12) # Dead code
+    rs_u[4] = round(missed * 0.28) # Safe call
+    rs_u[5] = round(missed * 0.16) # True miss
 
     print ('rs_u', rs_u)
 

### 2017-04-28 paper/jfp-submission/evaluation.tex
update discussion section

@@
 %
-(2) the inability to handle certain instances of infinite types;
+(2) dealing with an explosion in the size of generated traces;
 %
-(3) dealing with an explosion in the size of generated traces; and
+%(3) the inability to handle certain instances of infinite types; and
+(3) our use of a non-parametric function type; and
 %
@@
 
-\paragraph{Infinite Types}
-Our implementation does check for infinite types inside \forcesym, but
-there are some degenerate cases where it is unable to detect
-them. Consider, the following buggy @replicate@
-%
-\begin{code}
-  let rec replicate n x =
-    if n <= 0 then
-      []
-    else
-      replicate (n-1) [x]
-\end{code}
-%
-This code produces a nested list (with @n@ levels of nesting) containing
-a single copy of @x@, instead of a list with @n@ copies of @x@. \ocaml\
-detects a cyclic \hbox{@'a = 'a list@} constraint in the recursive call
-and throws a type error, whereas \nanomaly\ happily % recurses @n@ times to
-produces the nested list.  Strictly speaking, this function itself cannot
-``go wrong'', the program would not get stuck until a \emph{client}
-attempted to use the result expecting a flat list. But this is not very
-satisfying as @replicate@ is clearly to blame. Furthermore, in our
-experience, infinite-type errors are often difficult to %some of the more difficult ones to
-debug (and to explain to novices), so better support for this scenario
-would be useful.
+% \paragraph{Infinite Types}
+% Our implementation does check for infinite types inside \forcesym, but
+% there are some degenerate cases where it is unable to detect
+% them. Consider the following buggy @replicate@.
+% %
+% \begin{code}
+%   let rec replicate n x =
+%     if n <= 0 then
+%       []
+%     else
+%       replicate (n-1) [x]
+% \end{code}
+% %
+% This code produces a nested list (with @n@ levels of nesting) containing
+% a single copy of @x@, instead of a list with @n@ copies of @x@. \ocaml\
+% detects a cyclic \hbox{@'a = 'a list@} constraint in the recursive call
+% and throws a type error, whereas \nanomaly\ happily recurses @n@ times to
+% produces the nested list.  Strictly speaking, this function itself cannot
+% ``go wrong'', the program would not get stuck until a \emph{client}
+% attempted to use the result expecting a flat list. But this is not very
+% satisfying as @replicate@ is clearly to blame. Furthermore, in our
+% experience, infinite-type errors are often difficult to %some of the more difficult ones to
+% debug (and to explain to novices), so better support for this scenario
+% would be useful.
 
@@
 
-\paragraph{Ad-hoc Polymorphism}
-Our approach can only support ad-hoc polymorphism (\eg\ type-classes in
-\haskell\ or polymorphic comparison functions in \ocaml) in limited cases
-where we have enough typing information at the call-site to resolve the
-overloading. For example, consider the @n <= 0@ test in our @fac@ example.
-@<=@ is polymorphic in \ocaml, but in this case we can make progress because
-the literal @0@ is not. If we parameterized @fac@ by a lower bound, \eg
+\paragraph{Non-Parametric Function Type}
+As we discussed in \S~\ref{sec:how-safe} some ill-typed programs
+lack a witness in our semantics due to our use of a non-parametric
+type $\tfun$ for functions.
 %
-\begin{code}
-  let rec fac n m =
-    if n <= m then
-      1
-    else
-      n * fac (n - 1) m
-\end{code}
+These programs cannot ``go wrong'', strictly speaking, but would be very
+difficult to \emph{use} in practice.
 %
-and called @fac@ with two holes, we would get stuck at the @n <= m@
-test; not because of a type error, but because all we know about
-@n@ and @m@ at that point is that they must have the same (unknown)
-type.
+We also note that many of these programs induce cyclic typing constraints,
+causing infinite-type errors, which in our experience can be particularly
+difficult to debug (and to explain to novices).
 
-This issue is uncommon in \ocaml\
-(we detected it in around 5\% of our benchmarks),
-but it would surely be exacerbated
-by a language like \haskell, which makes heavy use of overloading. We
-suspect that dynamic-symbolic execution would allow us to handle ad-hoc
-polymorphism, but defer a proper treatment to future work.
+Better support for these programs would be welcome.
+%
+For example, we might track how the types of inputs change between
+recursive calls.
+%
+If we cannot find a traditional witness, we could then produce a trace
+expanded to show these particular steps.
+
+\paragraph{Ad-Hoc Polymorphism}
+% Our approach can only support ad-hoc polymorphism (\eg\ type-classes in
+% \haskell\ or polymorphic comparison functions in \ocaml) in limited cases
+% where we have enough typing information at the call-site to resolve the
+% overloading. For example, consider the @n <= 0@ test in our @fac@ example.
+% @<=@ is polymorphic in \ocaml, but in this case we can make progress because
+% the literal @0@ is not. If we parameterized @fac@ by a lower bound, \eg
+% %
+% \begin{code}
+%   let rec fac n m =
+%     if n <= m then
+%       1
+%     else
+%       n * fac (n - 1) m
+% \end{code}
+% %
+% and called @fac@ with two holes, we would get stuck at the @n <= m@
+% test; not because of a type error, but because all we know about
+% @n@ and @m@ at that point is that they must have the same (unknown)
+% type.
+
+Also discussed in \S~\ref{sec:how-safe}, our approach can only support
+ad-hoc polymorphism (\eg\ type-classes in \haskell\ or polymorphic
+comparison functions in \ocaml) in limited cases where we have enough
+typing information at the call-site to resolve the overloading. This
+issue is uncommon in \ocaml\ (we detected it in around 5\% of our
+benchmarks), but it would surely be exacerbated by a language like
+\haskell, which makes heavy use of overloading. We suspect that either
+dynamic-symbolic execution or speculative instantiation of holes would
+allow us to handle ad-hoc polymorphism, but defer a proper treatment to
+future work.
 

### 2017-04-28 paper/jfp-submission/safe.tex
use nate-style color boxes
update discussion section

@@
-\subsubsection{How Safe are the ``Safe'' Programs?}
+\subsubsection{How safe are the ``safe'' programs?}
 \label{sec:how-safe}
@@
 %
-or did \toolname simply fail to find the witness?
+or did \toolname simply fail to find a witness?
 %
@@
 
-\paragraph{Overloaded Functions}
+\paragraph{Ad-Hoc Polymorphism}
 %
@@
 %
-Consider, for example
+For example, consider the \hbox{@n <= 0@} test in our @fac@ example.
+%
+The @<=@ operator is polymorphic, but in this case we can make progress because
+the literal @0@ is not.
+%
+Suppose, however, we parameterized @fac@ by a lower bound, \eg
 %
 \begin{code}
-  let bad x y =
-    if x < y then
-      1 + true
+  let rec fac n m =
+    if n <= m then
+      true
     else
-      0
+      n * fac (n - 1) m
 \end{code}
 %
-When given @bad@, \toolname will generate two fresh holes
+When given @fac@, \toolname will generate two fresh holes
 $\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
-@x < y@ comparison.
+@n < m@ comparison.
 %
@@
 Thus, we cannot perform the comparison and proceed, and must give up our
-search for a witness, even though one obviously exists, any pair of |x|
-and |y| such that |x < y| is true.
+search for a witness, even though one obviously exists, any pair of |n|
+and |m| such that |n <= m| is false.
 
@@
 this issue, as we could then begin symbolically executing the program
-until we learn how to instantiate |x| and |y|.
+until we learn how to instantiate |n| and |m|.
 %
-Alternatively, we could \emph{speculatively} instantiate both |x| and
-|y| with some arbitrary type, and proceed with execution until we
+Alternatively, we could \emph{speculatively} instantiate both |n| and
+|m| with some arbitrary type, and proceed with execution until we
 discover a type error.
@@
 order to distinguish type errors that would have happened regardless, as
-in @bad@, from type errors that were caused by our instantiation.
+in @fac@, from type errors that were caused by our instantiation.
 
@@
 %
-For example, suppose we are given \pagebreak
+For example, suppose we are given
 %

### 2017-04-28 paper/jfp-submission/plots.py

@@ -21,7 +21,7 @@ ALL_D = [ ['U', 'O'], ['B'], ['D'], ['S', 'T', 'H']]
 ALL_DL = [ 'Witness', 'Unbound', 'Diverge', 'No Witness'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
 
 ALL_D_E = [ ['U', 'O', 'B', 'D'], ['H'], ['S', 'T']]
-ALL_DL_E = [ 'Witness Found', 'Overloaded Function', 'Non-Parametric Function *', 'Dead Code *', 'Safe Call *', 'Witness Exists *'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
+ALL_DL_E = [ 'Witness Found', 'Ad-Hoc Polymorphism', 'Non-Parametric Function *', 'Dead Code *', 'Safe Call *', 'Witness Exists *'] #   ['S', 'T'], ['U'], ['B'], ['D'] ]
 
 
 def read_csv(f):

### 2017-04-28 paper/jfp-submission/evaluation.tex
misc
misc tweaks

@@
 
-\begin{figure}[htp]
-\centerline{
-\begin{minipage}{1.2\textwidth}
-\includegraphics[width=0.5\linewidth]{coverage.png}
+\begin{figure}[t]
+% \centerline{
+% \begin{minipage}{1.2\textwidth}
+\centering
+\includegraphics[width=0.7\linewidth]{coverage.png}
 % \end{minipage}
 % \begin{minipage}{\linewidth}
-\includegraphics[width=0.5\linewidth]{distrib.png}
-\end{minipage}
-}
-\vspace{3ex}
-\caption{Results of our coverage testing and the distribution of test
-  outcomes. Our random search successfully finds witnesses for 76--82\% of
-  the programs in under one second, improving to 84--85\% in under
-  10 seconds. In both datasets we detect actual type errors at least 75\%
-  of the time, unbound variables or constructors 4--5\% of the time, and
-  diverging loops 3\% of the time. For the remaining 13--16\% of the
-  programs we are unable to provide any useful feedback.  }
+% \includegraphics[width=0.6\linewidth]{distrib.png}
+% \end{minipage}
+% }
+% \vspace{3ex}
+\caption{Results of our coverage testing. Our random search successfully
+  finds witnesses for 76--82\% of the programs in under one second,
+  improving to 84--85\% in under 10 seconds. }
 \label{fig:results-witness}
+\end{figure}
+\begin{figure}[t]
+\includegraphics[width=0.7\linewidth]{distrib.png}
+\caption{Distribution of test outcomes. In both datasets we detect
+  actual type errors at least 75\% of the time, unbound variables or
+  constructors 4--5\% of the time, and diverging loops 3\% of the
+  time. For the remaining 13--16\% of the programs we are unable to
+  provide any useful feedback. }
+\label{fig:results-distrib}
 \end{figure}
@@
 The results of our experiments are summarized in
-Figure~\ref{fig:results-witness}.
+Figures~\ref{fig:results-witness}~and~\ref{fig:results-distrib}.
 %
@@
 \begin{figure}[t]
-\centerline{
-\begin{minipage}{1.2\textwidth}
-\includegraphics[width=0.5\linewidth]{trace_size_step.png}
-\includegraphics[width=0.5\linewidth]{trace_size_jump.png}
-\end{minipage}
-}
-\vspace{3ex}
+% \centerline{
+% \begin{minipage}{1.2\textwidth}
+\includegraphics[width=0.7\linewidth]{trace_size_step.png}
+\includegraphics[width=0.7\linewidth]{trace_size_jump.png}
+% \end{minipage}
+% }
+% \vspace{3ex}
 \caption{Complexity of the generated traces. Over 80\% of the combined traces
@@
 %
-In both datasets 80\% or more traces have at most 10 jumps.
+In both datasets 60\% of traces have at most 5 jumps, and 80\% or more
+have at most 10 jumps.
 
@@
 Unfortunately the student has used the list-append
-operator |@| instead of \texttt{+} to compute the sum.
+operator |@| instead of \texttt{+}. % to compute the sum.
 %
@@
 %
-\toolname\ produces a trace showing how the evaluation of
-@sqsum [1]@ gets stuck.
-%
 \begin{center}
@@
 \end{center}
+%
+\toolname\ produces the above trace showing how the evaluation of
+@sqsum [1]@ gets stuck.
 %
@@
 \begin{figure}[t]
-\centerline{
-\begin{minipage}{1.2\textwidth}
-\includegraphics[width=0.5\linewidth]{user-study-reason.png}
-\includegraphics[width=0.5\linewidth]{user-study-fix.png}
-\end{minipage}
-}
-\vspace{3ex}
+% \centerline{
+% \begin{minipage}{1.2\textwidth}
+\includegraphics[width=0.7\linewidth]{user-study-reason.png}
+\includegraphics[width=0.7\linewidth]{user-study-fix.png}
+% \end{minipage}
+% }
+% \vspace{3ex}
 \caption{A classification of students' explanations and fixes for type
@@
 %
-(2) with traces that have a low average complexity of 7 jumps;
+(2) with traces that have a low median complexity of 5 jumps;
 %
@@
 %
-(2) dealing with an explosion in the size of generated traces;
+(2) dealing with explosions in the size of generated traces;
 %
@@
 difficult to debug (and to explain to novices).
-
+%
 Better support for these programs would be welcome.

### 2017-04-28 paper/jfp-submission/plots.py

@@ -233,7 +233,7 @@ def plot_trace_size(seminal, ucsd):
     print('avg/med/max:\t{}\t{}\t{}'.format(np.mean(step_u), np.median(step_s), np.max(step_u)))
     p2 = plt.bar(ind + width, c_step_u, label=UCSD, width=width, color=COLORS[1])
     plt.legend((p1[0],p2[0]), ('UW',UCSD), loc='lower right', fontsize=16)
-    # plt.title('Trace Complexity', fontsize=24)
+    plt.title('Step Complexity', fontsize=24)
     plt.xlabel('Total Steps', fontsize=20)
     plt.ylabel('Traces (%)', fontsize=20)
     # ax.set_xlim(0,6)
@@ -269,7 +269,7 @@ def plot_trace_size(seminal, ucsd):
     print('avg/med/max:\t{}\t{}\t{}'.format(np.mean(jump_u), np.median(jump_s), np.max(jump_u)))
     p2 = plt.bar(ind + width, c_jump_u, label=UCSD, width=width, color=COLORS[1])
     plt.legend((p1[0],p2[0]), ('UW',UCSD), loc='lower right', fontsize=16)
-    # plt.title('Complexity of Traces (in jumps)', fontsize=24)
+    plt.title('Jump Complexity', fontsize=24)
     plt.xlabel('Total Jumps', fontsize=20)
     # plt.xlabel('Jumps', fontsize=20)
     plt.ylabel('Traces (%)', fontsize=20)
@@ -342,7 +342,7 @@ def plot_distrib(seminal, ucsd):
 
     #plt.tight_layout()
 
-    plt.suptitle('Distribution of Results', fontsize=24, y=0.9)
+    plt.suptitle('Distribution of Test Outcomes', fontsize=24, y=0.9)
     plt.figlegend(p1[0], ALL_DL, 'lower center', fontsize=18, ncol=2)
 
 

### 2017-04-28 paper/jfp-submission/locating.tex
misc tweaks
tweak
minor

@@
 We compare \toolname's witness-based predictions against a baseline of
-the \ocaml compiler as well as the state-of-the-art tools \sherrloc
-and \mycroft.
+the \ocaml compiler as well as the state-of-the-art %type error
+localization tools \sherrloc and \mycroft.
 %
-\sherrloc attempts to predict the most likely source of a type error
-by searching the typing constraint graph for constraints that participate
-in many unsatisfiable paths and few satisfiable paths.
+\sherrloc~\cite{Zhang2014-lv} attempts to predict the most likely source
+of a type error by searching the typing constraint graph for constraints
+that participate in many unsatisfiable paths and few satisfiable paths.
 %
-\mycroft reduces the localization problem to MaxSAT by searching for a
-minimal subset of constraints that can be removed, such that the
-resulting system is satisfiable.
+\mycroft~\cite{Loncaric2016-uk} reduces the localization problem to
+MaxSAT by searching for a minimal subset of constraints that can be
+removed, such that the resulting system is satisfiable.
 %
@@
 We discard all such programs in our evaluation to level the playing
-field, leaving us with a benchmark set of 1,569 programs.
+field, around 15\% for each tool, leaving us with a benchmark set of
+1,569 programs.
 
 \begin{figure}[t]
-\includegraphics[width=0.8\linewidth]{blame.png}
+\includegraphics[width=0.7\linewidth]{blame.png}
 \caption{Accuracy of type error localization. \toolname's witness-based

### 2017-04-28 paper/jfp-submission/evaluation.tex
tweak

@@
 \begin{enumerate}
-\item \emphbf{Witness Coverage}
-      How many ill-typed programs can we find witnesses for?
-\item \emphbf{Witness Complexity}
+\item \emphbf{Witness Coverage} (\S~\ref{sec:eval:witness-coverage},~\ref{sec:how-safe})
+      How many ill-typed programs \emph{admit} witnesses?
+\item \emphbf{Witness Complexity} (\S~\ref{sec:trace-complexity})
       How \emph{complex} are the traces produced by the witnesses?
-\item \emphbf{Witness Utility}
+\item \emphbf{Witness Utility} (\S~\ref{sec:advantage-traces},~\ref{sec:user-study})
       How \emph{helpful} %(qualitatively and quantitatively)
-      are the witnesses and traces in debugging type errors?
-\item \emphbf{Witness-based Blame}
+      are the witnesses in debugging type errors?
+\item \emphbf{Witness-based Blame} (\S~\ref{sec:locating})
       Can witnesses be used to \emph{locate} the source
-      of a type error?
+      of an error?
 \end{enumerate}

### 2017-04-28 paper/jfp-submission/evaluation.tex
add a tldr results section

@@
 
-\paragraph{Benchmarks}
+In the sequel we present our experimental methodology (\S~\ref{sec:methodology})
+and then answer the above questions.
+%
+However, for the impatient reader, we begin with a quick summary of our main results:
+
+\paragraph{1. Most Type Errors Admit Witnesses}
+Our prime result is that the vast majority of static type errors, around
+85\%, do in fact admit a dynamic witness.
+%
+Further, \toolname efficiently synthesizes witnesses with its randomized search;
+it can synthesize a witness for over 75\% of programs in under one second, \ie
+fast enough to be integrated into the edit-compile-debug cycle.
+%
+
+\paragraph{2. Jump-Compressed Traces Are Small}
+We find that our jump-compression heuristic effectively abstracts the
+pedestrian details of computation, compressing the median trace with
+14--15 single-step reductions to only 4 jumps.
+%
+Over 80\% of programs have a jump-compressed trace with at most 10
+jumps, providing a bird's-eye view from which we can launch a more
+in-depth investigation.
+
+\paragraph{3. Witnesses Help Novices}
+A witness should also help programmers \emph{understand} and
+\emph{fix} type errors.
+%
+We use a set of ill-typed student programs to show that \toolname's
+witnesses effectively demonstrate the runtime error that the type
+system prevented.
+%
+Furthermore, we find, in a study of undergraduate students, that
+\toolname's witnesses lead to more accurate diagnoses and fixes of type
+errors than \ocaml's type error messages.
+
+\paragraph{4. Witnesses Assign Blame}
+Finally, we find that a simple heuristic allows us to use witnesses
+to \emph{automatically} assign blame for type errors.
+%
+We treat the values inside the stuck term as \emph{sources} of typing
+constraints and the stuck term itself as a \emph{sink}, producing
+a small slice of the program that likely contains the error.
+%
+Using this heuristic, \toolname's witness are competitive with the
+state-of-the-art type error localization tools \mycroft and \sherrloc.
+
+\subsection{Methodology}
+\label{sec:methodology}
 We answer the first two questions on two sets of ill-typed programs,
@@
 %
-For each ill-typed program we identify the student's \emph{fix}
-by searching for the first type-correct program that follows it in
-the interaction trace.
+For each ill-typed program in a student's interaction trace, we identify
+the student's \emph{fix} by searching for the first type-correct program
+that follows it in the trace.
 %
@@
   let rec sumList xs = match xs with
-    | []    -> (*@\hlSherrloc{hello}@*)
+    | []    -> (*@\hlSherrloc{[]}@*)
     | y::ys -> y + (*@\hlOcaml{sumList ys}@*)

### 2017-04-28 paper/jfp-submission/locating.tex
todo
add a tldr results section

@@
 
-As discussed at the beginning of \S~\ref{sec:evaluation}, we recorded
+As discussed in \S~\ref{sec:methodology}, we recorded
 each interaction of our students with the \ocaml top-level system.
@@
 %
-We identify fixed programs by searching for the first type-correct
-program that follows the ill-typed program in the interaction trace.
+For each ill-typed program in a student's interaction trace, we identify
+the student's fix by searching for the first type-correct program
+that follows it in the trace.
 %
@@
 
+\ES{TODO: threats to validity}
+
 \begin{figure}[t]

### 2017-04-29 paper/jfp-submission/evaluation.tex
add threats for localization

@@
 \#140608),
-from which we extracted \ucsdsize\ distinct, ill-typed \ocaml\ programs.
+from which we extracted \ucsdsize\ distinct, ill-typed \ocaml\ programs
+from a cohort of 46 students..
 %
@@
 %
+\toolname\ produces a trace showing how the evaluation of
+@sqsum [1]@ gets stuck.
+%
 \begin{center}
@@
 \end{center}
-%
-\toolname\ produces the above trace showing how the evaluation of
-@sqsum [1]@ gets stuck.
 %

### 2017-04-29 paper/jfp-submission/locating.tex
add threats for localization

@@
 
-\ES{TODO: threats to validity}
+\paragraph{Threats to Validity}
+Our benchmarks were drawn from students in an undergraduate course
+at \ucsdbench\ and may not be representative of other student bodies.
+%
+We mitigate this threat with a large empirical evaluation of 1,569
+programs, drawn from a cohort of 46 students.
+%
+A similar threat is that students are not industrial programmers, thus
+our results may not translate to large-scale software engineering.
+%
+In particular, \toolname's reliance on running the program suggests that
+scaling it to larger programs would be more difficult than static
+methods.
+%
+However, in our experience programmers are able to construct a mental
+model of type systems after sufficient exposure, at which point
+traditional error reports may suffice.
+%
+We are thus particularly interested in aiding novice programmers as
+they learn to work with the type system.
+
+Our definition of the next well-typed program as the intended ground
+truth answer is another threat to validity. Students might submit
+multiple well-typed ``rewrites'' between the initial ill-typed program
+and the final intended answer.
+%
+Our approach to discarding outliers is intended to mitigate this threat.
+%
+A similar threat is our removal of programs where any of the tools could
+not produce an answer.
+%
+It may be, for example, that \mycroft and \sherrloc are particularly
+effective on programs that do not admit dynamic witnesses.
+%
+Finally, our use of student fixes as oracles for the source of type
+errors assumes that students are able to correctly identify the source.
+%
+As the students are in the process of learning \ocaml and the type
+system, this assumption may be faulty, \emph{expert} users may disagree
+with the student fixes.
+%
+We believe, however, that it is reasonable to use student fixes as
+oracles, as the student is the best judge of what she \emph{intended} to
+do.
 

### 2017-04-29 paper/jfp-submission/abstract.tex
mention localization in abs/intro

--- 2016-09-08 paper/icfp16-cameraready/abstract.tex
@@
   student programs.
+  %
   Our technique is able to generate witnesses for
-  88\% of the programs, and our reduction graph
-  yields small counterexamples for 81\% of the witnesses.
+  around 85\% of the programs, our reduction graph
+  yields small counterexamples for over 80\% of the witnesses,
+  and a simple heuristic allows us to use witnesses to
+  locate the source of type errors with around 70\% accuracy.
   %

### 2017-04-29 paper/jfp-submission/evaluation.tex
mention localization in abs/intro

@@
 %
-However, for the impatient reader, we begin with a quick summary of our main results:
+However, for the impatient reader, we first summarize our main results:
 
@@
 it can synthesize a witness for over 75\% of programs in under one second, \ie
-fast enough to be integrated into the edit-compile-debug cycle.
+fast enough for interactive use. %to be integrated into the edit-compile-debug cycle.
 %
@@
 \paragraph{4. Witnesses Assign Blame}
-Finally, we find that a simple heuristic allows us to use witnesses
+Finally, we present a simple heuristic that allows us to use witnesses
 to \emph{automatically} assign blame for type errors.
@@
 constraints and the stuck term itself as a \emph{sink}, producing
-a small slice of the program that likely contains the error.
+a slice of the program that likely contains the error.
 %
 Using this heuristic, \toolname's witness are competitive with the
-state-of-the-art type error localization tools \mycroft and \sherrloc.
+state-of-the-art localization tools \mycroft and \sherrloc.
 
@@
     else
-      append ((*@\hlSherrloc{digitsOfInt (n / 10)}@*))
-             [(*@\hlOcaml{n mod 10}@*)]
+      append ((*@\hlSherrloc{digitsOfInt (n / 10)}@*)) [(*@\hlOcaml{n mod 10}@*)]
 \end{ecode}

### 2017-04-29 paper/jfp-submission/evaluation.tex
significance

@@
 
+Finally, we used a one-sided Mann-Whitney $U$ test~\cite{Mann1947-fd} to
+determine the significance of our results.
+%
+The null hypothesis was that the responses from students given
+\toolname's witnesses were drawn from the same distribution as those
+given \ocaml's errors, \ie \toolname had no effect.
+%
+Since we used a one-sided test, the alternative to the null hypothesis
+is that \toolname had a \emph{positive} effect on the responses.
+%
+We reject the null hypothesis in favor of the alternative if the test
+produces a significance level $p < 0.05$, a standard threshold for
+determining statistical significance.
+
 \paragraph{Threats to Validity}
@@
 %
-We assigned students randomly to two groups. The first group was given
+We assigned students randomly to two groups. The first was given
 \ocaml's errors for @append@ and \hbox{@digitsOfInt@,} and \toolname's trace
-for @sumList@ and @wwhile@. The second group was given the opposite
-assignment of errors and traces. This assignment ensured that (1) each
-student was given \ocaml and \toolname problems, and (2) each student
+for @sumList@ and \hbox{@wwhile@;} the second was given the opposite
+assignment of errors and traces. This assignment ensured that: (1) each
+student was given \ocaml and \toolname problems; and (2) each student
 was given an ``easy'' and ``hard'' problem for both \ocaml and
@@
 %
-Collecting more responses per test pair was not possible, as it
+Our results are statistically significant in only 2 out of 8 tests; however,
+collecting more responses per test pair was not possible as it
 would require having students answer the same problem twice (once with
@@
 \caption{A classification of students' explanations and fixes for type
-  errors, given either \ocaml's error % message
-  or \toolname's
+  errors, given either \ocaml's error message or \toolname's
   jump-compressed trace. The students given \toolname's jump-compressed
-  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
-  error.}
+  trace consistently scored better ($\ge 10\%$) than those given
+  \ocaml's type error. We report the result of a one-sided Mann-Whitney
+  $U$ test for statistical significance in parentheses.}
 \label{fig:results-user-study}
@@
 \paragraph{Results}
+%
+The measured kappa values were $\kappa = 0.72$ for the explanations and
+$\kappa = 0.83$ for the fixes; while there is no formal notion for what
+consititutes strong agreement~\cite{Krippendorff2012-wd}, kappa values
+above $0.60$ are often called ``substantial''
+agreement~\cite{Landis1977-ey}.
 %
@@
 the students who had access to \toolname's traces had a better
-understanding of the type errors.
+understanding of the type errors;
 %
-The measured kappa values were $\kappa = 0.72$ for the explanations and
-$\kappa = 0.83$ for the fixes; while there is no formal notion for what
-consititutes strong agreement~\cite{Krippendorff2012-wd}, kappa values
-above $0.60$ are often called ``substantial''
-agreement~\cite{Landis1977-ey}.
-%
+however, only the @append@ tests were statistically significant at
+$p < 0.05$.
 %

### 2017-04-29 paper/jfp-submission/plots.py

@@ -132,7 +132,7 @@ def plot_user_study():
     plt.title('Explanation',fontsize=24)
     # plt.xlabel('Problem', fontsize=20)
     plt.ylabel('% Correct', fontsize=20)
-    plt.xticks(ind + width, ['sumList', 'append', 'digitsOfInt', 'wwhile'], fontsize='large')
+    plt.xticks(ind + width, ['sumList\n(p = 0.061)', 'append\n(p = 0.018)', 'digitsOfInt\n(p = 0.12)', 'wwhile\n(p = 0.14)'], fontsize='large')
     plt.legend(('OCaml', 'NanoMaLy'), loc='lower right', fontsize=16)
     # autolabel(plt, p_o)
     # autolabel(plt, p_n)
@@ -174,7 +174,7 @@ def plot_user_study():
     plt.title('Fix',fontsize=24)
     # plt.xlabel('Problem', fontsize=20)
     plt.ylabel('% Correct', fontsize=20)
-    plt.xticks(ind + width, ['sumList', 'append', 'digitsOfInt', 'wwhile'], fontsize='large')
+    plt.xticks(ind + width, ['sumList\n(p = 0.067)', 'append\n(p = 0.038)', 'digitsOfInt\n(p = 0.083)', 'wwhile\n(p = 0.20)'], fontsize='large')
     plt.legend(('OCaml', 'NanoMaLy'), loc='lower right', fontsize=16)
     # autolabel(plt, p_o)
     # autolabel(plt, p_n)

### 2017-04-29 paper/jfp-submission/study.R

@@ -97,108 +97,132 @@ print(table(fixes[,4]))
 print(kappam.fleiss(fixes,exact=TRUE,detail=TRUE))
 #print(kappam.fleiss(fixes[1:3],detail=TRUE))
 
-
 significance <- function(x,y) {
-  wilcox.test(x, y, alternative="t")
+  print(x)
+  print(y)
+  wilcox.test(x, y, alternative="g")
 }
 
 print("append reason")
-#print(table(a3$X1..append.reason))
-#print(table(b3$X5..append.reason))
-print(significance(a3$X1..append.reason[a3$X1..append.reason != -1],
-             b3$X5..append.reason[b3$X5..append.reason != -1]))
+print(significance(
+        b3$X5..append.reason[b3$X5..append.reason != -1],
+        a3$X1..append.reason[a3$X1..append.reason != -1]
+))
+
 print("digistofint reason")
-#print(table(a3$X3..digitsofint.reason))
-#print(table(b3$X7..digitsofint.reason))
-print(significance(a3$X3..digitsofint.reason[a3$X3..digitsofint.reason != -1],
-             b3$X7..digitsofint.reason[b3$X7..digitsofint.reason != -1]))
+print(significance(
+        b3$X7..digitsofint.reason[b3$X7..digitsofint.reason != -1],
+        a3$X3..digitsofint.reason[a3$X3..digitsofint.reason != -1]
+))
+
 print("sumlist reason")
-#print(table(b3$X1..sumlist.reason))
-#print(table(a3$X5..sumlist.reason))
-print(significance(b3$X1..sumlist.reason[b3$X1..sumlist.reason != -1],
-             a3$X5..sumlist.reason[a3$X5..sumlist.reason != -1]))
+print(significance(
+        a3$X5..sumlist.reason[a3$X5..sumlist.reason != -1],
+        b3$X1..sumlist.reason[b3$X1..sumlist.reason != -1]
+))
+
 print("wwhile reason")
-#print(table(b3$X3..wwhile.reason))
-#print(table(a3$X7..wwhile.reason))
-print(significance(b3$X3..wwhile.reason[b3$X3..wwhile.reason != -1],
-             a3$X7..wwhile.reason[a3$X7..wwhile.reason != -1]))
+print(significance(
+        a3$X7..wwhile.reason[a3$X7..wwhile.reason != -1],
[…]

### 2017-04-29 paper/jfp-submission/study.R
add effect size calculation according to Arcuri (2014)

@@ -101,6 +101,10 @@ significance <- function(x,y) {
   print(x)
   print(y)
   wilcox.test(x, y, alternative="g")
+  R1=sum(rank(c(x,y))[seq_along(x)])
+  m=length(x)
+  n=length(y)
+  A12 = (R1/m - (m+1)/2)/n
 }
 
 print("append reason")

### 2017-04-30 paper/jfp-submission/evaluation.tex
add user-study to appendix

@@
 For each problem the students were given the ill-typed program and
-either \ocaml's error or \toolname's jump-compressed trace.
+either \ocaml's error or \toolname's jump-compressed trace;
+the full user study is available in \S~\ref{sec:user-study}.
 %

### 2017-04-30 paper/jfp-submission/locating.tex
minor

@@
 %
-There were a number of student programs where \mycroft or \sherrloc
+There were a number of programs where \mycroft or \sherrloc
 encountered an unsupported language feature or timed out after one
@@
 %
-In particular, \toolname's reliance on running the program suggests that
-scaling it to larger programs would be more difficult than static
-methods.
+% In particular, \toolname's reliance on running the program suggests that
+% scaling it to larger programs would be more difficult than static
+% methods.
 %

### 2017-04-30 paper/jfp-submission/evaluation.tex
get rid of interaction trace language, seems to bother RJ

@@
 %
-For each ill-typed program in a student's interaction trace, we identify
-the student's \emph{fix} by searching for the first type-correct program
-that follows it in the trace.
+% For each ill-typed program in a student's interaction trace, we identify
+% the student's \emph{fix} by searching for the first type-correct program
+% that follows it in the trace.
+For each ill-typed program compiled by a student, we identify the student's
+\emph{fix} by searching for the first type-correct program that the student
+subsequently compiled.
 %

### 2017-04-30 paper/jfp-submission/interactive-short.tex
Wes pass over the the entire document. Minor editing changes only.

@@
 Note that jump compressed traces are not quite the same as
-stack-traces as they show \emph{all} function calls, including
+stack traces as they show \emph{all} function calls, including
 those that returned successfully.

### 2017-04-30 paper/jfp-submission/introduction.tex
mention localization in abs/intro
Wes pass over the the entire document. Minor editing changes only.

@@
 A jump-compressed trace includes only function
-calls and returns, for example the trace in the bottom-left of
+calls and returns. For example, the trace in the bottom-left of
 Figure~\ref{fig:factorial} is jump-compressed.
 %
-Jump-compressed traces are similar to stack traces, both show a
-sequence of function calls that lead to a crash, but the jump-compressed
-trace also shows the return values of successful calls, which can be
+Jump-compressed traces are similar to stack traces in that both show a
+sequence of function calls that lead to a crash. However, jump-compressed
+traces also show the return values of successful calls, which can be
 useful in understanding why a particular path was taken.
@@
 of the programs, and at most 10 steps for over 80\% of the programs.
+%
+We can even use witnesses to \emph{localize} type errors with a simple
+heuristic that treats the values in a ``stuck'' term as \emph{sources}
+of typing constraints and the term itself as a \emph{sink},
+achieving around 70\% accuracy in locating the source of the error.
 

### 2017-04-30 paper/jfp-submission/safe.tex
misc tweaks
tweak
Wes pass over the the entire document. Minor editing changes only.

@@
-\subsubsection{How safe are the ``safe'' programs?}
+\subsection{How safe are the ``safe'' programs?}
 \label{sec:how-safe}
@@
 %\centering
-\includegraphics[width=0.8\linewidth]{distrib_ext.png}
+\includegraphics[width=0.7\linewidth]{distrib_ext.png}
 \vspace{-0.75cm}
@@
 %
-This speculative instantiation is of course unsound; we would have to
+This speculative instantiation is, of course, unsound; we would have to
 take care to avoid reporting frivolous type errors that were caused by
@@
 %
-We would need to track which holes were instantiated speculatively in
-order to distinguish type errors that would have happened regardless, as
+We would need to track which holes were instantiated speculatively 
+to distinguish type errors that would have happened regardless, as
 in @fac@, from type errors that were caused by our instantiation.
@@
 %
-Since our instantiation was speculative, we can still view @x@ and @y@
-\emph{morally} as holes, and thus treat the @*.@ operator as a normal
+Since our instantiation was speculative, and @x@ and @y@
+% WRW thinks "morally" is too informal for this venue. 
+were \emph{originally} hole, we can treat the @*.@ operator as a normal
 narrowing point with two holes.
@@
 %
-This seems like a modest extension to our semantics, and would be
+This represents only a modest extension to our semantics, and would be
 interesting to explore further.

### 2017-05-16 paper/jfp-submission/evaluation.tex

@@
 %
-Using this heuristic, \toolname's witness are competitive with the
+Using this heuristic, \toolname's witnesses are competitive with the
 state-of-the-art localization tools \mycroft and \sherrloc.
@@
 idioms including (tail) recursive functions, higher-order functions,
-polymorphic and algebraic data types. % and expression evaluators.
+and polymorphic and algebraic data types. % and expression evaluators.
 

### 2017-05-16 paper/jfp-submission/locating.tex

@@
 %
-In addition to collecting ill-typed programs, we collected
-subsequent, fixed versions of the programs.
+This means that, in addition to collecting ill-typed programs, we
+collected subsequent, fixed versions of the same programs.
 %
-For each ill-typed program in a student's interaction trace, we identify
-the student's fix by searching for the first type-correct program
-that follows it in the trace.
+% For each ill-typed program in a student's interaction trace, we identify
+% the student's fix by searching for the first type-correct program
+% that follows it in the trace.
+For each ill-typed program compiled by a student, we identify the student's
+\emph{fix} by searching for the first type-correct program that the student
+subsequently compiled.
 %

### 2017-05-16 paper/jfp-submission/overview.tex

--- 2016-05-24 paper/icfp16-cameraready/overview.tex
@@
 %
-In other words, for \emph{any} type you might
-assign to the function, there exist inhabitants
+In other words, for \emph{any} types you might
+assign to the function's inputs, there exist values
 that will cause the function to go wrong.
@@
   \includegraphics[height=2in]{simple.png}
-  \caption{The reduction graph for \texttt{1+2+3}. The two edges
-    produced by the transition from \texttt{1+2+3} to \hbox{\texttt{3+3}}
-    are highlighted.}
+  \caption{The reduction graph for \texttt{1+2+3}, highlighting the edges
+    produced by reducing \texttt{1+2+3} to \hbox{\texttt{3+3}}.}
 \label{fig:simple-reduction-hi}
@@
 operations \emph{post-hoc}: ``step into'' to zoom in on a specific
-function call, and ``step over'' to skip over an uninteresting
-sub-computation.
+function call, and ``step over'' to skip over uninteresting
+computations.
 

### 2017-06-21 paper/jfp-submission/evaluation.tex
update some numbers, still need investigation of failed witness

@@
 from which we extracted \ucsdsize\ distinct, ill-typed \ocaml\ programs
-from a cohort of 46 students..
+from a cohort of 46 students.
 %
@@
 \caption{Results of our coverage testing. Our random search successfully
-  finds witnesses for 76--82\% of the programs in under one second,
+  finds witnesses for 76--83\% of the programs in under one second,
   improving to 84--85\% in under 10 seconds. }
@@
 \caption{Distribution of test outcomes. In both datasets we detect
-  actual type errors at least 75\% of the time, unbound variables or
-  constructors 4--5\% of the time, and diverging loops 3\% of the
-  time. For the remaining 13--16\% of the programs we are unable to
+  actual type errors at least 77\% of the time, unbound variables or
+  constructors 4\% of the time, and diverging loops 2--3\% of the
+  time. For the remaining 15--16\% of the programs we are unable to
   provide any useful feedback. }
@@
 compile-time check. If we extend our tolerance to a 10 second timeout,
-we reach 84--85\% coverage, and if we allow a 60 second search,
-we hit a maximum of 84--87\% coverage.
+we reach 84\% coverage, and if we allow a 60 second search,
+we hit a maximum of 84--85\% coverage.
 %
 Interestingly, while the vast majority of witnesses corresponded to a
-type-error, as expected, 4--5\% triggered an unbound variable error (even
+type-error, as expected, 4\% triggered an unbound variable error (even
 though \ocaml\ reported a type error) and 3\% triggered an infinite
@@
 %
-For the remaining 13--16\% of programs we were unable to provide any useful
+For the remaining 15--16\% of programs we were unable to provide any useful
 feedback as they either completed 1,000 tests successfully, or timed out
@@
 %
-The average number of single-step reductions per trace is 30 for the
-\ucsdbench\ dataset (45 for the \uwbench\ dataset) with a maximum of
-2,745 (\resp 982) and a median of 14.5 (\resp 14.5).
+The average number of single-step reductions per trace is 17 for the
+\ucsdbench\ dataset (42 for the \uwbench\ dataset) with a maximum of
+2,745 (\resp 982) and a median of 15 (\resp 15).
 %
-The average number of jumps per trace is 7 (\resp 10) with a
+The average number of jumps per trace is 7 (\resp 9) with a
 maximium of 353 (\resp 221) and a median of 4 (\resp 4).
 %
-In both datasets 60\% of traces have at most 5 jumps, and 80\% or more
+In both datasets about 60\% of traces have at most 5 jumps, and 80\% or more
 have at most 10 jumps.

### 2017-06-21 paper/jfp-submission/locating.tex
update some numbers, still need investigation of failed witness

@@
 standard deviation above the mean, establishing a diff threshold of
-45\%.
+40\%.
 %
 This accounts for roughly 14\% of programs pairs we discovered, leaving
-us with 2,425 program pairs.
+us with 2,710 program pairs.
 
@@
 field, around 15\% for each tool, leaving us with a benchmark set of
-1,569 programs.
+1,759 programs.
 
@@
 %
-We mitigate this threat with a large empirical evaluation of 1,569
+We mitigate this threat with a large empirical evaluation of 1,759
 programs, drawn from a cohort of 46 students.
@@
 \caption{Accuracy of type error localization. \toolname's witness-based
-  predictions outperform \ocaml by 22 points, and are competitive
+  predictions outperform \ocaml by 21 points, and are competitive
   with the state-of-the-art tools \mycroft and \sherrloc.}
@@
 locations per program, \mycroft and \sherrloc have a long tail with a
-maximum of 22 (\resp 12) locations, while \toolname's maximum is 5
+maximum of 22 (\resp 11) locations, while \toolname's maximum is 5
 locations.

### 2017-06-21 paper/jfp-submission/plots.py

@@ -404,7 +404,7 @@ def plot_blame():
 
     # FIXME: load these numbers from csv...
     tools = ['OCaml', 'NanoMaLy', 'Mycroft', 'SHErrLoc']
-    accs  = [49.0,    70.7,       71.0,      78.9      ]
+    accs  = [47.4,    68.3,       73.2,      75.8      ]
 
     N = len(tools)
     ind = np.arange(N)    # the x locations for the groups
@@ -430,7 +430,7 @@ def plot_blame():
 
 if __name__ == '__main__':
     seminal = read_csv('../../seminal.csv')
-    ucsd = read_csv('../../ucsd.csv')
+    ucsd = read_csv('../../ucsd-sp14.csv')
 
     plot_distrib(seminal, ucsd)
     plot_distrib_extended(ucsd)

### 2017-06-21 paper/jfp-submission/study.R

@@ -82,10 +82,10 @@ fixes = data.frame(fixes0, fixes1, fixes2, fixes3)
 fixes = fixes[fixes$fixes0 != -1 & fixes$fixes1 != -1 & fixes$fixes2 != -1 & fixes$fixes3 != -1,]
 
 print("REASONS")
-print(table(reasons[,1]))
-print(table(reasons[,2]))
-print(table(reasons[,3]))
-print(table(reasons[,4]))
+print(reasons[,1])
+print(reasons[,2])
+print(reasons[,3])
+print(reasons[,4])
 print(kappam.fleiss(reasons,exact=TRUE,detail=TRUE))
 #print(kappam.fleiss(reasons[1:3],detail=TRUE))
 
@@ -99,7 +99,9 @@ print(kappam.fleiss(fixes,exact=TRUE,detail=TRUE))
 
 significance <- function(x,y) {
   print(x)
+  print(mean(x))
   print(y)
+  print(mean(y))
   wilcox.test(x, y, alternative="g")
 }
 

### 2017-06-21 paper/jfp-submission/safe.tex
typo
numbers

@@
 
-An immediate question arises regarding the 13--16\% of programs for
+An immediate question arises regarding the 15--16\% of programs for
 which we could not synthesize a witness:
@@
 
-To answer this question, we investigated the 720 \ucsdbench programs for
+To answer this question, we investigated the 732 \ucsdbench programs for
 which we failed to find a witness.
@@
 %
-This left us with 473 programs that required manual coding; we selected
+This left us with 504 programs that required manual coding; we selected
 a random sample of 50 programs to investigate, and will report results
@@
 % WRW thinks "morally" is too informal for this venue. 
-were \emph{originally} hole, we can treat the @*.@ operator as a normal
+were \emph{originally} holes, we can treat the @*.@ operator as a normal
 narrowing point with two holes.

### 2017-06-22 paper/jfp-submission/evaluation.tex
woops, dedup labels

@@
 either \ocaml's error or \toolname's jump-compressed trace;
-the full user study is available in \S~\ref{sec:user-study}.
+the full user study is available in \S~\ref{sec:user-study-exams}.
 %

### 2017-06-22 paper/jfp-submission/evaluation.tex
reference

--- 2017-06-21 paper/jfp-submission/evaluation.tex
@@
 either \ocaml's error or \toolname's jump-compressed trace;
-the full user study is available in \S~\ref{sec:user-study}.
+the full user study is available in \autoref{sec:user-study-exams}.
 %

### 2017-07-12 paper/jfp-submission/evaluation.tex
oops

--- 2017-06-21 paper/jfp-submission/evaluation.tex
@@
 either \ocaml's error or \toolname's jump-compressed trace;
-the full user study is available in \S~\ref{sec:user-study}.
+the full user study is available in Appendix~\ref{sec:user-study-exams}.
 %

### 2017-12-10 paper/jfp-submission/response.txt
start responding to reviews

We thank the reviewers for their detailed comments and suggestions.

## Questions about the user study

Reviewer 1
> I was not completely convinced by the metrics used for the statistical
> analysis; I would have found it more convincing to measure the speed
> at which an incorrect program may be fixed by a student sitting in front
> of a machine. That said, some statistical analysis is better than none,
> and I am greatful that the authors attempted some evaluation.

Reviewer 2
> 7. why did you do a user study as part of an exam, and not as an
> independent study? I don't think this is a good idea.

We agree that it would be preferable to have timing data in addition to
correctness data. We did attempt a study that would have collected
timing data, but unfortunately were not able to recruit enough
participants to draw any conclusions. We have updated the text in the
"Constuct" subsection of the "Threats to Validity" to mention lack of
timing data as a threat.

Reviewer 2
> 2. you compare only to OCAML in your student user study, which seems
> to be to be an easy target. Why not Sherrloc or Mycroft which at least
> get the location right more often? If OCAML much more often gives the
> wrong location, the message is not likely to be helpful. Indeed, I
> would expect you to only compare the systems for those cases that both
> NANOMALY and OCAML blame the same location.

Reviewer 2
> 3. does your tool help students to learn about the type system, or is
> it simply a better crutch to help them along. Did you investigate this
> at all?

These are valid concerns, we have updated the "Threats to Validity" to
mention them.

Reviewer 3
> The discussion of inter-rater reliability is strange. Why do we need
> to know if the exams are graded similarly, except to check that UCSD
> students are fairly graded? Also, why is IRR so low, given that
> these are questions with a correct answer? Given that there are only
> 60 exams in question, having a single evaluator with a clear rubric
> would eliminate both this potential source of inconsistency as well
> as the need to discuss IRR at all.

Our use of IRR is intended to account for potential bias in the
raters. One such form of bias (conscious or not) could be bias in favor
of NanoMaLy since it is new and interesting, and the raters (even if
they are not directly connected to the project) may want to see it
succeed. Due to the graphical nature of NanoMaLy's traces it is not easy
to hide whether a given answer was assisted by OCaml or NanoMaLy from
the raters. A further form of potential bias is inconsistency in
evaluating the student answers, which were long-form, English
explanations of the error/fix. There is a surprising amount of
subjectivity involved in judging, based on a couple sentences, whether
the student truly understood the error.

For these reasons we feel that our use of IRR is justified to lend
credibility to the results.

(We have not added this discussion to the paper as the IRR test is
really meant to be a preliminary "sanity-check", but would be happy to
if the reviewers feel it would be helpful.)

## Reviewer 1

> The tool is restricted to a small subset of OCaml, in particular it does
> not support side-effects. While many FP courses begin by focusing on
> pure programs only, there are some intringuing type errors that students
> face only when working with effectful programs, e.g. forgetting to apply
> a function to the unit argument. I speculate that it would be /not/ so
> easy to generalize the present work to deal with side effects, because
> of the need to maintain the state and report it to the user.

TODO

> Random-test generation would likely fail to produce counter-examples
> on slightly more complex programs such as red-black trees; for such
> programs, more advanced techniques such as constraint-based test
> generation are probably required. I was not able to quite understand
> whether the approach presented in the paper could be extended to
> accomodate constraint-based test generation.

TODO

> page 23, regarding section "witness exists", I got puzzled. The study
> there is on 50 programs. It was found that 2% of the programs admit a
> witness. I deduce this must correspond to one single program. But
> then, the next sentence tells that "slightly over half of these
> programs involved had some particular property, while the rest had
> some other property."  There semms to be some inconcistency there.

Indeed, the wording here was confusing. The percentages in this section
are presented as a fraction of *all* programs, not just the random
sample of failing programs. We have updated the text at the start of 5.3
to explain this.

## Reviewer 2

> 1. you do not address the issue of bias, the fact that locations
> towards the end of the program are more likely to be crashed at. This
> is similar to the left-to-right-bias of algorithm W, although in your
> case the left-to-rightness is in the trace, not the syntax tree (as
> addressed for that case in work by Hage and Heeren (VODCA 2008) and
> the work of Erwig and Chen).  Does your implementation allow general
> datatypes so that I can have case with many arms?

TODO

> 5. your approximation of functions to fun is rather strong, and it is
> unclear to me how well that works. I guess since all executions start
> from the beginning of the programs keeps everything rather concrete,
> so that you do not need to invent functions. But I am not sure, and I
> would like to see a thorough discussion. I am fine with novice
> programs being short and simple since that is quite likely to be the
> case when they resort to Nanomaly, but when you do that, it should be
> clear that you do.  What may be related to this is the idea to deal
> with Currying. What about functions like foldr of which the result is
> a type variable? I know that some static work on error diagnosis
> sometimes runs into a problem of some sort and since you work at the
> dynamic level this may not be the case for you. But do say and explain
> so.

TODO

> 1. explain how things would be if applied to other functional
> languages, in particular Haskell. What about laziness? What about
> overloaded numerals?  It seems to me that these are quite
> problematic. For example, the 0 in your fac program will be Num a =>
> a.

TODO

> 2. what is the limit to the size of programs you can reasonably
> handle?  What if we have a large program, is it possible to have a
> different starting point? So does a trace have to start at the start
> of the program? I guess this is important since if you start with a
> higher order function then you need to dynamically invent a function,
> and then your approximation might well be in the way. This is
> something I am really curious about.  Now you have 5% Non-parametric
> function types. Does that change in this situation?

TOOD

### 2017-12-10 paper/jfp-submission/evaluation.tex
start responding to reviews

@@
 Using this heuristic, \toolname's witnesses are competitive with the
-state-of-the-art localization tools \mycroft and \sherrloc.
+state-of-the-art localization tools \mycroft~\cite{Loncaric2016-uk}
+and \sherrloc~\cite{Zhang2014-lv}.
 
@@
 
-\paragraph{Threats to Validity}
-Measuring understanding is a difficult task; the following summarize
-the threats to the validity of our results.
-
-\subparagraph{Construct.}
-%
-We used the correctness of the student's explanation of, and fix for,
-the type error as a proxy for her understanding, but it is possible
-that other metrics would produce different results.
-
-\subparagraph{Internal.}
-%
-We assigned students randomly to two groups. The first was given
-\ocaml's errors for @append@ and \hbox{@digitsOfInt@,} and \toolname's trace
-for @sumList@ and \hbox{@wwhile@;} the second was given the opposite
-assignment of errors and traces. This assignment ensured that: (1) each
-student was given \ocaml and \toolname problems; and (2) each student
-was given an ``easy'' and ``hard'' problem for both \ocaml and
-\toolname. Students without sufficient knowledge of \ocaml could affect
-the results, as could the time-constrained nature of an exam. For these
-reasons we excluded any answers left blank from our analysis.
-
-\subparagraph{External.}
-%
-Our experiment used students in the process of learning \ocaml,
-and thus may not generalize to all developers. The four
-programs were chosen manually, via a random selection and
-filtering of the programs in the \ucsdbench dataset. In some cases we made
-minor simplifying edits (\eg alpha-renaming, dead-code removal) to the
-programs to make them more understandable in the short timeframe of an
-exam; however, we never altered the resulting type-error. A different
-selection of programs may lead to different results.
-
-\subparagraph{Conclusion.}
-%
-We collected exams from 60 students, though due to the nature of the
-study not every student completed every problem.
-%
-The number of complete submissions ranges from 13 (for the \toolname
-version of @wwhile@) to 28 (for the \ocaml version of @sumList@), out of
-a maximum of 30 per program-tool pair.
-%
-Our results are statistically significant in only 2 out of 8 tests; however,
-collecting more responses per test pair was not possible as it
-would require having students answer the same problem twice (once with
-\ocaml and once with \toolname).
-
 \begin{figure}[t]
@@
 %
+
+\paragraph{Threats to Validity}
+Measuring understanding is a difficult task; the following summarize
+the threats to the validity of our results.
+
+\subparagraph{Construct.}
+%
+We used the correctness of the student's explanation of, and fix for,
+the type error as a proxy for her understanding, but it is possible that
+other metrics would produce different results. One such metric that is
+also surely relevant is time-to-completion, \ie a good error report
+should \emph{quickly} guide the student to a fix. Unfortunately, the
+in-class exam setting of our study did not admit the collection of
+timing data.
+
+Furthermore, one might object to our selection of \ocaml as
+the baseline comparison rather than \sherrloc or \mycroft, which also
+claim to produce more accurate error reports. This is indeed a
+limitation of our study, but we note that \sherrloc blames the same
+expression as \ocaml in both @wwhile@ and @append@, and in @digitsOfInt@
+both blame the wrong function.
+
+Finally, one might point out that our study investigates the use of
+\toolname as a \emph{debugging} aid rather than as a \emph{teaching}
+aid. That is, it may be that \toolname helps students solve their
+immediate problem, but does not help them build a lasting understanding
+of the type system. We have not attempted a longitudinal study of the
+long-term impact of using \toolname, but we agree that it would be an
+interesting future direction.
+
+\subparagraph{Internal.}
+%
+We assigned students randomly to two groups. The first was given
+\ocaml's errors for @append@ and \hbox{@digitsOfInt@,} and \toolname's trace
+for @sumList@ and \hbox{@wwhile@;} the second was given the opposite
+assignment of errors and traces. This assignment ensured that: (1) each
+student was given \ocaml and \toolname problems; and (2) each student
+was given an ``easy'' and ``hard'' problem for both \ocaml and
+\toolname. Students without sufficient knowledge of \ocaml could affect
+the results, as could the time-constrained nature of an exam. For these
+reasons we excluded any answers left blank from our analysis.
+
+\subparagraph{External.}
+%
+Our experiment used students in the process of learning \ocaml,
+and thus may not generalize to all developers. The four
+programs were chosen manually, via a random selection and
+filtering of the programs in the \ucsdbench dataset. In some cases we made
+minor simplifying edits (\eg alpha-renaming, dead-code removal) to the
+programs to make them more understandable in the short timeframe of an
+exam; however, we never altered the resulting type-error. A different
+selection of programs may lead to different results.
+
+\subparagraph{Conclusion.}
+%
+We collected exams from 60 students, though due to the nature of the
+study not every student completed every problem.
+%
+The number of complete submissions ranges from 13 (for the \toolname
+version of @wwhile@) to 28 (for the \ocaml version of @sumList@), out of
+a maximum of 30 per program-tool pair.
+%
+Our results are statistically significant in only 2 out of 8 tests; however,
+collecting more responses per test pair was not possible as it
+would require having students answer the same problem twice (once with
+\ocaml and once with \toolname).
 

### 2017-12-10 paper/jfp-submission/safe.tex
start responding to reviews

@@
 we note the classes that were based on the random sample with a ``*''.
+%
+Note that the percentages referenced in Figure~\ref{fig:no-witness} (and
+in the sequel) are with respect to the total number of programs in the
+\ucsdbench dataset, not only those were \toolname failed to find a
+witness.
 
@@
 %
-We found that for 5\% programs \toolname got stuck when it tried to
+We found that for 5\% of all programs \toolname got stuck when it tried to
 compare two holes.
@@
 %
-5\% of programs lack a witness in our semantics due to our
+5\% of all programs lack a witness in our semantics due to our
 non-parametric $\tfun$ type for functions.
@@
 %
-4\% of programs contained type errors that were unreachable, either
+4\% of all programs contained type errors that were unreachable, either
 because they were dead code, or because the student called the function
@@
 %
-We found that only 2\% of programs admit a witness that \toolname
+We found that only 2\% of all programs admit a witness that \toolname
 was unable to discover.

### 2017-12-10 paper/jfp-submission/introduction.tex
mention the extension

@@
 
+\paragraph{New Material}
+This paper is an extension of our ICFP '16 paper of the same
+name~\cite{Seidel2016-ul}.
+%
+In this extension we have focused on expanding our experimental evaluation.
+%
+First, in \S~\ref{sec:how-safe} we investigate the student programs for
+which we were unable to synthesize a witness. We group the failures into
+five categories, give representative examples, and suggest ways to
+improve our feedback in these cases. Interestingly, we find that in the
+majority of these failed cases, the programs do not actually admit a
+witness in our semantics.
+%
+Second, in \S~\ref{sec:locating} we attempt to use our witnesses to
+localize type errors with a simple heuristic. We treat the stuck term as
+a sink for typing constraints, and the values it contains as sources of
+constraints. We can then predict that either the stuck term or one of
+the terms that \emph{produced} a value it contains is likely at fault
+for the error. We compare our localizations to \ocaml and two
+state-of-the-art type error localization tools, and find that we are
+competitive with the state of the art.
+%
+Finally, we have also extended \S~\ref{sec:user-study} with an analysis
+of the statistical significance of our user study results.
+
 %%% Local Variables:

### 2017-12-10 paper/jfp-submission/response.txt
mention the extension
more response

@@
 
-## Questions about the user study
+As requested, we have added a paragraph at the end of the Introduction
+that describes the major changes from the ICFP paper.
+
+In the rest of this document, we will respond to the larger comments and
+concerns of the reviewers, and describe the changes we have made to
+address them.
+
+# Questions about the Evaluation
+
+## User Study
 
@@
 
-## Reviewer 1
+## Investigating the "safe" programs
 
-> The tool is restricted to a small subset of OCaml, in particular it does
-> not support side-effects. While many FP courses begin by focusing on
-> pure programs only, there are some intringuing type errors that students
-> face only when working with effectful programs, e.g. forgetting to apply
-> a function to the unit argument. I speculate that it would be /not/ so
-> easy to generalize the present work to deal with side effects, because
-> of the need to maintain the state and report it to the user.
-
-TODO
-
-> Random-test generation would likely fail to produce counter-examples
-> on slightly more complex programs such as red-black trees; for such
-> programs, more advanced techniques such as constraint-based test
-> generation are probably required. I was not able to quite understand
-> whether the approach presented in the paper could be extended to
-> accomodate constraint-based test generation.
-
-TODO
-
+Reviewer 1
 > page 23, regarding section "witness exists", I got puzzled. The study
@@
 
-## Reviewer 2
+# Questions about limitations
 
+## Traversal Bias
+
+Reviewer 2
 > 1. you do not address the issue of bias, the fact that locations
@@
 
-TODO
+Indeed, this was an oversight on our part. We have added a new paragraph
+to section 5.8 to discuss our bias.
 
+## Other language features
+
+Reviewer 1
+> The tool is restricted to a small subset of OCaml, in particular it does
+> not support side-effects. While many FP courses begin by focusing on
+> pure programs only, there are some intringuing type errors that students
+> face only when working with effectful programs, e.g. forgetting to apply
+> a function to the unit argument. I speculate that it would be /not/ so
+> easy to generalize the present work to deal with side effects, because
+> of the need to maintain the state and report it to the user.
+
+While side effects would certainly make the visualization more complex,
+we don't believe they present a fundamental limitation to our approach.
+Your example of the student forgetting to supply the unit argument is
+actually quite similar to the problem posed by curried functions. Either
+the function will be consumed where it's return value was expected,
+which would produce an error similar to our `wwhile` example, our it
+will be discarded entirely as the call was intended solely for its side
+effect. In the latter case, we could simply add a restriction that any
+such discarded values must be unit (it is commonly considered a best
+practice to explicitly discard the results of functions called just for
+their side effects with a `void` function).
+
+Reviewer 2
+> 1. explain how things would be if applied to other functional
+> languages, in particular Haskell. What about laziness? What about
+> overloaded numerals?  It seems to me that these are quite
+> problematic. For example, the 0 in your fac program will be Num a =>
+> a.
+
+Overloading is indeed problematic for us. We discuss this in the new
+section 5.3 (under the heading "Ad-Hoc Polymorphsim"), both describing
+the issue and presenting a possible solution. We return to the issue in
+section 5.8 and mention how a language like Haskell with pervasive
+overloading would exacerbate the problem.
+
+## Scalability
+
+Reviewer 1
+> Random-test generation would likely fail to produce counter-examples
+> on slightly more complex programs such as red-black trees; for such
+> programs, more advanced techniques such as constraint-based test
+> generation are probably required. I was not able to quite understand
+> whether the approach presented in the paper could be extended to
+> accomodate constraint-based test generation.
+
+Indeed, we discuss the limitations of random generation in section
+5.8. We have added a sentence there describing how we could replace
+random generation with constraint-based generation.
+
+Reviewer 2
 > 5. your approximation of functions to fun is rather strong, and it is
@@
 
-TODO
-
-> 1. explain how things would be if applied to other functional
-> languages, in particular Haskell. What about laziness? What about
-> overloaded numerals?  It seems to me that these are quite
-> problematic. For example, the 0 in your fac program will be Num a =>
-> a.
-
-TODO
-
 > 2. what is the limit to the size of programs you can reasonably
@@
 
-TOOD
+TODO

### 2017-12-10 paper/jfp-submission/evaluation.tex
more response

@@
 
-There are, of course, drawbacks to our approach. Four that stand out
+There are, of course, drawbacks to our approach. Five that stand out
 are:
@@
 %(3) the inability to handle certain instances of infinite types; and
-(3) our use of a non-parametric function type; and
+(3) our use of a non-parametric function type;
 %
-(4) handling ad-hoc polymorphism.
+(4) handling ad-hoc polymorphism; and
+%
+(5) traversal bias.
 
@@
 can address this short-coming by performing an exhaustive search for
-inputs (\emph{resp}.\ paths through the program). As our experiments
-show, however, novice programs do not appear to require more advanced
-search techniques, likely because they tend to be simple.
+inputs (\emph{resp}.\ paths through the program). Our approach does not
+rely on random generation, we could easily substitute it for
+dynamic-symbolic execution by extending the evaluation relation to
+maintain a path condition and replacing the \gensym function with a call
+to the constraint solver.
+%
+As our experiments show, however, novice programs do not appear to
+require more advanced search techniques, likely because they tend to be
+simple.
 
@@
 
+\paragraph{Traversal Bias}
+
+A common problem with typecheckers is that the order in which the
+typechecker traverses the abstract syntax tree \emph{biases} it in favor
+of blaming expressions that are seen later~\cite{McAdam1998-ub}. This
+usually takes the form of a left-to-right bias with respect to the
+source code (terms that appear later \emph{textually} are more likely to
+be blamed), but in our case the bias is with respect to the execution
+trace. Incorporating our notion of type sources from
+\S~\ref{sec:locating} into the visualization, \eg by including those
+reductions in the initial visualization, may help alleviate our bias.
+% in a similar manner to the proposals of McAdam~\shortcite{McAdam1998-ub},
+
+% Several solutions to the bias problem have been proposed.
+% %
+% Hage and Heeren~\shortcite{Hage2009-em} describe a system that allows
+% the compiler author to selectively control the bias and thus produce
+% better errors, while McAdam~\shortcite{McAdam1998-ub},
+
 % \begin{itemize}

### 2017-12-10 paper/jfp-submission/response.txt
more response

@@
 
-TODO
+We discuss the approximation of functions to `fun` in the
+"Non-Parametric Function Type" subsection of 5.3, it is the other major
+limitation of our work (the first being overloading). We do in fact have
+to invent functions sometimes (the definition of `gen` says that these
+must be constant functions), and this interacts very poorly with
+higher-order functions like `foldr`. We have added another example to
+the discussion in 5.3 that illustrates the issue.

### 2017-12-10 paper/jfp-submission/safe.tex
more response

@@
 
-\paragraph{Ad-Hoc Polymorphism}
+\paragraph{Ad-hoc Polymorphism}
 %
@@
 
+The $\tfun$ abstraction is also problematic when we have to generate new
+functions. Consider the following @pipe@ function that composes a list
+of functions.
+%
+\begin{code}
+  let pipe fs =
+    let f a x = a x in
+    let base a = a in
+    List.fold_left f base fs
+\end{code}
+%
+In the folding function @f@, the student has applied the accumulator @a@
+to the new function @x@, rather than composing the two.
+%
+The \ocaml compiler again detects a cyclic typing constraint and rejects
+the program, but \toolname is unable to produce a witness.
+%
+In this case the issue lies in the fact that the safety of a call to
+@pipe@ is determined by its arguments: the call @pipe [(fun x -> 1)]@ is
+safe, but the call @pipe [(fun x -> 1); (fun x -> 2); (fun x -> 3)]@
+will get stuck when we try to reduce @1 (fun x -> 3)@.
+%
+Unfortunately, \toolname is unable to synthesize such a witness because
+of the abstraction to $\tfun$.
+%
+Specifically, this abstraction forces our hand inside $\gensym$; we do
+not know the what the input and output types of the function should be,
+so the only safe thing to do is to generate a function that accepts any
+input and returns a value of a yet-to-be-determined type.
+%
+Thus, \toolname's lenient instantiation of holes prevents it from
+discovering a witness for @pipe@.
+
 \paragraph{Dead Code and ``Safe'' Function Calls *}
@@
 %
-Our investigation suggests that the vast majority of programs
+Our investigation suggests that the majority of programs
 for which we fail to find a witness do not, in fact, admit a
-witness.
+witness under \toolname's semantics.
 %

### 2017-12-10 paper/jfp-submission/response.txt
a word

@@
 to invent functions sometimes (the definition of `gen` says that these
-must be constant functions), and this interacts very poorly with
-higher-order functions like `foldr`. We have added another example to
-the discussion in 5.3 that illustrates the issue.
+must be constant functions), and this interacts poorly with higher-order
+functions like `foldr`. We have added another example to the discussion
+in 5.3 that illustrates the issue.

### 2017-12-22 paper/jfp-submission/response.txt
address wes comments

@@
 participants to draw any conclusions. We have updated the text in the
-"Constuct" subsection of the "Threats to Validity" to mention lack of
+"Constuct" subsection of the "Threats to Validity" to discuss the lack of
 timing data as a threat.
@@
 These are valid concerns, we have updated the "Threats to Validity" to
-mention them.
+discuss them.
 
@@
 
-(We have not added this discussion to the paper as the IRR test is
-really meant to be a preliminary "sanity-check", but would be happy to
-if the reviewers feel it would be helpful.)
+We have added a footnote to the discussion of IRR explaining our
+justification.
 
@@
 sample of failing programs. We have updated the text at the start of 5.3
-to explain this.
+to clarify this.
 
@@
 
+We have added a discussion of imperative features to section 6.
+
 Reviewer 2
@@
 overloading would exacerbate the problem.
+
+We have clarified the relevant text in the paper.
 

### 2017-12-22 paper/jfp-submission/evaluation.tex
address wes comments

@@
 variants --- in a tool called \nanomaly.
+\footnote{\url{https://github.com/ucsd-progsys/nanomaly}}
 %
@@
 degree to which the annotators consistently graded the exams.
+\footnote{Measuring IRR is an established practice to account for
+  potential bias among raters. The students were asked to explain the
+  errors in English; judging whether they truly understood the errors
+  invovled a surprising amount of subjectivity, and is thus subject to
+  rater bias.}
 %
@@
 
-There are, of course, drawbacks to our approach. Five that stand out
-are:
+There are, of course, drawbacks to our approach.
+% %
+% Five that stand out are:
+% %
+% (1) coverage limits due to random generation;
+% %
+% (2) dealing with explosions in the size of generated traces;
+% %
+% %(3) the inability to handle certain instances of infinite types; and
+% (3) our use of a non-parametric function type;
+% %
+% (4) handling ad-hoc polymorphism; and
+% %
+% (5) traversal bias.
 %
-(1) coverage limits due to random generation;
-%
-(2) dealing with explosions in the size of generated traces;
-%
-%(3) the inability to handle certain instances of infinite types; and
-(3) our use of a non-parametric function type;
-%
-(4) handling ad-hoc polymorphism; and
-%
-(5) traversal bias.
+In the sequel we discuss a selection of drawbacks, and how we might
+address them in future work.
 
@@
 benchmarks), but it would surely be exacerbated by a language like
-\haskell, which makes heavy use of overloading. We suspect that either
-dynamic-symbolic execution or speculative instantiation of holes would
-allow us to handle ad-hoc polymorphism, but defer a proper treatment to
-future work.
+\haskell, which overloads not only functions but also numeric literals,
+as well as strings and lists if one enables the respective language
+extensions.
+%
+We suspect that either dynamic-symbolic execution or speculative
+instantiation of holes would allow us to handle ad-hoc polymorphism, but
+defer a proper treatment to future work.
 
@@
 
+\paragraph{Side Effects}
+
+While our implementation does not currently support side effects, it
+would not be difficult to add support.
+%
+The search procedure is easily extended to support mutation by
+maintaining a store in the evaluation relation, the main complexity
+would be in extending the trace visualization to demonstrate mutation.
+%
+Incorporating mutation directly into our reduction-based visualization
+would be difficult, as mutation would have non-local effects on the
+expressions and may be difficult for students to follow.
+%
+Instead, we could follow the example of
+%
+\textsc{Python Tutor}~\cite{GuoSIGCSE2013}
+%
+and provide a separate visualization of the mutable store, with references
+visualized as constant pointers to changing objects.
+
 % \begin{itemize}

### 2017-12-22 paper/jfp-submission/introduction.tex
address wes comments

@@
 
-\paragraph{New Material}
-This paper is an extension of our ICFP '16 paper of the same
-name~\cite{Seidel2016-ul}.
-%
-In this extension we have focused on expanding our experimental evaluation.
+\paragraph{Contributions Relative to Prior Publications}
+This paper extends our ICFP '16 paper of the same
+name~\cite{Seidel2016-ul}, focusing on the experimental evaluation.
 %

### 2017-12-22 paper/jfp-submission/evaluation.tex
minor

@@
 By highlighting conflicting values, \ie\ the source and sink
-of the problem, and not making assumption about function correctness, \toolname\
+of the problem, and not making assumptions about function correctness, \toolname\
 focusses the user's attention on the piece of code that is

### 2017-12-22 paper/jfp-submission/related.tex
minor

@@
 %
-Many groups have explored techniques to pinpoint the true source
-of errors reported by static type checkers.
+Many groups have explored techniques to improve the error locations
+reported by static type checkers.
 % unification-based type inference procedures.
@@
 %
-Marceau~\etal~\shortcite{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
-messages in novice environments and present suggestions for improving
-their quality and consistency.
+Marceau~\etal~\shortcite{Marceau2011-ok,Marceau2011-cy} study the
+effectiveness of error messages in novice environments and present
+suggestions for improving their quality and consistency.
 %
-Hage~\&~Heeren~\shortcite{Hage2006-hc} identify a variety of general heuristics to improve
-the quality of type error messages, based on their teaching experience.
+Hage~\&~Heeren~\shortcite{Hage2006-hc} identify a variety of general
+heuristics to improve the quality of type error messages, based on their
+teaching experience.
+%
+Chargu{\'e}raud~\shortcite{Chargueraud2015-dc} presents a tabular format
+for type errors that can provide multiple explanations in a compact form.
 %
@@
 
-A function's type is a theorem about the its behavior.
+A function's type is a theorem about its behavior.
 Thus, \toolname's witnesses can be viewed as \emph{counter-examples},

### 2017-12-22 paper/jfp-submission/semantics.tex
minor

@@
 \begin{figure}[t]
-\[
+\[\def\arraystretch{1.5}
 \begin{array}{lll}
@@
   \triple{v}{\vsu}{\tsu'}    &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}, \\
-                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\end{array} \\
-  \triple{\stuck}{\vsu}{\tsu} &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}\end{array} \\
+                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\end{array} \\[0.16in]
+  \triple{\stuck}{\vsu}{\tsu} &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}\end{array} \\[0.08in]
   \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} &\ \textbf{if} \begin{array}{l} \tsu' = \unify{\{\thole, t\}}{\tsu}, \\ v = \gen{t}{\tsu'} \end{array}\\
-\end{cases}  \\
+\end{cases}  \\ %[0.2in]
 % \begin{cases}
@@
 %
-The types of (polymorphic) trees are obtained from the labels on their
-values, and the types of tuples directly from their values.
+The types of tuples are obtained directly from their values, and the
+types of (polymorphic) trees from their labels.
+%
+Note that the evaluation relation in Figure~\ref{fig:operational}
+guarantees that tree \emph{values} will be annotated with their type.
+%
+For nodes the type can be taken from the type of its value $v_1$, but
+for leaves the evaluation relation creates a new type hole $\thole$
+(this corresponds to polymorphic instantiation in a typechecker).
 
@@
 
-In order to convert the semantics into a \emph{procedure} for finding
-witnesses, we must address two challenges.
+We must address two challenges
+to convert the semantics into a \emph{procedure} for finding
+witnesses.
 %
@@
   If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$
-  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
+  then for all types $t_1 \ldots t_n$ there exist values $w_1:t_1\ \ldots\ w_n:t_n$ such that
   $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.

### 2017-12-22 paper/jfp-submission/related.tex
related work

@@
 thereby connecting it to work on using test cases to find
-counter-examples prior to starting a proof~\cite{ACL2Testing, Seidel15}.
+counter-examples prior to starting a proof~\cite{ACL2Testing, Nguyen2015-oo, Seidel15}.
 
@@
 to explore the resulting dynamic execution.
-%
+
+Clements~\etal~\shortcite{Clements2001-qj} present a reduction-based
+visualization of program execution similar to \toolname's, though their
+interaction model is closer to that of a traditional step-debugger,
+limited to taking single step forwards or backwards. In contrast,
+\toolname first presents an overview of the whole computation, and then
+allows the user to focus in on the interesting reductions.
+
 Perera~\etal~\shortcite{Perera2012-dy} present a tracing semantics

### 2018-01-13 paper/jfp-submission/safe.tex
minor formatting

@@
 
+\newpage
 Further, suppose that our speculative instantiation induces a frivolous
@@
 Thus, it is not particularly satisfying that \toolname fails to produce
-a witness here; a possible solution could be to track the types of the
+a witness here; one solution could be to track the types of the
 inputs, and demonstrate to the user how they change between recursive
@@
 %
-This represents only a modest extension to our semantics, and would be
+This represents only a modest extension to our semantics and would be
 interesting to explore further.
@@
 %
-Thus, \toolname's lenient instantiation of holes prevents it from
-discovering a witness for @pipe@.
+Thus, our lenient instantiation of holes prevents us from
+discovering a witness here.
 

### 2018-03-18 paper/jfp-submission/evaluation.tex
minor formatting
final reviewer comments

@@
 %
+\newpage
 \begin{ecode}
@@
   errors in English; judging whether they truly understood the errors
-  invovled a surprising amount of subjectivity, and is thus subject to
+  involved a surprising amount of subjectivity, and is thus subject to
   rater bias.}
@@
 be blamed), but in our case the bias is with respect to the execution
-trace. Incorporating our notion of type sources from
-\S~\ref{sec:locating} into the visualization, \eg by including those
-reductions in the initial visualization, may help alleviate our bias.
-% in a similar manner to the proposals of McAdam~\shortcite{McAdam1998-ub},
+trace.
+
+Incorporating our notion of type sources from \S~\ref{sec:locating} into
+the visualization, \eg by including those reductions in the initial
+visualization, may help alleviate our bias in a similar manner to
+McAdam's proposal. Hage and Heeren~\shortcite{Hage2009-em} offer another
+solution that allows the compiler author to selectively control the
+bias, and thus produce better errors, by prioritizing typing
+constraints. Unfortunately, due to \toolname's dynamic nature, providing
+this sort of control would likely require selectively changing the order
+of evalution; while sound for the pure subset of \ocaml that we address,
+this could nonetheless confuse newcomers to the language even more.
 
@@
 
-While our implementation does not currently support side effects, it
-would not be difficult to add support.
+While our implementation does not currently support side effects, we
+suspect it would not be difficult to add support.
 %
